### 2023-09-07 Stats Rethinking/Ch3/Notes.qmd
Took notes through 3.2

---
title: "Notes"
format: html
---

# Libraries

```{r}
library(rethinking)
```

# 3.1 - Sampling from a grid-approximate posterior

```{r}
#Add the prior dist which we set to a uniform dist
p_grid <- seq(from=0, to=1, length.out=1000)
prob_p <- rep(1,1000)

#Add the "data" probability
prob_data <- dbinom(6, size=9, prob=p_grid)

#Multiply the two to get the posterior dist
posterior <- prob_data * prob_p
posterior <- posterior / sum(posterior)
```

```{r}
#Grab (or sample) values from our posterior and graph them below
#1e4 = 10,000
samples <- sample(p_grid, prob=posterior, size = 1e4, replace = TRUE)
```

```{r}
plot(samples)
```

```{r}
dens(samples)
```

# 3.2 - Sampling to Summarize

We can use bounds to find information about our parameters

```{r}
#What is the probability that the proporiotn of water is less than 0.5?
sum(posterior[p_grid < 0.5])
```

### 2023-09-07 Stats Rethinking/Ch3/Notes.qmd
Finished Ch3 Notes

@@
 ```
+
+One such bound is the *compadability interval* (not to be confused with a confidence interval) which is an interval of posterior probability.
+
+However, there are infinite intervals from which to choose any density (given density \< 1).
+Therefore, we should use the *Highest Point Density Interval* (HPDI) which, for a given probability mass, will give us the range with the most (dense) observations.
+
+Still don't understand what a loss function is.
+
+# 3.3 - Sampling to simulate prediction

### 2023-09-08 Stats Rethinking/Ch3/Practice Problems.qmd
Homework for chapter 3 finished

---
title: "Practice Problems"
format: html
---

# Load Libraries

```{r}
library(rethinking)
library(tidyverse)
```

# Work with Dr.Parson

```{r}
N <- 1001
grid <- seq(0,1,length.out=N)
prior <- rep(1, N)
prior <- prior / N
```

```{r}
sams <- sample(grid, prob=prior, size=1e4, replace=TRUE)
```

```{r}
ggplot()+
  geom_density(aes(x=sams))
```

```{r}
dist <- rbinom(1e4, 15, sams)
```

```{r}
ggplot()+
  geom_histogram(aes(x=dist), bins=100)
```

```{r}
likelihood <- dbinom(8, 15, grid)
```

```{r}
posterior.raw <- likelihood*prior
posterior <- posterior.raw / sum(posterior.raw)
```

```{r}
sams2 <- sample(grid, prob=posterior, size=1e4, replace=TRUE)
```

```{r}
ggplot()+
  geom_density(aes(x=sams2))+
  geom_vline(aes(xintercept=8/15, color="red"))
```

```{r}
dist2 <- rbinom(1e4, 15, sams2)
```

```{r}
ggplot()+
  geom_histogram(aes(x=dist2), bins=100)
```

```{r}
dist3 <- rbinom(1e4, 15, prob=8/15)
```

```{r}
ggplot()+
  geom_histogram(aes(x=dist3), bins=100)
```

# 3E1

## Set up

```{r}
p_grid <- seq(from=0, to=1, length.out=1000)
prior <- rep(1,1000)
likelihood <- dbinom(6,size = 9, prob = p_grid)
posterior <- likelihood*prior
posterior <- posterior / sum(posterior)
set.seed(100)
samples <- sample(p_grid, prob=posterior, size=1e4, replace=TRUE)
```

How much posterior probability lies below p=0.2?

```{r}
sum(posterior[p_grid<.2])
```

# 3E2

How much of the posterior probability lies above p=0.8?

```{r}
sum(posterior[p_grid>.8])
```

# 3E3

How much of the posterior probability lies between p=0.2 and p=0.8?

```{r}
sum(posterior[p_grid<.8 & p_grid>.2])
```

# 3E4

20% of the posterior lies below which value of p?

```{r}
quantile(samples, .2)
```

# 3E5

20% of the posterior lies above which value of p?

```{r}
quantile(samples, .8)
```

# 3E6

Which values of p contain the narrowest interval equal to 66% to the posterior probability?

```{r}
HPDI(samples, prob=0.66)
```

# 3E7

Which vlaues of p contains 66% of the posterior probability, assuming equal posterior probability both below and above the interval?

```{r}
quantile(samples, c(.166666,.833333))
```

# 3M1

Suppose the globe tossing data had turned out to be 8 water in 15 tosses. Construct the posterior distribution, using grid approximation. Use the same flat prior as before.

```{r}
#Did this with in the "Work with Dr.Parson" section above
likelihood2 <- dbinom(8, 15, prob = p_grid)

posterior2 <- likelihood2*prior
posterior2 <- posterior2/sum(posterior2)
```

# 3M2

Draw 10,000 samples form the grid approximation from above. Then use the samples to calculate the 90% HPDI for p.

```{r}
samples2 <- sample(p_grid, prob=posterior2, size=1e4, replace=T)
HPDI(samples2, prob=0.9)
```

# 3M3

Construct a posterior predictive check for this model and data. This means simulate the distribution of samples, averaging over the posterior uncertainty in p. What is the probability of observing 8 water in 15 tosses?

```{r}
sim <- rbinom(1e4, 15, prob=samples2)

sum(sim == 8) / length(sim)
```

```{r}
ggplot()+
  geom_histogram(aes(x=sim), bins=100)
```

The problem didn't call for a histogram but this should be a good check. Looking at the value for 8, it's about 1,400 which tells us that the probability should be about 14% since we have 10,000 samples. Which is what the above chunk is telling us so this should be right.

# 3M4

Using the posterior distribution constructed fromt the new (8/15) data, now calculate the probability of observing 6 waters in 9 tosses.

```{r}
sim <- rbinom(1e4, 9, prob=samples2)

sum(sim == 6) / length(sim)
```

```{r}
ggplot()+
  geom_histogram(aes(x=sim), bins=100)
```

This tells us that the probability of seeing 6 water in 9 tosses is about 17% given our posterior sample which has p of 8/15.

# 3M5

Start over at 3M1, but now use a prior that is zero below p=0.5 and a constant above p=0.5. This corresponds to prior information that a majority of Earth's surface is water. Repeat each problem above and compare the inferences. What differences does the better prior make? If it helps, compare inferences (using both priors) to the true value p=0.7.

```{r}
#3M1
prior2 <- c(rep(0,500), rep(1, 500))

posterior3 <- likelihood2*prior2
posterior3 <- posterior3/sum(posterior3)
```

```{r}
#3M2
samples3 <- sample(p_grid, prob=posterior3, size=1e4, replace=T)
HPDI(samples3, prob=0.9)
```

```{r}
#3M3
sim <- rbinom(1e4, 15, prob=samples3)

sum(sim == 8) / length(sim)

ggplot()+
  geom_histogram(aes(x=sim), bins=100)
```

```{r}
#3M4
sim <- rbinom(1e4, 9, prob=samples3)
sum(sim == 6) / length(sim)

ggplot()+
  geom_histogram(aes(x=sim), bins=100)
```

This prior makes it more likely for us to observe the 6 waters in 9 tosses compared to the 8 waters in 15 tosses. Which doesn't make much sense and tells me that I've messed up somewhere. In face, shouldn't our prior with a 0 going all the way up to .5 make it so that we see very low values on the left half of the posteriors?

# 3M6

Suppose you want to estimate the Earth's proportion of water very precisely. Specifically, you want the 99% percentile interval of the posterior distribution of p to be only 0.05 wide. This means the distance between the upper and lower bound of the interval should be 0.05. How many times will you have to toss the globe to do this?

```{r}
#I think we should use the flat prior again for this
#However, our likelihood function should change here, but we're going to use a likelihood which reflects the true p of Earth
p_grid2 <- seq(from=0, to=1, length.out=1e4)

likelihood3 <- dbinom(7,10,prob=p_grid2)

posterior4 <- likelihood3*prior
posterior4 <- posterior4/sum(posterior4)

quantile(posterior4, c(.05, .995))
```

This is clearly wrong but I don't know where I'm messing up.

# 3H1

Using grid approximation, compute the posterior distribution for the probability of a birth being a boy. Assume a uniform prior probability. Which parameter value maximizes the posterior probability.

```{r}
#Set up for the problems
data(homeworkch3)
```

```{r}
b_grid <- seq(from=0, to=1, length.out=200)
b_prior <- rep(1, 200)

b_likelihood <- dbinom(111,200,prob=b_grid)

b_posterior <- b_likelihood*b_prior
b_posterior <- b_posterior/sum(b_posterior)
```

```{r}
b_grid[which.max(b_posterior)]
```

Our parameter value of .55 maximizes this probability, which makes sense since in the 200 births we saw there were 111 boys, which was just over half. Furthermore, with a flat prior the data we observe (the likelihood function) is going to heavily influence our posterior.

# 3H2

Using the sample function, draw 10,000 random parameter values from the posterior distribution you calculated above. Use these samples to estimate the 50%, 89%, and 97% highest posterior density intervals.

```{r}
b_samples <- sample(b_grid, prob=b_posterior, size = 1e4, replace = TRUE)
```

```{r}
HPDI(b_samples, prob=.5)
HPDI(b_samples, prob=.89)
HPDI(b_samples, prob=.97)
```

# 3H3

Use rbinom to simulate 10,000 replicates of 200 births. You should end up with 10,000 numbers, each one a count of boys out of 200 births. Compare the distributions of predicted numbers of boys to the actual count in the data (111 out of 200 births). Does it look like the model fits the data well? That is, doees the distribution of predictions include the actual observation as a central, likely outcome?

```{r}
b_sim <- rbinom(1e4, 200, prob = b_samples)

ggplot()+
  geom_histogram(aes(x=b_sim), bins=100)+
  geom_vline(xintercept = 111, color="red")
```

I would say this model does fit the data well if we only consider the central fit of it. That is, we see a peak of observations around 110 like the question was asking for.

# 3H4

Now compare 10,000 counts of boys from 100 simulated first borns only to the number of first births, birth1. How does the model look in this light?

```{r}
#Number of first born males
sum(birth1)
```

```{r}
b2_grid <- seq(from=0, to=1, length.out=100)
b2_prior <- rep(1, 100)

b2_likelihood <- dbinom(51,100,prob=b2_grid)

b2_posterior <- b2_likelihood*b2_prior
b2_posterior <- b2_posterior/sum(b2_posterior)

b2_samples <- sample(b2_grid, prob=b2_posterior, size = 1e4, replace = TRUE)

b2_sim <- rbinom(100, 100, prob = b2_samples)

ggplot()+
  geom_histogram(aes(x=b2_sim), bins=75)+
  geom_vline(xintercept = 51, color="red")
```

# 3H5

```{r}
b3_grid <- seq(from=0, to=1, length.out=100)
b3_prior <- rep(1, 100)

b3_likelihood <- dbinom(49,100,prob=b3_grid)

b3_posterior <- b3_likelihood*b3_prior
b3_posterior <- b3_posterior/sum(b3_posterior)

b3_samples <- sample(b3_grid, prob=b3_posterior, size = 1e4, replace = TRUE)

b3_sim <- rbinom(1e4, 100, prob = b3_samples)

ggplot()+
  geom_histogram(aes(x=b3_sim), bins=75)+
  geom_vline(xintercept = 49, color="red")
```

### 2023-09-08 Stats Rethinking/Ch3/Dr_Parson_ch_3_problems.Rmd
Create Dr_Parson_ch_3_problems.Rmd
Share my work on Chapter 3 practice problems.

---
title: "Chapter 3 practice problems"
author: "James Parson"
date: '2023-09-07'
output: html_document
---

```{r}
library(tidyverse)
library(rethinking)
```

# Sampling from prior

```{r}
p_grid <- seq( from=0 , to=1 , length.out=1000 )
prior <- rep( 1 / 1000 , 1000 )
samples <- sample( p_grid , prob=prior , size=1e4 , replace=TRUE )
```

```{r}
simulated_prior <- rbinom(n = 1e4, size = 15, prob = samples)

ggplot() +
  geom_histogram(mapping = aes(x = simulated_prior), bins = 100)
```

# Set-up for E problems

```{r}
p_grid <- seq( from=0 , to=1 , length.out=1000 )
prior <- rep( 1 , 1000 )
likelihood <- dbinom( 6 , size=9 , prob=p_grid )
posterior <- likelihood * prior
posterior <- posterior / sum(posterior)
set.seed(100)
samples <- sample( p_grid , prob=posterior , size=1e4 , replace=TRUE )
```

# 3E1

```{r}
sum(samples <= 0.2) / 1e4
```

#3e2

```{r}
sum(samples >= 0.8) / 1e4
```

#3e3

```{r}
sum(samples > 0.2 & samples < 0.8) / 1e4
```

#3e4
```{r}
quantile(samples, 0.2)
```

#3e5
```{r}
quantile(samples, 0.8)
```

#3e6

```{r}
HPDI(samples, prob = 0.66)
```

```{r}
0.7737738 - 0.5085085
```

#3e7

```{r}
quantile(samples, c(0.5 - 0.66/2, 0.5 + 0.66/2))
```

```{r}
0.7697698 - 0.5025025
```

A slightly wider interval!

# 3m1

```{r}
N <- 1e3
grid <- seq(from = 0, to = 1, length.out = N)
prior <- rep(1, N)
likelihood <- dbinom(x = 8, size = 15, prob = grid)
posterior.raw <- prior * likelihood
posterior <- posterior.raw / sum(posterior.raw)

ggplot() +
  geom_line(mapping = aes(x = grid, y = posterior * N))
```
I scaled the probabilities in the posterior by N for the graph to match the vertical scale of the density below.

# 3m2

```{r}
samples <- sample(grid, size = 1e4, replace = TRUE, prob = posterior)

ggplot() +
  geom_density(mapping = aes(x = samples))
```

```{r}
HPDI(samples, prob = 0.9)
```

```{r}
water <- rbinom(1e4, 15, prob = samples)

ggplot() +
  geom_histogram(mapping = aes(x = water), bins = 100)
```

```{r}
sum(water == 8) / 1e4
```

Comparing Bayesian result with maximum-likelihood model.

```{r}
dbinom(8, 15, prob = 8/15)
```

Heavier tails in the predictive distribution from the Bayesian model.

```{r}
max_likelihood <- rbinom(1e4, size = 15, prob = 8 / 15)

ggplot() +
  geom_freqpoly(mapping = aes(x = water), bins = 100) +
  geom_freqpoly(mapping = aes(x = max_likelihood), bins = 100, color = "red")
```

#3m4

```{r}
water2 <- rbinom(1e4, size = 9, prob = samples)
```

```{r}
sum(water2 == 6) / 1e4
```

#3m5

```{r}
N <- 1e3
grid <- seq(from = 0, to = 1, length.out = N)
prior <- as.numeric(grid >= 0.5)
likelihood <- dbinom(x = 8, size = 15, prob = grid)
posterior.raw <- prior * likelihood
posterior <- posterior.raw / sum(posterior.raw)

ggplot() +
  geom_line(mapping = aes(x = grid, y = posterior * N))
```

```{r}
samples <- sample(grid, size = 1e5, replace = TRUE, prob = posterior)

ggplot() +
  geom_density(mapping = aes(x = samples))
```

```{r}
HPDI(samples, prob = 0.9)
```

```{r}
water <- rbinom(1e4, 15, prob = samples)

ggplot() +
  geom_histogram(mapping = aes(x = water), bins = 100)
```

```{r}
sum(water == 8) / 1e4
```

Comparing again with maximum-likelihood model.

```{r}
dbinom(8, 15, prob = 8/15)
```

```{r}
max_likelihood <- rbinom(1e4, size = 15, prob = 8 / 15)

ggplot() +
  geom_freqpoly(mapping = aes(x = water), bins = 100) +
  geom_freqpoly(mapping = aes(x = max_likelihood), bins = 100, color = "red")
```

```{r}
water2 <- rbinom(1e4, size = 9, prob = samples)
```

```{r}
sum(water2 == 6) / 1e4
```

# 3m6

Not sure what he is getting at! I'll have to come back to this one.

# 3h1

```{r}
data(homeworkch3)
```

```{r}
N <- 1e3
prior <- rep(1, N)
grid <- seq(from = 0, to = 1, length.out = N)
likelihood <- dbinom(x = 111, size = 200, prob = grid)
posterior.raw = prior * likelihood
posterior = posterior.raw / sum(posterior.raw)

ggplot() +
  geom_line(mapping = aes(x = grid, y = posterior * N))

```
```{r}
grid[which.max(posterior)]
```

```{r}
111/200
```

As expected! Maximum posterior probability (flat prior) matches maximum likelihood...

#3h2

```{r}
samples <- sample(grid, 1e4, replace = TRUE, prob = posterior)
```

```{r}
HPDI(samples, prob = c(0.5, 0.89, 0.97))
```

#3h3

```{r}
simulated <- rbinom(1e4, size = 200, prob = samples)
```

```{r}
ggplot() +
  geom_density(mapping = aes(x = simulated))
```
#3h4

```{r}
simulated_100 <- rbinom(1e4, size = 100, prob = samples)
```

```{r}
ggplot() +
  geom_density(mapping = aes(x = simulated_100))
```

```{r}
sum(birth1) / 100
```

Seeing at most 51 male births out of 100 is somewhat unusual but not terribly strange.

```{r}
sum(simulated_100 <= 51) / 1e4
```

#3h5

```{r}
births <- tibble(birth1 = birth1, birth2 = birth2)
```

```{r}
births %>% filter(birth1 == 0, birth2 == 1) %>% count()
```
```{r}
births %>% filter(birth1 == 0) %>% count()
```
```{r}
simulate_girl_first <- rbinom(n = 1e4, size = 49, prob = samples)
```

```{r}
ggplot() +
  geom_density(mapping = aes(x = simulate_girl_first)) +
  geom_vline(mapping = aes(xintercept = 39), color = "red")
```

```{r}
sum(simulate_girl_first >= 39) / 1e4
```

Seeing so many male second births given female first birth is very surprising given the model. Sexes of first and second birth do not appear to be independent.

### 2023-09-14 Stats Rethinking/Ch3/Dr_Parson_ch_3_problems.Rmd
Notes through 4.3.4

@@
 #3e4
+
 ```{r}
@@
 #3e5
+
 ```{r}
@@
 ```
+
 I scaled the probabilities in the posterior by N for the graph to match the vertical scale of the density below.
@@
 ```
+
 ```{r}
@@
 ```
+
 #3h4

### 2023-09-14 Stats Rethinking/Ch4/Notes.qmd
Notes through 4.3.4

---
title: "Notes"
format: html
---

# Import Libraries

```{r}
library(rethinking)
library(tidyverse)
```

# 4.1 - Why normal dists are normal

```{r}
#Simulate flipping a coin 16 times and walking depending on the outcome

pos <- replicate(1000, sum(runif(16,-1,1)))

hist(pos)
```

## 4.1.2 Normality by multiplication

```{r}
growth <- replicate(10000, prod(1+runif(12,0,0.1)))

dens(growth, norm.comp=TRUE)
```

## 4.1.3 Normal by log-multiplication

```{r}
log.big <- replicate(10000, log(prod(1+runif(12,0,0.5))))
dens(log.big)
```

# 4.2 - A language for describing models

1.  We recognize a set of variables to work with. Some of these are observable, and are called *data.* Others are unobservable things like rates and averages, which are called *parameters*.
2.  We define each variable either in terms of the other variables or in terms of a probability distribution.
3.  The combination of variables and their probability distributions defines a *joint generative model* that can be used both to simulate hypothetical observations as well as analyze real ones.

W \~ Binomial(*N,p*) - Read as, "The count *W* is distributed binomially with sample size *N* and probability *p*.

p \~ Uniform(0,1) - Read as, "The prior for *p* is assumed to be uniform between zero and one.

Both of the above lines in the model are stochastic, as indicated by the \~. A stochastic relationship is just a mapping of a variable or parameter onto a distribution. It is stochastic because no single instance of he variable on the left is known with certainty.

# 4.3 - Gaussian model of height

```{r}
data(Howell1)
d <- Howell1
str(d)
#height - cm's
#weight - kg's
#age - years
#gender - 1=Male, 0=Female
```

```{r}
precis(d)
```

```{r}
#Book only wants to work with adults
d2 <- d[d$age > 18,]
```

## 4.3.2 The model

```{r}
dens(d2$height)
```

Height is approximately normal so we can describe it as: $h_i \sim$ Normal($\mu,\sigma$).

We need priors which the book gives us as: $\mu\sim Normal(178,20)$ and $\sigma\sim Uniform(0,50)$.

```{r}
#Plot the priors
curve(dnorm(x, 178, 20), from=100, to=250)
```

```{r}
curve(dunif(x,0,50), from=-10, to=60)
```

```{r}
#Prior Predictive simulation
sample_mu <- rnorm(1e4, 178, 20)
sample_sigma <- runif(1e4, 0, 50)
prior_h <- rnorm(1e4, sample_mu, sample_sigma)
dens(prior_h)
```

## 4.3.3 Grid approximation of the posterior

```{r}
#Grid approximation of the posterior
mu.list <- seq(from=150, to=160, length.out=100)
sigma.list <- seq(from=7, to=9, length.out=100)
post <- expand.grid(mu=mu.list, sigma=sigma.list)
post$LL <- sapply(1:nrow(post), function(i) sum(
  dnorm(d2$height, post$mu[i], post$sigma[i], log=TRUE)))
post$prod <- post$LL + dnorm(post$mu, 178, 20, TRUE) + dunif(post$sigma, 0, 50, TRUE)
post$prob <- exp(post$prod - max(post$prod))
image_xyz(post$mu, post$sigma, post$prob)
```

## 4.3.4 Sampling from the posterior

### 2023-09-14 Stats Rethinking/Ch4/Practice Problems.qmd
Completed Easy Practice Problems

---
title: "Practice Problems"
format: html
---

# Load Libraries

```{r}
library(rethinking)
library(tidyverse)
```

# 4E1

In the model definition, which is the likelihood?

$$
y_i\sim Normal(\mu,\sigma)\\
\mu\sim Normal(0,10)\\
\sigma\sim Exponential(1)
$$

The likelihood is, $y_i\sim Normal(\mu,\sigma)$

# 4E2

In the model definition above, how many parameters are in the posterior distribution?

There are two parameters, $\mu$ and $\sigma$.

# 4E3

Using the model definition above, write down the appropriate form of Bayes' theorem that includes the proper likelihood and priors

$$
Pr(\mu,\sigma|y_i)=\frac{\prod_i Normal(y_i|\mu,\sigma)Normal(\mu|0,10)Exponential(\sigma|1)}{\int\int\prod_i Normal(y_i|\mu,\sigma)Normal(\mu|0,10)Exponential(\sigma|1)d\mu d\sigma}
$$

(Used formatting of Bayes on page 84)

# 4E4

In the model definition, which line is the linear model?

$$
y_i\sim Normal(\mu, \sigma)\\
\mu_i=\alpha+\beta x_i\\
\alpha\sim Normal(0,10)\\
\beta\sim Normal(0,1)\\
\sigma\sim Exponential(2)
$$

In this model, the $\mu_i=\alpha+\beta x_i$ is the linear model.

# 4E5

In the model definition for 4E4, how many parameters are in the posterior distribution?

There are 3 parameters in the posterior dist.

# 4M1

# 4M2

# 4M3

# 4M4

# 4M5

# 4M6

# 4M7

# 4M8

# 4H1

# 4H2

# 4H3

# 4H4

# 4H5

# 4H6

# 4H7

# 4H8

### 2023-09-14 Stats Rethinking/Ch4/Dr_Parson_ch_4_problems.qmd
Create Dr_Parson_ch_4_problems.qmd
This is work in progress! I will continue to develop it before we meet on 9/15.

---
title: "Chapter 4 practice"
format: html
editor: visual
---

```{r}
library(tidyverse)
```

# E problems

## 4E1

The likelihood is y_i \~ Normal(mu, sigma)

## 4E2

There are two parameters in the posterior distribution, mu and sigma.

## 4E3

Bayes' theorem gives us the following formula for the posterior joint density of $\mu$ and $\sigma$.

$$
Pr(\mu, \sigma | y_1, \dots, y_n) = \frac{\prod_{i = 1}^n Pr(y_i | \mu, \sigma) \operatorname{Normal}(\mu | 0, 10) \operatorname{Exponential}(\sigma | 1)}
{\int\int \prod_{i = 1}^n Pr(y_i | \mu, \sigma) \operatorname{Normal}(\mu | 0, 10)\operatorname{Exponential}(\sigma | 1) d\mu d\sigma}
$$

## 4E4

The linear model is $\mu_i = \alpha + \beta x_i$.

## 4E5

There are three parameters in the posterior distribution.

# M problems

## 4M1

We have the following model:

y_i \~ Normal(mu, sigma)

mu \~ Normal(0, 10)

sigma \~ Exponential(1)

To simulate data from the prior distribution, we first draw 10,000 samples (say) for mu and sigma from the given prior distributions and then draw 10,000 samples (using our mu and sigma samples) from the normal distribution.

```{r}
N <- 1e4
mus <- rnorm(N, mean = 0, sd = 10)
sigmas <- rexp(N, rate = 1)
ys <- rnorm(N, mean = mus, sd = sigmas)
```

We can make a density graph of the simulated y values. For comparison, I have overlaid a normal distribution with the same mean and standard deviation as the list of simulated y values. It looks pretty close! (Checking whether the prior distribution of the y values is actually a normal distribution requires some probability theory. Maybe something to do at the end of MATH 351...)

```{r}
ggplot() +
  geom_density(mapping = aes(x = ys)) +
  geom_function(fun = dnorm, args = list(mean = mean(ys), sd = sd(ys)), color = "red")
```

## 4M2

In `quap` notation, the model looks like this:

```         
y ~ dnorm(mu, sigma),
mu ~ dnorm(0, 10),
sigma ~ dexp(1)
```

## 4M3

The following mathematical model definition matches the given `quap` formula:

y_i \~ Normal(mu, sigma)

mu_i = a + b x_i

a \~ Normal(0, 10)

b \~ Uniform(0, 1)

sigma \~ Exp(1)

## 4M4

| Student | year | height |
|---------|------|--------|
| 1       | 0    | 60     |
| 1       | 1    | 61     |
| 1       | 2    | 62     |
| 2       | 0    | 65     |
| 2       | 1    | 67     |
| 2       | 2    | 69     |
| 3       | 0    | 55     |
| 3       | 1    | 60     |
| 3       | 2    | 65     |

In our model, we ignore the "student" column. We model the height column using the year column as predictor. The subscript i indicates the row in the data table. The parameter a is something like the average height in year 0. The parameter b is something like the typical increase in height per year.

height_i \~ Normal(mu_i, sigma)

m_i = a + b year_i

sigma \~ Exp(2)

a \~ Normal(5, 0.5)

b \~ Exp(6)

Likelihood: heights for the students could be normally distributed or a mixture of normal distributions.

Prior on sigma: I am imagining students in middle school (6th, 7th, 8th). Typical differences in height might be 6 inches (1/2 foot). We'll use an exponential distribution.

Prior on a: This would be something like average height among 6th graders. Maybe 5 feet give or take 6 inches to be safe.

Prior on b: This would be something like average height increase per year. This parameter could not be negative! Maybe 2 inches (1/6 foot) would be a reasonable guess for middle school. We'll use an exponential distribution to guarantee positive values. But we could try a log-normal distribution instead.

A richer (multilevel) model would allow for different slopes for each student with pooling. Something like this might work. (But I haven't thought about this at all!)

height_ij \~ Normal(mu_ij, sigma)

mu_ij = a_i + b_i year_j

sigma \~ Exp(2)

a_i \~ Normal(5, 0.5)

b_i \~ Exp(6)

```{r}
N <- 1e4
sigmas <- rexp(N, 2)
as <- rnorm(N, 5, 0.5)
bs <- rexp(N, 6)

student_simulation <- tibble(year = rep(c(0,1,2), each = N),
                             height = )

heights <- rnorm(N, mean = as + bs * , sigma = sigmas)
```

```{r}
ggplot() +
  geom_density()
```

### 2023-09-14 Stats Rethinking/Ch4/Notes.qmd

@@
 ## 4.3.4 Sampling from the posterior
+
+```{r}
+sample.rows <- sample(1:nrow(post), size=1e4, replace=TRUE, prob=post$prob)
+sample.mu <- post$mu[sample.rows]
+sample.sigma <- post$sigma[sample.rows]
+plot(sample.mu, sample.sigma, cex=0.5, pch=16, col=col.alpha(rangi2,0.1))
+```
+
+```{r}
+#Marginal posterior density of mu
+dens(sample.mu)
+```
+
+```{r}
+#Marginal posterior density of sigma
+dens(sample.sigma)
+```
+
+```{r}
+PI(sample.mu)
+PI(sample.sigma)
+```
+
+## 4.3.5 Finding the posterior distribution with `quap`
+
+```{r}
+flist <- alist(
+  height ~ dnorm(mu, sigma),
+  mu ~ dnorm(178, 20),
+  sigma ~ dunif(0,50)
+)
+
+m4.1 <- quap(flist, data=d2)
+
+precis(m4.1)
+```
+
+These numbers provide Gaussian approximations for each parameter's marginal distributions.
+This means the plausibility of each value of $\mu$ after averaging over the plausibilities of each value of $\sigma$, is given by a Gaussian distribution with mean 154.6 and standard deviation of 0.4.
+
+We can change our prior of $\mu$ to be much more concentrated around our mean, signalling that it is much stronger.
+
+```{r}
+m4.2 <- quap(
+  alist(
+    height ~ dnorm(mu, sigma),
+    mu ~ dnorm(178, 0.1),
+    sigma ~ dunif(0,50)
+  ), data = d2
+)
+
+precis(m4.2)
+```
+
+Our estimate for $\mu$ has hardly changed off the prior now because the prior was so much more concentrated.
+However, our estimate for $\sigma$ has changed a lot even though we didn't change its prior.
+This is because once the model (a metaphorical golem) is certain the mean is near 178 - as the prior insists - then the model has to estimate a $\sigma$ conditional on that fact.
+Which results in a different posterior for $\sigma$.
+
+## 4.3.6 Sampling from `quap`
+
+```{r}
+diag(vcov(m4.1))
+cov2cor(vcov(m4.1))
+```
+
+The two element vector (listed first) is the list of variances.
+Taking the square root of the vector will give the standard deviations shown in the `precis` output of `m4.1`.
+While the two-by-two matrix represents the correlation (bounded by +1 and -1) between the pairs of parameters.
+The estimates near 0 for both tell us that learning $\mu$ tells us nothing about $\sigma$ and vice versa.
+
+```{r}
+#Sampling from multidimensional posterior
+post <- extract.samples(m4.1, n=1e4)
+precis(post)
+```
+
+# 4.4 - Linear prediction
+
+```{r}
+plot(d2$height, d2$weight)
+```
+
+## 4.4.1.3 *Priors*
+
+```{r}
+set.seed(2971)
+N <- 100
+a <- rnorm(N, 178, 20)
+b <- rnorm(N, 0, 10)
+```
+
+```{r}
+plot(NULL, xlim=range(d2$weight), ylim=c(-100,400), xlab="weight", ylab="height")
+abline(h=0, lty=2)
+abline(h=272, lty=1, lwd=0.5)
+mtext("b ~ dnorm(0,10)")
+xbar <- mean(d2$weight)
+for(i in 1:N) curve(a[i] + b[i]*(x-xbar), from=min(d2$weight), to=max(d2$weight), add=TRUE, col=col.alpha("black", 0.2))
+```
+
+```{r}
+#Experimenting with b being log-normal
+b <- rlnorm(1e4, 0, 1)
+dens(b, xlim=c(0,5), adj=0.1)
+```
+
+```{r}
+set.seed(2971)
+N <- 100
+a <- rnorm(N, 178, 20)
+b <- rlnorm(N, 0, 1)
+
+plot(NULL, xlim=range(d2$weight), ylim=c(-100,400), xlab="weight", ylab="height")
+abline(h=0, lty=2)
+abline(h=272, lty=1, lwd=0.5)
+mtext("log(b) ~ dnorm(0,1)")
+xbar <- mean(d2$weight)
+for(i in 1:N) curve(a[i] + b[i]*(x-xbar), from=min(d2$weight), to=max(d2$weight), add=TRUE, col=col.alpha("black", 0.2))
+```
+
+## 4.4.2 Finding the posterior distribution
+
+```{r}
+#Load data again
+data("Howell1"); d <- Howell1; d2 <- d[d$age > 18,]
+
+#defin average weight
+xbar <- mean(d2$weight)
+
+#fit model
+m4.3 <- quap(
+  alist(
+    height ~ dnorm(mu, sigma),
+    mu <- a + b*(weight - xbar),
+    a ~ dnorm(178,20),
+    b ~ dlnorm(0,1),
+    sigma ~ dunif(0,50)
+  ), data=d2
+)
+```
+
+### 4.4.3.1 Tables of marginal distributions
+
+```{r}
+precis(m4.3)
+```
+
+The first row gives us the quadratic approximation for $\alpha$, the second row $\beta$, and the third row $\sigma$.
+
+```{r}
+round(vcov(m4.3),3)
+```
+
+### 4.4.3.2 Plotting posterior inference against the data
+
+```{r}
+plot(height ~ weight, data=d2, col=rangi2)
+post <- extract.samples(m4.3)
+a_map <- mean(post$a)
+b_map <- mean(post$b)
+curve(a_map + b_map*(x-xbar), add=TRUE)
+```
+
+### 4.4.3.4 Plotting regression intervals and contours
+
+General recipe for generating predictions and intervals from the posterior of a fit model:
+
+1.
+Use `link` to generate distributions of posterior values for $\mu$.
+The default behavior of `link` is to use the original data, so you have to pass it a list of new horizontal axis values you want to plot posterior predictions across.
+2.  Use summary functions like `mean` or `PI` to find averages and lower and upper bounds of $\mu$ for each value of the predictor variable
+3.  Finally, use plotting functions like `lines` and shade to draw the lines and intervals. Or plot the distributions of predictions, or do further numerical calculations with them.
+
+# 4.5 Curves from lines
+
+## 4.5.1 Polynomial regression
+
+```{r}
+plot(height ~ weight, d)
+```
+
+```{r}
+#Very different than in the book
+data("Howell1")
+d <- Howell1
+
+d$weight_s <- (d$weight - mean(d$weight))/sd(d$weight)
+d$weight_s2 <- d$weight^2
+
+m4.5 <- quap(
+  alist(
+    height ~ dnorm(mu, sigma),
+    mu <- a + b1*weight_s + b2*weight_s2,
+    a ~ dnorm(178, 20),
+    b1 <- dlnorm(0,1),
+    b2 ~ dnorm(0,1),
+    sigma ~ dunif(0,50)
+  ), data=d
+)
+precis(m4.5)
+```
+
+```{r}
+weight.seq <- seq(from=-2.2, to=2, length.out=30)
+pred_dat <- list(weight_s=weight.seq, weight_s2=weight.seq^2)
+mu <- link(m4.5, data=pred_dat)
+mu.mean <- apply(mu, 2, mean)
+mu.PI <- apply(mu, 2, PI, prob=0.89)
+sim.height <- sim(m4.5, data=pred_dat)
+height.PI <- apply(sim.height, 2, PI, prob=0.89)
+```
+
+```{r}
+#Where is the line?
+plot(height ~ weight, d, col=col.alpha(rangi2, 0.5))
+lines(weight.seq, mu.mean)
+shade(mu.PI, weight.seq)
+shade(height.PI, weight.seq)
+```

### 2023-09-14 Stats Rethinking/Ch4/Practice Problems.qmd
Attempted all easy and medium problems from chpt 4

@@
 
+For the model definition below, simulate observed $y$ values from the prior
+
+$$
+y_i\sim Normal(\mu,\sigma)\\
+\mu\sim Normal(0,10)\\
+\sigma\sim Exponential(1)
+$$
+
+```{r}
+mu <- rnorm(1e4, 0, 10)
+sigma <- rexp(1e4)
+prior_y <- rnorm(1e4, mu, sigma)
+dens(prior_y)
+```
+
 # 4M2
+
+Translate the above model into a `quap` formula
+
+```{r}
+#Should there be a data arguement provided?
+quap(
+  alist(
+    y <- dnorm(mu, sigma),
+    mu <- dnorm(0,10),
+    sigma <- dexp(1)
+  )
+)
+```
 
@@
 
+Translate the `quap` model below into a mathematical model definition
+
+```{r}
+#| eval: false
+
+y~dnorm(mu,sigma),
+mu <- a + b*x,
+a ~ dnorm(0,10),
+b ~ dunif(0,1),
+sigma ~ dexp(1)
+```
+
+$$
+y_i\sim Normal(\mu,\sigma)\\
+\mu=\alpha+\beta x\\
+\alpha\sim Normal(0,10)\\
+\beta\sim Uniform(0,1)\\
+\sigma\sim Exponential(1)
+$$
+
 # 4M4
+
+A sample of students is measured for height each year for 3 years.
+After the third year, you want to fit a linear regression predicting height using year as a predictor.
+Write down the mathematical model definition for this regression, using any variable names and priors you choose.
+Be prepared to defend your choice of priors.
+
+$$
+Height_i\sim Normal(\mu,\sigma)\\
+\mu=\alpha+\beta x\\
+\alpha\sim Normal(48,4)\\
+\beta\sim Exponential(1)\\
+\sigma\sim Uniform(0,12)
+$$
+
+What my model says is that height, accross all 3 years is normally distributed.
+However, the mean height, $\mu$, is dependent the average height, $\alpha$, and the growth each year by the population, $\beta$.
+The average height is normally distributed centered at 48in with a standard deviation of 4in.
+Our growth change in the population is exponential because we assume kids grow at non-linear rates which allows us to catch this.
+However, I don't know how exponential functions work in this context so I just put a 1.
+The standard deviation is uniformly distributed from the lower bound 0, to the upper bound 12, representing that we think \~95% of the population lies between 12 inches of the average, 48 inches.
 
@@
 
+Now, suppose I remind you that every student got taller each year. Does this information lead you to change your choice of priors? How?
+
+I feel like my priors should stay the same. I already factored in students growing with the exponential function.
+
 # 4M6
+
+Now suppose I tell you that the variance among heights for students of the same age is never more than 64cm. How does this lead you to revise your priors?
+
+64 inches is about 2 feet, which I think my $\sigma$ covers.
 
@@
 
+Refit the model `m4.3` from the chapter, but omit the mean weight `xbar` this time.
+Compare the new model's posterior to that of the original model.
+In particular, look at the covariance among the parameters.
+What is different?
+Then compare the posterior predictions of both models.
+
+```{r}
+#Replicate m4.3
+#Load data again
+data("Howell1"); d <- Howell1; d2 <- d[d$age > 18,]
+
+#defin average weight
+xbar <- mean(d2$weight)
+
+#fit model
+m4.3 <- quap(
+  alist(
+    height ~ dnorm(mu, sigma),
+    mu <- a + b*(weight - xbar),
+    a ~ dnorm(178,20),
+    b ~ dlnorm(0,1),
+    sigma ~ dunif(0,50)
+  ), data=d2
+)
+
+precis(m4.3)
+round(vcov(m4.3),3)
+```
+
+```{r}
+#Now replicate without xbar
+
+#fit model
+m4.3.nobar <- quap(
+  alist(
+    height ~ dnorm(mu, sigma),
+    mu <- a + b*(weight),
+    a ~ dnorm(178,20),
+    b ~ dlnorm(0,1),
+    sigma ~ dunif(0,50)
+  ), data=d2
+)
+
+precis(m4.3.nobar)
+round(vcov(m4.3.nobar),3)
+```
+
+Immediately noticeable is that our means changed.
+It changed by about 39, which is close to our value of `xbar~45`.
+It did not, however, change our measurement of `b` or `sigma` almost at all.
+Interestingly though, in our correlation matrix, our `b` and `a` are much more related.
+I don't know how to interpret that though.
+
 # 4M8
+
+In the chapter, we used 15 knots with the cherry blossom spline.
+Increase the number of knots and observe what happens to the resulting spline.
+Then adjust also the width of the prior on the weights-change the standard deviation of the prior and watch what happens.
+What do you think the combination of knot number and the prior on the weights controls?
+
+This is where I got lost so bear with me.
+
+```{r}
+library(splines)
+```
+
+```{r}
+#Plot the 15 knot splines
+data("cherry_blossoms")
+dc <- cherry_blossoms
+
+dc2 <- dc[complete.cases(dc$doy),]
+num_knots <- 15
+knot_list <- quantile(dc2$year, probs=seq(0,1,length.out=num_knots))
+
+B <- bs(dc2$year,
+        knots = knot_list[-c(1,num_knots)],
+        degree=3, intercept=TRUE)
+
+plot(NULL, xlim=range(dc2$year), ylim=c(0,1), xlab="year", ylab="basis")
+for (i in 1:ncol(B)) lines(dc2$year, B[,i])
+```
+
+```{r}
+#Change knots to 30
+num_knots <- 30
+knot_list <- quantile(dc2$year, probs=seq(0,1,length.out=num_knots))
+
+B <- bs(dc2$year,
+        knots = knot_list[-c(1,num_knots)],
+        degree=3, intercept=TRUE)
+
+plot(NULL, xlim=range(dc2$year), ylim=c(0,1), xlab="year", ylab="basis")
+for (i in 1:ncol(B)) lines(dc2$year, B[,i])
+```
+
+Adding more knots makes it wiggle more?
+
+```{r}
+m4M8 <- quap(
+  alist(
+    D ~ dnorm(mu, sigma),
+    mu <- a + B %*% w,
+    a ~ dnorm(100,10),
+    w ~ dnorm(0,10),
+    sigma ~ dexp(1)
+  ), data=list(dc2$doy, B=B),
+  start = list(w=rep(0,ncol(B)))
+)
+```
+
+Don't know why it's throwing this error
 

### 2023-09-15 Stats Rethinking/Ch4/Dr_Parson_ch_4_problems.qmd
Add more problems
That's all for tonight! I can finish these up later.

@@
 ```{r}
-N <- 1e4
+N <- 1e1
 sigmas <- rexp(N, 2)
@@
 
-student_simulation <- tibble(year = rep(c(0,1,2), each = N),
-                             height = )
-
-heights <- rnorm(N, mean = as + bs * , sigma = sigmas)
+student_simulation <- tibble(replication = rep(1:N, times = 3),
+                             year = rep(c(0,1,2), each = N),
+                             a = rep(as, times = 3),
+                             b = rep(bs, times = 3),
+                             sigma = rep(sigmas, times = 3),
+                             height = rnorm(3*N, mean = a + b*year, sd = sigma))
 ```
@@
 ```{r}
+ggplot(data = student_simulation) +
+  geom_point(mapping = aes(x = factor(year), y = height)) +
+  geom_line(mapping = aes(x = factor(year), y = height, group = replication))
+```
+
+Huh! Sometimes students shrink, according to this model. To make this work better, one might consider a time-series model instead of a generic regression model with time as a predictor.
+
+## 4M5
+
+The model above already includes a prior for the slope that does not allow negative slopes.
+The problem with the model is that it does not remember how tall students are from year to year.
+The random component of the prediction (likelihood) is intended to reflect between-student variability in heights.
+But that leads to occasional shrinking when we track one person.
+
+I guess that's not how we should use the model.
+We cannot model the growth of a singular person with this set-up.
+What we are modeling is the heights of sixth graders, seventh graders, and eighth graders as groups.
+
+The following chunks simulate 100 sixth graders, 100 seventh graders, and 100 eighth graders. We can see that the prior is not great, since it has too much variability.
+
+```{r}
+N <- 1e2
+sigmas <- rexp(3*N, 2)
+as <- rnorm(3*N, 5, 0.5)
+bs <- rexp(3*N, 6)
+
+student_simulation2 <- tibble(year = rep(c(0,1,2), each = N),
+                             a = as,
+                             b = bs,
+                             sigma = sigmas,
+                             height = rnorm(3*N, mean = a + b*year, sd = sigma))
+```
+
+```{r}
+ggplot(data = student_simulation2) +
+  geom_jitter(mapping = aes(x = factor(year), y = height), height = 0, width = 0.1, alpha = 0.2)
+```
+
+The confusing part of the problem statement is the tracking of heights over three years. It might be clearer framed as data about students in a middle school.
+
+## 4M6
+
+The word "variance" is a strange choice!
+I think what McElreath has in mind, though, is that we could limit sigma above in some way by 64 cm.
+I'm not sure exactly what "variance" means here.
+Maybe we could use sigma \~ Uniform(0, 64 / (2.54 \* 12)).
+
+## 4M7
+
+```{r}
+library(rethinking)
+data(Howell1)
+```
+
+```{r}
+d2 <- Howell1 %>% filter(age >= 18)
+
+xbar <- d2 %>% summarize(m = mean(weight)) %>% pull(m)
+```
+
+```{r}
+m4.3 <- quap(
+  alist(
+    height ~ dnorm( mu , sigma ) ,
+    mu <- a + b*( weight - xbar ) ,
+    a ~ dnorm( 178 , 20 ) ,
+    b ~ dlnorm( 0 , 1 ) ,
+    sigma ~ dunif( 0 , 50 )
+    ) , data=d2 )
+
+precis(m4.3)
+```
+
+```{r}
+post <- extract.samples(m4.3)
+
+post %>% pull(a) %>% quantile(c(0.05, 0.95))
+```
+
+```{r}
+PI(post %>% pull(a), prob = 0.9)
+```
+
+```{r}
+m4.3v2 <- quap(
+  alist(
+    height ~ dnorm( mu , sigma ) ,
+    mu <- a + b * weight ,
+    a ~ dnorm( 178 , 20 ) ,
+    b ~ dlnorm( 0 , 1 ) ,
+    sigma ~ dunif( 0 , 50 )
+    ) , data=d2 )
+```
+
+```{r}
+post2 <- extract.samples(m4.3v2)
+vcov(m4.3v2)
+```
+
+```{r}
+sim.height <- sim(m4.3, data = list(weight = 50))
 ggplot() + 
-  geom_density()
+  geom_density(mapping = aes(x = sim.height))
 ```
+
+```{r}
+quantile(sim.height, c(0.025, 0.975))
+```
+
+```{r}
+sim.heightv2 <- sim(m4.3v2, data = list(weight = 50))
+ggplot() + 
+  geom_density(mapping = aes(x = sim.heightv2))
+```
+
+```{r}
+quantile(sim.heightv2, c(0.025, 0.975))
+```
+
+## 4M8
+
+# H problems
+
+## 4H1
+
+```{r}
+newdata <- c(46.95, 43.72, 64.78, 32.59, 54.63)
+sim.height.new <- sim(m4.3v2, data = list(weight = newdata))
+
+for(i in 1:5){
+  print(round(mean(sim.height.new[, i]), 1))
+  print(round(PI(sim.height.new[,i]), 1))
+}
+
+```

### 2023-09-17 Stats Rethinking/Ch4/Dr_Parson_ch_4_problems.qmd
Finish Chapter 4 problems
I finished the chapter 4 problem, including all the splines questions.

@@
 
+Refitting model changes `a`, as one would expect, but the model predictions are the same.
+
+*Regression and Other Stories* (<https://avehtari.github.io/ROS-Examples/>) suggests centering predictors so that one can assign meaningful priors.
+
 ```{r}
@@
     a ~ dnorm( 178 , 20 ) ,
-    b ~ dlnorm( 0 , 1 ) ,
+    b ~ dnorm( 1 , 2 ) ,
     sigma ~ dunif( 0 , 50 )
@@
 
+```{r}
+m4.3v3 <- quap(
+  alist(
+    height ~ dnorm( mu , sigma ) ,
+    mu <- a + b*( weight - xbar ) + c * male ,
+    mu2 <- a + b*(50 - xbar),
+    a ~ dnorm( 178 , 20 ) ,
+    b ~ dlnorm( 0 , 1 ) ,
+    c ~ dnorm( 0, 1 ),
+    sigma ~ dunif( 0 , 50 )
+    ) , data=d2 )
+
+post <- extract.samples(m4.3v3)
+
+post %>% pull(c) %>% quantile(c(0.05, 0.95))
+```
+
 ## 4M8
+
+I'll have to come back to this one later! I am feeling too lazy to study the syntax for fitting spline models at the moment.
+
+For general information on splines (and polynomial fits) see Chapter 7 of *Statistical Learning* (<https://www.statlearning.com>).
+
+```{r}
+data("cherry_blossoms")
+
+ggplot(data = cherry_blossoms) +
+  geom_point(mapping = aes(x = year, y = doy))
+```
+
+```{r}
+cherry_full <- cherry_blossoms %>% drop_na(doy)
+num_knots <- 15
+knot_list <- quantile( cherry_full %>% pull(year) , probs=seq(0, 1, length.out=num_knots) )
+```
+
+```{r}
+library(splines)
+B <- bs(cherry_full %>% pull(year),
+        knots = knot_list[-c(1,num_knots)], degree=3 , intercept=TRUE )
+cherry_full <- cherry_blossoms %>% drop_na(doy)
+num_knots <- 15
+knot_list <- quantile( cherry_full %>% pull(year) , probs=seq(0, 1, length.out=num_knots) )
+```
+
+The following chunk is copied from the textbook.
+
+```{r}
+plot( NULL , xlim=range(cherry_full$year) , ylim=c(0,1) , xlab="year" , ylab="basis" )
+for ( i in 1:ncol(B) ) lines( cherry_full$year , B[,i] )
+```
+
+```{r}
+m4.7 <- quap(
+  alist(
+    D ~ dnorm( mu , sigma ) ,
+    mu <- a + B %*% w ,
+    a ~ dnorm(100,10),
+    w ~ dnorm(0,100),
+    sigma ~ dexp(1)
+    ),
+  data=list( D= cherry_full %>% pull(doy), B = B ) ,
+  start=list( w=rep( 0 , ncol(B) ) ) )
+```
+
+```{r}
+post <- extract.samples( m4.7 )
+w <- apply( post$w , 2 , mean )
+plot( NULL , xlim = range(cherry_full$year) ,
+      ylim=c(-6,6) , xlab="year" , ylab="basis * weight" )
+for ( i in 1:ncol(B) ) lines( cherry_full$year , w[i]*B[,i] )
+```
+
+```{r}
+mu <- link( m4.7 )
+mu_PI <- apply(mu,2,PI,0.97)
+plot( cherry_full$year , cherry_full$doy , col=col.alpha(rangi2, 0.3) , pch=16, 
+      shade( mu_PI , cherry_full$year , col=col.alpha("black",0.5) ))
+```
+
+With 100 knots, the model is very flexible and matches data fluctuations on small scales.
+
+With 5 knots, the model is not flexible at all and only reflects long-term trends. (I like 5 knots!)
+
+Increasing the standard deviation on the prior for the weights (say from 10 to 100) also increases the flexibility of the model but not as dramatically as increasing the number of knots.
 
@@
 ## 4H1
+
+There should be a better way to organize displaying this work! If only the commands were made to give tidy outputs...
 
@@
 }
+```
 
+This question is somewhat silly. Who measures the weight of an adult to within 10 grams? Why bother when the resulting prediction intervals for height are so wide?
+
+## 4H2
+
+### Part a
+
+We should do all the manipulations using `tidyverse`! I recommend barring `[ ] $`.
+
+```{r}
+howell_young <- Howell1 %>% filter(age < 18)
 ```
+
+```{r}
+young_fit <- quap(
+  alist(height ~ dnorm(mu, sigma),
+        mu <- a + b * (weight - mean(weight)),
+        a ~ dnorm(100, 20),
+        b ~ dlnorm(1, 1),
+        sigma ~ dunif(0, 50)
+        ), data = howell_young
+)
+```
+
+```{r}
+precis(young_fit)
+```
+
+For comparison, we can fit a traditional `lm` model.
+The coefficients and standard errors are essentially the same!
+Since we have so much data, the priors do not much matter.
+(And the `quap` approximation also works like `lm`.)
+
+```{r}
+howell_young <- howell_young %>% mutate(c.weight = weight - mean(weight))
+lm(height ~ c.weight, data = howell_young) %>% summary
+```
+
+A 10 kg difference in weight is associated with a 27 cm difference in height.
+
+### Part b
+
+```{r}
+xbar <- howell_young %>% summarize(m = mean(weight)) %>% pull(m)
+```
+
+```{r}
+samples <- extract.samples(young_fit)
+N <- 20
+xs <- seq(from = 4, to = 45, length.out = N) - xbar
+mu_min <- rep(0, times = N)
+mu_max <- rep(0, times = N)
+pred_min <- rep(0, times = N)
+pred_max <- rep(0, times = N)
+
+for (i in 1:20){
+  q <- samples %>% mutate(mu = a + b * xs[i],
+                     pred = rnorm(nrow(samples), mu, sigma)) %>%
+    summarize(q_mu_low = quantile(mu, 0.055),
+              q_mu_high = quantile(mu, 0.945),
+              q_pred_low = quantile(pred, 0.055),
+              q_pred_high = quantile(pred, 0.945))
+  mu_min[i] <- q %>% pull(q_mu_low)
+  mu_max[i] <- q %>% pull(q_mu_high)
+  pred_min[i] <- q %>% pull(q_pred_low)
+  pred_max[i] <- q %>% pull(q_pred_high)
+}
+
+strips <- tibble(mu_min = mu_min, mu_max = mu_max,
+                 pred_min = pred_min, pred_max = pred_max)
+```
+
+```{r}
+ggplot(data = howell_young) +
+  geom_ribbon(data = strips,
+              mapping = aes(x = xs + xbar, ymin = mu_min, ymax = mu_max),
+              alpha = 0.2) +
+  geom_ribbon(data = strips,
+              mapping = aes(x = xs + xbar, ymin = pred_min, ymax = pred_max),
+              alpha = 0.3) +
+  geom_point(mapping = aes(x = weight, y = height), shape = 1, color = "blue", alpha = 0.5) +
+  geom_function(fun = function(x){108.31 + 2.72 * (x - xbar)}) +
+  labs(x = "weight (kg)", y = "height (cm)") +
+  theme_bw()
+```
+
+This model does not fit the data well.
+The scatterplot shows clear non-linearity.
+Many observations at the low and high ends of weight fall outside of the prediction interval.
+The conditional means clearly do not fall in the (very narrow) interval for the means.
+
+### Part c
+
+We need to use a non-linear model!
+
+## 4H3
+
+I did not know what to do for the prior on this one.
+My first choice was horrible, and the posterior model did not fit the data well.
+I checked the model from `lm` (matching a "non-informative prior") and adjusted my prior to be much more flexible.
+This model now essentially reproduces `lm`.
+
+```{r}
+log_fit <- quap(
+  alist(height ~ dnorm(mu, sigma),
+        mu <- a + b * log(weight),
+        a ~ dnorm(0, 20),
+        b ~ dnorm(0, 40),
+        sigma ~ dunif(0, 50)),
+  data = Howell1
+)
+
+precis(log_fit)
+```
+
+```{r}
+ggplot(data = Howell1) +
+  geom_point(mapping = aes(x = weight, y = height)) +
+  geom_function(fun = function(x)(-23.70 + 47.05 * log(x)))
+```
+
+```{r}
+lm(height ~ log(weight), data = Howell1) %>% summary
+```
+
+The following code produces the 97% interval for the mean and the 97% interval for the predicted heights.
+
+```{r}
+samples <- extract.samples(log_fit)
+N <- 20
+xs <- seq(from = 4, to = 63, length.out = N)
+mu_min <- rep(0, times = N)
+mu_max <- rep(0, times = N)
+pred_min <- rep(0, times = N)
+pred_max <- rep(0, times = N)
+
+for (i in 1:20){
+  q <- samples %>% mutate(mu = a + b * log(xs[i]),
+                     pred = rnorm(nrow(samples), mu, sigma)) %>%
+    summarize(q_mu_low = quantile(mu, 0.015),
+              q_mu_high = quantile(mu, 0.985),
+              q_pred_low = quantile(pred, 0.015),
+              q_pred_high = quantile(pred, 0.985))
+  mu_min[i] <- q %>% pull(q_mu_low)
+  mu_max[i] <- q %>% pull(q_mu_high)
+  pred_min[i] <- q %>% pull(q_pred_low)
+  pred_max[i] <- q %>% pull(q_pred_high)
+}
+
+strips <- tibble(mu_min = mu_min, mu_max = mu_max,
+                 pred_min = pred_min, pred_max = pred_max)
+```
+
+```{r}
+ggplot(data = Howell1) +
+  geom_ribbon(data = strips,
+              mapping = aes(x = xs, ymin = mu_min, ymax = mu_max),
+              alpha = 0.4) +
+  geom_ribbon(data = strips,
+              mapping = aes(x = xs, ymin = pred_min, ymax = pred_max),
+              alpha = 0.3) +
+  geom_point(mapping = aes(x = weight, y = height), shape = 1, color = "blue", alpha = 0.5) +
+  geom_function(fun = function(x)(-23.70 + 47.05 * log(x))) +
+  labs(x = "weight (kg)", y = "height (cm)") +
+  theme_bw()
+```
+
+This model looks much better, though the prediction intervals seem too wide for low weights and to narrow for high weights. Perhaps `sigma` could vary with the weight...
+
+## 4H4
+
+The quadratic model has the following likelihood:
+
+height \~ Normal(mu, sigma)
+
+mu = a + b_1 weight_s + b_2 weight_s\^2,
+
+where weight_s is standardized (z-score) weight.
+
+The following chunk simulates data for various choices of prior.
+
+```{r}
+N = nrow(Howell1)
+howell_quadratic <- Howell1 %>% mutate(weight_s = (weight - mean(weight)) / sd(weight),
+                                       weight_s2 = weight_s^2,
+                                       a = rnorm(N, 178, 20),
+                                       b1 = rlnorm(N, 2.5, 1),
+                                       b2 = rnorm(N, -20, 1),
+                                       sigma = runif(N, 0, 10),
+                                       mu = a + b1 * weight_s + b2 * weight_s^2,
+                                       height_sim = rnorm(N, mu, sigma)
+                                       )
+```
+
+```{r}
+ggplot(data = howell_quadratic) +
+  geom_point(mapping = aes(x = weight, y = height_sim))
+```
+
+One could keep on fiddling... but the text asks us not to fit a model by hand.
+
+## 4H5
+
+```{r}
+cherry_temp <- cherry_blossoms %>% drop_na(doy, temp)
+
+ggplot(data = cherry_temp) +
+  geom_point(mapping = aes(x = temp, y = doy))
+```
+
+```{r}
+linear_fit <- quap(
+  alist(doy ~ dnorm(mu, sigma),
+        mu <- a + b * (temp - mean(temp)),
+        a ~ dnorm(100, 20),
+        b ~ dnorm(10, 10),
+        sigma ~ dexp(1/10)),
+  data = cherry_temp
+)
+
+precis(linear_fit)
+```
+
+Perhaps I should have started with a prior predictive check.
+
+```{r}
+N <- nrow(cherry_temp)
+prior_predictive <- cherry_temp %>% mutate(a = rnorm(N, 100, 20),
+                                           b = rnorm(N, 10, 10),
+                                           sigma = rexp(N, 1/10),
+                                           mu = a + b * (temp - mean(temp)),
+                                           sim_day = rnorm(N, mu, sigma))
+
+ggplot(data = prior_predictive) +
+  geom_point(mapping = aes(x = temp, y = sim_day))
+```
+
+This looks fine, I think, for a prior model. There is too much variability---sometimes we predict blossom before day 0!\--but it looks like a reasonable starting point.
+
+We can try a posterior predictive check as well.
+
+```{r}
+post <- extract.samples(linear_fit)
+
+posterior_predictive <- cherry_temp %>% cbind(slice_head(post, n = N)) %>%
+  mutate(mu = a + b * (temp - mean(temp)),
+         doy_sim = rnorm(N, mu, sigma))
+
+ggplot(data = posterior_predictive) +
+  geom_point(mapping = aes(x = temp, y = doy_sim))
+```
+
+```{r}
+samples <- extract.samples(linear_fit)
+N <- 20
+xs <- seq(from = 4.5, to = 8.5, length.out = N)
+mu_min <- rep(0, times = N)
+mu_max <- rep(0, times = N)
+pred_min <- rep(0, times = N)
+pred_max <- rep(0, times = N)
+xbar <- cherry_temp %>% summarize(m = mean(temp)) %>% pull(m)
+
+for (i in 1:20){
+  q <- samples %>% mutate(mu = a + b * (xs[i] - xbar),
+                     pred = rnorm(nrow(samples), mu, sigma)) %>%
+    summarize(q_mu_low = quantile(mu, 0.015),
+              q_mu_high = quantile(mu, 0.985),
+              q_pred_low = quantile(pred, 0.015),
+              q_pred_high = quantile(pred, 0.985))
+  mu_min[i] <- q %>% pull(q_mu_low)
+  mu_max[i] <- q %>% pull(q_mu_high)
+  pred_min[i] <- q %>% pull(q_pred_low)
+  pred_max[i] <- q %>% pull(q_pred_high)
+}
+
+strips <- tibble(mu_min = mu_min, mu_max = mu_max,
+                 pred_min = pred_min, pred_max = pred_max)
+```
+
+```{r}
+ggplot(data = cherry_temp) +
+  geom_ribbon(data = strips,
+              mapping = aes(x = xs, ymin = mu_min, ymax = mu_max),
+              alpha = 0.4) +
+  geom_ribbon(data = strips,
+              mapping = aes(x = xs, ymin = pred_min, ymax = pred_max),
+              alpha = 0.3) +
+  geom_point(mapping = aes(x = temp, y = doy), shape = 1, color = "blue", alpha = 0.5) +
+  geom_function(fun = function(x)(104.92 + (-2.98) * (x - xbar))) +
+  labs(x = "March temperature (C)",
+       y = "Blossom date (day of year") +
+  theme_bw()
+```
+
+I am content with this linear model!
+
+## 4H6
+
+```{r}
+cherry_full <- cherry_blossoms %>% drop_na(doy)
+num_knots <- 15
+knot_list <- quantile( cherry_full %>% pull(year) , probs=seq(0, 1, length.out=num_knots) )
+```
+
+```{r}
+library(splines)
+B <- bs(cherry_full %>% pull(year),
+        knots = knot_list[-c(1,num_knots)], degree=3 , intercept=TRUE )
+cherry_full <- cherry_blossoms %>% drop_na(doy)
+```
+
+```{r}
+cherry_refit <- quap(
+  alist(
+    D ~ dnorm( mu , sigma ) ,
+    mu <- a + B %*% w ,
+    a ~ dnorm(100,10),
+    w ~ dnorm(0,10),
+    sigma ~ dexp(1)
+    ),
+  data=list( D= cherry_full %>% pull(doy), B = B ) ,
+  start=list( w=rep( 0 , ncol(B) ) ) )
+
+precis(cherry_refit)
+```
+
+```{r}
+post <- extract.samples( cherry_refit )
+w <- apply( post$w , 2 , mean )
+plot( NULL , xlim = range(cherry_full$year) ,
+      ylim=c(-6,6) , xlab="year" , ylab="basis * weight" )
+for ( i in 1:ncol(B) ) lines( cherry_full$year , w[i]*B[,i] )
+```
+
+```{r}
+mu <- link( cherry_refit )
+mu_PI <- apply(mu,2,PI,0.97)
+plot( cherry_full$year , cherry_full$doy , col=col.alpha(rangi2, 0.3) , pch=16, 
+      shade( mu_PI , cherry_full$year , col=col.alpha("black",0.5) ))
+```
+
+Not sure what this question is getting at. When I change the mean on the prior for the weights, the fitting procedure compensates by adjusting a. I end up with essentially the same final model.
+
+## 4H7
+
+What happened to 4H7?
+
+## 4H8
+
+We refit the model without an intercept term. To do that, we change the call to "bs", using `intercept = FALSE` and including all the knots (instead of dropping the first).
+
+```{r}
+cherry_full <- cherry_blossoms %>% drop_na(doy)
+num_knots <- 15
+knot_list <- quantile( cherry_full %>% pull(year) , probs=seq(0, 1, length.out=num_knots) )
+```
+
+```{r}
+library(splines)
+B <- bs(cherry_full %>% pull(year),
+        knots = knot_list, degree=3 , intercept = FALSE )
+cherry_full <- cherry_blossoms %>% drop_na(doy)
+```
+
+Removing the intercept means that we need to change the priors. I'm not sure what to expect, but we can set the means for the spline terms to the old mean for the intercept.
+
+```{r}
+cherry_no_intercept <- quap(
+  alist(
+    D ~ dnorm( mu , sigma ) ,
+    mu <- B %*% w ,
+    w ~ dnorm(100,100),
+    sigma ~ dexp(1)
+    ),
+  data=list( D= cherry_full %>% pull(doy), B = B ) ,
+  start=list( w=rep( 100 , ncol(B) ) ) )
+
+precis(cherry_no_intercept, depth = 2)
+```
+
+Here are the spline terms, with large weights, as expected.
+
+```{r}
+post <- extract.samples( cherry_no_intercept )
+w <- apply( post$w , 2 , mean )
+plot( NULL , xlim = range(cherry_full$year) ,
+      ylim=c(0,120) , xlab="year" , ylab="basis * weight" )
+for ( i in 1:ncol(B) ) lines( cherry_full$year , w[i]*B[,i] )
+```
+
+And here is the model, the same as before.
+
+```{r}
+mu <- link( cherry_no_intercept )
+mu_PI <- apply(mu,2,PI,0.97)
+plot( cherry_full$year , cherry_full$doy , col=col.alpha(rangi2, 0.3) , pch=16, 
+      shade( mu_PI , cherry_full$year , col=col.alpha("black",0.5) ))
+```

### 2023-09-21 Stats Rethinking/Ch7/Practice Problems.qmd
Created Practice Problems Sheet

--- 2023-09-14 Stats Rethinking/Ch4/Practice Problems.qmd
@@
 
-# 4E1
-
-In the model definition, which is the likelihood?
-
-$$
-y_i\sim Normal(\mu,\sigma)\\
-\mu\sim Normal(0,10)\\
-\sigma\sim Exponential(1)
-$$
-
-The likelihood is, $y_i\sim Normal(\mu,\sigma)$
-
-# 4E2
-
-In the model definition above, how many parameters are in the posterior distribution?
-
-There are two parameters, $\mu$ and $\sigma$.
-
-# 4E3
-
-Using the model definition above, write down the appropriate form of Bayes' theorem that includes the proper likelihood and priors
-
-$$
-Pr(\mu,\sigma|y_i)=\frac{\prod_i Normal(y_i|\mu,\sigma)Normal(\mu|0,10)Exponential(\sigma|1)}{\int\int\prod_i Normal(y_i|\mu,\sigma)Normal(\mu|0,10)Exponential(\sigma|1)d\mu d\sigma}
-$$
-
-(Used formatting of Bayes on page 84)
-
-# 4E4
-
-In the model definition, which line is the linear model?
-
-$$
-y_i\sim Normal(\mu, \sigma)\\
-\mu_i=\alpha+\beta x_i\\
-\alpha\sim Normal(0,10)\\
-\beta\sim Normal(0,1)\\
-\sigma\sim Exponential(2)
-$$
-
-In this model, the $\mu_i=\alpha+\beta x_i$ is the linear model.
-
-# 4E5
-
-In the model definition for 4E4, how many parameters are in the posterior distribution?
-
-There are 3 parameters in the posterior dist.
-
-# 4M1
-
-# 4M2
-
-# 4M3
-
-# 4M4
-
-# 4M5
-
-# 4M6
-
-# 4M7
-
-# 4M8
-
-# 4H1
-
-# 4H2
-
-# 4H3
-
-# 4H4
-
-# 4H5
-
-# 4H6
-
-# 4H7
-
-# 4H8
+# 7E1

### 2023-09-21 Stats Rethinking/Ch7/Practice Problems.qmd
Finished Easy Problems

@@
 # 7E1
+
+State the three motivating criteria that define information entropy. Try to express each in your own words.
+
+1.  Uncertainty measures should be continuous so that arbitrarily small changes in probabilities do not correspond to massive shifts in uncertainty.
+2.  The measure of uncertainty should increase as the number of possible events increases.
+3.  The measure of uncertainty should be additive.
+
+# 7E2
+
+Suppose a coin is weighted such that, when it is tossed and lands on a table, it comes up heads 70% of the time. What is the entropy of this coin?
+
+```{r}
+p_coin <- c(0.7, 0.3)
+-sum(p_coin * log(p_coin))
+```
+
+# 7E3
+
+Suppose a four-sided die is loaded such that, when tossed onto a table, it shows "1" 20%, "2" 25%, "3" 25%, and "4" 30% of the time. What is the entropy of this die?
+
+```{r}
+p_die1 <- c(0.2, 0.25, 0.25, 0.3)
+-sum(p_die1 * log(p_die1))
+```
+
+# 7E4
+
+Suppose another four-sided die is loaded such that it never shows "4". The other three sides show equally often. What is the entropy for of this die?
+
+```{r}
+p_die2 <- c(1/3, 1/3, 1/3)
+-sum(p_die2 * log(p_die2))
+```
+
+# 7M1
+
+Write down and compare the definitions of AIC and WAIC. Which of these criteria is most general? Which assumptions are required to transform the more general criteria into a less general one?
+
+# 7M2
+
+Explain the differences between model selection and model comparison. What information is lost under model selection.
+
+Model selection is where you choose a model which has the lowest specified criterion value and then discarding the other models.
+While model comparison is where you keep all models but you weight what the models say based on their criterion value.
+You do not discard any of them.
+
+Under model selection you lose the information about relative model accuracy contained between the different CV/PSIS/WAIC values.
+Sometimes this is large and sometimes this is small, but discarding the information prevents this step from being performed.
+
+# 7M3\*
+
+When comparing models with an information criterion, why must all models be fit to exactly the same observations?
+What would happen to the information criterion values, if the models were fit to different numbers of observations?
+Provide some examples, if you are not sure.
+
+# 7M4
+
+What happens to the effective nymber of parameters, as measured by PSIS or WAIC, as a prior becomes more concentrated? Why? Perform some experiments, if you are not sure.
+
+# 7M5
+
+Provide an informal explanation of why informative priors reduce overfitting.
+
+When priors are not informative, our models get really excited about the data and basically only have this data to go off of in estimation.
+Thus, when using small sample sizes where these observations may happen to be wrong it can lead to the model overfit to just the small number of observations provided.
+Giving informative priors means that the model will be more skeptical when being given observations and lead to it incorporating these observations more conservatively into its posterior and predictions.
+
+# 7M6
+
+Provide and informal explanation of why overly informative priors result in underfitting.
+
+For a similar reasons above as to why informative priors reduce overfitting.
+Making priors too skepticals will mean that they ignore the data given to them which in turn will lead to the model being underfit.
+In other words the model just won't pay attention to the information given to it and will insist that its priors are the best.
+
+# 7H1
+
+# 7H2
+
+# 7H3
+
+# 7H4
+
+# 7H5

### 2023-09-21 Stats Rethinking/Ch7/Practice Problems.qmd
Finished through medium

@@
 
+$$
+AIC=-2*LLPD+2p\\
+WAIC(y,\Theta)=-2(LLPD-\sum var_{\theta}logp(y_i|\theta))
+$$
+
+Of these information criterion WAIC is more general. In order to transform WAIC into AIC you need the following conditions:
+
+1.  The priors are flat or overwhelmed by the likelihood.
+2.  The posterior distribution is approximately multivariate Gaussian.
+3.  The sample size $N$ is much greater than the number of parameters $k$.
+
 # 7M2
@@
 
-# 7M4
+Is this a trick question? It's obvious that if models are fit to different training data then they are different models entirely and not just variations on assumptions.
 
-What happens to the effective nymber of parameters, as measured by PSIS or WAIC, as a prior becomes more concentrated? Why? Perform some experiments, if you are not sure.
+# 7M4\*
+
+What happens to the effective number of parameters, as measured by PSIS or WAIC, as a prior becomes more concentrated? Why? Perform some experiments, if you are not sure.
+
+Shouldn't it become smaller? Since as you get more data they will drown out the noise of overfitting created by the extra parameters.
 
@@
 
+In 2007, *The Wall Street Journal* published an editorial with a graph
+
 # 7H2

### 2023-09-21 Stats Rethinking/Ch7/Practice Problems.qmd
Finished first hard problem

@@
 
-In 2007, *The Wall Street Journal* published an editorial with a graph
+In 2007, *The Wall Street Journal* published an editorial with a graph of corporate tax rates in 29 countries plotted against tax revenue.
+A badly fit curve was drawn in, seemingly by hand, to make the argument that the relationship between tax rate and tax revenue increases and then declines, such that higher tax rates can actually produce less tax revenue.
+I want you to actually fit a curve to these data, found in `data(Laffer)`.
+Consider models that use tax rate to predict tax revenue.
+Compare, using WAIC or PSIS, a straight line model to any curved models you like.
+What do you conclude about the relationship between tax rate and tax revenue?
+
+```{r}
+#Load in and observe data
+data(Laffer)
+head(Laffer)
+```
+
+```{r}
+#Make models
+#Quickly checked Laffer to see what my priors should be. Not sure if I was supposed to do that
+m7H1.1 <- quap(
+  alist(
+    tax_revenue ~ dnorm(mu, exp(sigma)),
+    mu <- a + b*tax_rate,
+    a ~ dnorm(20, 10),
+    b ~ dnorm(0, 10),
+    sigma ~ dnorm(0,1)
+  ), data=Laffer
+)
+
+m7H1.2 <- quap(
+  alist(
+    tax_revenue ~ dnorm(mu, exp(sigma)),
+    mu <- a + b[1]*tax_rate + b[2]*tax_rate,
+    a ~ dnorm(20, 10),
+    b ~ dnorm(0, 10),
+    sigma ~ dnorm(0,1)
+  ), data=Laffer, start = list(b=rep(0,2))
+)
+```
+
+```{r}
+#See what these curves should generally looks like
+ggplot(data=Laffer)+
+  aes(x=tax_rate, y=tax_revenue)+
+  geom_point()+
+  geom_smooth(method="lm", color="red", se=F)+
+  stat_smooth(method="lm", formula = y ~ x + I(x^2), color="blue", se=F)
+```
+
+```{r}
+#I can't read either of these outputs
+compare(m7H1.1, m7H1.2, func = WAIC)
+```
+
+```{r}
+compare(m7H1.1, m7H1.2, func = PSIS)
+```
 

### 2023-09-28 Stats Rethinking/Ch7/Practice Problems.qmd
Updated Prac Problems
Update Practice Problems.qmd

@@
 
+```{r}
+#Load his data on cars and create model with "quap"
+data(cars)
+m <- quap(
+  alist(
+    dist ~ dnorm(mu,sigma),
+    mu <- a + b*speed,
+    a ~ dnorm(0,100),
+    b ~ dnorm(0,10),
+    sigma ~ dexp(1)
+  ) , data=cars )
+set.seed(94)
+post <- extract.samples(m,n=1000)
+```
+
+```{r}
+n_samples <- 1000
+logprob <- sapply( 1:n_samples ,
+  function(s) {
+    mu <- post$a[s] + post$b[s]*cars$speed
+    dnorm( cars$dist , mu , post$sigma[s] , log=TRUE )
+  } )
+
+n_cases <- nrow(cars)
+lppd <- sapply( 1:n_cases , function(i) log_sum_exp(logprob[i,]) - log(n_samples) )
+
+pWAIC <- sapply( 1:n_cases , function(i) var(logprob[i,]) )
+
+sum(pWAIC)
+```
+
+```{r}
+lm(data=cars, dist ~ speed) %>% 
+  summary()
+```
+
+```{r}
+m <- quap(
+  alist(
+    dist ~ dnorm(mu,sigma),
+    mu <- a + b*speed,
+    a ~ dnorm(-17.6, 1),
+    b ~ dnorm(3.9,0.1),
+    sigma ~ dnorm(15,2)
+  ) , data=cars )
+set.seed(94)
+post <- extract.samples(m,n=1000)
+
+n_samples <- 1000
+logprob <- sapply( 1:n_samples ,
+  function(s) {
+    mu <- post$a[s] + post$b[s]*cars$speed
+    dnorm( cars$dist , mu , post$sigma[s] , log=TRUE )
+  } )
+
+n_cases <- nrow(cars)
+lppd <- sapply( 1:n_cases , function(i) log_sum_exp(logprob[i,]) - log(n_samples) )
+
+pWAIC <- sapply( 1:n_cases , function(i) var(logprob[i,]) )
+
+sum(pWAIC)
+```
+
 # 7M5

### 2023-09-28 Stats Rethinking/Ch8/Practice Problems.qmd
Practice problems written

---
title: "Practice Problems"
format: html
---

# Load Libraries

```{r}
library(tidyverse)
library(rethinking)
```

# 8E1

For each of the causal relationship below, name a hypothetical third variable that would lead to an interaction effect.

1.  Bread dough rises because of yeast.
    1.  Temperature - Too cold or hot an the bacteria die, somewhere there is a sweet-spot for it
2.  Education leads to higher income.
    1.  
3.  Gasoline makes a car go.

# 8E2

Which of the following explanations invokes an interaction?

1.  Caramelizing onions requires cooking over low heat and making sure the onions do not dry out.
2.  A car will go faster when it has more cylinders or when it has a better fuel injector.
3.  Most people acquire their political beliefs from their parents, unless they get them instead from their friends.
4.  Intelligent animal species tend to either be highly social or have manipulative appendages (hands, tentacles, etc.)

# 8E3

For each of the explanations in **8E2**, write a linear model that expresses the stated relationship.

# 8M1

Recall the tulips example from the chapter. Suppose another set of treatments adjusted the temperature in the greenhouse over two levels: cold and hot. The data in the chapter were collected at the cold temperature. You find none of the plants grown under the hot temperature developed any blooms at all, regardless of the water and shade levels. Can you explain this result in terms of interactions between water, shade, and temperature?

# 8M2

Can you invent a regression equation that would make the bloom size zero, whenever the temperature is hot?

# 8M3

In parts of North America, ravens depend upon wolves for their food. This is because ravens are carnivorous but cannot usually kill or open carcasses of prey. Wolves however can and do kill and tear open animals, and they tolerate ravens co-feeding at their kills. This species relationship is generally described as a "species interaction." Can you invent a hypothetical set of data on raven populations size in which this relationship would manifest as a statistical interaction? Do you think the biological interaction could be linear? Why or why not?

# 8M4

Repeat the tulips analysis, but this time use priors that constrain the effect of water to be positive and the effect of shade to be negative. Use prior predictive simulation. What do these prior assumptions mean for the interaction prior, if anything?

# 8H1

# 8H2

# 8H3

# 8H4

# 8H5

# 8H6

### 2023-09-28 Stats Rethinking/Ch8/Practice Problems.qmd
Finished through medium

@@
 2.  Education leads to higher income.
-    1.  
+    1.  Gender
 3.  Gasoline makes a car go.
+    1.  Horsepower
 
@@
 1.  Caramelizing onions requires cooking over low heat and making sure the onions do not dry out.
+    1.  No interaction
 2.  A car will go faster when it has more cylinders or when it has a better fuel injector.
+    1.  Interaction
 3.  Most people acquire their political beliefs from their parents, unless they get them instead from their friends.
+    1.  No interaction
 4.  Intelligent animal species tend to either be highly social or have manipulative appendages (hands, tentacles, etc.)
+    1.  Interaction
 
-# 8E3
+# 8E3\*
 
 For each of the explanations in **8E2**, write a linear model that expresses the stated relationship.
+
+1\)
+
+$$
+O_i\sim Normal(\mu_i, \sigma)\\
+\mu_i=\alpha+\beta(H_i-\bar{H})\\
+\alpha\sim Normal(1,1)\\
+\beta\sim Normal(1,1)\\
+\sigma\sim Exponential(1)
+$$
+
+2\)
+
+$$
+S_i\sim Normal(\mu_i, \sigma)\\
+\mu_i=\alpha+(\gamma_{C,i})C_i+\beta_{FI}FI_i\\
+\gamma_{C,i}=\beta_C+\beta_{CFI}FI_i\\
+\alpha\sim Normal(0.5,1)\\
+\beta_C\sim Normal(0,1)\\
+\beta_{FI}\sim Normal(0,1)\\
+\sigma\sim Exponential(1)
+$$
+
+3\)
+
+I have no idea how to model this
+
+$$
+P_i\sim Normal(\mu_i,\sigma)\\
+\mu_i=\alpha+\beta
+$$
+
+4\)
+
+I have no idea how to model this one either
 
@@
 
+Could we just change the equation to $\mu_i=C_i(\alpha+\gamma_{W,i}W_i+\beta_SS_i)$ where $\gamma_{W,i}=\beta_W+\beta_{WS}S_i$ and $C_i$ is a dummy which represents if it was cold in the room (1 yes, 0 no).
+This would make the model usually predict a 0 for the buds but then could leave a possibility of it blooming.
+
 # 8M2
@@
 Can you invent a regression equation that would make the bloom size zero, whenever the temperature is hot?
+
+See 8M1.
 
@@
 
+```{r}
+data("tulips")
+d <- tulips
+
+d$blooms_std <- d$blooms / max(d$blooms)
+d$water_cent <- d$water - mean(d$water)
+d$shade_cent <- d$shade - mean(d$shade)
+
+a <- rnorm(1e4, 0.5, 1); sum(a<0 | a>1)/length(a)
+a <- rnorm(1e4, 0.5, 0.25); sum(a<0 | a>1)/length(a)
+```
+
+```{r}
+m8.5 <- quap(
+  alist(
+    blooms_std ~ dnorm(mu, sigma),
+    mu <- bw*water_cent + bs*shade_cent + bws*water_cent*shade_cent,
+    a ~ dnorm(0.5, 0.25),
+    bw ~ dnorm(0, 0.25),
+    bs ~ dnorm(0, 0.25),
+    bws ~ dnorm(0, 0.25),
+    sigma ~ dexp(1)
+  ), data=d
+)
+
+m8.5.2 <- quap(
+  alist(
+    blooms_std ~ dnorm(mu, sigma),
+    mu <- bw*water_cent + bs*shade_cent + bws*water_cent*shade_cent,
+    a ~ dnorm(0.5, 0.25),
+    bw ~ dnorm(0.5, 0.25),
+    bs ~ dnorm(-0.5, 0.25),
+    bws ~ dnorm(0, 0.25),
+    sigma ~ dexp(1)
+  ), data=d
+)
+```
+
+```{r}
+precis(m8.5)
+```
+
+```{r}
+precis(m8.5.2)
+```
+
+```{r}
+compare(m8.5, m8.5.2, func=WAIC)
+```
+
+```{r}
+compare(m8.5, m8.5.2, func=PSIS)
+```
+
+Doesn't appear that they change the interaction effect at all.
+
 # 8H1
+
+Return to the `data(tulips)` example in the chapter.
+Now include the `bed` variable as a predictor in the interaction model.
+Don't interact `bed` with the other predictors; just include it a s a main effect.
+Not that `bed` is categorical.
+So to use it properly, you will need to either construct dummy variables or rather an index variable, as explained in Chapter 5.
 

### 2023-09-29 Lahman Analysis/quap/RoughAnalysis.qmd
Added Lahman Analysis folder and documents

---
title: "Rough Analysis"
format: html
---

# Load Libraries

## Installing Lahman

```{r}
#| eval: false

#Installing Lahman is easy and simply requires running this code chunk

install.packages("Lahman")
library(Lahman)
```

## Loading Lahman

```{r}
#There are more datasets in Lahman than we care about so we'll select "Batting"

Batting <- Lahman::Batting
```

The `Batting` data set describes hitting outcomes for each MLB player each year from 1871-present. The explanation of the variable abbreviations is available on page 14 of [this document](https://cran.r-project.org/web/packages/Lahman/Lahman.pdf).

## Load Other Libraries

```{r}
#Of course we need the other two main libraries
library(tidyverse)
library(rethinking)
```

# Quap

### 2023-09-29 Stats Rethinking/Ch8/Practice Problems.qmd
Update Practice Problems.qmd
Added Lahman Analysis folder and documents

@@
     blooms_std ~ dnorm(mu, sigma),
-    mu <- bw*water_cent + bs*shade_cent + bws*water_cent*shade_cent,
+    mu <- a + bw*water_cent + bs*shade_cent + bws*water_cent*shade_cent,
     a ~ dnorm(0.5, 0.25),
@@
 
+How do you make a index variable?
+
+```{r}
+#We'll make a bed into an integer
+d <- d %>% 
+  mutate(bed_id = as.integer(bed), .after=bed)
+```
+
+```{r}
+#Recreate the m8.4 from chapter
+m8.4 <- quap(
+  alist(
+    blooms_std ~ dnorm(mu, sigma),
+    mu <- a + bw*water_cent + bs*shade_cent,
+    a ~ dnorm(0.5, 0.25),
+    bw ~ dnorm(0, 0.25),
+    bs ~ dnorm(0, 0.25),
+    sigma ~ dexp(1)
+  ), data=d
+)
+```
+
+```{r}
+m8h1 <- quap(
+  alist(
+    blooms_std ~ dnorm(mu, sigma),
+    mu <- a[bed_id] + bw*water_cent + bs*shade_cent,
+    a[bed_id] ~ dnorm(0.5, 0.25),
+    bw ~ dnorm(0, 0.25),
+    bs ~ dnorm(0, 0.25),
+    sigma ~ dexp(1)
+  ), data=d
+)
+```
+
+```{r}
+precis(m8h1, depth=2)
+```
+
+```{r}
+precis(m8.4)
+```
+
+```{r}
+precis(m8.5)
+```
+
 # 8H2
+
+```{r}
+compare(m8h1, m8.4, m8.5, func=WAIC)
+```
+
+```{r}
+plot(compare(m8h1, m8.4, m8.5, func=WAIC))
+```
 

### 2023-10-03 Stats Rethinking/Ch9/Practice Problems.qmd
Created chpt 9 practice doc

--- 2023-09-28 Stats Rethinking/Ch8/Practice Problems.qmd
@@
 
-# 8E1
-
-For each of the causal relationship below, name a hypothetical third variable that would lead to an interaction effect.
-
-1.  Bread dough rises because of yeast.
-    1.  Temperature - Too cold or hot an the bacteria die, somewhere there is a sweet-spot for it
-2.  Education leads to higher income.
-    1.  
-3.  Gasoline makes a car go.
-
-# 8E2
-
-Which of the following explanations invokes an interaction?
-
-1.  Caramelizing onions requires cooking over low heat and making sure the onions do not dry out.
-2.  A car will go faster when it has more cylinders or when it has a better fuel injector.
-3.  Most people acquire their political beliefs from their parents, unless they get them instead from their friends.
-4.  Intelligent animal species tend to either be highly social or have manipulative appendages (hands, tentacles, etc.)
-
-# 8E3
-
-For each of the explanations in **8E2**, write a linear model that expresses the stated relationship.
-
-# 8M1
-
-Recall the tulips example from the chapter.
-Suppose another set of treatments adjusted the temperature in the greenhouse over two levels: cold and hot.
-The data in the chapter were collected at the cold temperature.
-You find none of the plants grown under the hot temperature developed any blooms at all, regardless of the water and shade levels.
-Can you explain this result in terms of interactions between water, shade, and temperature?
-
-# 8M2
-
-Can you invent a regression equation that would make the bloom size zero, whenever the temperature is hot?
-
-# 8M3
-
-In parts of North America, ravens depend upon wolves for their food.
-This is because ravens are carnivorous but cannot usually kill or open carcasses of prey.
-Wolves however can and do kill and tear open animals, and they tolerate ravens co-feeding at their kills.
-This species relationship is generally described as a "species interaction." Can you invent a hypothetical set of data on raven populations size in which this relationship would manifest as a statistical interaction?
-Do you think the biological interaction could be linear?
-Why or why not?
-
-# 8M4
-
-Repeat the tulips analysis, but this time use priors that constrain the effect of water to be positive and the effect of shade to be negative.
-Use prior predictive simulation.
-What do these prior assumptions mean for the interaction prior, if anything?
-
-# 8H1
-
-# 8H2
-
-# 8H3
-
-# 8H4
-
-# 8H5
-
-# 8H6
+# 9E1

### 2023-10-03 Stats Rethinking/Ch9/Practice Problems.qmd
Finished through easy

@@
 # 9E1
+
+Which of the following is a requirement of the simple Metropolis algorithm?
+
+1.  The parameters must be discrete
+2.  The likelihood function must be Gaussian
+3.  Th proposal distribution must be symmetric
+
+1 definitely must be true. 2 I would say not. And 3 is probably not.
+
+# 9E2
+
+Gibbs sampling is more efficient than the Metropolis algorithm. How does it achieve this extra efficiency? Are there any limitations to the Gibbs sampling strategy?
+
+It is more efficient because it uses adaptive proposals from conjugate pairs.
+It runs into problems because it has a high concentration of measure, meaning it has a had time exploring without going wildly in the wrong direction.
+
+# 9E3
+
+Which sort of parameters can Hamiltonian Monte Carlo not handle? Can you explain why?
+
+The HMC cannot handle discrete parameters, since it must be able to stop anywhere along gradient it travels.
+
+# 9E4
+
+Explain the difference between the effective number of samples, `n_eff` as calculated by Stan, and the actual number of samples.
+
+The effective number of samples is an estimate of the number of independent samples from the posterior distribution, in terms of some estimating some function like the posterior mean.
+However, since Markov chains are often correlated this will decreased the amount of independent samples.
+Thus, Stan provides a calculated amount of independent samples depending on the level of autocorrelation.
+
+# 9E5
+
+Which value should `Rhat` approach, when a chain is sampling the posterior distribution correctly?
+
+$\hat{R}$ should approach 1.
+
+# 9E6
+
+Sketch a good trace plot for a Markov chain, one that is effectively sampling from the posterior distribution.
+What is good about its shape?
+Then sketch a trace plot for malfunctioning Markov chain.
+What about its shape indicates malfunction?
+
+A good chain is one that is:
+
+1.  Stationary - The path of the chain says within the same high-probability proportion of the posterior distribution
+2.  Mixing well - Chain rapidly explores the full region
+3.  Converging - Multiple chains stick around the same region of high probability
+
+Violating any of these will result in a bad chain
+
+# 9E7
+
+Repeat the problem above but this time for a trace rank plot.
+
+The trank plot simply reduces the noise of a trace plot into a single line. Therefore the same conditions and looks as described above hold
+
+# 9M1

### 2023-10-04 Lahman Analysis/Rough Analysis/RoughAnalysis.qmd
Got stan set up

--- 2023-09-29 Lahman Analysis/quap/RoughAnalysis.qmd
@@
 
+Of course, we don't want to analyze all data from 1871-present since the game has changed considerable since then.
+So filter to only consider seasons past WW2 (1945).
+We can always cut down from here, but this will give us a huge sample to use where the rules (nor the game) has changed significantly since.
+
+```{r}
+Modern.Batting <- Batting %>% 
+  filter(yearID > 1945)
+```
+
 ## Load Other Libraries
@@
 # Quap
+
+# Ulam

### 2023-10-04 Stats Rethinking/Ch9/Practice Problems.qmd
Got stan set up

@@
 # 9M1
+
+Re-estimate the terrain ruggedness model from the chapter but now use a uniform prior for the standard deviation, `sigma`.
+The uniform prior should be `dunif(0,1)`.
+Use `ulam` to estimate the posterior.
+Does the different prior have any detectable influence on the posterior distribution of `sigma`?
+Why or why not?
+
+```{r}
+#Set up for model
+data("rugged")
+d <- rugged
+d$log_gdp <- log(d$rgdppc_2000)
+dd <- d[complete.cases(d$rgdppc_2000),]
+dd$log_gdp_std <- dd$log_gdp / mean(dd$log_gdp)
+dd$rugged_std <- dd$rugged / max(dd$rugged)
+dd$cid <- ifelse(dd$cont_africa==1, 1, 2)
+
+dat_slim <- list(
+  log_gdp_std = dd$log_gdp_std,
+  rugged_std = dd$rugged_std,
+  cid = as.integer(dd$cid)
+)
+
+#Run model
+m9.1 <- ulam(
+  alist(
+    log_gdp_std ~ dnorm(mu, sigma),
+    mu <- a[cid] + b[cid]*(rugged_std - 0.215),
+    a[cid] ~ dnorm(1, 0.1),
+    b[cid] ~ dnorm(0, 0.3),
+    sigma ~ dexp(1)
+  ), data=dat_slim, chains=1
+)
+```
+
+```{r}
+#Let's examine down here so we don't rerun that process a ton, it took a LONG time
+precis(m9.1, depth=2)
+```
+
+```{r}
+#Now we can do the problem
+m9M1 <- ulam(
+  alist(
+    log_gdp_std ~ dnorm(mu, sigma),
+    mu <- a[cid] + b[cid]*(rugged_std - 0.215),
+    a[cid] ~ dnorm(1, 0.1),
+    b[cid] ~ dnorm(0, 0.3),
+    sigma ~ dunif(0,1)
+  ), data=dat_slim, chains=1
+)
+```
+
+```{r}
+precis(m9M1, depth=2)
+```
+
+# 9M2
+
+Modify the terrain ruggedness model again. This time, change the prior for `b[cid]` to `dexp(0.3)`. What does this do to the posterior distribution? Can you explain it?
+
+# 9M3
+
+Re-estimate one of the Stan models from the chapter, but at different number of `warmup iterations`.
+Be sure to use the same number of sampling iterations in each case.
+Compare the `n_eff` values.
+How much warm-up is enough?
+
+# 9H1
+
+# 9H2
+
+# 9H3
+
+# 9H4
+
+# 9H5
+
+# 9H6
+
+# 9H7

### 2023-10-04 Stats Rethinking/Ch9/Practice Problems.qmd
Finished through medium

@@
 
+This did not appear to have an effect on `sigma`. My guess would be that there is so much data that this change to a flat prior simply got overridden by the data.
+
 # 9M2
@@
 Modify the terrain ruggedness model again. This time, change the prior for `b[cid]` to `dexp(0.3)`. What does this do to the posterior distribution? Can you explain it?
+
+```{r}
+m9M2 <- ulam(
+  alist(
+    log_gdp_std ~ dnorm(mu, sigma),
+    mu <- a[cid] + b[cid]*(rugged_std - 0.215),
+    a[cid] ~ dnorm(1, 0.1),
+    b[cid] ~ dexp(0.3),
+    sigma ~ dunif(0,1)
+  ), data=dat_slim, chains=1
+)
+```
+
+```{r}
+precis(m9M2, depth=2)
+```
+
+This lessened our estimated effect of `b[2]` considerable from -0.14 to 0.02.
+Meanwhile, almost all other values remained the same.
+This should mean that our guess for the prior would have a large influence on the output of `b[2]` possibly suggesting the estimates aren't very good for it.
 
@@
 
+```{r}
+#Recreate m9.1 but change warmup iterations
+
+m9M3 <- ulam(
+  alist(
+    log_gdp_std ~ dnorm(mu, sigma),
+    mu <- a[cid] + b[cid]*(rugged_std - 0.215),
+    a[cid] ~ dnorm(1, 0.1),
+    b[cid] ~ dnorm(0, 0.3),
+    sigma ~ dexp(1)
+  ), data=dat_slim, chains=1, warmup = 900
+)
+```
+
+```{r}
+precis(m9M3, depth=2)
+```
+
+With 900 warm-ups we start to see that our `n_eff` value drops off sharply.
+As the textbook explains, we only need about 200 and for the most part our observations are falling around that range so it's not impacting out model too much.
+However, with much more then we will probably start to see this model begin to deteriorate.
+
+```{r}
+m9M3.2 <- ulam(
+  alist(
+    log_gdp_std ~ dnorm(mu, sigma),
+    mu <- a[cid] + b[cid]*(rugged_std - 0.215),
+    a[cid] ~ dnorm(1, 0.1),
+    b[cid] ~ dnorm(0, 0.3),
+    sigma ~ dexp(1)
+  ), data=dat_slim, chains=1, warmup = 100
+)
+```
+
+```{r}
+precis(m9M3.2, depth=2)
+```
+
+In contrast here we only give the model 100 to warm up.
+This drastically increases our `n_eff` value and doesn't seem to effect our $\hat{R}$ value much, if at all.
+Thus, for this model it seems that we could get away with having much less warm-up than the model uses.
+
 # 9H1
+
+Run the model below and then inspect the posterior distribution and explain what it is accomplishing.
+
+```{r}
+#Cell provided for problem
+
+mp <- ulam(
+  alist(
+    a ~ dnorm(0,1),
+    b ~ dcauchy(0,1)
+  ), data=list(y=1), chains=1
+)
+```
+
+```{r}
+precis(mp, depth=2)
+```
 

### 2023-10-04 Stats Rethinking/Ch9/Practice Problems.qmd
Finished some hards

@@
 
-# 9H1
+# 9H1\*
 
 Run the model below and then inspect the posterior distribution and explain what it is accomplishing.
+
+Compare this samples for the parameters `a` and `b`.
+Can you explain the different trace plots?
+If you are unfamiliar with Cauchy distribution, you should look it up.
+The key feature to attend to is that it has no expected value/ Can you connect this fact to the trace plot?
 
@@
 
+```{r}
+traceplot(mp)
+```
+
+I'm not going to lie, I have no idea what this is doing or trying to accomplish. It also does look like for `b` that there is an expected value that it settles to, at least for this run.
+
 # 9H2
@@
 
+Modify the Metropolis algorithm code from the chapter to handle the case that the island populations have a different distribution than the island labels.
+This means that the island's number will not be the same as its population.
+
 # 9H6

### 2023-10-06 Lahman Analysis/Rough Analysis/RoughAnalysis.qmd
Updated Model

@@
 install.packages("Lahman")
-library(Lahman)
 ```
@@
 #There are more datasets in Lahman than we care about so we'll select "Batting"
+library(tidyverse)
+library(Lahman)
 
-Batting <- Lahman::Batting
+Teams <- Lahman::Teams
 ```
@@
 ```{r}
-Modern.Batting <- Batting %>% 
+Modern.Teams <- Teams %>% 
   filter(yearID > 1945)
@@
 ```{r}
-#Of course we need the other two main libraries
-library(tidyverse)
+#| warning: false
+#Of course we need the other main libraries
 library(rethinking)
@@
 
+# What are we modeling?
+
+For this project, I want to model how many homeruns a team hits in a season. We know that this should probably be normally distributed and we can derive our expected mean $\mu$ with a little math.
+
+In the more modern game, having a 25HR would probably be considered about average.
+With 9 guys on the team with some above and some below then we could expect about $9\cdot25=225$.
+So we would expect teams to hit about 225 homeruns in a year.
+However, our sample is of teams from 1945 to 2022 which is decidedly not comprised of modern teams.
+Furthermore, in the "good-ol-days" batting average was stressed much more than power; hence players did not hit as many home runs.
+Therefore, lets guess that each player averaged about 15HR a season.
+This would mean we could expect each team to hit $9\cdot15=135$ homeruns each year.
+
+Now with our general guess of the mean we can use a prior. However, this problem is shaping up as one that might lend itself not to a Gaussian prior, but a Poisson one.
+
+Furthermore, we want to model this, not just match distributions.
+So, I would argue (since we're keeping this simple) that the number of team AB's has a lot to do with how many home runs the team hits.
+More AB's implies more chances to hit HR's and suggests the team is just better at not getting out overall.
+So, let's use AB's as our predictor.
+
 # Quap
 
+```{r}
+#Calculate mean AB's now so it's easy to access later
+mean_AB <- mean(Modern.Teams$AB)
+```
+
+```{r}
+q1 <- quap(
+  alist(
+    HR ~ dnorm(mu, sigma),
+    mu <- a + b*(AB - mean_AB),
+    #Standard deviation of 20 seems fair to me for home runs in a year
+    a ~ dnorm(135, 20),
+    #Want to make a very non descriptive prior here since I don't know what the value of b could be
+    b ~ dnorm(1, 1),
+    sigma ~ dexp(1)
+  ), data = Modern.Teams
+)
+```
+
+```{r}
+precis(q1, depth=2)
+```
+
 # Ulam
+
+```{r}
+u1 <- ulam(
+  alist(
+    HR ~ dnorm(mu, sigma),
+    mu <- a + b*(AB - mean_AB),
+    #Standard deviation of 20 seems fair to me for home runs in a year
+    a ~ dnorm(135, 20),
+    #Want to make a very non descriptive prior here since I don't know what the value of b could be
+    b ~ dnorm(1, 1),
+    sigma ~ dexp(1)
+  ), data = Modern.Teams, chains=1
+)
+```
+
+```{r}
+precis(u1)
+```
+
+# Comparison

### 2023-10-06 Lahman Analysis/Rough Analysis/RoughAnalysis.qmd
Fixed Ulam

@@
 
+We can see from the `precis` that our `sigma` prior is WAY too small. `precis` tells us its about 40 and we ahve it at 1!
+
 # Ulam
+
+```{r}
+slim <- list(
+  AB = Modern.Teams$AB,
+  HR = Modern.Teams$HR
+)
+
+str(slim)
+```
 
@@
     HR ~ dnorm(mu, sigma),
-    mu <- a + b*(AB - mean_AB),
+    mu <- a + b*(AB - 5368),
     #Standard deviation of 20 seems fair to me for home runs in a year
@@
     sigma ~ dexp(1)
-  ), data = Modern.Teams, chains=1
+  ), data = slim, chains=4, cores = 4
 )
@@
 # Comparison
+
+```{r}
+traceplot(u1)
+```
+
+```{r}
+trankplot(u1)
+```

### 2023-10-11 Stats Rethinking/Ch11/Practice Problems.qmd
Added space for chpt 11 practice problems

--- 2023-10-04 Stats Rethinking/Ch9/Practice Problems.qmd
@@
 
-# 9E1
+# 11E1
 
-Which of the following is a requirement of the simple Metropolis algorithm?
+# 11E2
 
-1.  The parameters must be discrete
-2.  The likelihood function must be Gaussian
-3.  Th proposal distribution must be symmetric
+# 11E3
 
-1 definitely must be true. 2 I would say not. And 3 is probably not.
+# 11E4
 
-# 9E2
+# 11M1
 
-Gibbs sampling is more efficient than the Metropolis algorithm. How does it achieve this extra efficiency? Are there any limitations to the Gibbs sampling strategy?
+# 11M2
 
-It is more efficient because it uses adaptive proposals from conjugate pairs.
-It runs into problems because it has a high concentration of measure, meaning it has a had time exploring without going wildly in the wrong direction.
+# 11M3
 
-# 9E3
+# 11M4
 
-Which sort of parameters can Hamiltonian Monte Carlo not handle? Can you explain why?
+# 11H1
 
-The HMC cannot handle discrete parameters, since it must be able to stop anywhere along gradient it travels.
+# 11H2
 
-# 9E4
+# 11H3
 
-Explain the difference between the effective number of samples, `n_eff` as calculated by Stan, and the actual number of samples.
-
-The effective number of samples is an estimate of the number of independent samples from the posterior distribution, in terms of some estimating some function like the posterior mean.
-However, since Markov chains are often correlated this will decreased the amount of independent samples.
-Thus, Stan provides a calculated amount of independent samples depending on the level of autocorrelation.
-
-# 9E5
-
-Which value should `Rhat` approach, when a chain is sampling the posterior distribution correctly?
-
-$\hat{R}$ should approach 1.
-
-# 9E6
-
-Sketch a good trace plot for a Markov chain, one that is effectively sampling from the posterior distribution.
-What is good about its shape?
-Then sketch a trace plot for malfunctioning Markov chain.
-What about its shape indicates malfunction?
-
-A good chain is one that is:
-
-1.  Stationary - The path of the chain says within the same high-probability proportion of the posterior distribution
-2.  Mixing well - Chain rapidly explores the full region
-3.  Converging - Multiple chains stick around the same region of high probability
-
-Violating any of these will result in a bad chain
-
-# 9E7
-
-Repeat the problem above but this time for a trace rank plot.
-
-The trank plot simply reduces the noise of a trace plot into a single line. Therefore the same conditions and looks as described above hold
-
-# 9M1
-
-Re-estimate the terrain ruggedness model from the chapter but now use a uniform prior for the standard deviation, `sigma`.
-The uniform prior should be `dunif(0,1)`.
-Use `ulam` to estimate the posterior.
-Does the different prior have any detectable influence on the posterior distribution of `sigma`?
-Why or why not?
-
-```{r}
-#Set up for model
-data("rugged")
-d <- rugged
-d$log_gdp <- log(d$rgdppc_2000)
-dd <- d[complete.cases(d$rgdppc_2000),]
-dd$log_gdp_std <- dd$log_gdp / mean(dd$log_gdp)
-dd$rugged_std <- dd$rugged / max(dd$rugged)
-dd$cid <- ifelse(dd$cont_africa==1, 1, 2)
-
-dat_slim <- list(
-  log_gdp_std = dd$log_gdp_std,
-  rugged_std = dd$rugged_std,
-  cid = as.integer(dd$cid)
-)
-
-#Run model
-m9.1 <- ulam(
-  alist(
-    log_gdp_std ~ dnorm(mu, sigma),
-    mu <- a[cid] + b[cid]*(rugged_std - 0.215),
-    a[cid] ~ dnorm(1, 0.1),
-    b[cid] ~ dnorm(0, 0.3),
-    sigma ~ dexp(1)
-  ), data=dat_slim, chains=1
-)
-```
-
-```{r}
-#Let's examine down here so we don't rerun that process a ton, it took a LONG time
-precis(m9.1, depth=2)
-```
-
-```{r}
-#Now we can do the problem
-m9M1 <- ulam(
-  alist(
-    log_gdp_std ~ dnorm(mu, sigma),
-    mu <- a[cid] + b[cid]*(rugged_std - 0.215),
-    a[cid] ~ dnorm(1, 0.1),
-    b[cid] ~ dnorm(0, 0.3),
-    sigma ~ dunif(0,1)
-  ), data=dat_slim, chains=1
-)
-```
-
-```{r}
-precis(m9M1, depth=2)
-```
-
-# 9M2
-
-Modify the terrain ruggedness model again. This time, change the prior for `b[cid]` to `dexp(0.3)`. What does this do to the posterior distribution? Can you explain it?
-
-# 9M3
-
-Re-estimate one of the Stan models from the chapter, but at different number of `warmup iterations`.
-Be sure to use the same number of sampling iterations in each case.
-Compare the `n_eff` values.
-How much warm-up is enough?
-
-# 9H1
-
-# 9H2
-
-# 9H3
-
-# 9H4
-
-# 9H5
-
-# 9H6
-
-# 9H7
+# 11H4

### 2023-10-12 Stats Rethinking/Ch11/Practice Problems.qmd
Wrote through medium questions

@@
 
+If an event has probability 0.35, what are the log-odds for this event?
+
 # 11E2
+
+If and event has log-odds 3.2, what is the probability of this event?
 
@@
 
+Suppose that a coefficient in a logistic regression has a value 1.7. What does this imply about the proportional change in odds of the outcome?
+
 # 11E4
+
+Why do Poisson regressions sometimes require the use of an offset? Provide an example.
 
@@
 
+As explained in the chapter, binomial data can be organized in aggregated and disaggregated forms, without any impact on inference.
+But the likelihood of the data does change when the data are converted between the two formats.
+Can you explain why?
+
 # 11M2
+
+If a coefficient in a Poisson regression has a value of 1.7, what does that imply about the change in the outcome?
 
@@
 
+Explain why the logit ink is appropriate for a binomial generalized linear model.
+
 # 11M4
+
+Explain why the log link is appropriate for a Poisson generalized linear model.
+
+# 11M5
+
+What would it imply to use a logit link for the mean of a Poisson generalized linear model? Can you think of a scenario for which this would make sense?
+
+# 11M6
+
+State the constraints for which the binomial and Poisson distributions have maximum entropy. Are the constraints different at all for binomial and Poisson? Why or why not?
+
+# 11M7
+
+Use `quap` to construct a quadratic approximate posterior distribution for the chimpanzee model that includes a unique intercept for each actor `m11.4`.
+Compare the quadratic approximation to the posterior distribution produced instead from MCMC/ Can you explain both the differences and the similarities between the approximate and the MCMC distributions?
+Relax the prior on the actor intercepts to Normal(0,10).
+Re-estimate the posterior using both `ulam` and `quap`.
+Do the differences increase or decrease?
+Why?
+
+# 11M8
+
+Use `WAIC` or `PSIS` to compare the chimpanzee model that includes a unique intercept for each actor, `m11.4`, to the simpler models fit in the same section. Interpret the results.
 
@@
 # 11H4
+
+# 11H5
+
+# 11H6

### 2023-10-12 Lahman Analysis/Rough Analysis/BinomialRoughAnalysis.qmd
Added space for Binomial regression analysis

---
title: "Binomial Rough Analysis"
format: html
---

# Load Libraries

```{r}
library(tidyverse)
library(rethinking)
library(Lahman)
```

# Clean Lahman

```{r}
Batting <- Lahman::Batting

Batting %>% 
  filter(yearID >= 1945)
```

# Quap

# Ulam

# Comparison

### 2023-10-12 Lahman Analysis/Rough Analysis/BinomialRoughAnalysis.qmd
Finished through medium practice problems

@@
 
-Batting %>% 
-  filter(yearID >= 1945)
+Batting <- Batting %>% 
+  filter(yearID >= 1945) %>% 
+  select(playerID, yearID, lgID, AB, AB, HR) %>% 
+  mutate(lg = ifelse(lgID=="NL",1,2))
 ```
@@
 
+```{r}
+head(Batting)
+```
+
+```{r}
+slim <- list(
+  AB = Batting$AB,
+  HR = Batting$HR,
+  LG = Batting$lg
+)
+
+str(slim)
+```
+
+```{r}
+u1 <- ulam(
+  alist(
+    HR ~ binomial(p),
+    logit(p) = ,
+  ), data=slim, chains=4, cores=4
+)
+```
+
 # Comparison

### 2023-10-12 Stats Rethinking/Ch11/Practice Problems.qmd
Finished most easy problems
Finished through medium practice problems

@@
 
+```{r}
+logit(.35)
+```
+
 # 11E2
@@
 If and event has log-odds 3.2, what is the probability of this event?
+
+```{r}
+inv_logit(3.2)
+```
 
@@
 
+```{r}
+inv_logit(1.7)
+```
+
+A one unit change in the independent variable results in a 0.85 change in the dependent variable.
+
 # 11E4
@@
 Why do Poisson regressions sometimes require the use of an offset? Provide an example.
+
+Having an offset allows us to analyze the same count variable over different time frames.
+An example of this in the chapter was when the original monastery was doing so well that the author was considering buying a second one.
+Both produced books but the original monastery tracked by day while the second tracked by week.
+Thus, using the offset factor $\tau$, we can compare the two monasteries as if they were recorded at the same rate.
 
@@
 
-# 11M2
+Aggregated data will have a larger probability simply because it contains an extra factor in the log-probabilities not present for disaggregated data (339).
+
+# 11M2\*
 
@@
 
-# 11M3
+He never explicitly states coefficient interpretation in 11.2.
 
-Explain why the logit ink is appropriate for a binomial generalized linear model.
+# 11M3\*
 
-# 11M4
+Explain why the logit link is appropriate for a binomial generalized linear model.
+
+Logistic regression makes sense for binomial since in a logistic regression you can only have either a 0 or 1 for the final outcome. Thus modeling a binomial will follow this 0 or 1 trend.
+
+# 11M4\*
 
 Explain why the log link is appropriate for a Poisson generalized linear model.
+
+Possibly because a logarithmic function implies that the growth of the model parameter, this case $\lambda$ (346), will grow exponentially.
 
@@
 
+When priors are flat, then both distributions have flat priors
+
 # 11M7
@@
 Use `quap` to construct a quadratic approximate posterior distribution for the chimpanzee model that includes a unique intercept for each actor `m11.4`.
-Compare the quadratic approximation to the posterior distribution produced instead from MCMC/ Can you explain both the differences and the similarities between the approximate and the MCMC distributions?
+Compare the quadratic approximation to the posterior distribution produced instead from MCMC.
+Can you explain both the differences and the similarities between the approximate and the MCMC distributions?
 Relax the prior on the actor intercepts to Normal(0,10).
@@
 
-# 11M8
+## Prepare m11.4
+
+```{r}
+data(chimpanzees)
+d <- chimpanzees
+d$treatment <- 1+d$prosoc_left + 2*d$condition
+```
+
+```{r}
+dat_list <- list(
+  pulled_left = d$pulled_left,
+  actor = d$actor,
+  treatment = as.integer(d$treatment)
+)
+```
+
+```{r}
+m11.4 <- ulam(
+  alist(
+    pulled_left ~ dbinom(1,p),
+    logit(p) <- a[actor] + b[treatment],
+    a[actor] ~ dnorm(0,1.5),
+    b[treatment] ~ dnorm(0,0.5)
+  ), data=dat_list, chains=4, cores=4, log_lik = TRUE
+)
+```
+
+```{r}
+precis(m11.4, depth=2)
+```
+
+## Part 1
+
+```{r}
+m11.4.2 <- quap(
+  alist(
+    pulled_left ~ dbinom(1,p),
+    logit(p) <- a[actor] + b[treatment],
+    a[actor] ~ dnorm(0,1.5),
+    b[treatment] ~ dnorm(0,0.5)
+  ), data=dat_list
+)
+```
+
+```{r}
+precis(m11.4.2, depth=2)
+```
+
+In general, it appears that the coefficients of "a" generally became lower while the coefficients for "b" became higher.
+
+## Part 2
+
+```{r}
+m11.4.3 <- ulam(
+  alist(
+    pulled_left ~ dbinom(1,p),
+    logit(p) <- a[actor] + b[treatment],
+    a[actor] ~ dnorm(0,10),
+    b[treatment] ~ dnorm(0,0.5)
+  ), data=dat_list, chains=4, cores=4, log_lik = TRUE
+)
+```
+
+```{r}
+precis(m11.4.3, depth=2)
+```
+
+```{r}
+m11.4.4 <- quap(
+  alist(
+    pulled_left ~ dbinom(1,p),
+    logit(p) <- a[actor] + b[treatment],
+    a[actor] ~ dnorm(0,10),
+    b[treatment] ~ dnorm(0,0.5)
+  ), data=dat_list
+)
+```
+
+```{r}
+precis(m11.4.4, depth=2)
+```
+
+This greatly exaggerated the differences seen between the models. However, it seems to affect `quap` more than it did for `ulam`.
+
+# 11H1
 
@@
 
-# 11H1
+```{r}
+#Creat m11.3
+
+m11.3 <- quap(
+  alist(
+    pulled_left ~ dbinom(1,p),
+    logit(p) <- a + b[treatment],
+    a ~ dnorm(0,1.5),
+    b[treatment] ~ dnorm(0,0.5)
+  ), data=dat_list
+)
+```
+
+```{r}
+compare(m11.4, m11.3, func=WAIC)
+```
+
+Obviously, the MCMC model does much better than the `quap` models.
 
@@
 # 11H5
-
-# 11H6

### 2023-10-12 Lahman Analysis/Rough Analysis/BinomialRoughAnalysis.qmd
Began replication of Fisher Fellingham

@@
 
-# Quap
-
 # Ulam
@@
   HR = Batting$HR,
-  LG = Batting$lg
+  LG = Batting$lg,
+  YEAR = Batting$yearID -1944
 )
@@
 ```{r}
+#Recommended model format from Dr.Parson on 10/6/23
 u1 <- ulam(
   alist(
-    HR ~ binomial(p),
-    logit(p) = ,
+    HR ~ dbinom(AB,p),
+    logit(p) <- a[LG],
+    a[LG] ~ dnorm(0,10)
   ), data=slim, chains=4, cores=4
@@
 
-# Comparison
+```{r}
+precis(u1, depth=2)
+```
+
+```{r}
+#Since the mode says -3, lets refine the model a little bit
+u2 <- ulam(
+  alist(
+    HR ~ dbinom(AB,p),
+    logit(p) <- a[LG],
+    a[LG] ~ dnorm(0,2)
+  ), data=slim, chains=2, cores=2
+)
+```
+
+```{r}
+precis(u2, depth=2)
+```
+
+These "a" represent the logit probability change in hitting a homerun from each AB.
+To find this in probability terms, it would be `inv_logit(a[i])`, which becomes about 2%.
+And when running an LM we get a coefficient of about 0.03.
+Thus, this result is believable but on the higher end of what I would have expected.
+
+```{r}
+#Let's include year as a parameter on p since I've already assumed it mattered!
+u3 <- ulam(
+  alist(
+    HR ~ dbinom(AB,p),
+    logit(p) <- a[LG] + b[YEAR],
+    a[LG] ~ dnorm(0,2),
+    b[YEAR] ~ dnorm(0,10)
+  ), data=slim, chains=2, cores=2
+)
+```
+
+```{r}
+precis(u3, depth=2)
+```
+
+Is there a way I can extract this and view the walk of `b[i]` as it changes to see how home runs have changed over the years?
+
+```{r}
+trankplot(u3)
+```
+
+# Replicate
+
+The use the following model:
+
+$$
+HR_{i,j}\sim Binom(AB_{i,j}, p_{i,j})\\
+logit(p_{i,j})=\beta_1 Age+\beta_2 Ballpark+\beta_3 Year
+$$
+
+```{r}
+r1 <- ulam(
+  alist(
+    HR ~ dbinom(AB,p),
+    logit(p) <- a[LG],
+    a[LG] ~ dnorm(0,2)
+  ), data=slim, chains=2, cores=2
+)
+```
+
+One question I have here, why only choose one ballpark per player? Why not consider all ballparks the player played at in a season.
+
+I would want to run the following model
+
+$$
+HR_{i,j}\sim Binom(AB_{i,j}, p_{i,j})\\
+logit(p_{i,j})=\beta_1 Age+\beta_2 Ballpark + \beta_3Year+\beta_4SB
+$$
+
+Since stolen bases suggest a different archetype of player. One who's inherent underlying probability of hitting a homerun is lower.

### 2023-10-13 Lahman Analysis/Rough Analysis/BinomialRoughAnalysis.qmd
Added notes
From Meeting with Dr.Parson

@@
 library(Lahman)
+library(scales)
 ```
@@
   mutate(lg = ifelse(lgID=="NL",1,2))
+
+Batting2 <- Batting %>% 
+  filter(yearID >=1973 & yearID <=2021)
+```
+
+# From Meeting
+
+```{r}
+rate <- Batting %>% 
+  group_by(yearID) %>% 
+  summarize(sHR = sum(HR),
+            sAB = sum(AB)) %>% 
+  mutate(p = sHR / sAB,
+         se = sqrt(p*(1-p)/sAB))
+```
+
+```{r}
+ggplot(data=rate)+
+  aes(x=yearID, y=p)+
+  geom_point()+
+  scale_y_continuous(labels=label_percent(), limits = c(0,NA))
+```
+
+```{r}
+#rate2 <- 
+Batting2 %>% 
+  group_by(lgID) %>% 
+  summarize(sHR = sum(HR),
+            sAB = sum(AB)) %>% 
+  mutate(p = sHR / sAB,
+         se = sqrt(p*(1-p)/sAB))
+```
+
+```{r}
+rate2 <- Batting2 %>% 
+  group_by(yearID, lgID) %>% 
+  summarize(sHR = sum(HR),
+            sAB = sum(AB)) %>% 
+  mutate(p = sHR / sAB,
+         se = sqrt(p*(1-p)/sAB)) %>% 
+  select(yearID, lgID, p) %>% 
+  ungroup() %>% 
+  pivot_wider(names_from = lgID, values_from = p) %>% 
+  mutate(diff = AL - NL)
+```
+
+```{r}
+ggplot(data=rate2)+
+  aes(x=yearID, y=diff)+
+  geom_point()
+```
+
+```{r}
+ggplot(data=rate2)+
+  aes(x=diff)+
+  geom_histogram()
+```
+
+```{r}
+Batting3 <-  Batting %>% 
+  group_by(yearID, lgID) %>% 
+  summarize(sAB = sum(AB), sHR = sum(HR)) %>% 
+  mutate(year = yearID-1944,
+         lg = case_when(lgID=="NL" ~ 1,
+                        lgID=="AL" ~ 2))
+```
+
+```{r}
+slim2 <- list(
+  N = nrow(Batting3),
+  z = n_distinct(Batting3$yearID),
+  YEAR = Batting3 %>% pull(year),
+  AB = Batting3 %>% pull(sAB),
+  HR = Batting3 %>% pull(sHR),
+  LG = Batting3 %>% pull(lg)
+)
+
+str(slim2)
+```
+
+# Stan
+
+See Stan file
+
+```{r}
+fit <- stan(
+  file = "RoughBinomAnalysis.stan",
+  data = slim2,
+  chains = 4,
+  warmup = 1000,
+  iter= 5000,
+  cores = 4,
+  refresh=100
+)
+```
+
+```{r}
+plot(fit)
+```
+
+```{r}
+print(fit)
+```
+
+```{r}
+traceplot(fit)
+```
+
+```{r}
+saveRDS(fit, file="fit.rds")
+readRDS("fit.rds")
 ```
@@
   LG = Batting$lg,
+  #Need to subtract by 1944 to make it an index variable instead of a number
   YEAR = Batting$yearID -1944

### 2023-10-18 Lahman Analysis/Parson example/logistic_examples.qmd
Create example file with binomial models in Stan

---
title: "Logistic examples with Stan"
format: html
editor: visual
---

# Introduction

The bulk of this file explores fitting a baseball data set with a Bayesian method. The whole project is rather ridiculous, since there does not seem to be a good reason to fit the model. Simple work with proportions and classical binomial standard errors would be a better way to answer the questions.

Nonetheless, there may be some value in playing around this way. The present situation is simple and thus easy to understand, but it shows some issues that one may encounter in more complicated models.

In our model, we look at home-runs by season pooled by league. We use a binomial model for home runs, with a probability of an at-bat's resulting in a home run depending on the season and the league. The difficulty in fitting the model is that, in classical terms, the parameters are not identifiable: infinitely many sets of parameters give exactly the same probabilities for home runs. In Bayesian statistics, we have the extra tool of the prior, and using priors guided by our expectations about home-run probabilities we can (sort of) identify the parameters.

# Set-up

```{r}
library(tidyverse)
library(rstan)

library(Lahman)
```

It may be worthwhile to learn the `tidybayes` package, which systematizes a lot of graphing and data manipulation that we do when using Stan. It probably does automatically everything that I coded below by hand.

The following chunk has some recommended Stan options:

```{r}
options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)
```

The `rethinking` package includes an `inv_logit` function as below. It is handy to have around!

```{r}
inv_logit <- function(p){
  1 / (1 + exp(-p))
}
```

In this analysis of home runs, we focus on the seasons between 1973 and 1996, when only the AL had designated hitters, and when there was no regular interleague play. We fit a binomial-logistic model for home runs (specified further below) with separate constants for each league and for each season. We expect to see a slightly higher probability of home runs in the AL during these seasons.

When fitting the model, we pool the at-bats and home runs among all teams in the same league for each season. Separating the teams makes not difference for the posterior model, since the likelihood is the same whether we work one team at a time or after pooling the teams. Here is the basic calculation:

```         
p^d1 * (1 - p)^(n1 - d1) * p^d2 * (1 - p) ^ (n2 - d2) =
p^(d1 + d2) * (1 - p) ^ (n1 + n2 - (d1 + d2))
```

```{r}
Teams <- Lahman::Teams

DH.Teams <- Teams %>% 
  filter(yearID >= 1973 & yearID <= 1996)

DH.summary <- DH.Teams %>%
  group_by(yearID, lgID) %>%
  summarize(AB = sum(AB), HR = sum(HR)) %>%
  mutate(year = yearID - 1972,
         league = case_when(lgID == "NL" ~ 1,
                            lgID == "AL" ~ 2))
```

Here is the model that we will use:

```         
p, a[1], a[2], b[1], ..., b[24] parameters

for i in 1:24, HR[i] ~ binomial(AB[i], inv_logit(a[league[i]] + b[year[i]]))

a[1] ~ normal(-3.5, 0.1)
a[2] ~ normal(-3.5, 0.1)

for i in 1:24, b[i] ~ normal(0, 0.1)
```

The difficulty with the model is that if we shift the `a` parameters by a constant `c` and the `b` parameters by `-c`, we get the same binomial probabilities. In classical statistics, there would be no way to separate (or "identify") the parameters, since the likelihood is constant along these shifts. We can imagine an infinite ridge of maximum likelihood---no way to pick out a single combination of parameter values maximizing the likelihood.

The way to fix this problem in classical statistics is to reduce the number of parameters. Pick a reference level, maybe AL in the first season. Introduce an overall constant for the probability at the reference level, and use 24 comparison parameters for the NL versus AL and for the other seasons versus season 1. Following McElreath's advice (and the advice of some other Bayesians), we will keep what looks like one parameter too many from the classical perspective.

The Bayesian solution to the identification problem is to use priors. The priors that we have specified will help with this problem (but not solve it completely).

The priors require some explanation:

-   We use the league coefficients `a[1]` (NL) and `a[2]` (AL) as the main terms. We know that home runs are rare, maybe around 3% of at-bats. We use `logit(0.03)` or -3.5 for the center of the prior distribution. Then we allow a rather wide range for this overall probability. On the probability, our prior spans from around `inv_logit(-3.7)` (about 0.024) to around `inv_logit(-3.3)` (about 0.036).

-   We think of the season coefficients `b[i]` as adjustments to the overall league average. To help us think about these priors, we recall a bit about odd and logits:

    -   For rare events (small probability `p`) probability `p` and odds `p / (1 - p)` are roughly equal.

    -   An adjustment `b[i]` to `logit(p)` translates to scaling the odds by `exp(b[i])`. For small `b[i]`, this means scaling the odds by a factor of about `1 + b[i]` or by `100b[i]%`. When the probability `p` is small, this means (roughly) scaling the probability by `100b[i]%`.

-   With this in mind, we set prior on the `b[i]` to allow for variations in the HR probability of +/- 20%. Of course, the model will learn from the data! We'll see if the prior is too strong by testing it on some data.

This model would be a good candidate for a *multilevel* treatment. The coefficients `b[i]` come as a batch and might profitably be modeled that way. We can try that too.

# Running the model on fake data

Before we fit the model to our real data set, let's simulate some data (with known parameters) to see how well the model does recovering the known parameters.

In the chunk to simulate data, our `a` values match what we expect from the real data (and the priors of our model). The `b` values, however, have a bit more variability than we put in our priors. We can see what happens!

I had to work a bit to get the column of fake binomial data. My first attempts generated errors, which is why I ended up with `add_column`.

```{r}
fake_a <- rep(NA, 2)
fake_a[1] <- -3.6
fake_a[2] <- -3.5
fake_b <- rnorm(24, 0, 0.3)

summary_fake <- DH.summary %>%
  mutate(probs = inv_logit(fake_a[league] + fake_b[year]))

summary_fake <- summary_fake %>%
  add_column(fake_HR = rbinom(nrow(DH.summary), summary_fake$AB, summary_fake$probs))

test_data <- list(
  N = nrow(summary_fake),
  z = n_distinct(summary_fake %>% pull(year)),
  AB = summary_fake %>% pull(AB),
  year = summary_fake %>% pull(year),
  HR = summary_fake %>% pull(fake_HR),
  league = summary_fake %>% pull(league)
)
```

The Stan program `HR1.stan` has the model that I described above.

```{r}
fake_fit1 <- stan(
  file = "HR1.stan",
  data = test_data,
  chains = 4,
  warmup = 1000,
  iter = 5000,
  cores = 4,
  refresh = 0
)
```

We can look at the posterior distributions for the parameters:

```{r}
print(fake_fit1)
```

And at the trace plot:

```{r}
traceplot(fake_fit1)
```

We can make graphs of the known parameters and of the fitted values.

```{r}
post_fake1 <- fake_fit1 %>% (rstan :: extract)
fake_fit_as <- post_fake1$a %>% as_tibble %>% rename(NL = V1, AL = V2)

fake_as_sum <- fake_fit_as %>%
  pivot_longer(c(AL, NL), names_to = "league", values_to = "sample") %>%
  group_by(league) %>% summarize(med = median(sample), low = quantile(sample, probs = 0.025),
                                 high = quantile(sample, probs = 0.975))

ggplot(data = fake_as_sum) +
  geom_pointrange(mapping = aes(x = med, y = league, xmin = low, xmax = high)) +
  geom_point(mapping = aes(x = c(-3.6, -3.5), y = c(2, 1)), color = "red", size = 3)
```

```{r}
fake_fit_bs <- post_fake1$b %>% as_tibble

fake_bs_sum <- tibble(coeff = factor(1:24),
                      true_param = fake_b,
                      low = rep(NA, 24),
                      med = rep(NA, 24),
                      high = rep(NA, 24))

for(i in 1:24){
  bb <- post_fake1$b[,i]
  fake_bs_sum$low[i] <- quantile(bb, 0.025)
  fake_bs_sum$med[i] <- median(bb)
  fake_bs_sum$high[i] <- quantile(bb, 0.975)
}

ggplot(data = fake_bs_sum) +
  geom_pointrange(mapping = aes(x = med, y = coeff, xmin = low, xmax = high)) +
  geom_point(mapping = aes(x = fake_b, y = coeff), color = "red", size = 3) +
  scale_x_continuous(breaks = c(-0.75, -0.5, -0.25, 0, 0.25, 0.5, 0.75))
```

At least in some runs, the model looks overconfident. The true parameter values seem to often to be outside of the ranges of typical samples.

We see from the graphs (which depend on the randomly assigned values for the `b` coefficients) that this model cannot fully separate the `a` and `b` coefficients. Either it underestimates both `a` coefficients and overestimates all `b` coefficients (by the same amount) or vice versa. Nonetheless, it recovers `a + b`, as we see in the following graphs.

Note in the graphs that our model seems to regularize, pulling the estimates toward the center.

```{r}
fake_NLbs_sum <- tibble(coeff = factor(1:24),
                      true_param = fake_b,
                      low = rep(NA, 24),
                      med = rep(NA, 24),
                      high = rep(NA, 24))

for(i in 1:24){
  bb <- post_fake1$a[,1] + post_fake1$b[,i]
  fake_NLbs_sum$low[i] <- quantile(bb, 0.025)
  fake_NLbs_sum$med[i] <- median(bb)
  fake_NLbs_sum$high[i] <- quantile(bb, 0.975)
}

ggplot(data = fake_NLbs_sum) +
  geom_pointrange(mapping = aes(x = med, y = coeff, xmin = low, xmax = high)) +
  geom_point(mapping = aes(x = fake_b - 3.6, y = coeff), color = "red", size = 3) +
  labs(x = "", y = "season", title = "Fitted middle 95% (black) and true parameter (red)",
       subtitle = "logit probability of HR for NL at-bats")
```

```{r}
fake_ALbs_sum <- tibble(coeff = factor(1:24),
                      true_param = fake_b,
                      low = rep(NA, 24),
                      med = rep(NA, 24),
                      high = rep(NA, 24))

for(i in 1:24){
  bb <- post_fake1$a[,2] + post_fake1$b[,i]
  fake_ALbs_sum$low[i] <- quantile(bb, 0.025)
  fake_ALbs_sum$med[i] <- median(bb)
  fake_ALbs_sum$high[i] <- quantile(bb, 0.975)
}

ggplot(data = fake_ALbs_sum) +
  geom_pointrange(mapping = aes(x = med, y = coeff, xmin = low, xmax = high)) +
  geom_point(mapping = aes(x = fake_b - 3.5, y = coeff), color = "red", size = 3) +
  labs(x = "", y = "season", title = "Fitted middle 95% (black) and true parameter (red)",
       subtitle = "logit probability of HR for AL at-bats")
```

One way to understand how the coefficients combine to something more certain is to graph samples for one of the `b` coefficients versus samples for one of the `a` coefficients.

```{r}
ggplot() +
  geom_point(mapping = aes(x = post_fake1$a[,1], y = post_fake1$b[,1])) +
  labs(x = "samples for a[1] (NL)", y = "samples for b[1] (season 1)")
```

We see a strong correlation, showing the problematic multicolinearity.

The fact that the model recovers well the true `a + b` parameter values means that it will generate predictions similar to the process that generated the original data: only the sums `a + b` in the model appear in the formulas that we would use to generate new data.

Judging from this simulation, we will not take the parameter values for such a model too seriously and we will expect the sort of multicolinearity visible above. But we can trust the model to make reasonable predictions.

## A multilevel model

As a variant, we can use a *multilevel model*, whose structure matches the way that we created our fake data. We model the `b` coefficients as random draws from a single `normal(mu, sigma)` distribution. We use the data to fit the *hyperparameters* (really just more parameters) `mu` and `sigma` as well as the `b` coefficients.

Here is the model:

```         
p, a[1], a[2], mu, sigma, b[1], ..., b[24] parameters

a[i] ~ normal(-3.5, 0.1)

mu ~ normal(0, 0.1)
sigma ~ exponential(10)

b[i] ~ normal(mu, sigma)

HR[i] ~ binomial(AB[i], inv_logit(a[league[i]] + b[year[i]]))
```

The prior for `mu` matches (I think!) our informal reasoning above. This kind of model is harder for Stan to sample.

```{r}
fit_multilevel_fake <- stan(
  file = "HR_multi.stan",
  data = test_data,
  chains = 4,
  warmup = 1000,
  iter = 7000,
  cores = 4,
  refresh = 1000
)
```

```{r}
print(fit_multilevel_fake)
```

Note that `mu` and `sigma` do roughly match the parameters that I used to generate the `b` coefficients.

We can make the same kind of graphs as before. The model fit does not seem better, and the posterior distributions are not as tight. We have the same issues coming from multicolinearity.

```{r}
post_multi <- fit_multilevel_fake %>% (rstan :: extract)
fake_multi_as <- post_multi$a %>% as_tibble %>% rename(NL = V1, AL = V2)

fake_multi_as_sum <- fake_multi_as %>%
  pivot_longer(c(AL, NL), names_to = "league", values_to = "sample") %>%
  group_by(league) %>% summarize(med = median(sample), low = quantile(sample, probs = 0.025),
                                 high = quantile(sample, probs = 0.975))

ggplot(data = fake_multi_as_sum) +
  geom_pointrange(mapping = aes(x = med, y = league, xmin = low, xmax = high)) +
  geom_point(mapping = aes(x = c(-3.6, -3.5), y = c(2, 1)), color = "red", size = 3)
```

```{r}
fake_multi_bs <- post_multi$b %>% as_tibble

fake_multi_bs_sum <- tibble(coeff = factor(1:24),
                      true_param = fake_b,
                      low = rep(NA, 24),
                      med = rep(NA, 24),
                      high = rep(NA, 24))

for(i in 1:24){
  bb <- post_multi$b[,i]
  fake_multi_bs_sum$low[i] <- quantile(bb, 0.025)
  fake_multi_bs_sum$med[i] <- median(bb)
  fake_multi_bs_sum$high[i] <- quantile(bb, 0.975)
}

ggplot(data = fake_multi_bs_sum) +
  geom_pointrange(mapping = aes(x = med, y = coeff, xmin = low, xmax = high)) +
  geom_point(mapping = aes(x = fake_b, y = coeff), color = "red", size = 3) +
  scale_x_continuous(breaks = c(-0.75, -0.5, -0.25, 0, 0.25, 0.5, 0.75))
```

The following graph shows the multicollinearity that makes it hard to recover individual coefficients. As before, though, we recover the sums `a + b` that contribute directly to observable quantities (numbers of home runs).

```{r}
ggplot() +
  geom_point(mapping = aes(x = post_multi$a[,1], y = post_multi$b[,1])) +
  labs(x = "samples for a[1] (NL)", y = "samples for b[1] (season 1)")
```

Probably the AL graph is sufficient:

```{r}
fake_ALbs_msum <- tibble(coeff = factor(1:24),
                      true_param = fake_b,
                      low = rep(NA, 24),
                      med = rep(NA, 24),
                      high = rep(NA, 24))

for(i in 1:24){
  bb <- post_multi$a[,2] + post_multi$b[,i]
  fake_ALbs_msum$low[i] <- quantile(bb, 0.025)
  fake_ALbs_msum$med[i] <- median(bb)
  fake_ALbs_msum$high[i] <- quantile(bb, 0.975)
}

ggplot(data = fake_ALbs_msum) +
  geom_pointrange(mapping = aes(x = med, y = coeff, xmin = low, xmax = high)) +
  geom_point(mapping = aes(x = fake_b - 3.5, y = coeff), color = "red", size = 3) +
  labs(x = "", y = "season", title = "Fitted middle 95% (black) and true parameter (red)",
       subtitle = "logit probability of HR for AL at-bats")
```

# Trying the model on the actual data

```{r}
real_data <- list(
  N = nrow(DH.summary),
  z = n_distinct(DH.summary %>% pull(year)),
  AB = DH.summary %>% pull(AB),
  year = DH.summary %>% pull(year),
  HR = DH.summary %>% pull(HR),
  league = DH.summary %>% pull(league)
)

fit1 <- stan(
  file = "HR1.stan",
  data = real_data,
  chains = 4,
  warmup = 1000,
  iter = 4000,
  cores = 4,
  refresh = 1000
)
```

```{r}
print(fit1)
```

Stan model objects have plot methods. It may even be possible to use these methods to create the kinds of graphs that I made by hand above. (The documentation suggests that one can get ggplot objects from this command and modify them as needed.)

```{r}
plot(fit1, pars = "a")
```

```{r}
plot(fit1, pars = "b")
```

```{r}
post1 <- (rstan :: extract)(fit1)
```

As in our analysis of the fake data, we see strong correlations between the posterior distributions. The multicollinearity limits our ability to estimate individual parameters.

```{r}
ggplot() +
  geom_point(mapping = aes(x = post1$a[,1], y = post1$b[,1]))
```

## The DH effect

Oh, yes! We should look at the average designated-hitter effect in this model. (But see below for how one should actually do this using simpler models.)

```{r}
ggplot() +
  geom_histogram(mapping = aes(x = post1$a[,2] - post1$a[,1],
                               y = ..density..), bins = 50) +
  labs(x = "posterior difference in logit HR probability (AL - NL)")
```

Using the principles for interpreting differences on the logit scale mentioned above, we can say that in these seasons where only the AL allowed designated hitters, the home-run probability was about 16% higher (give or take 1 percentage point) for AL.

```{r}
quantile(post1$a[,2] - post1$a[,1], c(0.025, 0.25, 0.5, 0.75, 0.975)) %>% round(3)
```

We could be a bit more scientific by calculating the posterior intervals as above, but I don't think it adds much.

Instead of using the approximations to move from differences on the logit scale to ratios of probabilities, we could calculate directly with probabilities, either using differences, as the first histogram, or ratios, as in the second histogram.

```{r}
ggplot() +
  geom_histogram(mapping = aes(x = inv_logit(post1$a[,2]) - inv_logit(post1$a[,1]),
                               y = ..density..), bins = 50) +
  labs(x = "posterior difference in HR probability (AL - NL)")
```

```{r}
ggplot() +
  geom_histogram(mapping = aes(x = inv_logit(post1$a[,2]) / inv_logit(post1$a[,1]),
                               y = ..density..), bins = 50) +
  labs(x = "posterior ratio of HR probability (AL / NL)")
```

As expected, the histogram for ratios matches closely the histogram for differences on the logit scale. There is a shift of 1 in the horizontal scale, since a difference of x on the logit scale corresponds roughly to scaling the (small) probability by a factor of 1 + x.

## Multilevel model

We can also fit the multilevel model using the real data.

```{r}
fit_multilevel <- stan(
  file = "HR_multi.stan",
  data = real_data,
  chains = 4,
  warmup = 1000,
  iter = 10000,
  cores = 4,
  refresh = 1000
)
```

```{r}
print(fit_multilevel)
```

```{r}
plot(fit_multilevel, pars = "a")
```

```{r}
plot(fit_multilevel, pars = "b")
```

```{r}
plot(fit_multilevel, pars = c("mu", "sigma"))
```

# Enough with all this silliness...

All of that work is rather ridiculous. Let's try some more straightforward things, for which we barely need models.

If our goal is simply to compare AL and NL home-run probabilities during this period, we could pool all the data for the 24 years as follows.

```{r}
total_pooled <- DH.summary %>% group_by(league) %>%
  summarize(HR = sum(HR), AB = sum(AB))
```

Our model is the following:

```         
p[1], p[2] parameters constrainted to be between 0 and 1.

HR[1] ~ binomial(AB[1], p[1])
HR[2] ~ binomial(AB[2], p[2])
```

Because we are just estimating probabilities, we do not need to use the `logit` transformation, though it would not hurt to do so.

Since we have so much data, uniform priors---not necessary to specify in Stan---seem fine. If we wanted to set a prior, we could, say, use a normal distribution centered at 0.03 with standard deviation 0.01 and truncated at 0 and 1. Or we could look up other options such as beta distributions. (Beta distributed random variables have range \[0, 1\]. One can look them up on Wikipedia or in many probability textbooks.)

```{r}
pooled_data <- list(
  AB = total_pooled %>% pull(AB),
  HR = total_pooled %>% pull(HR)
)
```

For the following Stan call, I just use the defaults. They work here! It's good to remember that one can make a simple call like this without typing everything. It may actually be possible to remember these two mandatory options...

```{r}
fit_pooled <- stan(
  file = 'pooled_model.stan',
  data = pooled_data
)
```

We can summarize the marginal posterior distributions (i.e. separate distributions for the two parameters) in a table with `summary` and with `plot`:

```{r}
summary(fit_pooled, pars = "p")$summary %>% signif(3)
```

```{r}
plot(fit_pooled)
```

In this case, we can get the same estimates using classical statistical methods. We estimate the probabilities as home-run proportions by league and use binomial standard errors (`sqrt(p (1 - p) / n)`). The numbers are practically the same as what we estimate from our MCMC samples. (We could even get analytic formulas for the priors in this simple situation.) The Bayesians would still advocate (of course) for Bayesian methods, since it is more straightforward to interpret the results.

```{r}
total_pooled %>% group_by(league) %>%
  mutate(prop = HR / AB, se = sqrt(prop * (1 - prop) / AB)) %>%
  select(league, prop, se) %>% signif(3)
```

We can also use the MCMC samples to summarize the comparison between AL and NL home-run probabilities in our model. The posterior distribution is similar to what we saw before when we compared overall AL and NL home-run probabilities using our more-complex model. Note that the *precision* for this parameter (ratio of probabilities) is no better here than it was in the earlier calculations, even though we can estimate the individual probabilities better. (The posterior distributions of the two probabilities are strongly correlated, and so MCMC samples with high AL probability, for example, also tend to have high NL probability, and so the ratios of the probabilities are more consistent among samples than the individual probabilities.)

```{r}
post_pooled <- as.data.frame(fit_pooled, pars = "p")

ggplot() +
  geom_histogram(mapping = aes(x =
                                 post_pooled %>% pull("p[2]") /
                                 post_pooled %>% pull("p[1]"),
                               y = ..density..),
                 bins = 50) +
  labs(x = "posterior ratio of HR probability (AL / NL)")
```

To get a better sense for the difference in home-run probabilities, let's fit a new model with two probability parameters for each season. We'll set up the data as follows:

```{r}
year_by_year_data <- list(
  al_ab = DH.summary %>% filter(league == 2) %>% pull(AB),
  nl_ab = DH.summary %>% filter(league == 1) %>% pull(AB),
  al_hr = DH.summary %>% filter(league == 2) %>% pull(HR),
  nl_hr = DH.summary %>% filter(league == 1) %>% pull(HR)
)
```

And we fit the model using defaults, which should be fine here.

```{r}
fit_year_by_year <- stan(
  file = "year_by_year.stan",
  data = year_by_year_data
)
```

The following (overwhelming) summary describes the fitted model. In the test run that I made, I got many Rhat values less than 1, which is odd. I read, though, on the Stan forum that we should not worry about it.

```{r}
summary(fit_year_by_year, pars = c("p_al", "p_nl"))$summary %>% round(4)
```

We can also summarize with Stan's default parameter plots.

```{r}
plot(fit_year_by_year, pars = "p_nl")
```

```{r}
plot(fit_year_by_year, pars = "p_al")
```

The next chunk calculates posterior samples for the ratio of HR probabilities (AL / NL) year by year and summarize the posteriors using median and quartiles. The following chunk graphs these summary statistics.

```{r}
post_year_by_year <- (rstan :: extract)(fit_year_by_year)

diffs <- post_year_by_year$p_al / post_year_by_year$p_nl

diffs2 <- as_tibble(diffs) %>%
  pivot_longer(cols = everything(), names_prefix = "V",
               names_to = "season") %>%
  mutate(season = as.integer(season))

mean_diffs <- diffs2 %>% group_by(season) %>%
  summarize(median = median(value),
            Q1 = quantile(value, 0.25),
            Q3 = quantile(value, 0.75))
```

```{r}
ggplot(data = mean_diffs) +
  geom_pointrange(mapping = aes(x = season, y = median, ymin = Q1, ymax = Q3)) +
  labs(x = "season", y = "ratio of HR probability (AL / NL)")
```

The year-by-year graph shows an interesting pattern. The ratio of probabilities grew in the first 10 seasons to about 1.35. Te pattern of growth was quite consisten between seasons 4 and season 10. The ration then declined through season 17 and was variable, centered around 1.1, between seasons 17 and 24.

To understand better what is going on, I would expand the analysis, including a few more seasons on either side, marking the start and end of this 24-season period in all my graphs. The context of those earlier seasons could help us interpret the probabilities in this DH period. I am also curious what would happen if we fit a multilevel model here, modeling the home-run probabilities as draws from a common distribution (normal, let's say) whose parameters we fit from the data instead of unrelated quantities. That kind of model could smooth the fit somewhat. Not really sure what to expect.

But I'll stop here for now!

### 2023-10-19 Lahman Analysis/Rough Analysis 2/Analysis2.qmd
Started more complex model

---
title: "Analysis2"
format: html
---

# Load Libraries

```{r}
#| warning: false
library(tidyverse)
library(rstan)
library(Lahman)
```

# Clean Lahman

```{r}
#| message: false

#Creates a DF called "Batting" which has players stats from 1973-2019

#Load the DF's from Lahman directly
Batting <- Lahman::Batting
Teams <- Lahman::Teams
People <- Lahman::People

Batting <- Batting %>% 
  #Minimum year = DH was introduced; maximum year = before COVID
  filter(yearID >= 1973 & yearID <= 2019)

#Filter teams for years considered and select only necessary columns
Teams <- Teams %>% 
  filter(yearID>=1973 & yearID <= 2019) %>% 
  select(yearID, teamID, park)

#Combine current "Batting" and "Teams" DF to associate player with stadium played at
Batting <- Batting %>% 
  right_join(Teams, by = join_by(yearID, teamID))

#Add player's age to the "Batting" DF
Batting <- Batting %>% 
  right_join(Lahman::People, by = join_by(playerID)) %>% 
  mutate(Age = yearID - birthYear, .after=yearID) %>%
  #Don't consider the people who don't have a birthday in the database
  filter(!is.na(Age))

#Make the lgID variable into an index variable
Batting <- Batting %>% 
  mutate(lg = case_when(lgID == "NL" ~ 1,
                        lgID == "AL" ~ 2), .after=lgID)
```

```{r}
#We need to make the data here "slim" like is suggested by McElreath
slim <- list(
  N = nrow(Batting),
  z = n_distinct(Batting %>% pull(yearID)),
  Year = as.integer(Batting %>% pull(yearID) - 1972),
  Age = as.integer(Batting %>% pull(Age)-17),
  LG = as.integer(Batting %>% pull(lg)),
  AB = Batting %>% pull(AB),
  HR = Batting %>% pull(HR),
  SB = Batting %>% pull(SB),
  BirthYear = as.integer(Batting %>% pull(birthYear)-1922)
)
```

# Theoretical Model

The model used by Fellingham & Fisher (2017) is structured as,

$$
HR_{i,j}|p_{i,j}\sim Binom(AB_{i,j},p_{i,j})\\
logit(p_{i,j})=x_{i,j}\beta_i+\theta_i+\delta_j+\xi_p
$$

Where $\theta_i$ is the effect of decade of birth, $\delta_j$ is the effect of season of play, and $\xi_p$ is the effect of home ballpark.

However, I think that this model could be improved and should be structured as,

$$
HR_{i,j}\sim Binom(AB_{i,j},p_{i,j})\\
logit(p_{i,j})=\beta_1Age+\beta_2Ballpark+\beta_3Year+\beta_4SB+\beta_5(Age\cdot Year)
$$

All of these variables should influence the underlying probability of a player to hit a home run. In addition, I think that Age and Year played will have some sort of interaction effect. Possibly not during the small time frame that we consider here, however over a longer study I believe it should.

# Empirical Model

```{r}
fit1 <- stan(
  file = "HR21.stan",
  data = slim,
  chains = 4,
  warmup = 1000,
  iter = 4000,
  cores = 4,
  refresh = 1000
)
```

### 2023-10-19 Lahman Analysis/Rough Analysis/CleanLahman.qmd
Started more complex model

---
title: "Clean Lahman"
format: html
---

# Load Libraries

```{r}
#| message: false

library(tidyverse)
library(Lahman)
```

# Clean

```{r}
#| message: false

#Creates a DF called "Batting" which has players stats from 1973-2019

#Load the DF's from Lahman directly
Batting <- Lahman::Batting
Teams <- Lahman::Teams
People <- Lahman::People

Batting <- Batting %>% 
  #Minimum year = DH was introduced; maximum year = before COVID
  filter(yearID >= 1973 & yearID <= 2019)

#Filter teams for years considered and select only necessary columns
Teams <- Teams %>% 
  filter(yearID>=1973 & yearID <= 2019) %>% 
  select(yearID, teamID, park)

#Combine current "Batting" and "Teams" DF to associate player with stadium played at
Batting <- Batting %>% 
  right_join(Teams, by = join_by(yearID, teamID))

#Add player's age to the "Batting" DF
Battin <- Batting %>% 
  right_join(Lahman::People, by = join_by(playerID)) %>% 
  mutate(Age = yearID - birthYear, .after=yearID) %>%
  #Don't consider the people who don't have a birthday in the database
  filter(!is.na(Age))
```

# Cleaning

```{r}
Batting <- Lahman::Batting

Batting <- Batting %>% 
  #Minimum year = DH was introduced; maximum year = before COVID
  filter(yearID >= 1973 & yearID <= 2019)

Batting
```

```{r}
Teams <- Lahman::Teams

Teams <- Teams %>% 
  filter(yearID>=1973 & yearID <= 2019) %>% 
  select(yearID, teamID, park)
```

```{r}
Batting <- Batting %>% 
  right_join(Teams, by = join_by(yearID, teamID))
```

```{r}
Batting %>% 
  right_join(Lahman::People, by = join_by(playerID)) %>% 
  mutate(Age = yearID - birthYear, .after=yearID) %>%
  #Don't consider the people who don't have a birthday in the database
  filter(!is.na(Age))
```

# Dead Ends

```{r}
#Delets the Wrigley field in LA

Parks <- Lahman::Parks %>% 
  filter(park.key != "LOS02")
```

```{r}
#This does not work. Stupid things like "Fenway II" are in there

#TeamID <- 
Teams %>% 
  filter(yearID >= 1973 & yearID <= 2019) %>% 
  select(yearID, teamID, name, park) %>% 
  left_join(Parks, by = join_by(park == park.name)) %>% 
  filter(is.na(state))
```

```{r}
#Manually replace the fields so merge with Parks is possible
Teams <- Lahman::Teams %>% 
  filter(yearID >= 1973 & yearID <= 2019) %>% 
  select(yearID, teamID, name, park) %>% 
  mutate(park = case_when(park=="Fenway Park II" ~ "Fenway Park",
                          park=="Anaheim Stadium" ~ "Angel Stadium of Anaheim",
                          park=="Comiskey Park" ~ "Comiskey Park I",
                          park=="Riverfront Stadium" ~ "Cinergy Field",
                          park=="Royals Stadium" ~ "Kauffman Stadium",
                          park=="Jarry Park" ~ "Parc Jarry",
                          park=="Oakland Coliseum" ~ "Oakland-Alameda County Coliseum",
                          park=="Jack Murphy Stadium" ~ "Qualcomm Stadium",
                          park=="Hubert H Humphrey Metrodome" ~ "Hubert H. Humphrey Metrodome",
                          park=="Exhibition Stadium /Skydome" ~ "Exhibition Stadium",
                          park=="Skydome" ~ "Rogers Centre",
                          park=="Comiskey Park II" ~ "Guaranteed Rate Field",
                          park=="Joe Robbie Stadium" ~ "Sun Life Stadium",
                          park=="Jacobs Field" ~ "Progressive Field",
                          park=="The Ballpark at Arlington" ~ "Rangers Ballpark in Arlington",
                          park=="3Com Park" ~ "Candlestick Park",
                          .default = park))
```

```{r}
Batting %>% 
  right_join(Lahman::Teams %>% filter(yearID>=1973 & yearID <= 2019) %>% select(yearID, teamID, park), by = join_by(yearID, teamID))
```

```{r}
Lahman::People
```

### 2023-10-20 Lahman Analysis/Rough Analysis 2/Analysis2.qmd
More complex model made
Took an hour to make

@@
 
+```{r}
+#Define logit for convenience
+
+inv_logit <- function(p){
+  1 / (1 + exp(-p))
+}
+```
+
 # Theoretical Model
@@
   warmup = 1000,
-  iter = 4000,
+  iter = 1000,
   cores = 4,
@@
 ```
+
+```{r}
+plot(fit1, pars = "a")
+```
+
+```{r}
+plot(fit1, pars = "b")
+```
+
+```{r}
+fit2 <- stan(
+  file = "HR22.stan",
+  data = slim,
+  chains = 1,
+  warmup = 1000,
+  iter = 4000,
+  cores = 1,
+  refresh = 1000
+)
+```
+
+```{r}
+plot(fit2, pars = "a")
+```
+
+```{r}
+plot(fit2, pars = "b")
+```
+
+```{r}
+plot(fit2, pars = "c")
+```

### 2023-10-20 Lahman Analysis/Rough Analysis 2/Analysis2.qmd
Add simple model (which compiles!) to file
I added a simple model with just `Age` as a predictor so that we could compile something. Before fitting the model, I made a graph for guidance. It would be good to explore this simple model further, including a prior predictive check and a posterior predictive check. At very least, there should be a graph of the model. (Maybe I'll add that next.)

@@
 library(Lahman)
+```
+
+```{r}
+options(mc.cores = parallel::detectCores())
+rstan_options(auto_write = TRUE)
 ```
@@
   #Don't consider the people who don't have a birthday in the database
-  filter(!is.na(Age))
+  drop_na(Age)
 
@@
 
+# Graphical exploration
+
+Before building a model, we explore the data by graphing the home-run proportion with classical binomial standard errors.
+
+```{r}
+Batting_summary <- Batting %>% group_by(Age) %>%
+  summarize(HR = sum(HR), AB = sum(AB)) %>%
+  mutate(prop = HR / AB, se = sqrt(prop*(1-prop)/AB))
+
+ggplot(data = Batting_summary) +
+  geom_pointrange(mapping = aes(x = Age, y = prop, ymin = prop - se,
+                                ymax = prop + se))
+```
+Let's graph again, restricting to players aged 20 to 40.
+
+```{r}
+ggplot(data = Batting_summary %>% filter(Age >= 20 & Age <= 40)) +
+  geom_pointrange(mapping = aes(x = Age, y = prop, ymin = prop - se,
+                                ymax = prop + se))
+```
+Our first model will include a linear term for age and will be restricted to this age range.
+
+# Simple model
+
+```{r}
+Batting_summary_restricted <- Batting_summary %>%
+  filter(Age >= 20 & Age <= 40) %>% mutate(c.Age = Age - 30)
+
+simple_data <-   list(
+  N = nrow(Batting_summary_restricted),
+  Age = Batting_summary_restricted %>% pull(c.Age),
+  AB = Batting_summary_restricted %>% pull(AB),
+  HR = Batting_summary_restricted %>% pull(HR)
+)
+```
+
+```{r}
+glm(cbind(HR, AB - HR) ~ c.Age, data = Batting_summary_restricted, family = binomial)
+```
+
+```{r}
+fit_simple <- stan(
+  file = "HR2_simple.stan",
+  data = simple_data
+)
+```
+
+```{r}
+summary(fit_simple, pars = c("beta0", "beta1"))$summary %>% signif(3)
+```
+```{r}
+plot(fit_simple, pars = c("beta0", "beta1"))
+```
+
 # Empirical Model

### 2023-10-20 Lahman Analysis/Rough Analysis 2/Analysis2.qmd
Add some explanation for simple model and graph of posterior fit
I added a graph of the posterior fit for the simple model that I proposed. (It's bad.) There may be a package (tidybayes?) that automates making this sort of graph. I planned to make a graph on the probability scale (instead of logit scale), but, as far as I could tell, geom_function only plots one function at a time. The geom_abline layer, on the other hand takes vectors of slopes and intercepts.

@@
 
+We start with a frequentist fit using maximum likelihood.
+
 ```{r}
 glm(cbind(HR, AB - HR) ~ c.Age, data = Batting_summary_restricted, family = binomial)
+```
+And now let's fit using the following simple model:
+
+```
+data:
+N = number of ages (21 between 20 and 40)
+AB = at-bats, pooled for each age
+HR = home runs, pooled for each age
+Age = age, recentered at 30 for easier interpretation of intercept
+
+parameters:
+beta0 (intercept for logistic model)
+beta1 (slope for logistic model)
+
+model:
+
+For n in 1:N,
+
+HR[n] ~ binomial(AB[n], inv_logit(beta0 + beta1 * Age[n]))
 ```
@@
 
+The fitted parameters match the frequentist maximum likelihood parameters. (But there now no mystery about the meaning of the standard errors.)
+
 ```{r}
@@
 ```
+
+We can make a plot of the parameters, but there is not much to see, since we can estimate the parameters quite precisely thanks to the huge data set.
+
 ```{r}
 plot(fit_simple, pars = c("beta0", "beta1"))
+```
+
+We can get a sense for the fitted model by sampling from the posterior.
+We take 10 random rows and plot the model lines with the empirical logits of the proportions.
+Yuck!
+This model does not fit the data well at all.
+Room for improvement!
+(Quadratic model...
+Splines...)
+
+```{r}
+logit <- function(p) log( p / (1 - p))
+
+post_simple <- as.data.frame(fit_simple) %>% select(beta0, beta1) %>%
+  slice_sample(n = 10)
+
+ggplot(data = Batting_summary_restricted) +
+  geom_point(mapping = aes(x = c.Age, y = logit(prop))) +
+  geom_abline(slope = post_simple$beta1, intercept = post_simple$beta0,
+              alpha = 0.3)
 ```

### 2023-10-20 Lahman Analysis/Rough Analysis 2/Analysis2.qmd
Add logistic-binomial model HR23 with varying intercept (by year) and constant slope (for age)

@@
 ```
+
+```{r}
+data_23 <- list(
+  N = nrow(Batting),
+  n_year = n_distinct(Batting$yearID),
+  AB = Batting %>% pull(AB),
+  HR = Batting %>% pull(HR),
+  Year = Batting %>% pull(yearID) - 1972,
+  Age = Batting %>% pull(Age) - 30
+)
+```
+
+```{r}
+fit_23 <- stan(
+  file = "HR23.stan",
+  data = data_23
+)
+```
+```{r}
+summary(fit_23)$summary %>% signif(3)
+```
+
+```{r}
+plot(fit_23, pars = "b")
+```
+
+```{r}
+saveRDS(fit_23, "HR23_new.rds")
+```
+
+```{r}
+traceplot(fit_23)
+```
+
+```{r}
+sum_2001 <- Batting %>% filter(yearID == 2001) %>% group_by(Age) %>%
+  summarize(HR = sum(HR), AB = sum(AB)) %>% mutate(prop = HR / AB)
+```
+
+```{r}
+post_sample <- as.data.frame(fit_23, pars = c("b[29]", "beta1")) %>% slice_sample(n = 10)
+
+ggplot(data = sum_2001) + 
+  geom_point(mapping = aes(x = Age - 30, y = logit(prop))) +
+  geom_abline(slope = post_sample$beta1, intercept = post_sample$"b[29]")
+```
+Try a new version of the model:
+
+```{r}
+sum_all_years <- Batting %>% group_by(yearID, Age) %>%
+  summarize(HR = sum(HR), AB = sum(AB))
+```
+
+Fit a model first with varying intercepts (different intercept for each year) but constant slope.
+
+Start with just 200 iterations to see if the model works at all.

### 2023-10-24 Lahman Analysis/Rough Analysis 2/Analysis2.qmd
Added comments to Stan and added analyses

@@
 
+```{r}
+#Create the summarized data frame
+
+Sum.Batting <- Batting %>% 
+  group_by(yearID, Age) %>% 
+  summarize(AB = sum(AB),
+            HR = sum(HR))
+```
+
 # Theoretical Model
@@
 ```
+
 Let's graph again, restricting to players aged 20 to 40.
@@
 ```
+
 Our first model will include a linear term for age and will be restricted to this age range.
@@
 ```
+
 And now let's fit using the following simple model:
@@
 ```
+
 Try a new version of the model:
@@
 sum_all_years <- Batting %>% group_by(yearID, Age) %>%
-  summarize(HR = sum(HR), AB = sum(AB))
+  summarize(HR = sum(HR), AB = sum(AB), SB = sum(SB))
 ```
@@
 Start with just 200 iterations to see if the model works at all.
+
+```{r}
+data23b <- list(
+  N = nrow(sum_all_years),
+  n_year = n_distinct(sum_all_years$yearID),
+  AB = sum_all_years %>% pull(AB),
+  HR = sum_all_years %>% pull(HR),
+  Year = sum_all_years %>% pull(yearID) -1972,
+  Age = sum_all_years %>% pull(Age) -30
+)
+```
+
+```{r}
+#I find that 500 works and 200 doesn't
+fit23b <- stan(
+  file = "HR23b.stan",
+  data = data23b,
+  iter = 500
+)
+```
+
+```{r}
+summary(fit23b)$summary %>% signif(3)
+```
+
+Fitting the model but with SB's considered as a proxy or player archetype
+
+```{r}
+data24 <- list(
+  N = nrow(sum_all_years),
+  n_year = n_distinct(sum_all_years$yearID),
+  AB = sum_all_years %>% pull(AB),
+  HR = sum_all_years %>% pull(HR),
+  Year = sum_all_years %>% pull(yearID) -1972,
+  Age = sum_all_years %>% pull(Age) -30,
+  SB = sum_all_years %>% pull(SB)
+)
+```
+
+```{r}
+fit23b <- stan(
+  file = "HR24.stan",
+  data = data24,
+  iter = 500
+)
+```
+
+Iterating using both 500 and 1000 loops leaves an error. Suggesting that using SB probably is not a good idea. Or at least we should talk about it more before implementation.

### 2023-10-26 Lahman Analysis/Rough Analysis 2/Analysis2.qmd
Did a simple multilevel model

@@
 
-Fitting the model but with SB's considered as a proxy or player archetype
+We should try to see if we can implement player archetype into the model.
+We can use SB to represent players archetype.
+Basically, we have the big slugger who doesn't steal bags and the skinny fast player who does steal bags.
+
+```{r}
+no_age_sum_all_years <- sum_all_years %>% 
+  group_by(yearID) %>% 
+  summarize(HR = sum(HR),
+            AB = sum(AB),
+            SB = sum(SB)) %>% 
+  mutate(p = HR / AB)
+```
+
+```{r}
+#Lets check graphically
+
+ggplot(data=no_age_sum_all_years)+
+  aes(x=SB, y=HR)+
+  geom_point()
+```
+
+When ignoring age, we can see a general downward trend for HR proportion and age over the time series we are considering.
+
+```{r}
+ggplot(data = no_age_sum_all_years)+
+  aes(x=yearID, y=p)+
+  geom_line()
+```
+
+```{r}
+ggplot(data=no_age_sum_all_years)+
+  aes(x=yearID, y=SB)+
+  geom_line()
+```
+
+This confirms my hypothesis that players are starting to hit home runs with more frequency at the trade off of stolen bases.
+Those who remain stealing bags are may not be hitting home runs and thus we should at least try adding SB's into the model to see how it does.
 
@@
 Iterating using both 500 and 1000 loops leaves an error. Suggesting that using SB probably is not a good idea. Or at least we should talk about it more before implementation.
+
+## Multilevel Model
+
+Maybe we can start with the model we used in "HR23.stan,"
+
+$$
+HR\sim Binom(AB,p)\\
+p=b(Year)+\beta_1Age\\
+b\sim Normal(-3.5,0.1)\\
+\beta_1\sim Normal(0,1)
+$$
+
+And change it to something like
+
+$$
+HR\sim Binom(AB,p)\\
+p=b(Year)+\beta_1Age\\
+b\sim Normal(\bar{b},\sigma_b)\\
+\bar{b}\sim Normal(-3.5,0.1)\\
+\sigma_b\sim Exp(1)\\
+\beta_1\sim Normal(0,1)
+$$
+
+This mirrors the differences in the models used on pages 402-403.
+
+```{r}
+#We don't need to import any more data than is already present in HR23 so we can use that dataset again.
+
+#After running the first time (which didn't take long. Yay!) it suggested that I should run some more iterations so I'll up it to 1000
+
+fit25 <- stan(
+  file = "HR25.stan",
+  data = data23b,
+  iter = 1000
+)
+```
+
+Doing 1000 iterations removed any problems I had with the model, and I believe that I was able to accurately do the (simple) multilevel model.
+
+```{r}
+summary(fit25)$summary %>% signif(3)
+```
+
+```{r}
+plot(fit25, pars = "b")
+```
+
+```{r}
+traceplot(fit25)
+```
+
+```{r}
+saveRDS(fit25, "HR25.rds")
+```
+
+Not sure that I did this right so I'm going to leave it here for now.

### 2023-10-26 Lahman Analysis/Rough Analysis 2/Analysis2.qmd
Added index variable for decade born for replication

@@
             HR = sum(HR))
+```
+
+```{r}
+#Create a column which described the decade a player was born in
+
+Batting %>% 
+  #Put the birth year down to the decade value. (i.e. 1936 becomes 193.6)
+  #Then force it to be an integer. R doesn't round so we are left with (1936=193) a decade
+  mutate(birthDecade = as.integer(birthYear/10), .before=birthYear) %>% 
+  #Then, index it by making the minimum year equal to 1
+  mutate(birthDecade = birthDecade - min(birthDecade) + 1)
 ```

### 2023-10-27 Lahman Analysis/Rough Analysis 2/Analysis2.qmd
Add new model using stolen bases
Added stolen bases to a model, focusing on individual players rather than total stolen bases by age.

@@
 ```{r}
-fit23b <- stan(
+fit24 <- stan(
   file = "HR24.stan",
   data = data24,
-  iter = 500
+  iter = 2000
 )
@@
 Iterating using both 500 and 1000 loops leaves an error. Suggesting that using SB probably is not a good idea. Or at least we should talk about it more before implementation.
+
+```{r}
+summary(fit24)$summary
+```
+```{r}
+plot(fit24, pars = c("beta1", "c"))
+```
+
+### Another try with SB
+
+Let's try a new version of the model above that separates players by SB as well as age.
+
+```{r}
+sum_all_years_SB <- Batting %>% group_by(yearID, Age, SB) %>%
+  summarize(HR = sum(HR), AB = sum(AB))
+```
+
+```{r}
+data24b <- list(
+  N = nrow(sum_all_years_SB),
+  n_year = n_distinct(sum_all_years_SB$yearID),
+  AB = sum_all_years_SB %>% pull(AB),
+  HR = sum_all_years_SB %>% pull(HR),
+  Year = sum_all_years_SB %>% pull(yearID) -1972,
+  Age = sum_all_years_SB %>% pull(Age) -30,
+  SB = sum_all_years_SB %>% pull(SB)
+)
+```
+
+```{r}
+fit24b <- stan(
+  file = "HR24b.stan",
+  data = data24b,
+  iter = 500
+)
+```
+```{r}
+print(fit24b)
+```
+```{r}
+plot(fit24b, pars = "b")
+```
+```{r}
+plot(fit24b, pars = c("beta1", "c"))
+```
+
+```{r}
+Batting_summary_SB <- Batting %>% group_by(SB) %>%
+  summarize(HR = sum(HR), AB = sum(AB)) %>%
+  mutate(prop = HR / AB, se = sqrt(prop*(1-prop)/AB))
+
+ggplot(data = Batting_summary_SB) +
+  geom_pointrange(mapping = aes(x = SB, y = prop, ymin = prop - se,
+                                ymax = prop + se))
+```
 

### 2023-10-28 Lahman Analysis/Rough Analysis 2/Analysis2.qmd
Add prior and posterior checks for model with SB
I added a simple evaluation of the prior for the model that I used including SB. I also tried out the bayesplot model to make graphical posterior predictive checks. (They immediately show problems with the model.)

@@
 library(rstan)
+library(bayesplot)
 library(Lahman)
@@
 Let's try a new version of the model above that separates players by SB as well as age.
+Before we do anything, we can look at home-run proportions when we group player seasons (cases in our data set) by the number of stolen bases.
+We first graph the proportions.
+
+```{r}
+Batting_summary_SB <- Batting %>% group_by(SB) %>%
+  summarize(HR = sum(HR), AB = sum(AB)) %>%
+  mutate(prop = HR / AB, se = sqrt(prop*(1-prop)/AB))
+
+ggplot(data = Batting_summary_SB) +
+  geom_pointrange(mapping = aes(x = SB, y = prop, ymin = prop - se,
+                                ymax = prop + se))
+```
+We can graph again on the logit scale, focusing on cases where the number of stolen bases is less than 50.
+The association looks plausibly linear, suggesting that a logit-binomial model may be reasonable.
+
+```{r}
+logit <- function(p){log(p / (1 - p))}
+
+ggplot(data = Batting_summary_SB %>% filter(SB < 50)) +
+  geom_pointrange(mapping = aes(x = SB, y = logit(prop), ymin = logit(prop - se),
+                                ymax = logit(prop + se))) +
+  geom_smooth(method = "lm", mapping = aes(x = SB, y = logit(prop)))
+```
+To prepare to fit the model, we summarize the data, grouping player-seasons by year, player age, and number of stolen bases.
+(As in other examples, this grouping will not affect the likelihood, and it seems to speed up the work of Stan considerably.)
 
@@
 sum_all_years_SB <- Batting %>% group_by(yearID, Age, SB) %>%
-  summarize(HR = sum(HR), AB = sum(AB))
+  summarize(HR = sum(HR), AB = sum(AB)) %>% ungroup
 ```
@@
 
+Before we fit the model, we can simulate data from the prior distributions. We just generate one simulated data set, since there is plenty of internal replication in the data set.
+
+```{r}
+prior24b <- stan(
+  file = "HR24b_prior_sim.stan",
+  data = data24b,
+  algorithm = "Fixed_param",
+  iter = 1,
+  chains = 1
+)
+```
+To get an overall sense for the simulated data, we look at simulated home-run proportions in the groups of players (by year, age, and stolen bases).
+
+```{r}
+HR_sim <- (rstan :: extract)(prior24b)$HR[1,]
+
+sum_all_years_SB <- sum_all_years_SB %>% mutate(HR_sim = HR_sim,
+                                                prop_sim = HR_sim / AB)
+```
+
+```{r}
+ggplot(data = sum_all_years_SB) +
+  geom_histogram(mapping = aes(x = prop_sim))
+```
+
+Huh!
+The Bayesians warn us about this phenomenon.
+(See Bayesian Workflow, Section 2.4 and Figure 3.) Weak independent priors on large numbers of parameters add up to something strong.
+The home-run proportions are either near 0% or 100%.
+If we did not have a lot of data to fit the model, we would have to use stronger priors to get something reasonable here.
+In our case, there is so much data that the prior distributions do not matter much (at least for this model).
+
+The next chunk fits the model.
+
 ```{r}
@@
 ```
+
+We have another version of the model that also generates replications.
+
+(Here is an RStan demo with some inspirational code: [http://avehtari.github.io/BDA_R_demos/demos_rstan/ppc/poisson-ppc.html])
+
 ```{r}
-print(fit24b)
-```
-```{r}
-plot(fit24b, pars = "b")
-```
-```{r}
-plot(fit24b, pars = c("beta1", "c"))
+fit24b_reps <- stan(
+  file = "HR24b_reps.stan",
+  data = data24b,
+  iter = 500
+)
 ```
@@
 ```{r}
-Batting_summary_SB <- Batting %>% group_by(SB) %>%
-  summarize(HR = sum(HR), AB = sum(AB)) %>%
-  mutate(prop = HR / AB, se = sqrt(prop*(1-prop)/AB))
+HR_rep <- as.matrix(fit24b_reps, pars = "HR_rep")
 
-ggplot(data = Batting_summary_SB) +
-  geom_pointrange(mapping = aes(x = SB, y = prop, ymin = prop - se,
-                                ymax = prop + se))
+dim(HR_rep)
 ```
+
+```{r}
+ppc_hist(sum_all_years_SB %>% pull(HR), HR_rep[1:8, ], binwidth = 1)
+```
+
+Yuck... A poor fit. The model doesn't reproduce low numbers of home runs as we see in the original data set.
+
+```{r}
+ppc_dens_overlay(sum_all_years_SB %>% pull(HR), HR_rep[1:50, ])
+```
+This family of density plots shows the same thing. A basic check of the model (histogram of home-run counts in our groups) shows a mismatch.
+
+The situation reminds me of the pictures in Section 26.2 of the Stan User's Guide [https://mc-stan.org/docs/2_26/stan-users-guide/plotting-multiples.html].
+In that section, the authors show an overdispersed count variable (negative binomial) fit with a binomial model.
+There is more on the same example in Section 26.3, using a test statistic and the p-value idea to quantify the poor fit.
+
+Perhaps I will add more here, trying some better posterior plots from the bayesplot package.
+See, for example, this page [https://mc-stan.org/bayesplot/reference/PPC-errors.html], which shows how to make the kind of binned-residual graphs for binomial models that we used in MATH 313.
 

### 2023-10-28 Lahman Analysis/Rough Analysis 2/Analysis2.qmd
Improve model with SB and model assessments
I added some more graphical evaluations of the model with SB. I have used proportion of home runs to evaluation the model.

@@
 ```
+Since we will use both stolen bases and age in the model, we can make the same sort of graph after grouping players by age. Let's look first at the distribution of age.
+
+```{r}
+ggplot(data = Batting) +
+  geom_histogram(mapping = aes(x = Age), binwidth = 1)
+```
+(What a lovely histogram!)
+
+We will split players into three roughly equally sized age groups, determined by quantiles.
+
+```{r}
+Batting %>% summarize(quantile(Age, 0.33), quantile(Age, 0.67))
+```
+Our three groups are "26 or younger", "27 to 30", and "over 30".
+
+```{r}
+Batting <- Batting %>% mutate(age_group = case_when(
+  Age <= 26 ~ "young",
+  Age <= 30 ~ "middle",
+  Age > 30 ~ "old"
+))
+```
+
+We repeat our graphical exploration, faceting by age group.
+The general pattern (negative association between SB and home-run probability) persists across age groups.
+We note, though, that the pattern looks a bit different for the youngest players
+
+```{r}
+Batting_summary_age_SB <- Batting %>% group_by(age_group, SB) %>%
+  summarize(HR = sum(HR), AB = sum(AB)) %>%
+  mutate(prop = HR / AB, se = sqrt(prop*(1-prop)/AB))
+
+ggplot(data = Batting_summary_age_SB) +
+  geom_pointrange(mapping = aes(x = SB, y = prop, ymin = prop - se,
+                                ymax = prop + se)) +
+  facet_wrap(~age_group)
+```
+
+```{r}
+ggplot(data = Batting_summary_age_SB) +
+  geom_pointrange(mapping = aes(x = SB, y = logit(prop), ymin = prop - se,
+                                ymax = prop + se)) +
+  facet_wrap(~age_group)
+```
+
 To prepare to fit the model, we summarize the data, grouping player-seasons by year, player age, and number of stolen bases.
@@
 sum_all_years_SB <- Batting %>% group_by(yearID, Age, SB) %>%
-  summarize(HR = sum(HR), AB = sum(AB)) %>% ungroup
+  summarize(HR = sum(HR), AB = sum(AB)) %>% mutate(prop = HR / AB) %>% ungroup
 ```
@@
 
+Here is the actual distribution of HR proportions in the data. We shouldn't be trying to fit the data with the prior, but we can see how far off things are from this second histogram.
+
+```{r}
+ggplot(data = sum_all_years_SB) +
+  geom_histogram(mapping = aes(x = prop))
+```
+
 Huh!
@@
 In our case, there is so much data that the prior distributions do not matter much (at least for this model).
+
+We can repeat setting stronger priors.
+
+```{r}
+prior24b2 <- stan(
+  file = "HR24b_prior_sim2.stan",
+  data = data24b,
+  algorithm = "Fixed_param",
+  iter = 1,
+  chains = 1
+)
+```
+
+```{r}
+HR_sim2 <- (rstan :: extract)(prior24b2)$HR[1,]
+
+sum_all_years_SB2 <- sum_all_years_SB %>% mutate(HR_sim = HR_sim2,
+                                                prop_sim = HR_sim / AB)
+```
+
+```{r}
+ggplot(data = sum_all_years_SB2) +
+  geom_histogram(mapping = aes(x = prop_sim))
+```
+
+That prior distribution of simulated HR proportions looks better than what we got with the very weak priors on `beta1` and `c`. But perhaps these priors would be too strong...
 
@@
 
-(Here is an RStan demo with some inspirational code: [http://avehtari.github.io/BDA_R_demos/demos_rstan/ppc/poisson-ppc.html])
+(Here is an RStan demo with some inspirational code for working with the replications: [http://avehtari.github.io/BDA_R_demos/demos_rstan/ppc/poisson-ppc.html].
+Here is another vignette with examples of posterior predictive checks: [https://cran.r-project.org/web/packages/bayesplot/vignettes/graphical-ppcs.html])
+
+A better way to do this would be to remove the rows in the data table for playes with 0 at-bats.
+We could then have Stan calculate proportions (without divide-by-0 errors).
+I got around those rows in an ugly way below.
 
@@
 ```{r}
-ppc_hist(sum_all_years_SB %>% pull(HR), HR_rep[1:8, ], binwidth = 1)
+ppc_hist(sum_all_years_SB %>% filter(AB > 0) %>% pull(prop) ,
+         HR_rep[1:8, sum_all_years_SB$AB > 0] / (sum_all_years_SB %>% filter(AB > 0) %>% pull(AB))) +
+  xlim(0, 0.3)
 ```
 
-Yuck... A poor fit. The model doesn't reproduce low numbers of home runs as we see in the original data set.
+Yuck...
+A poor fit.
+The model replications are quite consistent---and don't look much like the original data.
+In the model, home-run percentages over 10% and as high as 20% sometimes occur.
+These never happen in the real data.
+The distribution of proportions in the simulated data has mode at close to 0.
+In the real data, the model appears closer to 3%.
 
 ```{r}
-ppc_dens_overlay(sum_all_years_SB %>% pull(HR), HR_rep[1:50, ])
+ppc_dens_overlay(sum_all_years_SB %>% filter(AB > 0) %>% pull(prop) ,
+         HR_rep[1:50, sum_all_years_SB$AB > 0] / (sum_all_years_SB %>% filter(AB > 0) %>% pull(AB))) +
+  xlim(0, 0.3)
 ```
-This family of density plots shows the same thing. A basic check of the model (histogram of home-run counts in our groups) shows a mismatch.
-
-The situation reminds me of the pictures in Section 26.2 of the Stan User's Guide [https://mc-stan.org/docs/2_26/stan-users-guide/plotting-multiples.html].
-In that section, the authors show an overdispersed count variable (negative binomial) fit with a binomial model.
-There is more on the same example in Section 26.3, using a test statistic and the p-value idea to quantify the poor fit.
+This family of density plots shows the same thing as the histograms. The data from the model are quite consistent---and quite different from the original data.
 

### 2023-11-02 Lahman Analysis/Formal Analysis/Draft 1.qmd
Added age by year, age polynomial, and stadium indicies

---
title: "Draft 1"
format: html
---

# Fellingham & Fisher (2017)

To apply the methods discussed in *Statistical Rethinking* by Richard McElreath, Dr. Parson and I decided to replicate *Predicting Home Run Production in Major League Baseball Using a Bayesian Semiparametric Model* by Gilbert Fellingham and Jared Fisher. This article in particular has drawn our attention because of some unique and interesting modeling choices made by the authors. Indeed, both Dr. Parson and I believe that while the initial investigation by Fellingham and Fisher (2017) is interesting, there are some questionable choices made by the authors throughout the paper. For example, there are few independent variables and their associated priors are very weak. The paper does a good job with displaying plenty of visualizations for the reader, yet, under careful examination one is left with the impression that the authors have cherry-picked certain players to focus on rather than examining players at random to examine the validity of their model. These reasons in tandem bring their findings into question. Thus, Dr. Parson and I have embarked to investigate the same research question, can we forecast home run potential using a Bayesian model which accounts for age, home ballpark, and other factors.

## Background

In Major League Baseball (MLB) the team on offense is allowed to send one player, at a time, up to home plate to "bat." Any time a player is sent up to bat this counts as a plate appearance (PA) for said player. Depending on its outcome the PA may or may not be considered an at-bat (AB). If the PA ends in the player being walked (BB \[base on balls\]) then it is not recorded as an AB because it is thought the player did not receive a fair chance at hitting the baseball. On the other hand, if the PA ends in the player getting a hit or getting out, then it is assumed that he had a fair chance at hitting the ball and the PA is recorded as an AB.

The principle concern of this paper is how to model the amount of home runs (HR) a player will hit in a given season. Home runs are an outcome which is recorded as an AB where the player is able to touch all four bases before being "tagged out," where the defense is able to touch the player with the baseball. The primary way this occurs, and is the one that most readers will be familiar with, is by hitting the ball over the fence in the outfield. However, it is just as correct for the ball to remain in the boundaries of the field and for the player to reach all bases without being tagged out. Due to HR's being recorded as AB's both ourselves and Fellingham and Fisher (2017) incorporate AB's in our models rather than PA's.

Another interesting quirk of baseball is that, like soccer, stadiums differ in dimensions. This is not universal; baseball fields and stadiums must have the same measurements for their infields. Yet, for the most part baseball stadiums are allowed to configure their outfields as they please. This gives baseball stadiums a variety of unique shapes ranging in outfield sizes. For example, Fenway Park (1923-) is known to have the closest outfield wall with their "Green Monster" in left field being only 305 ft. from home plate. While the Polo Grounds (1891-1963) boasted one of the furthest walls from home play at 483 ft. in center field. Stadiums also differ in location. Marlins Park (2012-2020) sits at 15 ft. above sea level while Coors Field (1995-) sits at 5183 ft. above sea level. As illustrated there is not only vastly different dimensions that baseball fields can take on but also different environments which affect a player's ability to hit home runs. Thus, any model which attempts to model players' ability to hit HR's needs to consider ballpark factors.

It is obvious from examining the years that stadiums were used to infer that baseball has been played for a long time. Indeed, MLB as we know it has been played since about 1871 and the origins of the game go back even father than that. Over this time the some rules of the game have undergone incredible transformations which vastly change the nature of the game. For example, from 1900-1920 baseball was in the "deadball era." During this period the materials used to create baseballs made it so that the ball absorbed much more force than it does today. This had a plethora of implications for the game and how it was played; HR's were much less common and since they were harder to hit. In contrast, the period from 1994-2004 is known as the "steroid era" because of the rampant use of performance enhancing drugs (PED) used by players. This, in combination with other factors, led to players obliterating previous records, such as the career and season home runs records. It is obvious from these facts that decade played for player's had a drastic impact on their ability to produce outcomes. In fact, this can be visualized by considering all the runs scored during a given season from 1871-2019 displayed in @fig-RunsByYear.

Finally, it does not take an advanced degree to understand that age is an important factor in athletic ability. To an extent in MLB, seniority can act both as a blessing and a curse. Older players are less able to hit home runs because their body is starting to lose its peak athletic performance it had when it was, say, 30. Yet, these players also have more experience than younger players and thus may be better able to hit home runs from strategies they have found during their time playing. Further we can visualize this by looking at the median home runs hit by each player from 1871-2022, which is displayed in

## Model

# Figures

```{r}
#| message: false
#| warning: false

library(Lahman)
library(tidyverse)
```

```{r}
Runs.By.Year <- Lahman::Teams %>% 
  group_by(yearID) %>% 
  summarize(R = sum(R))
```

```{r}
#| label: fig-RunsByYear

ggplot(data = Runs.By.Year)+
  aes(x=yearID, y=R)+
  geom_line()+
  geom_rect(inherit.aes = FALSE, aes(xmin=1900, xmax=1920, ymin=0, ymax=26000), color="transparent", fill="red", alpha=0.005)+
  geom_rect(inherit.aes = FALSE, aes(xmin=1994, xmax=2004, ymin=0, ymax=26000), color="transparent", fill="orange", alpha=0.005)+
  scale_y_continuous(limits =c(0,26000), breaks = seq(from=0, to=25000, by=5000))+
  scale_x_continuous(breaks = seq(from=1870, to=2020, by=10))+
  labs(title = "Runs Scored in MLB by Year",
       x = "Year",
       y = "Runs Scored")
```

### 2023-11-02 Lahman Analysis/Rough Analysis 2/Analysis2.qmd
Added age by year, age polynomial, and stadium indicies

@@
             HR = sum(HR))
+```
+
+```{r}
+#Make an index for the parks to add them to the model
+
+#The index goes from 1-88. No idea how it came up with them but I kept the park name in Batting so anyone should be able to reference to see how the parks impact player performance
+
+Batting <- Batting %>% 
+  mutate(park = as.factor(park)) %>% 
+  mutate(park_index = as.integer(park), .after = park)
 ```
@@
 ```
+
 We can graph again on the logit scale, focusing on cases where the number of stolen bases is less than 50.
@@
 ```
+
 Since we will use both stolen bases and age in the model, we can make the same sort of graph after grouping players by age. Let's look first at the distribution of age.
@@
 ```
+
 (What a lovely histogram!)
@@
 ```
+
 Our three groups are "26 or younger", "27 to 30", and "over 30".
@@
 ```
+
 To get an overall sense for the simulated data, we look at simulated home-run proportions in the groups of players (by year, age, and stolen bases).
@@
 
-(Here is an RStan demo with some inspirational code for working with the replications: [http://avehtari.github.io/BDA_R_demos/demos_rstan/ppc/poisson-ppc.html].
-Here is another vignette with examples of posterior predictive checks: [https://cran.r-project.org/web/packages/bayesplot/vignettes/graphical-ppcs.html])
+(Here is an RStan demo with some inspirational code for working with the replications: \[http://avehtari.github.io/BDA_R_demos/demos_rstan/ppc/poisson-ppc.html\].
+Here is another vignette with examples of posterior predictive checks: \[https://cran.r-project.org/web/packages/bayesplot/vignettes/graphical-ppcs.html\])
 
@@
 ```
+
 This family of density plots shows the same thing as the histograms. The data from the model are quite consistent---and quite different from the original data.
@@
 Perhaps I will add more here, trying some better posterior plots from the bayesplot package.
-See, for example, this page [https://mc-stan.org/bayesplot/reference/PPC-errors.html], which shows how to make the kind of binned-residual graphs for binomial models that we used in MATH 313.
+See, for example, this page \[https://mc-stan.org/bayesplot/reference/PPC-errors.html\], which shows how to make the kind of binned-residual graphs for binomial models that we used in MATH 313.
 
@@
 Not sure that I did this right so I'm going to leave it here for now.
+
+I also wanted to investigate how park factors can play into a hitter's ability.
+So, let's add them into the model here.
+As a note, we don't need to create a diferent coefficient for every year since the ballpark doesn't move.
+
+```{r}
+sum_all_years_park <- Batting %>% group_by(yearID, park, park_index, Age) %>%
+  summarize(HR = sum(HR), AB = sum(AB), SB = sum(SB))
+```
+
+```{r}
+data26 <- list(
+  N = nrow(sum_all_years_park),
+  n_year = n_distinct(sum_all_years_park$yearID),
+  AB = sum_all_years_park %>% pull(AB),
+  HR = sum_all_years_park %>% pull(HR),
+  Year = sum_all_years_park %>% pull(yearID) -1972,
+  Age = sum_all_years_park %>% pull(Age) -30,
+  park = sum_all_years_park %>% pull(park_index)
+)
+```
+
+```{r}
+fit26 <- stan(
+  file = "HR26.stan",
+  data = data26,
+  iter = 500
+)
+```
+
+Even 500 iterations did not work and took 15 minutes so I think it's safe to say that adding the stadium might not be the greatest idea until we get more computing power.
+However, I still want to save this.
+
+```{r}
+saveRDS(fit26, "HR26.rds")
+```
+
+Of course, the error that is being reported is one of poor priors, so I'll try and go in and give it some more confident priors.
+
+```{r}
+fit26b <- stan(
+  file = "HR26b.stan",
+  data = data26,
+  iter = 400
+)
+```
+
+```{r}
+saveRDS(fit26b, "HR26b.rds")
+```
+
+Still get the warning after 400 iterations AND making the priors more informative.
+Not sure why this is behaving so poorly, there is such good theory to support this hypothesis.
+Of course, this (along with the problems Dr.
+Parson was facing) could explain why Fellingham and Fisher had to do over 200,000 iterations (I believe it was that crazy amount).
+
+## Graphical Exploration pt.2
+
+Do we see evidence of probability changing with age and year?
+
+```{r}
+p_year_age <- Batting %>% 
+  group_by(yearID, Age) %>% 
+  summarize(HR = sum(HR), AB = sum(AB)) %>% 
+  mutate(p = HR / AB)
+```
+
+```{r}
+ggplot(data = p_year_age)+
+  aes(x=yearID, y=p, color=as.factor(Age))+
+  geom_line()
+```
+
+Clearly, we aren't getting anything from this graph.
+So, let's graph two more, one with the player's age rounded to the it's 10 (ie.
+32-\> 30 and 36 -\> 30) and one where we use the cut-offs from the prior examination (below 27, 27 to 30, and above 30).
+
+```{r}
+p_year_age <- p_year_age %>% 
+  mutate(age_10 = Age / 10, .after = Age) %>% 
+  mutate(age_10 = as.integer(age_10)) %>% 
+  mutate(age_group = case_when(Age < 27 ~ "young",
+                               Age >= 27 & Age <= 30 ~ "medial",
+                               Age > 30 ~ "old") , .after = age_10) %>% 
+  mutate(age_group = as.factor(age_group)) %>% 
+  mutate(age_10 = as.factor(age_10))
+```
+
+```{r}
+ggplot(data = p_year_age)+
+  aes(x = yearID, y = p, color = age_10)+
+  geom_line()+
+  geom_point()
+```
+
+```{r}
+ggplot(data = p_year_age)+
+  aes(x = yearID, y = p, color = age_group)+
+  geom_line()+
+  geom_point()
+```
+
+Based on these graphs, there does not seem to be good evidence to justify fitting a different slope for each age in each year.
+They all tend to the same amounts.
+Other than the 50+ players and the 18 and 19 year-olds.
+However, I suspect that this is due much more to a lack of a representitive sample than anything else.
+
+Another things we wanted to check was the fit of a polynomial with degree 4 to see what Fellingham and Fisher were getting at with their Detrich process.
+We already have a graph but it's at the top so let's bring it down to see it again.
+
+```{r}
+Batting_summary2 <- Batting_summary %>% 
+  filter(Age >=20 & Age <= 40)
+```
+
+```{r}
+ggplot(data = Batting_summary2) +
+  geom_pointrange(mapping = aes(x = Age, y = prop, ymin = prop - se,
+                                ymax = prop + se))
+```
+
+We can run an lm to get a general sense of what these coefficients should be
+
+```{r}
+quart <- lm(data = Batting_summary2, prop ~ Age + I(Age^2) + I(Age^3) + I(Age^4))
+
+quart %>% 
+  summary()
+```
+
+Not that it means much but we do get significance on each coefficient. I find this interesting because I've never seen a degree 4 polynomial have such significance.
+
+```{r}
+ggplot(data = Batting_summary2) +
+  geom_pointrange(mapping = aes(x = Age, y = prop, ymin = prop - se,
+                                ymax = prop + se))+
+  geom_function(fun = function(x) (2.364e-01) + (-3.334e-02)*x + (1.871e-03)*(x^2) + (-4.463e-05)*(x^3) + (3.873e-07)*(x^4), color="red")
+```
+
+So, very clearly this has a much better fit to the age variable. How does it look for the full data set we considered (not just the 20-40 year-olds).
+
+```{r}
+ggplot(data = Batting_summary) +
+  geom_pointrange(mapping = aes(x = Age, y = prop, ymin = prop - se,
+                                ymax = prop + se))+
+  geom_function(fun = function(x) (2.364e-01) + (-3.334e-02)*x + (1.871e-03)*(x^2) + (-4.463e-05)*(x^3) + (3.873e-07)*(x^4), color="red")
+```
+
+This is actually pretty funny.
+We lose all the fit past the age of 40 when considering the whole data set.
+Since we can't even justify the 4th degree and this doesn't fit past 40 it probably isn't a good idea to use this fit.
+Of course, if all we care about is fit, then we could just keep addign degrees until all years considered are accounted for.

### 2023-11-02 Lahman Analysis/Formal Analysis/Draft 1.qmd
Continued writing formal analysis

@@
 ---
+
+```{r}
+#| message: false
+#| warning: false
+
+library(tidyverse)
+library(rstan)
+library(Lahman)
+library(bayesplot)
+```
+
+```{r}
+#Set Stan to read the amount of cores that your computer can handel
+options(mc.cores = parallel::detectCores())
+rstan_options(auto_write = TRUE)
+```
+
+```{r}
+#| message: false
+
+#Creates a DF called "Batting" which has players stats from 1973-2019
+
+#Load the DF's from Lahman directly
+Batting <- Lahman::Batting
+Teams <- Lahman::Teams
+People <- Lahman::People
+
+Batting <- Batting %>% 
+  #Minimum year = DH was introduced; maximum year = before COVID
+  filter(yearID >= 1973 & yearID <= 2019)
+
+#Filter teams for years considered and select only necessary columns
+Teams <- Teams %>% 
+  filter(yearID>=1973 & yearID <= 2019) %>% 
+  select(yearID, teamID, park)
+
+#Combine current "Batting" and "Teams" DF to associate player with stadium played at
+Batting <- Batting %>% 
+  right_join(Teams, by = join_by(yearID, teamID))
+
+#Add player's age to the "Batting" DF
+Batting <- Batting %>% 
+  right_join(Lahman::People, by = join_by(playerID)) %>% 
+  mutate(Age = yearID - birthYear, .after=yearID) %>%
+  #Don't consider the people who don't have a birthday in the database
+  drop_na(Age)
+
+#Make the lgID variable into an index variable
+Batting <- Batting %>% 
+  mutate(lg = case_when(lgID == "NL" ~ 1,
+                        lgID == "AL" ~ 2), .after=lgID)
+
+#Make an index for the parks to add them to the model
+
+#The index goes from 1-88. No idea how it came up with them but I kept the park name in Batting so anyone should be able to reference to see how the parks impact player performance
+
+Batting <- Batting %>% 
+  mutate(park = as.factor(park)) %>% 
+  mutate(park_index = as.integer(park), .after = park)
+```
 
@@
 To an extent in MLB, seniority can act both as a blessing and a curse.
-Older players are less able to hit home runs because their body is starting to lose its peak athletic performance it had when it was, say, 30.
-Yet, these players also have more experience than younger players and thus may be better able to hit home runs from strategies they have found during their time playing.
-Further we can visualize this by looking at the median home runs hit by each player from 1871-2022, which is displayed in
+On the one hand, older players are less able to hit home runs because their body is starting to lose its peak athletic performance it had when it was, say, 30.
+On the other hand, these older players also have more experience than younger players and thus may be better able to hit home runs from strategies they have found during their time playing.
+Further we can visualize this by looking at the median home runs hit by each player from 1871-2022, which is displayed in @fig-HR.By.Age.All.
+However, this graph does not depict some of the nuances of the data well.
+For example, we have almost no data on how often 18 -year-olds hit HR's because they play in MLB infrequently.
+As can be seen with the standard error bars on the ages above 40, older players suffer a similar problem of small sample size.
+Instead, consider @fig-HR.By.Age.20to40 which shows only considers players who are 20-40 years old during the season.
+This graph clearly shows the complex non-linear relationship between player performance and age.
+It shows that players are generally improving in their ability to hit HR's from 20-30 years old.
+Then, however, we see a slight downwards trend followed by an upward trend with increasing error bars.
+This shows that as players age past 30 their abilities start to decline.
+However, the jump afterwards can probably be attributed to players who are very good, continuing their careers.
+If a player is very good then they are going to be able to perform at above average when they are, say, 35.
+However, if the player is not generally considered good, then a team will likely opt to try out a new young rookie who may not be as good in the moment but is more projectable.
+Thus, we see less players playing past 35 which explains the higher standard errors, but they are of a higher caliber than any other age group which is why they are still playing, thus we see their proportion as higher.
 
@@
 ```{r}
-#| message: false
-#| warning: false
-
-library(Lahman)
-library(tidyverse)
-```
-
-```{r}
+#Make the dataframe needed for Runs Scored by year graph
 Runs.By.Year <- Lahman::Teams %>% 
@@
   scale_x_continuous(breaks = seq(from=1870, to=2020, by=10))+
-  labs(title = "Runs Scored in MLB by Year",
+  labs(title = "Cumulative Runs Scored in MLB by Year",
        x = "Year",
@@
 ```
+
+```{r}
+#Make DF which describes the HR by age across the timeline considered
+HR.Prop.Age <- Batting %>% 
+  group_by(Age) %>% 
+  summarize(HR = sum(HR), AB = sum(AB)) %>% 
+  mutate(prop = HR / AB, prop.se = sqrt(prop*(1-prop)/AB))
+```
+
+```{r}
+#| label: fig-HR.By.Age.All
+
+ggplot(data = HR.Prop.Age)+
+  geom_pointrange(mapping = aes(x = Age, y = prop, ymin = prop - prop.se,
+                                ymax = prop + prop.se))+
+  labs(title = "Proportion of Home Runs per At-Bat by Age",
+       x = "Age",
+       y = "Proportion of HR per AB")
+```
+
+```{r}
+#| label: fig-HR.By.Age.20to40
+
+ggplot(data = HR.Prop.Age %>% filter(Age >=20 & Age <= 40))+
+  geom_pointrange(mapping = aes(x = Age, y = prop, ymin = prop - prop.se,
+                                ymax = prop + prop.se))+
+  labs(title = "Proportion of Home Runs per At-Bat by Age",
+       x = "Age",
+       y = "Proportion of HR per AB")
+```

### 2023-11-03 Lahman Analysis/Rough Analysis 2/Analysis2.qmd
Progress from meeting

@@
 
+### 11/3 meeting
+
+```{r}
+by_park <- Batting %>% 
+  group_by(yearID, park) %>% 
+  summarize(HR = sum(HR), AB = sum(AB)) %>% 
+  mutate(park_index = as.integer(park))
+```
+
+```{r}
+data26c <- list(
+  N = nrow(by_park),
+  n_year = n_distinct(by_park$yearID),
+  n_park = n_distinct(by_park$park_index),
+  AB = by_park %>% pull(AB),
+  HR = by_park %>% pull(HR),
+  park = by_park %>% pull(park_index),
+  Year = by_park %>% pull(yearID) -1972
+)
+```
+
+```{r}
+fit26c <- stan(
+  file = "HR26c.stan",
+  data = data26c,
+  iter = 2000
+)
+```
+
+```{r}
+print(fit26c, pars = "b")
+```
+
+```{r}
+traceplot(fit26c)
+```
+
+```{r}
+saveRDS(fit26c, "HR26c.rds")
+```
+
+```{r}
+year_and_stadium_age <- Batting %>% 
+  group_by(yearID, park, Age) %>% 
+  summarize(HR = sum(HR), AB = sum(AB)) %>% 
+  mutate(park_index = as.integer(park)) %>% 
+  mutate(cAge = Age - 30, .after = Age)
+```
+
+```{r}
+data26d <- list(
+  N = nrow(year_and_stadium_age),
+  n_year = n_distinct(year_and_stadium_age$yearID),
+  n_park = n_distinct(year_and_stadium_age$park_index),
+  AB = year_and_stadium_age %>% pull(AB),
+  HR = year_and_stadium_age %>% pull(HR),
+  park = year_and_stadium_age %>% pull(park_index),
+  Year = year_and_stadium_age %>% pull(yearID) -1972,
+  cAge = year_and_stadium_age %>% pull(cAge)
+)
+```
+
+```{r}
+fit26d <- stan(
+  file = "HR26d.stan",
+  data = data26d,
+  iter = 200
+)
+```
+
+```{r}
+summary(fit26d, 'beta1')$summary
+```
+
 ## Graphical Exploration pt.2
@@
 Of course, if all we care about is fit, then we could just keep addign degrees until all years considered are accounted for.
+
+```{r}
+ggplot(sum_all_years_park)+
+  geom_histogram(aes(x = park))
+```

### 2023-11-04 Lahman Analysis/Rough Analysis 2/Analysis2.qmd
Tried fitting individual intercepts for players

@@
 Since we can't even justify the 4th degree and this doesn't fit past 40 it probably isn't a good idea to use this fit.
-Of course, if all we care about is fit, then we could just keep addign degrees until all years considered are accounted for.
+Of course, if all we care about is fit, then we could just keep adding degrees until all years considered are accounted for.
+
+## Player intercepts
 
 ```{r}
-ggplot(sum_all_years_park)+
-  geom_histogram(aes(x = park))
+player_index_small <- Batting %>% 
+  #Consider only two years to get a managable amount of players
+  filter(yearID == 2018 | yearID == 2019) %>% 
+  #Number the players for the multilevel model
+  mutate(player_index = as.factor(playerID), .after = playerID) %>% 
+  mutate(player_index = as.integer(player_index)) %>% 
+  mutate(cAge = Age - 30, .after = Age) %>% 
+  mutate(year_index = yearID - min(yearID) + 1, .after = yearID)
 ```
+
+```{r}
+<- list(
+  N = nrow(player_index_small),
+  n_year = n_distinct(player_index_small$yearID),
+  n_park = n_distinct(player_index_small$park_index),
+  n_player = n_distinct(player_index_small$player_index),
+  AB = player_index_small %>% pull(AB),
+  HR = player_index_small %>% pull(HR),
+  park = player_index_small %>% pull(park_index),
+  Year = player_index_small %>% pull(year_index),
+  cAge = player_index_small %>% pull(cAge),
+  Player = player_index_small %>% pull(player_index)
+)
+```
+
+```{r}
+data27 <- list(
+  N = nrow(player_index_small),
+  n_player = n_distinct(player_index_small$player_index),
+  AB = player_index_small %>% pull(AB),
+  HR = player_index_small %>% pull(HR),
+  Player = player_index_small %>% pull(player_index)
+)
+```
+
+```{r}
+fit27 <- stan(
+  file = "HR27.stan",
+  data = data27,
+  iter = 200
+)
+```

### 2023-11-04 Lahman Analysis/Rough Analysis 2/Analysis2.qmd
Worked up to adding park effects

@@
   mutate(cAge = Age - 30, .after = Age) %>% 
-  mutate(year_index = yearID - min(yearID) + 1, .after = yearID)
+  mutate(year_index = yearID - min(yearID) + 1, .after = yearID) %>%
+  #For this test of conecpt we'll need to redefine the stadium's index since not all stadiums in the original dataset were used in 2018 and 2019
+  select(!park_index) %>% 
+  mutate(park_index = as.integer(park), .after = park)
 ```
@@
   data = data27,
+  iter = 1000
+)
+```
+
+```{r}
+plot(fit27, pars = "a")
+```
+
+This plot is very packed but from it we can see that there is a fair amount of variance among the players and their home run rates.
+However, even with 1000 iterations we still get the warnings of that the posterior medians and medians may be unreliable.
+We can get a sense of this from the large standard error bars on the parameters.
+On the logit scale -5 represents a probability of 0.0067, while a -2 represents 0.119.
+I'm not sure if there's anything that we can do to solve this problem.
+So, maybe we can reduce the standard errors if we include more factors such as year.
+
+```{r}
+data27b <- list(
+  N = nrow(player_index_small),
+  n_player = n_distinct(player_index_small$player_index),
+  n_year = n_distinct(player_index_small$yearID),
+  AB = player_index_small %>% pull(AB),
+  HR = player_index_small %>% pull(HR),
+  Player = player_index_small %>% pull(player_index),
+  Year = player_index_small %>% pull(year_index)
+)
+```
+
+```{r}
+fit27b <- stan(
+  file = "HR27b.stan",
+  data = data27b,
+  iter = 3000
+)
+```
+
+Normally, I would stop the model at 1,000 simulations to save time; however, with this one I saw that we were getting really close to having the Rhat getting to 1.
+And only having one error is better to me than having two.
+
+```{r}
+print(fit27b, pars = "a")
+```
+
+```{r}
+plot(fit27b, pars="a")
+```
+
+Including the year parameter removed a little bit of of the extremeness from the plot.
+
+```{r}
+traceplot(fit27b)
+```
+
+And in general the trace plots look pretty good. Maybe a\[4\] isn't great or a\[5\] but they're not terrible either.
+
+So, now we can probably move into including a park factor. I think this will play a larger role than the age of a player.
+
+```{r}
+data27c <- list(
+  N = nrow(player_index_small),
+  n_player = n_distinct(player_index_small$player_index),
+  n_year = n_distinct(player_index_small$yearID),
+  n_park = n_distinct(player_index_small$park_index),
+  AB = player_index_small %>% pull(AB),
+  HR = player_index_small %>% pull(HR),
+  Player = player_index_small %>% pull(player_index),
+  Year = player_index_small %>% pull(year_index),
+  park = player_index_small %>% pull(park_index)
+)
+```
+
+```{r}
+fit27c <- stan(
+  file = "HR27c.stan",
+  data = data27c,
   iter = 200

### 2023-11-05 Lahman Analysis/Rough Analysis 2/Analysis2.qmd
Fix coding for park in small sample of players and add lower bound on sd in HR27c

@@
   #For this test of conecpt we'll need to redefine the stadium's index since not all stadiums in the original dataset were used in 2018 and 2019
-  select(!park_index) %>% 
+  select(!park_index) %>% mutate(park = as.character(park)) %>%
+  mutate(park = as.factor(park)) %>%
   mutate(park_index = as.integer(park), .after = park)

### 2023-11-08 Lahman Analysis/Rough Analysis 2/Analysis2.qmd
Finished through linear age effect
Next thing to consider is the nonlinear effect that age could have. I will probably use the quartic model that Fellingham and Fisher used.

@@
   data = data27c,
-  iter = 200
-)
-```
+  iter = 3500
+)
+```
+
+```{r}
+saveRDS(fit27c, "HR27c.rds")
+```
+
+Needed to run 3500 iterations in order to reduce the Rhat values to an acceptable levels.
+
+```{r}
+plot(fit27c, pars="a")
+```
+
+Very interestingly we finally see a big reduction in the logit probabilities here.
+It's we will investigate the standard errors below since the graph doesn't lend itself well to this, however, we can see that their means and confidence intervals are much more closely centered on -3.5 suggesting that a lot of what we originally contriubted to player ability is actually just a ballpark effect.
+
+```{r}
+print(fit27c)
+```
+
+We can see from this table that the standard errors are not super huge so, we don't get much out of this exploration just as we didn't with the plot above.
+We should look into a better way of displaying them.
+
+```{r}
+plot(fit27c, pars="b")
+```
+
+There's still an interesting year effect here but not nearly as large as we see from player to player.
+
+```{r}
+plot(fit27c, pars="c")
+```
+
+Interestingly, we can see that there is some variation between stadiums.
+On the logit scale they go between (-0.2, 0.2) which on the non logit scale is (0.45,0.55).
+So a 10% change in probabilities between say, stadium 15 and stadium 32.
+I wonder, if we changede the prior to be less descriptive if we would see this resulting in even larger staium effects.
+It's certainly possible but 10% is a hugh leap anyways so I somewhat soubt that we would see much of a larger jump from a less informative prior.
+
+```{r}
+data27d <- list(
+  N = nrow(player_index_small),
+  n_year = n_distinct(player_index_small$yearID),
+  n_park = n_distinct(player_index_small$park_index),
+  n_player = n_distinct(player_index_small$player_index),
+  AB = player_index_small %>% pull(AB),
+  HR = player_index_small %>% pull(HR),
+  park = player_index_small %>% pull(park_index),
+  Year = player_index_small %>% pull(year_index),
+  cAge = player_index_small %>% pull(cAge),
+  Player = player_index_small %>% pull(player_index)
+)
+```
+
+```{r}
+fit27d <- stan(
+  file = "HR27d.stan",
+  data = data27d,
+  iter = 3000
+)
+```
+
+```{r}
+#Since this took a while let's save it quickly
+saveRDS(fit27d, "HR27d.rds")
+```
+
+So, this took a lot longer than the other simulations. Let's see how it did.
+
+```{r}
+plot(fit27d, pars="a")
+```
+
+This didn't have much of an impact compared with what it looked like before with *HR27c*.
+
+```{r}
+plot(fit27d, pars="b")
+```
+
+Agan not much of a difference.
+
+```{r}
+plot(fit27d, pars="c")
+```
+
+Still not much of a change.
+
+```{r}
+plot(fit27d, pars="beta1")
+```
+
+We can see the age factor has an almost negligable effect on player performance, at least when we only allow the effect to be linear.
+Fellingham and Fisher allowed for it to be non-linear so the next steps should be focused on investigating that relationship to see if it adds to our model at all.
+
+START EXPLORING NON-LINEAR EFFECT OF AGE.

### 2023-11-09 Lahman Analysis/Rough Analysis 2/Analysis2.qmd
I think it wants me to reupload everything
Started 10 year analysis

@@
 ```{r}
+#| eval: false
 fit27d <- stan(
@@
 
-We can see the age factor has an almost negligable effect on player performance, at least when we only allow the effect to be linear.
+We can see the age factor has an almost negligible effect on player performance, at least when we only allow the effect to be linear.
 Fellingham and Fisher allowed for it to be non-linear so the next steps should be focused on investigating that relationship to see if it adds to our model at all.
 
-START EXPLORING NON-LINEAR EFFECT OF AGE.
+### Nonlinear effects of Age
+
+Before diving into the full quadratic that Fellingham and Fisher used, we should first check out what it looks like by simply making it non-linear.
+
+```{r}
+data28 <- list(
+  N = nrow(player_index_small),
+  n_year = n_distinct(player_index_small$yearID),
+  n_park = n_distinct(player_index_small$park_index),
+  n_player = n_distinct(player_index_small$player_index),
+  AB = player_index_small %>% pull(AB),
+  HR = player_index_small %>% pull(HR),
+  park = player_index_small %>% pull(park_index),
+  Year = player_index_small %>% pull(year_index),
+  Player = player_index_small %>% pull(player_index),
+  cAge = player_index_small %>% pull(cAge),
+  cAgeSq = (player_index_small %>% pull(cAge))**2
+)
+```
+
+```{r}
+fit28 <- stan(
+  file = "HR28.stan",
+  data = data28,
+  iter = 4000
+)
+```
+
+```{r}
+saveRDS(fit28, "HR28.rds")
+```
+
+Now we can plot to see what the effect is on the different intercepts.
+
+```{r}
+plot(fit28, pars="a")
+```
+
+Nothing really has changed here in the plot compared to the other previous ones.
+
+```{r}
+plot(fit28, pars="b")
+```
+
+```{r}
+plot(fit28, pars="c")
+```
+
+```{r}
+summary(fit28, 'beta1')$summary
+summary(fit28, 'beta2')$summary
+```
+
+We can see from these that the beta1 and beta2 are much lower than we expected with our priors.
+So, we can probably make their priors more confident in the remaining cubic and quadratic terms of the next model.
+
+```{r}
+data28b <- list(
+  N = nrow(player_index_small),
+  n_year = n_distinct(player_index_small$yearID),
+  n_park = n_distinct(player_index_small$park_index),
+  n_player = n_distinct(player_index_small$player_index),
+  AB = player_index_small %>% pull(AB),
+  HR = player_index_small %>% pull(HR),
+  park = player_index_small %>% pull(park_index),
+  Year = player_index_small %>% pull(year_index),
+  Player = player_index_small %>% pull(player_index),
+  cAge = player_index_small %>% pull(cAge),
+  cAgeSq = (player_index_small %>% pull(cAge))**2,
+  cAgeCb = (player_index_small %>% pull(cAge))**3
+)
+```
+
+```{r}
+fit28b <- stan(
+  file = "HR28b.stan",
+  data = data28b,
+  iter = 2000
+)
+```
+
+So that took a super long time and I was still running into the error of log(0), and that was after taking out the quadratic term.
+
+```{r}
+summary(fit28b, 'beta1')$summary
+summary(fit28b, 'beta2')$summary
+summary(fit28b, 'beta3')$summary
+```
+
+So, these are huge with big standard errors suggesting that we need more iterations.
+
+Chains aren't mixing even after 2000 iterations, it took forever to run, and the RDS is too big for GitHub to handle so I think we can chalk this up as a failure.
+
+## 10 Year Examination
+
+At the recommendation of Dr.
+Parson, I think that it would be really interesting to consider a 10 year sample of MLB seasons rather than just doing to the 2-year one done above.
+I liked how the HR27d analysis came out so we'll copy that down here.
+
+```{r}
+ten_year_batting <- Batting %>% 
+  #Consider only two years to get a managable amount of players
+  filter(yearID >= 2010 & yearID <= 2019) %>% 
+  #Number the players for the multilevel model
+  mutate(player_index = as.factor(playerID), .after = playerID) %>% 
+  mutate(player_index = as.integer(player_index)) %>% 
+  mutate(cAge = Age - 30, .after = Age) %>% 
+  mutate(year_index = yearID - min(yearID) + 1, .after = yearID) %>%
+  #For this test of conecpt we'll need to redefine the stadium's index since not all stadiums in the original dataset were used in 2018 and 2019
+  select(!park_index) %>% mutate(park = as.character(park)) %>%
+  mutate(park = as.factor(park)) %>%
+  mutate(park_index = as.integer(park), .after = park)
+```
+
+```{r}
+data29 <- list(
+  N = nrow(ten_year_batting),
+  n_year = n_distinct(ten_year_batting$yearID),
+  n_park = n_distinct(ten_year_batting$park_index),
+  n_player = n_distinct(ten_year_batting$player_index),
+  AB = ten_year_batting %>% pull(AB),
+  HR = ten_year_batting %>% pull(HR),
+  park = ten_year_batting %>% pull(park_index),
+  Year = ten_year_batting %>% pull(year_index),
+  cAge = ten_year_batting %>% pull(cAge),
+  Player = ten_year_batting %>% pull(player_index)
+)
+```
+
+```{r}
+fit29 <- stan(
+  file = "HR29.stan",
+  data = data29,
+  iter = 500
+)
+```

### 2023-11-10 Lahman Analysis/Rough Analysis 2/Analysis2.qmd
11/10 Meeting

@@
 
+### 11/10 Meeting
+
+```{r}
+N_player <- Batting %>% 
+  distinct(playerID) %>% 
+  pull(playerID) %>% 
+  sample(size = 100, replace = FALSE)
+```
+
+```{r}
+Samp_Batting <-  Batting %>% 
+  filter(playerID %in% N_player)
+```
+
+```{r}
+Samp_Batting <- Samp_Batting %>% 
+  select(!park_index) %>% 
+  mutate(park = as.character(park)) %>% 
+  mutate(park = as.factor(park)) %>% 
+  mutate(park_index = as.integer(park), .after = park) %>% 
+  mutate(player_index = as.factor(playerID), .after = playerID) %>% 
+  mutate(player_index = as.integer(player_index))
+```
+
+```{r}
+data28aa <- list(
+  N = nrow(Samp_Batting),
+  n_player = n_distinct(Samp_Batting$player_index),
+  AB = Samp_Batting %>% pull(AB),
+  HR = Samp_Batting %>% pull(HR),
+  Player = Samp_Batting %>% pull(player_index),
+  cAge = Samp_Batting %>% pull(Age) -30
+)
+```
+
+```{r}
+fit28aa <- stan(
+  file = "HR28aa.stan",
+  data = data28aa,
+  iter = 200
+)
+```
+
+```{r}
+plot(fit28aa, pars="beta0")
+```
+
+```{r}
+plot(fit28aa, pars="beta1")
+```
+
+```{r}
+data28ab <- list(
+  N = nrow(Samp_Batting),
+  n_player = n_distinct(Samp_Batting$player_index),
+  AB = Samp_Batting %>% pull(AB),
+  HR = Samp_Batting %>% pull(HR),
+  Player = Samp_Batting %>% pull(player_index),
+  cAge = Samp_Batting %>% pull(Age) -30,
+  cAgeSq = (Samp_Batting %>% pull(Age) -30)**2
+)
+```
+
+```{r}
+fit28ab <- stan(
+  file = "HR28ab.stan",
+  data = data28ab,
+  iter = 500
+)
+```
+
+```{r}
+plot(fit28ab, pars="beta0")
+plot(fit28ab, pars = "beta1")
+plot(fit28ab, pars = "beta2")
+```
+
+Start plotting individual player's home run "trajectories" to see how the model is doing.
+Should be a parabola from the means of the posterior.
+This would show the logit of player home runs.
+Prefer players who have lots of seasons to start and then do it for players who have very little seasons.
+
+After, consider either more than 100 players or add in other effects like park effects.
+
 ## 10 Year Examination
@@
 ```
+
+200 iterations took 15 minutes. So, we should wait until we have some time to render this or really try and get this to work on my PC.

### 2023-11-13 Lahman Analysis/Parson multilevel/splines_test.qmd
Explain orthogonal polynomials and spline fits and create multilevel model
In this Quarto file, I discuss "poly", the command that produces the kind of orthogonal polynomial fits that Fellingham and Fisher use (I think). I also discuss B-splines, which work essentially the same way. Finally, I create a multilevel model fitting a quartic in age for each player's logit HR probability.

---
title: "Untitled"
format: html
editor: visual
---

In this document, we demonstrate fitting polynomial and splines models for home-run probability using age as predictor. Our purpose is to demonstrate how to use `poly` with orthogonal polynomials and how to use `bs` for B-splines.

# Load packages and Lahman data

```{r}
#| warning: false
library(tidyverse)
library(rstan)
library(bayesplot)
library(Lahman)
library(splines)
```

```{r}
options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)
```

We'll load the data here to match Fellingham and Fisher:

-   Players with at least 6 seasons and 50 at-bats each season.

-   Cut off seasons at age 45.

-   Fellingham and Fisher consider seasons from 1871 to 2008. We'll look at 1973 - 2019.

```{r}
#| message: false

#Creates a DF called "Batting" which has players stats from 1973-2019

#Load the DF's from Lahman directly
Batting <- Lahman::Batting
Teams <- Lahman::Teams
People <- Lahman::People

Batting <- Batting %>% 
  #Minimum year = DH was introduced; maximum year = before COVID
  filter(yearID >= 1973 & yearID <= 2019)

#Filter teams for years considered and select only necessary columns
Teams <- Teams %>% 
  filter(yearID>=1973 & yearID <= 2019) %>% 
  select(yearID, teamID, park)

#Combine current "Batting" and "Teams" DF to associate player with stadium played at
Batting <- Batting %>% 
  right_join(Teams, by = join_by(yearID, teamID))

#Add player's age to the "Batting" DF and remove seasons with age > 45
Batting <- Batting %>% 
  right_join(Lahman::People, by = join_by(playerID)) %>% 
  mutate(Age = yearID - birthYear, .after=yearID) %>%
  #Don't consider the people who don't have a birthday in the database
  drop_na(Age) %>% filter(Age <= 44)

#Make the lgID variable into an index variable
Batting <- Batting %>% 
  mutate(lg = case_when(lgID == "NL" ~ 1,
                        lgID == "AL" ~ 2), .after=lgID)

players <- Batting %>% group_by(playerID, Age) %>%
  summarize(AB = sum(AB),
            HR = sum(HR)) %>% group_by(playerID) %>%
  filter(min(AB) >= 50, n() >= 6) %>% ungroup
```

```{r}
logit <- function(p){
  log(p / (1 - p))
}
```

```{r}
Samp_Batting <-  players %>% 
  select(AB, HR, Age) %>% group_by(Age) %>%
  summarize(AB = sum(AB), HR = sum(HR)) %>%
  mutate(not_HR = AB - HR, p = HR / AB, lp = logit(p))

ages <- seq(19, 44)
```

# Polynomial model

We start with a polynomial model for the home-run probability. To illustrate how the `poly` command (a shortcut that also gives us access to orthogonal polynomials) works, we start with a classical model and move on later to a Bayesian model.

## Fitting a quartic model by hand

We begin with a quartic model, fitting it as we would have in MATH 313.

In the formula in the `glm` command below, we include the constant term `1 +` explicitly, even though it is not necessary. Since we have to include it when we write Stan code for our models, it may be helpful to see it in the classical models as well.

```{r}
fit1_poly <- glm(cbind(HR, not_HR) ~ 1 + Age + I(Age^2) + I(Age^3) + I(Age^4),
                 family = "binomial",
                 data = Samp_Batting)

summary(fit1_poly)

```

We can graph the fitted model as follows. In this graph---and in the other graphs of binomial logistic models---I add a scatterplot of empirical logits to help us assess the fit. This fit does not look bad, especially for players under 40. It suggests that, on average, players in this group had their highest home-run probabilities (about 3.3%) around age 30. The home run probabilities in the fitted model increase quickly among younger players from about 2.2% at age 19 to a maximum around age 30 and then decline gradually after 30, reaching about 2.8% at age 40.

```{r}
ggplot() +
  geom_line(mapping = aes(x = ages,
                           y = fit1_poly$coefficients[1] +
                            fit1_poly$coefficients[2] * ages +
                            fit1_poly$coefficients[3] * ages^2 +
                            fit1_poly$coefficients[4] * ages^3 +
                            fit1_poly$coefficients[5] * ages^4)
             ) +
  geom_point(data = Samp_Batting, mapping = aes(x = Age, y = lp)) +
  labs(x = "Age",
       y = "log odds of HR",
       title = "Quartic model fit by hand")
```

## Fitting a quartic model with `poly` (raw polynomial)

The following chunk fits the model using `poly`, which automates the cubic regression that we set up by hand above. As when we set the model up by hand, we include the explicit constant term `1 +`, since we will need to include it in our Stan code later. Note that the coefficients and standard errors in this model are identical to the coefficients and standard errors in the model that we fit by hand.

```{r}
fit2_poly <- glm(cbind(HR, not_HR) ~ 1 + poly(Age, degree = 4, raw = T),
                 family = "binomial",
                 data = Samp_Batting)

summary(fit2_poly)
```

We graph the model that we fit using `poly` with the following code. It is the same as the model that we fit by hand!

```{r}
B_raw <- poly(Samp_Batting$Age, degree = 4, raw = T)

# Add a column at the start for intercept. Probably a better way to do this!

C_raw <- rep(1, length(ages)) %>% cbind(B_raw %>% predict(newdata = ages))

ggplot() +
  geom_line(mapping = aes(x = ages,
                          y = C_raw %*% fit2_poly$coefficients)) +
  geom_point(data = Samp_Batting, mapping = aes(x = Age, y = lp)) +
  labs(x = "Age",
       y = "log odds of HR",
       title = "Quartic model fit with poly (raw terms)")
```

### How does `poly` work?

Let's look at the matrix `B_raw` that `poly` constructs:

```{r}
B_raw
```

Each row has the form `Age Age^2 Age^3 Age^4`, where `Age` is one of the ages (ranging from 19 to 44) from our data. We add a column of 1s at the start to make `C_raw`:

```{r}
C_raw
```

When I made `C_raw`, I did something mysterious with `predict`. That step is not really necessary here. I will come back to it later!

To use `C_raw`, we multiply it by a column vector of length 5. If that vector is `c(w1, w2, w3, w4, w5)`, we get a new column vector of length 26 whose entries are `w1 + w2*Age + w3*Age^2 + w4 * Age^3 + w5 * Age^4`, where `Age` runs from 19 to 44. The entries of that vector are the predicted ages for the quartic model with coefficients `w1, w2, w3, w4, w5`.

In particular, to see the predictions (of logit probability of home runs) for the fitted model, we multiply by `c(-9.70, 0.629, -0.0234, 0.000398, -0.00000272)`

## Fitting a quartic model with `poly` (orthogonal)

We can also fit a model using "orthogonal polynomials" customized to our predictor `Age`. Before fitting the model, let me show you how those polynomials look. The following chunk graphs the five special polynomials that we use for our quartic model. The first polynomial is just the constant function `f0(x) = 1`. We also have a linear polynomial `f1(x)`, quadratic polynomial `f2(x)`, cubic polynomial `f3(x)`, and quartic polynomial `f4(x)`. We use these four polynomials in place of `x`, `x^2`, `x^3`, and `x^4`, which are what we used when fitting a quartic polynomial by hand and using `poly` with the `raw = TRUE` option.

The "orthogonality" property of these new polynomials includes a sort of centering and some normalization. The four polynomials all have about the same magnitude on the interval \[19, 44\] of ages that we consider. This is a big contrast with how the "raw" polynomials `x`, `x^2`, `x^3`, and `x^4` behave on that interval. Using orthogonal polynomials in the Bayesian setting may make it easier to set priors.

```{r}
ages_fine <- seq(19, 44, by = 0.1)

B_ortho <- poly(Samp_Batting$Age, degree = 4)

C_ortho <- cbind("0" = rep(1, length(ages_fine)),
                     B_ortho %>% predict(newdata = ages_fine))

ggplot(mapping = aes(x = ages_fine)) +
  geom_line(mapping = aes(y = C_ortho[, 1])) +
    geom_line(mapping = aes(y = C_ortho[, 2])) +
    geom_line(mapping = aes(y = C_ortho[, 3])) +
    geom_line(mapping = aes(y = C_ortho[, 4])) +
    geom_line(mapping = aes(y = C_ortho[, 5])) +
  labs(y = "", x = "Age", title = "Five orthogonal polynomials for Age")
```

We next fit a polynomial regression using these polynomials. The regression is looking for the choices of weights `w0`, `w1`, `w2`, `w3`, `w4` so that the polynomial function `w0 + w1 * f1(x) + w2 * f2(x) + w3 * f3(x) + w4 * f4(x)` gives the best fit (maximum likelihood) for our binomial logistic model.

As above, we include the explicit constant term `1 +` to prepare ourselves for writing Stan code later.

```{r}
fit3_poly <- glm(cbind(HR, not_HR) ~ 1 + poly(Age, degree = 4),
                 family = "binomial",
                 data = Samp_Batting)

summary(fit3_poly)
```

We can graph this model as well. It matches what we found when we fit the binomial model in the two other ways! In other words, if we looked up the formulas for the four orthogonal polynomials `f1(x)`, `f2(x)`, `f3(x)`, `f4(x)` and used them to expand the model function `w0 + w1 * f1(x) + w2 * f2(x) + w3 * f3(x) + w4 * f4(x)` in powers of `x`, we would get exactly the same polynomial in `x` as in our earlier model.

```{r}
B_ortho <- poly(Samp_Batting$Age, degree = 4)

C_ortho <- cbind("0" = rep(1, length(ages)),
                     B_ortho %>% predict(newdata = ages))

ggplot() +
  geom_line(mapping = aes(x = ages,
                          y = C_ortho %*% fit3_poly$coefficients))  +
  geom_point(data = Samp_Batting, mapping = aes(x = Age, y = lp)) +
  labs(x = "Age",
       y = "log odds of HR",
       title = "Quartic model fit with orthogonal polynomials")
```

### Orthogonal polynomials and Stan

The othogonalization process that `poly` uses is a general method that could be applied to any collection of predictors. In a somewhat puzzling section, in fact, the Stan user's guide recommends this step as a standard part of processing before fitting a model. The discussion and sample code are in Section 1.2 \[<https://mc-stan.org/docs/stan-users-guide/QR-reparameterization.html>\], with the title "QR reparameterization." One might consider using that code as a replacement for pre-processing with `poly`.

# Spline fit

We learned about B-splines in Chapter 4 of *Statistical Rethinking*. We can use B-splines in place of a quartic polynomial model. In this case, the quartic model actually looks fine (to me at least!), but perhaps it's worth comparing with splines when one passes to Bayesian multilevel models.

We will try a more-complex spline model than our quartic models above. Those quartic models have 5 degrees of freedom, in the sense that there are 5 coefficients to fit in the model. We will fit a B-spline with 8 degrees of freedom (`df = 8` as an option in `bs`) plus 1 degree of freedom for a constant term. I picked 8 by stepping up from 5 degrees of freedom until the spline model looked like a very good fit---maybe an overfit!---for the empirical logits.

Before we fit the model, let's see how this spline basis looks. The `bs` command, which was saw in *Statistical Rethinking*, picks the knots using quantiles of `Age` when I tell it the number of degrees of freedom that I want (8) and the degree of the spline functions (3 for cubic splines, which are a standard choice). We add the constant function `1` to this B-spline basis.

An aside about the knots determined by `bs`: this R command just uses the list of ages, which run from 19 to 44, to determine the knots. It might be better to weight the ages by, say, the number of at-bats in our data set for each age or by the number of player-seasons of each age. If we did that, the knots would no longer be evenly spaced. We would have greater resolution at the ages that are more common in the data. I tried fitting splines using knots determined by the quantiles of ages weighted in this way, and it did not appear to make much of a difference. To keep things simple in this file, I discarded that work.

```{r}
ages_fine <- seq(19, 44, by = 0.1)

B_spline <- bs(Samp_Batting %>% pull(Age), df = 8, degree = 3)

C_spline <- rep(1, length(ages_fine)) %>%
  cbind(B_spline %>% predict(newx = ages_fine))

ggplot(mapping = aes(x = ages_fine)) +
  geom_line(mapping = aes(y = C_spline[, 1])) +
  geom_line(mapping = aes(y = C_spline[, 2])) +
  geom_line(mapping = aes(y = C_spline[, 3])) +
  geom_line(mapping = aes(y = C_spline[, 4])) +
  geom_line(mapping = aes(y = C_spline[, 5])) +
  geom_line(mapping = aes(y = C_spline[, 6])) +
  geom_line(mapping = aes(y = C_spline[, 7])) +
  geom_line(mapping = aes(y = C_spline[, 8])) +
  geom_line(mapping = aes(y = C_spline[, 9])) +
  labs(y = "", x = "Age", title = "B-splines for Age with df = 8 and constant")
```

## Aside on intercept

McElreath first fits a B-spline model using the option `intercept = TRUE`. I'm not convinced that he understands what that option does! Let's add that option and graph the resulting splines, omitting the constant function. We have to add one more degree of freedom for `bs`, since with this option, we do not need a separate intercept term in a spline model.

When we compare the graph of these nine spline functions with the graph of the eight spline functions above, we can see what is going on. With the default `intercept = FALSE` option, R drops the first B-spline from the set.

```{r}
ages_fine <- seq(19, 44, by = 0.1)

B_spline_int <- bs(Samp_Batting %>% pull(Age), df = 8 + 1,
               degree = 3, intercept = TRUE)

C_spline <- B_spline_int %>% predict(newx = ages_fine)

ggplot(mapping = aes(x = ages_fine)) +
  geom_line(mapping = aes(y = C_spline[, 1])) +
  geom_line(mapping = aes(y = C_spline[, 2])) +
  geom_line(mapping = aes(y = C_spline[, 3])) +
  geom_line(mapping = aes(y = C_spline[, 4])) +
  geom_line(mapping = aes(y = C_spline[, 5])) +
  geom_line(mapping = aes(y = C_spline[, 6])) +
  geom_line(mapping = aes(y = C_spline[, 7])) +
  geom_line(mapping = aes(y = C_spline[, 8])) +
  geom_line(mapping = aes(y = C_spline[, 9])) +
  labs(y = "", x = "Age",
       title = "B-splines for Age with df = 8 + 1 and `intercept = T`")
```

None of these spline basis functions looks like the constant function 1, which is what we normally think of as an intercept term. If we sum those 9 functions though, we get the constant 1, as the following chunk confirms.

```{r}
rowSums(B_spline_int)
```

We will compare below the coefficients for fitting a spline model with `intercept = FALSE` and `intercept = TRUE`.

In the classical context, the default is `intercept = FALSE`, since one may want to use splines for several predictor variables. If one used `intercept = TRUE` for splines for more than one predictor, one would run into a multicollinearity problem. In the Bayesian setting, one could work around that using priors, as we've seen in other situations. There may be an argument for using `intercept = TRUE` in that setting, especially if it makes working with the priors easier. McElreath's argument from symmetry for not using classical dummy coding in the Bayesian setting may also apply here (in favor of `intercept = TRUE`). One may have to experiment further to figure out how these things work...

## Fitting the B-spline model(s)

We next fit a regression using spline functions as above. We first use the default `intercept = FALSE` option and include a constant term in our model. We then fit a model using `intercept = TRUE` and remove the constant term.

```{r}
fit1_spline <- glm(cbind(HR, not_HR) ~ 1 + bs(Age, df = 8, degree = 3),
                 family = "binomial",
                 data = Samp_Batting)

summary(fit1_spline)
```

```{r}
fit2_spline <- glm(cbind(HR, not_HR) ~ bs(Age, df = 8 + 1, degree = 3,
                                              intercept = TRUE) - 1,
                 family = "binomial",
                 data = Samp_Batting)

summary(fit2_spline)
```

We'll confirm below that these two versions of the spline models give the same fitted function. Comparing the coefficients lists shows a simple relation, which follows from the fact that in the `intercept = TRUE` model, the splines functions sum to the constant 1:

-   The intercept coefficient in the first model is the same as the coefficient for the first spline basis element in the second model. It matches the model prediction for the log odds of a home run for a 19-year-old player. That's clear from the graphs of the spline functions. The only functions in those graphs that do not vanish at age 19 are the constant function 1 and the first spline basis function in the `intercept = TRUE` set. (We can also verify this below with an R calculation.)

-   The coefficients in the first model are the differences between the first coefficient in the second model and the later coefficients. For example, the sixth coefficient in the first model is 0.4311. This number is the difference -3.42470 - (-3.85579) between the sixth and first coefficients in the second model.

-   The coefficients in the second model are roughly the sizes of predicted log odds as we move across the range of ages. The coefficients in the first model are comparisons as we move across the range of ages to the log odds for a home run for 19-year-old players.

We verify that the first coefficient in both models is the predicted log odds of a home run for a 19-year-old player.

```{r}
fit1_spline %>% predict(newdata = data.frame(Age = 19))
```

```{r}
fit2_spline %>% predict(newdata = data.frame(Age = 19))
```

```{r}
fit1_spline$coefficients[1]
```

If we were in the Bayesian setting, we could add an intercept term to the `intercept = TRUE` model and use priors to make the model treat that intercept as the overall average and the spline coefficients as smaller age adjustments to that overall average. That is what McElreath is doing, I believe, in Chapter 4, but, as I mentioned above, I am not sure that he understands the options for `bs`.

Let's convince ourselves with a graph that these two models do the same thing.

```{r}
B_spline <- bs(Samp_Batting %>% pull(Age), df = 8, degree = 3)

C_spline <- rep(1, length(ages)) %>%
  cbind(B_spline %>% predict(newx = ages))

ggplot() +
  geom_line(mapping = aes(x = ages,
                          y = C_spline %*% fit1_spline$coefficients))  +
  geom_point(data = Samp_Batting, mapping = aes(x = Age, y = lp)) +
  labs(x = "Age",
       y = "log odds of HR",
       title = "Cubic spline model intercept = FALSE")
```

```{r}
B_spline_int <- bs(Samp_Batting %>% pull(Age), df = 8 + 1, degree = 3,
               intercept = TRUE)

C_spline_int <- B_spline_int %>% predict(newx = ages)

ggplot() +
  geom_line(mapping = aes(x = ages,
                          y = C_spline_int %*% fit2_spline$coefficients))  +
  geom_point(data = Samp_Batting, mapping = aes(x = Age, y = lp)) +
  labs(x = "Age",
       y = "log odds of HR",
       title = "Cubic spline model intercept = TRUE")
```

## How does `bs` work?

The `bs` command works just like the `poly` command. It creates a matrix ("design matrix") with as many columns as there are coefficients and as many rows as there are rows in the data set. Each row has the values of the spline basis for the corresponding row of the data set. Since we only use `Age` as the predictor, the rows are functions of `Age`.

As with `poly`, when we multiply the spline matrix by the vector of coefficients, we get a column vector of predictions for log odds of a home run for each row (case) in the original data set. (If we are using the `intercept = FALSE` version, we have to add a column of 1s at the left edge of the spline matrix to provide the intercept term.) For this work, the rows of our data set correspond to the ages 19 through 44, and so this column vector gives the predicted log odds for each age.

# Modeling with Stan

We next explore fitting these sorts of models with Stan.

## Fully pooled models

We can refit any of the above fully pooled models in Bayesian style using Stan. Let's try, for example, the degree-4 orthogonal polynomial fit. I called my Stan file `spline_model1.stan`, since I first thought to use it for fitting a B-spline model. The code is quite general, just using a design matrix (the `B`-type matrix from above) and adding an intercept. One would want to adjust the priors appropriately for different sorts of models.

```{r}
B_ortho <- poly(Samp_Batting %>% pull(Age), degree = 4)

data_ortho <- list(
  N = nrow(Samp_Batting),
  n_basis = ncol(B_ortho),
  AB = Samp_Batting %>% pull(AB),
  HR = Samp_Batting %>% pull(HR),
  B = B_ortho
)

fit_ortho <- stan(
  file = "spline_model1.stan",
  data = data_ortho,
  iter = 500
)
```

As we see below, the model parameters and standard errors in the Bayesian setting match those in the classical setting.

```{r}
print(fit_ortho)
```

We can get a sense for the model probabilities by drawing some samples (say, 50) of the parameters from the posterior and graphing the models' log odds values as functions of age.

```{r}
samples <- as.data.frame(fit_ortho) %>% slice_sample(n = 50) %>% select(-"lp__")
ms <- C_ortho %*% t(as.matrix(samples)) %>% as.data.frame() %>%
  mutate(Age = ages) %>%
  pivot_longer(cols = -Age, names_to = "group", values_to = "predicted")

ggplot(data = ms) +
  geom_line(mapping = aes(x = Age, y = predicted, group = group),
            alpha = 0.1) +
  geom_point(data = Samp_Batting, mapping = aes(x = Age, y = lp)) +
  labs(x = "Age",
       y = "log odds of HR",
       title = "Quartic model fit with orthogonal polynomials")
```

To evaluate how well the model fits (retrodicts!) the original data set, we could use a diagnostic graph such as binned residual plots. (But we would not need to bin.) I have already gotten carried away, and so I'll omit such graphs from this file. (If we were to make one, we might calculate residuals for a single draw from the posterior, since residuals are parameters in the Bayesian setting!)

## Multilevel models (partial pooling)

We can now fit a multilevel model, using a quartic polynomial for each player but with coefficients drawn from a common distribution. Warning: Fitting the model with only 200 iterations took about 15 minutes. Stan complained mightily about convergence problems.

This work is a start. One could pull a few players (in the style of Fellingham and Fisher) to look at how the model fits the data. I think the model imposed pretty strong pooling on the coefficients.

```{r}
player_IDs <- players %>% distinct(playerID) %>% pull(playerID)

batting_full <- Batting %>% filter(playerID %in% player_IDs) %>%
  select(AB, HR, Age, playerID) %>% group_by(playerID, Age) %>%
  summarize(AB = sum(AB), HR = sum(HR)) %>% ungroup
```

```{r}
B_ortho_full <- poly(batting_full %>% pull(Age), degree = 4)

data_ortho_full <- list(
  N = nrow(batting_full),
  n_player = length(player_IDs),
  n_basis = ncol(B_ortho_full),
  AB = batting_full %>% pull(AB),
  HR = batting_full %>% pull(HR),
  Player = batting_full %>% pull(playerID) %>% factor %>% as.integer,
  B = B_ortho_full
)

fit_ortho_full <- stan(
  file = "spline_model2.stan",
  data = data_ortho_full,
  iter = 200
)
```

Poking around, I see that the parameters are different from what I guessed with my priors. The fitted values for the weights vary between around -20 and 20, much wider than what my priors suggested. There was enough data to swamp those priors.

The overall model (determined by using `mu` for the intercept and `nu` for the weights) looks like the quartic model that I fit to the pooled data. Picking a couple of players, I see that the individual models do look different from this overall model. The individual models, for example, may predict later peaks for home-run probability. To explore further, I would add empirical logits to the graphs. Not sure what to expect!

```{r}
print(fit_ortho_full, pars = "weight")
```

```{r}
print(fit_ortho_full, pars = "a")
```

```{r}
print(fit_ortho_full, pars = "sigma")
```

```{r}
print(fit_ortho_full, pars = "tau")
```

```{r}
print(fit_ortho_full, pars = "nu")
```

```{r}
print(fit_ortho_full, pars = "mu")
```

### 2023-11-16 Lahman Analysis/Rough Analysis 2/Analysis2.qmd
Began modeling player's trajectories

@@
 ```{r}
+#Set the seed since the players chosen keep changing
+set.seed(300)
+
 N_player <- Batting %>% 
@@
   data = data28ab,
-  iter = 500
+  iter = 1000
 )
@@
 After, consider either more than 100 players or add in other effects like park effects.
+
+### Modeling Player "Trajectories" From Model Prediction
+
+Followed [this vignette](http://mjskay.github.io/tidybayes/articles/tidy-rstanarm.html) from tidybayes.
+
+```{r}
+#Start by installing tidybayes
+
+#This shows us what the variables are called in the model (we already know this from the stan code)
+get_variables(fit28ab)
+```
+
+```{r}
+#Grabs the beta1's from the model and shows first 10
+fit28ab %>% 
+  spread_draws(beta1[group]) %>% 
+  head(10)
+```
+
+```{r}
+#Shows the first 10 beta1's again but also gives medians
+fit28ab %>% 
+  spread_draws(beta1[group]) %>% 
+  median_qi() %>% 
+  head(10)
+```
+
+```{r}
+#Gives us a tidy df of all the outputs from when we print a model fit for a specific parameter
+fit28ab %>% 
+  spread_draws(beta1[group]) %>% 
+  summarise_draws()
+```
+
+```{r}
+#Though ugly, this output should be replicating the beta1 graph above.
+fit28ab %>%
+  spread_draws(beta1[group]) %>%
+  median_qi(.width = c(.95, .8)) %>%
+  ggplot(aes(y = group, x = beta1, xmin = .lower, xmax = .upper)) +
+  geom_pointinterval()
+```
+
+```{r}
+#Note: in the vingette, condition=HR for us and ABC=Samp_Batting
+Samp_Batting %>%
+  modelr::data_grid(HR) %>%
+  add_epred_draws(fit28ab) %>%
+  head(10)
+```
+
+```{r}
+grid = Samp_Batting %>% 
+  modelr::data_grid(HR)
+
+means = grid %>% 
+  add_epred_draws(fit28ab)
+```
+
+I'm getting som errors here, so I'm going to take a break from trying to use tidybayes and give this a go another way.
+
+We takled in the meeting on 11/10 about picking a player who had a lot of seasons so we could see how their trajectory looked over the course of their long career.
+Selecting arbitrarily, playerID "bayja01" played from 2003-2013.
+So a nice 11 year long career.
+(His index is 9)
+
+```{r}
+#Grab player 9's estimates for beta0, beta1, and beta2
+b0_i9 <- fit28ab %>% 
+  spread_draws(beta0[group]) %>% 
+  summarise_draws() %>% 
+  filter(group == 9)
+
+b1_i9 <- fit28ab %>% 
+  spread_draws(beta1[group]) %>% 
+  summarise_draws() %>% 
+  filter(group == 9)
+
+b2_i9 <- fit28ab %>% 
+  spread_draws(beta2[group]) %>% 
+  summarise_draws() %>% 
+  filter(group == 9)
+
+b0_i9
+b1_i9
+b2_i9
+```
+
+```{r}
+#Combine these into one tibble
+player_9_est <- b0_i9 %>% 
+  bind_rows(b1_i9, b2_i9)
+```
+
+```{r}
+#We want to plot this guy's fit compared to how he actually did
+player_9_Samp_Batting <- Samp_Batting %>% 
+  filter(player_index==9) %>% 
+  select(playerID, player_index, yearID, Age, AB, HR) %>% 
+  mutate(cAge = Age - 30, .after = Age)
+  mutate(p = HR / AB) %>% 
+  mutate(p_logit = log(p / (1-p)))
+```
+
+```{r}
+ggplot(data = player_9_Samp_Batting)+ 
+  aes(x = cAge, y=p_logit)+
+  geom_point()+
+  geom_function(fun = function(x) (-2.962388)	+ (-0.0443416*x) + (-0.004696772*(x**2)), color="red")
+```
+
+\
+\
+\
+\
 

### 2023-11-16 Lahman Analysis/Rough Analysis 2/Analysis2.qmd
Projected a player's trajectory

@@
 library(Lahman)
+library(tidybayes)
 ```
@@
 
-I'm getting som errors here, so I'm going to take a break from trying to use tidybayes and give this a go another way.
-
-We takled in the meeting on 11/10 about picking a player who had a lot of seasons so we could see how their trajectory looked over the course of their long career.
-Selecting arbitrarily, playerID "bayja01" played from 2003-2013.
+I'm getting some errors here, so I'm going to take a break from trying to use tidybayes and give this a go another way.
+
+We talked in the meeting on 11/10 about picking a player who had a lot of seasons so we could see how their trajectory looked over the course of their long career.
+Selecting arbitrarily, playerID "jenkige01" played from 2003-2013.
 So a nice 11 year long career.
-(His index is 9)
+(His index is 46)
 
@@
 #Grab player 9's estimates for beta0, beta1, and beta2
-b0_i9 <- fit28ab %>% 
+b0_i46 <- fit28ab %>% 
   spread_draws(beta0[group]) %>% 
   summarise_draws() %>% 
-  filter(group == 9)
-
-b1_i9 <- fit28ab %>% 
+  filter(group == 46)
+
+b1_i46 <- fit28ab %>% 
   spread_draws(beta1[group]) %>% 
   summarise_draws() %>% 
-  filter(group == 9)
-
-b2_i9 <- fit28ab %>% 
+  filter(group == 46)
+
+b2_i46 <- fit28ab %>% 
   spread_draws(beta2[group]) %>% 
   summarise_draws() %>% 
-  filter(group == 9)
-
-b0_i9
-b1_i9
-b2_i9
+  filter(group == 46)
+
+b0_i46
+b1_i46
+b2_i46
 ```
@@
 #Combine these into one tibble
-player_9_est <- b0_i9 %>% 
-  bind_rows(b1_i9, b2_i9)
+player_46_est <- b0_i46 %>% 
+  bind_rows(b1_i46, b2_i46)
 ```
@@
 #We want to plot this guy's fit compared to how he actually did
-player_9_Samp_Batting <- Samp_Batting %>% 
-  filter(player_index==9) %>% 
+player_46_Samp_Batting <- Samp_Batting %>% 
+  filter(player_index==46) %>% 
   select(playerID, player_index, yearID, Age, AB, HR) %>% 
-  mutate(cAge = Age - 30, .after = Age)
+  mutate(cAge = Age - 30, .after = Age) %>% 
   mutate(p = HR / AB) %>% 
@@
 ```{r}
-ggplot(data = player_9_Samp_Batting)+ 
+#This plot shows us how he actually did with the black points compared to how we would predict him to do, the red line
+ggplot(data = player_46_Samp_Batting)+ 
   aes(x = cAge, y=p_logit)+
   geom_point()+
-  geom_function(fun = function(x) (-2.962388)	+ (-0.0443416*x) + (-0.004696772*(x**2)), color="red")
-```
-
-\
+  geom_function(fun = function(x) (-2.967667)	+ (-0.0481313*x) + (-0.00995784*(x**2)), color="red")
+```
+
+```{r}
+#Extending the x-axis we can see how our model would predict him to do from the ages of 20-40 even though he only played from when he was 24-34.
+ggplot(data = player_46_Samp_Batting)+ 
+  aes(x = cAge, y=p_logit)+
+  geom_point()+
+  geom_function(fun = function(x) (-2.967667)	+ (-0.0481313*x) + (-0.00995784*(x**2)), color="red")+
+  xlim(-10, 10)
+```
+
+It is difficult to tell what the logit scale means, so in terms of probability, -3.5 would be about 3%.
+So, we would predict him to be an average home run hitter when he entered the league at 20.
+Then it peaks around -2.85 which is about 5.5%.
+And then when he's 40 he has an expected home run rate of -4.5 which is about 1.1%.
+
+This outlines the (somewhat laborious) process of how to extrapolate predictions from our model for specific players.
+Thus, I think the next thing to do would be to add in different factors, like park and year factors since I believe that they would give our model the strongest edge.
+I will do this on a new document though because this one is getting unwieldy.\
 \

### 2023-11-16 Lahman Analysis/Rough Analysis 2/Analysis3.qmd
Projected a player's trajectory

---
title: "Analysis 3"
format: html
---

# Load Libraries

### 2023-11-16 Lahman Analysis/Rough Analysis 2/Analysis3.qmd
Added park effects and tried quartic

@@
 # Load Libraries
+
+```{r}
+#| warning: false
+library(tidyverse)
+library(rstan)
+library(bayesplot)
+library(Lahman)
+library(tidybayes)
+```
+
+```{r}
+options(mc.cores = parallel::detectCores())
+rstan_options(auto_write = TRUE)
+```
+
+# Clean Lahman
+
+```{r}
+#| message: false
+
+#Creates a DF called "Batting" which has players stats from 1973-2019
+
+#Load the DF's from Lahman directly
+Batting <- Lahman::Batting
+Teams <- Lahman::Teams
+People <- Lahman::People
+
+Batting <- Batting %>% 
+  #Minimum year = DH was introduced; maximum year = before COVID
+  filter(yearID >= 1973 & yearID <= 2019)
+
+#Filter teams for years considered and select only necessary columns
+Teams <- Teams %>% 
+  filter(yearID>=1973 & yearID <= 2019) %>% 
+  select(yearID, teamID, park)
+
+#Combine current "Batting" and "Teams" DF to associate player with stadium played at
+Batting <- Batting %>% 
+  right_join(Teams, by = join_by(yearID, teamID))
+
+#Add player's age to the "Batting" DF
+Batting <- Batting %>% 
+  right_join(Lahman::People, by = join_by(playerID)) %>% 
+  mutate(Age = yearID - birthYear, .after=yearID) %>%
+  #Don't consider the people who don't have a birthday in the database
+  drop_na(Age)
+
+#Make the lgID variable into an index variable
+Batting <- Batting %>% 
+  mutate(lg = case_when(lgID == "NL" ~ 1,
+                        lgID == "AL" ~ 2), .after=lgID)
+```
+
+```{r}
+#We need to make the data here "slim" like is suggested by McElreath
+slim <- list(
+  N = nrow(Batting),
+  z = n_distinct(Batting %>% pull(yearID)),
+  Year = as.integer(Batting %>% pull(yearID) - 1972),
+  Age = as.integer(Batting %>% pull(Age)-17),
+  LG = as.integer(Batting %>% pull(lg)),
+  AB = Batting %>% pull(AB),
+  HR = Batting %>% pull(HR),
+  SB = Batting %>% pull(SB),
+  BirthYear = as.integer(Batting %>% pull(birthYear)-1922)
+)
+```
+
+```{r}
+#Define logit for convenience
+
+inv_logit <- function(p){
+  1 / (1 + exp(-p))
+}
+```
+
+```{r}
+#Create the summarized data frame
+
+Sum.Batting <- Batting %>% 
+  group_by(yearID, Age) %>% 
+  summarize(AB = sum(AB),
+            HR = sum(HR))
+```
+
+```{r}
+#Make an index for the parks to add them to the model
+
+#The index goes from 1-88. No idea how it came up with them but I kept the park name in Batting so anyone should be able to reference to see how the parks impact player performance
+
+Batting <- Batting %>% 
+  mutate(park = as.factor(park)) %>% 
+  mutate(park_index = as.integer(park), .after = park)
+```
+
+```{r}
+#Create a column which described the decade a player was born in
+
+Batting %>% 
+  #Put the birth year down to the decade value. (i.e. 1936 becomes 193.6)
+  #Then force it to be an integer. R doesn't round so we are left with (1936=193) a decade
+  mutate(birthDecade = as.integer(birthYear/10), .before=birthYear) %>% 
+  #Then, index it by making the minimum year equal to 1
+  mutate(birthDecade = birthDecade - min(birthDecade) + 1)
+```
+
+# Player Trajectories
+
+This is a continuation of where I left off `Analysis 2` because the document was getting unwieldy.
+When we left off, we were able to model the career HR probability of a certain player and extrapolate that to consider what his HR's would have looked like if he had played from 20-40 years-old.
+
+That model was a very simple multilevel model and is outlined below.
+In this part we seek to consider more factors to make our model more refined.
+Mainly, we seek to consider things such as park factors, year played, and putting higher order polynomials to better model the non-linear effect of age.
+
+Our current model as of starting this document:
+
+$$
+HR\sim Binomial(AB, p)\\
+p=\beta_0+\beta_1\cdot CenteredAge+\beta_2\cdot(CenteredAge)^2\\
+\beta_0\sim Normal(\mu_0,\sigma_0)\\
+\beta_1\sim Normal(\mu_1,\sigma_1)\\
+\beta_2\sim Normal(\mu_2,\sigma_2)\\
+\mu_0\sim Normal(-3.5, 0.1)\\
+\sigma_0\sim Exponential(1)\\
+\mu_1\sim Normal(0, 0.1)\\
+\sigma_1\sim Exponential(10)\\
+\mu_2\sim Normal(0, 0.01)\\
+\sigma_2\sim Exponential(100)
+$$
+
+## Clean Random Sample
+
+We need to clean the random sample of 100 considered players in this smaller model.
+
+```{r}
+#Set a seed so that we don't keep grabbing random 100 players whenever this chunk re-runs
+set.seed(400)
+
+#Grab 100 random playerID's from our Batting df
+N_player <- Batting %>% 
+  distinct(playerID) %>% 
+  pull(playerID) %>% 
+  sample(size = 100, replace = FALSE)
+```
+
+```{r}
+#Make a df which keeps only the 100 players selected at random
+Samp_Batting <-  Batting %>% 
+  filter(playerID %in% N_player)
+```
+
+```{r}
+#Fully clean the Batting df to suit our future needs
+Samp_Batting <- Samp_Batting %>% 
+  #This part just re-indexes the parks to fit our subset
+  select(!park_index) %>% 
+  mutate(park = as.character(park)) %>% 
+  mutate(park = as.factor(park)) %>% 
+  mutate(park_index = as.integer(park), .after = park) %>% 
+  #Give the players a unique index
+  mutate(player_index = as.factor(playerID), .after = playerID) %>% 
+  mutate(player_index = as.integer(player_index)) %>% 
+  #Give the players a centered age (centered at 30)
+  mutate(cAge = Age-30, .after = Age)
+```
+
+## Consider Park Factors
+
+In my past explorations, I found that park factors can have huge influences on player's ability to hit HR's.
+So, the first way to improve this model would be incorporating park effects into this analysis.
+
+```{r}
+data3a <- list(
+  N = nrow(Samp_Batting),
+  n_player = n_distinct(Samp_Batting$player_index),
+  AB = Samp_Batting %>% pull(AB),
+  HR = Samp_Batting %>% pull(HR),
+  Player = Samp_Batting %>% pull(player_index),
+  cAge = Samp_Batting %>% pull(Age) -30,
+  cAgeSq = (Samp_Batting %>% pull(Age) -30)**2,
+  Park = Samp_Batting %>% pull(park_index)
+)
+```
+
+```{r}
+fit3a <- stan(
+  file = "HR3a.stan",
+  data = data3a,
+  iter = 1000
+)
+```
+
+Stan is complaining to me about R-hats and whatnot. I'll go back and iterate over this more on my PC but for the time being this is purely a test of concept, so we'll have to deal with imperfection.
+
+In the meantime we can explore how everything looks in the plots below:
+
+```{r}
+plot(fit3a, pars="beta0")
+plot(fit3a, pars = "beta1")
+plot(fit3a, pars = "beta2")
+plot(fit3a, pars = "beta5")
+```
+
+## Consider Year Factors
+
+As I ponder it, I realize that adding in a year effect at this point may not make much sense.
+I want the year's index to represent a change of 1 year.
+However, from my sample, I consider a lot of years and it is more than likely that there are "gaps" in my years considered from the players selected.
+That is, it is more than likely since I pick players at random that the years considered could follow a time interval such as \[1973-1980\] U \[1982-1990\].
+Notice in this interval we are missing the year 1981.
+However, from the index's point of view, there is no gap.
+It would give \[8\] to 1980 and \[9\] to 1982.
+But the jump from index \[8\] to \[9\] would not be missing the jump from 1980 to 1981 and the jump from 1981 to 1982.
+So, I think I'll leave the year effect out since it wouldn't describe what I want it to, nor would it be easy to interpret.
+
+## Consider Cubic and Quartic Polynomials.
+
+This then leaves us with considering the higher order of the polynomials.
+
+Adding these to the analysis will be really easy. Just continue the format for the previous betas for beta3 and beta4.
+
+```{r}
+data3b <- list(
+  N = nrow(Samp_Batting),
+  n_player = n_distinct(Samp_Batting$player_index),
+  AB = Samp_Batting %>% pull(AB),
+  HR = Samp_Batting %>% pull(HR),
+  Player = Samp_Batting %>% pull(player_index),
+  cAge = Samp_Batting %>% pull(Age) -30,
+  cAgeSq = (Samp_Batting %>% pull(Age) -30)**2,
+  cAgeCb = (Samp_Batting %>% pull(Age) -30)**3,
+  cAgeQa = (Samp_Batting %>% pull(Age) -30)**4,
+  Park = Samp_Batting %>% pull(park_index)
+)
+```
+
+```{r}
+fit3b <- stan(
+  file = "HR3b.stan",
+  data = data3b,
+  iter = 200
+)
+```
+
+I keep getting errors with this. I believe that I've gotten these problems before but I cant remember how we solved them.

### 2023-11-17 Lahman Analysis/Rough Analysis 2/Analysis3.qmd
Add 3aa model and update 3b model
The 3b still does not work, apparently because the initial conditions (random between -2 and 2) are too extreme for the quartic design. We could fix that by setting initial conditions or by using orthogonal polynomials in place of a raw quartic.

@@
 
+## Adjust 3a model
+
+In the 11/17 meeting, we changed how parks entered the model.
+
+```{r}
+data3aa <- list(
+  N = nrow(Samp_Batting),
+  n_player = n_distinct(Samp_Batting$player_index),
+  n_park = n_distinct(Samp_Batting$park_index),
+  AB = Samp_Batting %>% pull(AB),
+  HR = Samp_Batting %>% pull(HR),
+  Player = Samp_Batting %>% pull(player_index),
+  cAge = Samp_Batting %>% pull(Age) -30,
+  cAgeSq = (Samp_Batting %>% pull(Age) -30)**2,
+  Park = Samp_Batting %>% pull(park_index)
+)
+```
+
+```{r}
+fit3aa <- stan(
+  file = "HR3aa.stan",
+  data = data3aa,
+  iter = 2000
+)
+```
+
+```{r}
+print(fit3aa, pars = c("mu5", "sigma5"))
+```
+
 ## Consider Year Factors
@@
   n_player = n_distinct(Samp_Batting$player_index),
+  n_park = n_distinct(Samp_Batting$park_index),
   AB = Samp_Batting %>% pull(AB),

### 2023-11-27 Lahman Analysis/Formal Analysis/Draft 1.qmd
Work from Thanksgiving Break

@@
 ---
-title: "Draft 1"
-format: html
+title: "Bayesian Data Analysis - Rough Draft"
+author: "Sam Turner"
+date: today
+format: 
+  html:
+    toc: true
+    toc-location: left
+    toc-expand: true
+    embed-resources: true
+
+execute: 
+  echo: false
+  warning: false
 ---
@@
   mutate(park_index = as.integer(park), .after = park)
+
+#Find qualified players who have over 50 AB's in each of the minimum 6 seasons played from 1973-2019
+qualified_players <- Batting %>% group_by(playerID, Age) %>% 
+  summarize(AB = sum(AB),
+            HR = sum(HR)) %>% 
+  filter(min(AB) >= 50, n() >= 6)
+
+#Filter Batting DF for qualified players
+Batting <- Batting %>% 
+  filter(playerID %in% qualified_players$playerID)
+
+#Calculated each players HR proportion "p"
+Batting <- Batting %>% 
+  mutate(p = HR / AB, .after=AB)
 ```
 
-# Fellingham & Fisher (2017)
+# Introduction and Motivations
 
@@
 This article in particular has drawn our attention because of some unique and interesting modeling choices made by the authors.
-Indeed, both Dr.
-Parson and I believe that while the initial investigation by Fellingham and Fisher (2017) is interesting, there are some questionable choices made by the authors throughout the paper.
-For example, there are few independent variables and their associated priors are very weak.
-The paper does a good job with displaying plenty of visualizations for the reader, yet, under careful examination one is left with the impression that the authors have cherry-picked certain players to focus on rather than examining players at random to examine the validity of their model.
-These reasons in tandem bring their findings into question.
+Both Dr.
+Parson and I believe that the initial investigation by Fellingham and Fisher (2017) is interesting, but there are some questionable modeling choices made by the authors throughout the paper.
+For example, there are few predictor variables and their associated priors are very weak.
+The paper does a good job with displaying plenty of visualizations for the reader, yet, under careful examination one is left with the impression that the authors have chosen to display a non-random assortment of players.
+This leaves careful readers wondering if the validity of their model is constant across all players, or if they have chosen a certain subset because these special players justify the use of their model.
+These reasons in tandem bring Fellingham and Fisher's (2017) findings into question.
 Thus, Dr.
-Parson and I have embarked to investigate the same research question, can we forecast home run potential using a Bayesian model which accounts for age, home ballpark, and other factors.
+Parson and I have embarked to investigate the same research question as them: can we use a multilevel Bayesian model to predict a player's home run performance?
 
-## Background
+## Baseball Background
 
-In Major League Baseball (MLB) the team on offense is allowed to send one player, at a time, up to home plate to "bat." Any time a player is sent up to bat this counts as a plate appearance (PA) for said player.
-Depending on its outcome the PA may or may not be considered an at-bat (AB).
-If the PA ends in the player being walked (BB \[base on balls\]) then it is not recorded as an AB because it is thought the player did not receive a fair chance at hitting the baseball.
+In Major League Baseball (MLB) the team on offense is allowed to send one player, at a time, up to home plate to "bat." Any time a player is sent up to bat this constitutes a plate appearance (PA).
+Depending on the outcome of the PA it may or may not be considered an at-bat (AB).
+If the PA ends in the player being walked (BB \[Base on Balls\]) then it is not recorded as an AB because it is thought the player did not receive a fair chance at hitting the baseball.
 On the other hand, if the PA ends in the player getting a hit or getting out, then it is assumed that he had a fair chance at hitting the ball and the PA is recorded as an AB.
+This distinction to the average fan is mostly arbitrary.
+However, in a modeling context it is very important.
+When modeling a player's ability to hit home runs (HR), as this paper is concerned with, the distinction is important because it is unfair to consider all the PA's of a player because this set of outcomes includes outcomes where it is assumed that the player has not been given a fair chance to hit the ball.
+For similar reasons, considering the subset of outcomes where it is assumed that they player has been given a fair chance to hit the ball seems more appropriate, this subset is AB's.
+Thus, this paper uses AB's as the main predictor for home runs hit by a certain player.
+Fellingham and Fisher (2017) also agree with this reasoning and use AB's, rather than PA's, in their research.
 
-The principle concern of this paper is how to model the amount of home runs (HR) a player will hit in a given season.
-Home runs are an outcome which is recorded as an AB where the player is able to touch all four bases before being "tagged out," where the defense is able to touch the player with the baseball.
-The primary way this occurs, and is the one that most readers will be familiar with, is by hitting the ball over the fence in the outfield.
-However, it is just as correct for the ball to remain in the boundaries of the field and for the player to reach all bases without being tagged out.
-Due to HR's being recorded as AB's both ourselves and Fellingham and Fisher (2017) incorporate AB's in our models rather than PA's.
+Furthermore, it is logical to ask: why not consider both?
+Incorporating more predictors can make a model more accurate at prediction, which both Fellingham and Fisher (2017) and this paper are concerned with.
+The reasons this paper does not consider both is because, especially in a Bayesian context, researchers must be concerned with multicollinearity.
+Without a doubt there is a strong correlation between the two because one is a subset of the other.
+In fact, we can see that the correlation between the two is quite strong in (REFERENCE TO CORRELATION BETWEEN AB'S AND PA'S).
+When multicollinearity is introduced into a model, it can be difficult for the Markov Chain Monte Carlo (MCMC) algorithm to parse out which of the two variables is responsible for the changes in the dependent variable.
+Thus, we neglect PA's from our model both because theory stipulates it is an inferior predictor and used in tandem with our current predictor it would cause problems for the modeling software.
 
-Another interesting quirk of baseball is that, like soccer, stadiums differ in dimensions.
+An interesting quirk of baseball is that, like soccer, stadiums differ in dimensions.
 This is not universal; baseball fields and stadiums must have the same measurements for their infields.
 Yet, for the most part baseball stadiums are allowed to configure their outfields as they please.
-This gives baseball stadiums a variety of unique shapes ranging in outfield sizes.
+This becomes important because most HR's are achieved by hitting the baseball over the outfield fence.
+Thus, fences closer to home plate lead to more HR's from all hitters, ceteris paribus.
+The non-standardized nature of stadiums' fences gives them a variety of unique shapes and ranging outfield sizes.
 For example, Fenway Park (1923-) is known to have the closest outfield wall with their "Green Monster" in left field being only 305 ft.
 from home plate.
-While the Polo Grounds (1891-1963) boasted one of the furthest walls from home play at 483 ft.
+While the Polo Grounds (1891-1963) boasted one of the furthest walls from home plate at 483 ft.
 in center field.
@@
 above sea level.
-As illustrated there is not only vastly different dimensions that baseball fields can take on but also different environments which affect a player's ability to hit home runs.
+As illustrated there are not only vastly different dimensions that baseball fields can take on but also different environments which affect a player's ability to hit home runs.
 Thus, any model which attempts to model players' ability to hit HR's needs to consider ballpark factors.
 
-It is obvious from examining the years that stadiums were used to infer that baseball has been played for a long time.
-Indeed, MLB as we know it has been played since about 1871 and the origins of the game go back even father than that.
+MLB, as we know it, has been played since about 1871 and the origins of the game go back even father than that.
 Over this time the some rules of the game have undergone incredible transformations which vastly change the nature of the game.
 For example, from 1900-1920 baseball was in the "deadball era." During this period the materials used to create baseballs made it so that the ball absorbed much more force than it does today.
-This had a plethora of implications for the game and how it was played; HR's were much less common and since they were harder to hit.
-In contrast, the period from 1994-2004 is known as the "steroid era" because of the rampant use of performance enhancing drugs (PED) used by players.
-This, in combination with other factors, led to players obliterating previous records, such as the career and season home runs records.
+This had a plethora of implications for the game and how it was played; HR's were much less common because they were harder to hit.
+In contrast, the period from 1994-2004 is known informally as the "steroid era" because of the rampant use of performance enhancing drugs (PED) used by players.
+This, in combination with other factors such as harder materials used to make baseballs, led to players obliterating a variety of previous hitting records, such as the career and season home runs records.
 It is obvious from these facts that decade played for player's had a drastic impact on their ability to produce outcomes.
@@
 
-Finally, it does not take an advanced degree to understand that age is an important factor in athletic ability.
+Finally, age is an important factor in athletic ability.
 To an extent in MLB, seniority can act both as a blessing and a curse.
@@
 
+## Literature Review
+
+As mentioned, Fellingham and Fisher (2017) have both considered and published their thoughts on this research question.
+This research question concerns predicting player home run performance using Bayesian multilevel modeling.
+To do this, Fellingham and Fisher (2017) used data from the Lahman data set spanning 1871 to 2008 to estimate the parametric weights for their model.
+Over this time frame they only considered players who had played for at least 6 seasons and who, in each of those seasons, had at least 50 AB's.
+They only considered seasons where the players were 18-years-old to 45-years-old.
+
+To predict a player's HR's they used a binomial logistic model.
+This model predicts HR's using the AB's of a player for the $n$ trials and the underlying ability of a player to hit a HR, $\pi$ as their probability of success.
+Fellingham and Fisher (2017) estimate a player's ability to hit home runs from his age in a season, decade born, season played, and home ball park.
+This leaves Fellingham and Fisher with the following model:
+
+$$
+\begin{align}
+h_{ij}|\pi_{ij}&\sim Binomial(ab_{ij},\pi_{ij})\\
+\\
+logit(\frac{\pi_{ij}}{1-\pi_{ij}})&=x_{ij}\beta_{i}+\theta_i+\delta_j
++\xi_p
+\end{align}
+$$
+
+Where $h_{ij}$ is the HR by a player $i$ in year $j$, $\pi_{ij}$ is the probability of a HR for player $i$ in year $j$, $ab$ is the number of AB's for player $i$ in year $j$, $x_{ij}$ and $\beta_i$ are vectors of the associated weights of the orthogonal quartic polynomial for a player's age on HR probabilities, $\theta_i$ is the effect of decade of birth, $\delta_j$ is the effect of season of play, and $\xi_p$ is the effect of the home ballpark.
+
+What makes this model so strange are the priors that Fellingham and Fisher (2017) chose.
+Disregarding that there is no justification for these priors, Fellingham and Fisher (2017) set the following priors for these parameters:
+
+$$
+\begin{align}
+\theta_i&\sim Normal(\mu_d,\sigma^2_{d_i}), d\in \{1...14\}, i\in\{1,...,3735\}\\
+\\
+\mu_d&\sim Normal(-2.5,var=1), d_i\in\{1,...,14\}\\
+\\
+\sigma^2_{d_i}&\sim Gamma(3,3), d_i\in\{1,...14\} \\
+\\
+\delta_j&\sim Normal(-2.5, var=4),j\in \{1,...,138\}\\
+\\
+\xi_p&\sim Normal(-2.5, var=4), p\in \{1,...,222\}\\
+\end{align}
+$$
+
+These priors are quite strange because they are so broad.
+For example, consider the priors on either $\delta_j$ or $\xi_p$.
+The priors state that on a logistic scale, it is normal for the decade of birth and season of play parameters to vary along the interval \[-6.5, 1.5\].
+On the normal probability scale, this corresponds to probability fluctuations of about (0.0015, 0.81).
+Surely there must be better priors than this which Fellingham and Fisher (2017) could have used.
+Coincidentally, this is probably why in using MCMC to estimate their parameters they needed to use 200,000 iterations.
+Likely, their model could have estimated their parameters much more easily had they given stronger priors.
+This is one of the areas this paper seeks to improve Fellingham and Fisher's (2017) methods.
+
+Another questionable choice that Fellingham and Fisher (2017) make is that they estimated their model parameters using FORTRAN.
+Given the antiquity of FORTRAN, replication of Fellingham and Fisher (2017) is quite difficult.
+Not to mention that the 200,000 iterations makes the research so computationally expensive that it makes it nearly impossible for any other research team to replicate.
+This is disappointing because the scientific method relies on replication for the advancement of knowledge.
+Thus, our research seeks to make the algorithms used in this research more accessible because we chose to use R to interface with the modeling software Stan, which is written in the (more contemporary) C++ programming language.
+
+Finally, we see that Fellingham and Fisher (2017) project the performance of a number of players from their considered 3,735.
+However, upon closer inspection these players do not seem to be randomly selected because the selected players were pretty renown during their prime.
+Nor, does Fellingham and Fisher (2017) make the assurance that these players are randomly selected.
+This should leave readers with unease because it is unclear how well their model fits most players, which seemed to be what the paper was trying to investigate.
+Although unlikely, it is conceivable that they only selected players which make their model appear the most reliable.
+Instead, our paper seeks to remedy this by considering a random assortment of players to examine the fit of our model.
+
+DISCUSS THE OTHER ARTICLE THAT DR. PARSON SHARED WITH YOU ON SIMILAR/SAME RESEARCH
+
 ## Model
+
+Given all the previous problems with Fellingham and Fisher (2017) mentioned above, there is an argument to be made that a simple generalized linear model (GLM) could be used instead of their complicated Bayesian semiparametric model.
+While it is true that a GLM could be used to predict HR's of a player it would not nearly be as accurate as a Bayesian model.
+Nor would it be able to as accurately capture the non-linearities present in the data.
+This is because the greatest strength of using the Bayesian technique is that it allows for multilevel models to be created.
+That is, Bayesian models allow for models to "learn" from other players and group certain players when there is not as much data available for them.
+In turn, this leads to better predictions about players where data is lacking.
+This "learning" is known as partial pooling, where the model identifies that parameters for each player (or year etc.) probably come from the same distribution but may not share the exact same estimates.
+
+Exploring concretely how our model works will make it clearer than discussing it in the abstract. The current model that we have is the following:
+
+$$
+\begin{align}
+HR_n &\sim Binomial(AB_n,p)\\
+logit(p) &= \beta_{0}Player_n+\beta_1cAge_n+\beta_2cAge^2_n+\beta_3cAge^3_n+\beta_4cAge^4_n+\beta_5Park
+\end{align}
+$$
+
+$$
+\begin{align}
+\beta_0&\sim Normal(\mu_0, \sigma_0)\\
+\beta_1&\sim Normal(\mu_1, \sigma_1)\\
+\beta_2&\sim Normal(\mu_2, \sigma_2)\\
+\beta_3&\sim Normal(\mu_3, \sigma_3)\\
+\beta_4&\sim Normal(\mu_4, \sigma_4)\\
+\end{align}
+$$
+
+$$
+\begin{align}
+\mu_0 &\sim Normal(-3.5, 0.1)\\
+\sigma_0&\sim Exponential(1)
+\end{align}
+$$
+
+In this model, we say that the $HR$'s of player $n$ are binomialy distriubted with number of trials for player $n$ are $AB$.
+The probability of success (hitting a HR) denoted as $p$.
+Furthermore, we measure $p$ on the logistic scale.
+We note here that $p$ changes due to a variety of factors.
+First, we noticed in the data that most hitters hit a HR in about 3 out of every 100 AB's (REFERENCE TO FIGURE WHICH SHOWS AVERAGE HR PER AB IN MLB).
+Thus, we assume that the underlying probabilty of any player in the MLB for hitting a HR is about -3.5 on the logit scale, or 0.03 on the standard probability scale.
+
+We recognize that this depends on the hitter which is where the use of a multilevel model becomes useful.
+Instead of simply using the average probability as our intercept, which is what a dummy variable would do in a GLM, we are able to allow our model choose an intercept which agrees with the data but can adapt depending on the player.
+Thus, we let $\beta_0$ be normally distributed with mean $\mu_0$ (where $\mu_0$ is normally distributed with mean -3.5 and standard deviation 0.1) and standard deviation $\sigma_0$ (where the parameter $\sigma_0$ itself is exponentially distributed).
+
+EXPLORE THE DATA HERE AND WHY A MULTILEVEL MODEL WORKS BEST
 
@@
 ```
+
+```{r}
+ggplot(data = Batting)+
+  aes(x = p)+
+  geom_histogram(bins = 75)+
+  geom_vline(xintercept=median(Batting$p, na.rm = TRUE) , color="red")+
+  labs(title = "Distribution of Players' HR Proportions",
+       subtitle = "Seasons Considered: 1973-2019",
+       x = "Proportions (p)",
+       y = "Observations")
+```
+
+```{r}
+#I want this to have a little more breaks in the x-axis to highlight where the median is
+ggplot(data = Batting)+
+  aes(x = p)+
+  geom_histogram(bins = 75)+
+  geom_vline(xintercept=median(Batting$p, na.rm = TRUE) , color="red")+
+  xlim(x=c(0,0.15))+
+  labs(title = "Constricted View of Distribution of Players' HR Proportions",
+       subtitle = "Seasons Considered: 1973-2019",
+       x = "Proportions (p)",
+       y = "Observations")
+```

### 2023-11-29 Lahman Analysis/Rough Analysis 2/Analysis3.qmd
Work from Thanksgiving Break
Made model 3aa work
Noticed that model 3aa wasn't on the "Analysis3.qmd" file for my, but was able to parse through the Git history (thanks Dr.Parson) and figure out how we made it work and then put it onto the document.

--- 2023-11-16 Lahman Analysis/Rough Analysis 2/Analysis3.qmd
@@
 
+While these previous thoughts and conerns are fair, in my ultimate model I hope to use about 1000 players or so.
+Then, it will be much less likely that all years aren't considered.
+We could also just manipulate the seed such that we make sure that all years are covered from 1973-2019.
+This may violate the randomness of the sample though.
+There is at least an argument to be made.
+
+Moreover, I want to make this model including the year factor so I have something to reference back to if I want to include the year in the ultimate model.
+
+```{r}
+data3aaa <- list(
+  N = nrow(Samp_Batting),
+  n_player = n_distinct(Samp_Batting$player_index),
+  n_park = n_distinct(Samp_Batting$park_index),
+  n_year = n_distinct(Samp_Batting$yearID),
+  AB = Samp_Batting %>% pull(AB),
+  HR = Samp_Batting %>% pull(HR),
+  Player = Samp_Batting %>% pull(player_index),
+  cAge = Samp_Batting %>% pull(Age) -30,
+  cAgeSq = (Samp_Batting %>% pull(Age) -30)**2,
+  Park = Samp_Batting %>% pull(park_index),
+  Year = Samp_Batting %>% pull(yearID) -1972
+)
+```
+
+```{r}
+fit3aaa <- stan(
+  file = "HR3aaa.stan",
+  data = data3aaa,
+  iter = 200
+)
+```
+
+This probably wouldn't even run, so why waste time on it.
+I would probably end up having the problem I had with the stadiums a couple weeks ago where (since the reference indexes weren't consecutive) Stan threw a fit and wouldn't let anything run.
+
 ## Consider Cubic and Quartic Polynomials.
@@
 ```{r}
-data3b <- list(
+data3aa <- list(
   N = nrow(Samp_Batting),
   n_player = n_distinct(Samp_Batting$player_index),
+  n_park = n_distinct(Samp_Batting$park_index),
   AB = Samp_Batting %>% pull(AB),
@@
 fit3b <- stan(
-  file = "HR3b.stan",
-  data = data3b,
+  file = "HR3aa.stan",
+  data = data3aa,
   iter = 200

### 2023-11-29 Lahman Analysis/Formal Analysis/Draft 1.qmd
Randomly Selected Players for Formal
Randomly selected 500/657 players that met the qualifying criteria in the Formal analysis

@@
   mutate(p = HR / AB, .after=AB)
+#You can see this for yourself too, but there are 657 players in this dataframe now
+```
+
+```{r}
+#Choose 500 players at random to consider
+#If this doesn't take too long, then we can come back later and change this to possibly consider all players. Or just widen the initial time-frame considered and do 1000 players
+
+set.seed(500)
+
+N_player <- Batting %>% 
+  distinct(playerID) %>% 
+  pull(playerID) %>% 
+  sample(size=500, replace=FALSE)
+```
+
+```{r}
+#Make a dataframe that only keeps the 500 players
+Samp_Batting <- Batting %>% 
+  filter(playerID %in% N_player)
+
+head(Samp_Batting)
+#If curious, this seed and selection covers all years. So, it should be able to handle the year-to-year effects
 ```
@@
 
-In this model, we say that the $HR$'s of player $n$ are binomialy distriubted with number of trials for player $n$ are $AB$.
+In this model, we say that the $HR$'s of player $n$ are binomialy distributed with number of trials for player $n$ are $AB$.
 The probability of success (hitting a HR) denoted as $p$.
@@
 First, we noticed in the data that most hitters hit a HR in about 3 out of every 100 AB's (REFERENCE TO FIGURE WHICH SHOWS AVERAGE HR PER AB IN MLB).
-Thus, we assume that the underlying probabilty of any player in the MLB for hitting a HR is about -3.5 on the logit scale, or 0.03 on the standard probability scale.
+Thus, we assume that the underlying probability of any player in the MLB for hitting a HR is about -3.5 on the logit scale, or 0.03 on the standard probability scale.
 
@@
 #| label: fig-RunsByYear
+#| fig-cap: "Cumulative Runs Scored in MLB by Year"
 
@@
 #| label: fig-HR.By.Age.All
+#| fig-cap: "Proportion of Home Runs per At-Bat by Age"
 
@@
 #| label: fig-HR.By.Age.20to40
+#| fig-cap: "Proportion of Home Runs per At-Bat by Age"
 

### 2023-11-30 Lahman Analysis/Formal Analysis/Draft 1.qmd
Ran the model
Ran the big model and saved the RDS locally becasue GitHub won't let me upload it. I will use it to reference my figures which will mean that only I can view it, so we may have to find some way that I can transport it to Dr. Parson if he wants to take a peek.

@@
 #You can see this for yourself too, but there are 657 players in this dataframe now
+
+#Create an index for each player
+Batting <- Batting %>% 
+  mutate(player_index = as.factor(playerID), .after = playerID) %>% 
+  mutate(player_index = as.integer(player_index))
+
+#Create a centered age variable by subtracting 30 from everyone's age
+
+Batting <- Batting %>% 
+  mutate(cAge = Age -30, .after = Age)
 ```
@@
 ```{r}
+#| eval: false
 #Choose 500 players at random to consider
@@
 ```{r}
+#| eval: false
 #Make a dataframe that only keeps the 500 players
@@
 
-DISCUSS THE OTHER ARTICLE THAT DR. PARSON SHARED WITH YOU ON SIMILAR/SAME RESEARCH
-
-## Model
+# Model
 
@@
 
+```{r}
+#Code to see if HRPolyFit still runs
+stan(
+  file = "HRPolyFit.stan",
+  data = data.poly.fit,
+  iter = 100
+)
+```
+
 Exploring concretely how our model works will make it clearer than discussing it in the abstract. The current model that we have is the following:
@@
 \begin{align}
-HR_n &\sim Binomial(AB_n,p)\\
-logit(p) &= \beta_{0}Player_n+\beta_1cAge_n+\beta_2cAge^2_n+\beta_3cAge^3_n+\beta_4cAge^4_n+\beta_5Park
+HR_{ni} &\sim Binomial(AB_{ni},p_{ni})\\
+logit(\frac{p_{ni}}{1-p_{ni}}) &= \alpha_{n}+\beta_h+\eta_h+\delta_p+\xi_i
 \end{align}
@@
 \begin{align}
-\beta_0&\sim Normal(\mu_0, \sigma_0)\\
-\beta_1&\sim Normal(\mu_1, \sigma_1)\\
-\beta_2&\sim Normal(\mu_2, \sigma_2)\\
-\beta_3&\sim Normal(\mu_3, \sigma_3)\\
-\beta_4&\sim Normal(\mu_4, \sigma_4)\\
+\alpha_n&\sim Normal(\mu_0, \sigma_0)\\
+\beta_h&\sim Normal(\mu_1, \sigma_1)\\
+\eta_h&\sim Normal(\mu_2, \sigma_2)\\
+\delta_p&\sim Normal(\mu_5, \sigma_5)\\
+\xi_i&\sim Normal(\mu_6, \sigma_6)\\
 \end{align}
@@
 \mu_0 &\sim Normal(-3.5, 0.1)\\
-\sigma_0&\sim Exponential(1)
+\sigma_0&\sim Exponential(1)\\
+\\
+\mu_1&\sim Normal(0,0.01)\\
+\sigma_1&\sim Exponential(10)\\
+\\
+\mu_2&\sim Normal(0,0.01)\\
+\sigma_2&\sim Exponential(100)\\
+\\
+\mu_5&\sim Normal(0,0.01)\\
+\sigma_5&\sim Exponential(10)\\
+\\
+\mu_6&\sim Normal(0,0.25)\\
+\sigma_6&\sim Exponential(10)
 \end{align}
@@
 Instead of simply using the average probability as our intercept, which is what a dummy variable would do in a GLM, we are able to allow our model choose an intercept which agrees with the data but can adapt depending on the player.
-Thus, we let $\beta_0$ be normally distributed with mean $\mu_0$ (where $\mu_0$ is normally distributed with mean -3.5 and standard deviation 0.1) and standard deviation $\sigma_0$ (where the parameter $\sigma_0$ itself is exponentially distributed).
+Thus, we let $\alpha$ be normally distributed with mean $\mu_0$ (where $\mu_0$ is normally distributed with mean -3.5 and standard deviation 0.1) and standard deviation $\sigma_0$ (where the parameter $\sigma_0$ itself is exponentially distributed).
+
+$\beta_h$ the effect that a player being age $h$ has on HR probability.
+Similarly $\eta_h$ is the effect that a player's age squared has on the probability of a HR in an AB.
+This squared term captures the non-linearity of HR as player's age.
+
+$\delta_p$ is the effect that a player's home ballpark $p$ has on his underlying probability of hitting a home run.
+
+$\xi_i$ represents the change in the underlying probability of player hitting a HR in the year $i$
 
 EXPLORE THE DATA HERE AND WHY A MULTILEVEL MODEL WORKS BEST
+
+# Results
+
+```{r}
+#| eval: false
+
+#Since we aren't fitting anything when we render, we don't need this chunk to take up any extra time calculating usless vectors
+data.poly.fit <- list(
+  N = nrow(Batting),
+  n_player = n_distinct(Batting$player_index),
+  n_park = n_distinct(Batting$park_index),
+  n_year = n_distinct(Batting$yearID),
+  AB = Batting %>% pull(AB),
+  HR = Batting %>% pull(HR),
+  Player = Batting %>% pull(player_index),
+  cAge = Batting %>% pull(cAge),
+  cAgeSq = (Batting %>% pull(cAge))**2,
+  Park = Batting %>% pull(park_index),
+  Year = Batting %>% pull(yearID) -1972
+)
+```
+
+```{r}
+#| eval: false
+
+#We don't want the program, which takes my desktop 30-45 minutes to render, to run every time I need to render the docuement
+fit.poly <- stan(
+  file = "HRPolyFit.stan",
+  data = data.poly.fit,
+  iter = 1e4
+)
+```
+
+```{r}
+#| eval: false
+
+#This saves the fit as and RDS so that I can reference it later when rendering
+saveRDS(fit.poly, "PolyFit.rds")
+```
+
+```{r}
+plot(fit.poly, pars="xi")
+```
 

### 2023-11-30 Lahman Analysis/Formal Analysis/Draft 1.qmd
Did some more updates
Continued writing, added some graphs, and a correlation plot

@@
 library(bayesplot)
+library(corrplot)
 ```
@@
 ```{r}
+#Make the data needed for the correlation plot
+Batting_cor <- Batting %>%
+  mutate(PA = AB + BB + IBB + HBP + SH + SF) %>% 
+  select(AB, HR, PA, p, park_index, yearID, cAge) %>%
+  filter(!is.na(p))
+```
+
+```{r}
 #| eval: false
@@
 Without a doubt there is a strong correlation between the two because one is a subset of the other.
-In fact, we can see that the correlation between the two is quite strong in (REFERENCE TO CORRELATION BETWEEN AB'S AND PA'S).
+In fact, we can see that the correlation between the two is quite strong in @fig-BattingCor.
 When multicollinearity is introduced into a model, it can be difficult for the Markov Chain Monte Carlo (MCMC) algorithm to parse out which of the two variables is responsible for the changes in the dependent variable.
@@
 This becomes important because most HR's are achieved by hitting the baseball over the outfield fence.
-Thus, fences closer to home plate lead to more HR's from all hitters, ceteris paribus.
+Thus, fences closer to home plate generally lead to more HR's for all hitters, ceteris paribus.
 The non-standardized nature of stadiums' fences gives them a variety of unique shapes and ranging outfield sizes.
@@
 
-```{r}
-#Code to see if HRPolyFit still runs
-stan(
-  file = "HRPolyFit.stan",
-  data = data.poly.fit,
-  iter = 100
-)
-```
-
 Exploring concretely how our model works will make it clearer than discussing it in the abstract. The current model that we have is the following:
@@
 \begin{align}
-HR_{ni} &\sim Binomial(AB_{ni},p_{ni})\\
-logit(\frac{p_{ni}}{1-p_{ni}}) &= \alpha_{n}+\beta_h+\eta_h+\delta_p+\xi_i
+HR_{ni} &\sim Binomial(AB_{ni},\pi_{ni})\\
+logit(\frac{\pi_{ni}}{1-\pi_{ni}}) &= \alpha_{n}+\beta_h+\eta_h+\delta_p+\xi_i
 \end{align}
@@
 In this model, we say that the $HR$'s of player $n$ are binomialy distributed with number of trials for player $n$ are $AB$.
-The probability of success (hitting a HR) denoted as $p$.
-Furthermore, we measure $p$ on the logistic scale.
-We note here that $p$ changes due to a variety of factors.
-First, we noticed in the data that most hitters hit a HR in about 3 out of every 100 AB's (REFERENCE TO FIGURE WHICH SHOWS AVERAGE HR PER AB IN MLB).
+The probability of success (hitting a HR) denoted as $\pi$.
+Furthermore, we measure $\pi$ on the logistic scale.
+It is obvious from the model that $\pi$ changes due to a variety of factors.
+
+First we must noticed in the data that most hitters hit a HR in about 3 out of every 100 AB's, which can be seen in both @fig-HR.By.Age.20to40, @fig-pDist, and @fig-pDistSmall.
 Thus, we assume that the underlying probability of any player in the MLB for hitting a HR is about -3.5 on the logit scale, or 0.03 on the standard probability scale.
@@
 EXPLORE THE DATA HERE AND WHY A MULTILEVEL MODEL WORKS BEST
+
+CHOOSE A FEW PLAYERS TO SHOW HOW THE MODEL FIT CHANGES FOR EACH PLAYER
 
@@
 ```{r}
+#Lets me load in the model that I made. Will only work on my desktop.
+fit.poly <- readRDS("C:/Users/Sam/Desktop/RDS/PolyFit.rds", )
+```
+
+```{r}
 plot(fit.poly, pars="xi")
@@
 # Figures
+
+```{r}
+#| label: fig-BattingCor
+corrplot(cor(Batting_cor), type = "upper", order = "hclust", tl.col = "black", tl.srt = 45)
+```
 
@@
 ```{r}
+#| label: fig-pDist
 ggplot(data = Batting)+
@@
   geom_vline(xintercept=median(Batting$p, na.rm = TRUE) , color="red")+
+  scale_x_continuous(n.breaks = 10)+
   labs(title = "Distribution of Players' HR Proportions",
@@
 ```{r}
+#| label: fig-pDistSmall
 #I want this to have a little more breaks in the x-axis to highlight where the median is
@@
   geom_vline(xintercept=median(Batting$p, na.rm = TRUE) , color="red")+
-  xlim(x=c(0,0.15))+
+  scale_x_continuous(n.breaks = 16, limits = c(0,0.15))+
   labs(title = "Constricted View of Distribution of Players' HR Proportions",

### 2023-11-30 Lahman Analysis/Formal Analysis/Draft 1.qmd
Literally just added one sentance
Added a player trajectory

@@
 library(bayesplot)
+library(tidybayes)
 library(corrplot)
@@
 
-#Create a centered age variable by subtracting 30 from everyone's age
+#Create a centered age variable by subtracting 30 from everyone's age and square it for non-liner effect
 
 Batting <- Batting %>% 
-  mutate(cAge = Age -30, .after = Age)
+  mutate(cAge = Age -30, .after = Age) %>% 
+  mutate(cAgeSq = cAge**2, .after = cAge)
+
+#Make a year index because it's easier than having to create one everytime
+
+Batting <- Batting %>% 
+  mutate(year_index = yearID -1972, .after = yearID)
 ```
@@
 It is obvious from the model that $\pi$ changes due to a variety of factors.
+The first parameter which affects $\pi$ is $\alpha_n$.
 
@@
 ```
+
+```{r}
+#The selected player (id=651) is Robin Yount
+
+set.seed(1)
+
+select_player <- Batting %>% 
+  distinct(playerID) %>% 
+  pull(playerID) %>% 
+  sample(size=1, replace=FALSE)
+
+select_player
+```
+
+```{r}
+select_player_Batting <- Batting %>% 
+  filter(playerID %in% select_player) %>% 
+  select(playerID, player_index, yearID, year_index, Age, cAge, cAgeSq, AB, HR, p, park_index)
+
+select_player_Batting
+```
+
+```{r}
+#These first three effects (alpha, beta, and eta) are all corresponding to the 651st player (Robin Yount)
+alpha_i651 <- fit.poly %>% 
+  spread_draws(alpha[group]) %>% 
+  summarise_draws() %>% 
+  filter(group==651)
+
+beta_i651 <- fit.poly %>% 
+  spread_draws(beta[group]) %>% 
+  summarise_draws() %>% 
+  filter(group==651)
+
+eta_i651 <- fit.poly %>% 
+  spread_draws(eta[group]) %>% 
+  summarise_draws() %>% 
+  filter(group==651)
+
+#Meanwhile, this value corresponds to the park
+delta_i651 <- fit.poly %>% 
+  spread_draws(delta[group]) %>% 
+  summarise_draws() %>% 
+  filter(group==24)
+
+#And this value corresponds to the years that he played
+xi_i651 <- fit.poly %>% 
+  spread_draws(xi[group]) %>% 
+  summarise_draws() %>% 
+  filter(group >= 2 & group <= 21)
+```
+
+```{r}
+#This outputs the associated means for each index for player 651
+alpha_i651 %>% pull(mean)
+beta_i651 %>% pull(mean)
+eta_i651 %>% pull(mean)
+delta_i651 %>% pull(mean)
+xi_i651 %>% pull(mean)
+```
+
+```{r}
+select_player_Batting <- select_player_Batting %>% 
+  mutate(alpha = alpha_i651 %>% pull(mean), .after = player_index) %>%
+  mutate(beta = beta_i651 %>% pull(mean), .after = cAge) %>% 
+  mutate(eta = eta_i651 %>% pull(mean), .after = cAgeSq) %>% 
+  mutate(delta = delta_i651 %>% pull(mean), .after = park_index) %>% 
+  mutate(xi = xi_i651 %>% pull(mean), .after = year_index) %>% 
+  mutate(pred = alpha + (beta*cAge) + (eta*cAgeSq) + (delta) + (xi) , .after = p) %>% 
+  mutate(inv_logit_pred = inv.logit(pred), .after=pred)
+```
+
+```{r}
+ggplot(data = select_player_Batting)+
+  aes(x = cAge, y = p)+
+  geom_point()+
+  geom_smooth()+
+  labs(title = "Actual Trajectory for player #651")
+```
+
+```{r}
+ggplot(data = select_player_Batting)+
+  aes(x = cAge, y=inv_logit_pred)+
+  geom_point()+
+  geom_smooth()+
+  labs(title = "Predicted Career Trajectory for #651")
+```

### 2023-12-01 Lahman Analysis/Formal Analysis/Draft1.qmd
Rough Draft (mostly) done
Did as much as could with the rough draft before 12/1 meeting

--- 2023-11-30 Lahman Analysis/Formal Analysis/Draft 1.qmd
@@
 ---
-title: "Bayesian Data Analysis - Rough Draft"
+title: "Bayesian Data Analysis, Rough Draft"
 author: "Sam Turner"
@@
 library(corrplot)
+library(boot)
 ```
@@
 
-# Introduction and Motivations
+# Introduction and Motivations {#sec-intro}
 
@@
 
-## Baseball Background
+## Baseball Background {#sec-BaseBack}
 
@@
 
-## Literature Review
+## Literature Review {#sec-LitReview}
 
@@
 
-# Model
+# Model {#sec-Model}
 
@@
 
-In this model, we say that the $HR$'s of player $n$ are binomialy distributed with number of trials for player $n$ are $AB$.
-The probability of success (hitting a HR) denoted as $\pi$.
+In this model, we say that the $HR$'s of player $n$ follow a binomial distribution with number of trials for player $n$ in season $i$ are $AB_{ni}$.
+The probability of success (hitting a HR) denoted as $\pi_{ni}$ for player $n$ in season $i$.
 Furthermore, we measure $\pi$ on the logistic scale.
-It is obvious from the model that $\pi$ changes due to a variety of factors.
+
+As discussed in @sec-BaseBack, there are many factors that influence player $n$'s underlying probability of hitting a $HR$ in a given $AB$ during season $i$.
 The first parameter which affects $\pi$ is $\alpha_n$.
-
-First we must noticed in the data that most hitters hit a HR in about 3 out of every 100 AB's, which can be seen in both @fig-HR.By.Age.20to40, @fig-pDist, and @fig-pDistSmall.
-Thus, we assume that the underlying probability of any player in the MLB for hitting a HR is about -3.5 on the logit scale, or 0.03 on the standard probability scale.
-
-We recognize that this depends on the hitter which is where the use of a multilevel model becomes useful.
+In our model $\alpha_n$ can be generally thought of as the term which is used to account for most player's innate ability to hit HR's.
+This comes from our exploratory observations of the data.
+In @fig-HR.By.Age.20to40, @fig-pDist, and @fit-pDistSmall, we noticed that the average (median or mean) proportion of AB's which resulted in HR's to be about 0.03.
+This means that about 3 in every 100 AB's result in HR's.
+On the logit scale, which is what our model relies on, this is about -3.5.
+However, we recognize this depends on the hitter which is where multilevel modeling becomes useful.
 Instead of simply using the average probability as our intercept, which is what a dummy variable would do in a GLM, we are able to allow our model choose an intercept which agrees with the data but can adapt depending on the player.
 Thus, we let $\alpha$ be normally distributed with mean $\mu_0$ (where $\mu_0$ is normally distributed with mean -3.5 and standard deviation 0.1) and standard deviation $\sigma_0$ (where the parameter $\sigma_0$ itself is exponentially distributed).
+By allowing $\mu_0$ and $\sigma_0$ to have their own distributions, we are able to have the distribution of $\alpha_n$ adapt to the player's specific underlying probability of hitting HR's.
+That is, we are able to consider what can be most closely considered their innate ability because this estimate for $\alpha_n$ does not account for other considered factors, such as age, park factors, or year played.
 
-$\beta_h$ the effect that a player being age $h$ has on HR probability.
+The next parameter we considered which affects $\pi_{ni}$ was $\beta_h$ which represents the effect age $h$, on the centered scale, has on a player's ability to hit HR's.
+This topic was briefly touched upon in @sec-BaseBack, where we discussed how player age can impact their their strategy and physical prowess.
+These ideas are well demonstrated in @fig-HR.By.Age.20to40.
+In @fig-HR.By.Age.20to40, one can easily see that, in general, players ability to hit HR's improves over the course of their career.
+As a note, we centered the the age of the all players at 30 because not only was that roughly the mean and the median, but also because it makes interpreting the $\beta_h$ easier.
+The value of $\beta_h$ represents how changes away from the centered age affect player $n$'s probability of hitting a HR, $\pi_{ni}$.
+In normal GLM's, $\beta_h$ would be general to age, meaning that there would be one intercept for each age that fit all players in the data the best.
+However, as with $\alpha_n$, we have employed a multilevel model to allow for each $\beta_h$ to be specific to each player.
+It becomes more difficult to interpret the values of $\beta_h$ because they correspond to the multiplicative effect of the centered age index.
+That is, in calculating the value of $\beta_h$, our model considers the priors we have set with $\mu_1$ and $\sigma_1$ to create a distribution for $\beta_h$ and then multiplies this $\beta$ with our centered age value $h$.
+In essence, this means that our $\beta_h$ describes the multiplicative effect of being a certain distance from the centered age, and how that effects player's $\pi$.
+Another important note, is that although this effect of $\beta$ is multiplicative with respect to the player's index, $h$, it is additive with respect to $\alpha$ and therefore $\pi$.
+This means that the whatever the model predicts for $\beta_h$ and its weight, it must be added to $\alpha_n$ to generate the correct $\pi$.
+
 Similarly $\eta_h$ is the effect that a player's age squared has on the probability of a HR in an AB.
-This squared term captures the non-linearity of HR as player's age.
+The purpose of this squared term is to capture the non-linearity of HR as player's age.
+It is apparent in both @fig-HR.By.Age.20to40 and @fig-HR.By.Age.All that there is a non-linear effect of age on a player's ability to hit HR's.
+This ability appears to peak about when the player is 30-years-old.
+We include $\eta_h$ in the model to capture this downward trend that we see after a player reaches 30-years-old.
+It is nearly impossible to interpret what $\eta_h$ is doing because it is the multiplicative effect of the player's centered age squared.
+This begs the question, what even is a squared age?
+How does one interpret it?
+There is not a clear answer on this question, so it is best understood that this term simply accounts for the non-linearity we see in the data and not become too concerned with what the actual estimates are.
+However, it is easy to justify our choice for the priors on $\eta_h$, $\mu_2$ and $\sigma_2$.
+We chose $\mu_2\sim Normal(0,0.01)$ because we assumed that there would be a negative effect so small that it would be very close to zero.
+Further, any distinction to the slightly negative of zero would be hard to justify and, in essence, arbitrary.
+Therefore, we went with the mean of 0 and the standard deviation of 0.01 to account for any players who have a more or less extreme non-linearity to their trajectories.
+That is, higher or lower peaks and troughs throughout their career for their probability $\pi$.
+We chose $\sigma_2\sim Exponential(100)$ because standard deviations cannot be negative, which the exponential distribution lends itself nicely to.
+Furthermore, the exponential distribution with a parameter of 100 is very concentrated.
+Since we know that the coefficient of $\eta_h$ is likely to be very small since it is being multiplied by the age squared (which can be up over 100).
+Thus, we can be confident that our standard deviation will end up being small for our estimate.
 
-$\delta_p$ is the effect that a player's home ballpark $p$ has on his underlying probability of hitting a home run.
+Another parameter to consider the the effect that playing in certain ballparks has on HR probability $\pi$.
+This was discussed in @sec-BaseBack, where we highlighted that players in certain ballparks are more likely to hit HR's than in others because of the different dimensions of the outfield fences.
+Furthermore, we know that these parks will effect players on different teams differently because of how schedules are created.
+Players will play half of their games at their home-stadium and the other half at various other ballparks known as "on the road." Thus, the home stadium that player plays at is an important factor for their HR probability $\pi$.
+In our model, we have estimated the effect that each ballpark has on player's playing in that ballpark.
+Stadiums like Great American Ballpark (index=33) tend to have a better HR effect than stadiums such as Cleveland Stadium (index=19).
+We account for this difference in our model with the parameter $\delta_p$ where $\delta$ is the effect that park $p$ has on a player's HR probability $\pi$.
+Interpreting $\delta_p$ is much easier than the previous two parameters.
+It simply states, on the logit scale, how the park effects the underlying probability of hitting a HR, $\pi$.
+Similar to $\alpha_n$ this effect is additive and can act almost like an intercept in the model.
+It also takes advantage of multilevel modeling since we have set priors which can adapt to the data on it, $\mu_5$ and $\sigma_5$.
+We have purposely set the parameter for $\mu_5\sim Normal(0,0.01)$ with mean 0 because we assume that parks in the aggregate will not have an effect on a player's ability to hit HR's.
+That is, for every Cleveland Stadium there exists a Great American Ballpark to counteract this.
+There is no formal justification for this, it is just an assumption that we make based on our preliminary exploration and our understanding that ownership (who controls the dimensions of the stadium) are not particularly biased to any ballpark size which means that we should see good variation between parks.
+Further, we assume that a fair standard deviation for $\mu_5$ is 0.01 because on the logistic scale, this means that we could see about a two percent change in the probabilities between any given stadiums.
+At most we would see a 4% difference.
+We chose a standard deviation for $\delta_p$ of $\sigma_5\sim Exponential(10)$ for similar reasons to all of the other $\sigma$'s.
+The exponential distribution cannot be negative, which we need for standard deviations.
+Further, we know that the effect of each park will probably not be huge because, although there is deviation, there is not extreme deviation between parks that makes some "unfair." Thus, with our parameter of 10 for the exponential distribution we are being fairly confident with our model about our estimates of the priors.
 
-$\xi_i$ represents the change in the underlying probability of player hitting a HR in the year $i$
+Our final parameter, $\xi_i$ represents the change in the underlying probability of player hitting a HR in the year $i$.
+This stems mostly from the observation in @fig-HRByYear which shows a steep increase in the amount of HR's hit per year across any subset of years considered.
+This means that there is an underlying change in the $\pi$ depending on the year considered.
+The reason for this change is irrelevant.
+It could be because batters are becoming better or pitchers are getting worse, or some combination of the two.
+Regardless, we can see that there is an increase and we need to account for this in our estimates for $\pi$ of player $n$.
+Thus, we include $\xi_i$ to represent the effect that playing in a certain year will have on the player's ability to hit a HR.
+We set the mean of $\xi_i$ as $\mu_6\sim Normal(0, 0.25)$ because we can see that the HR proportion change year over year in @fig-pByYear by only fractions of a percent.
+Thus, when we set our prior with mean 0 and standard deviation of 0.25, we are saying that we expect most year of year variation to be from \[-4, -3\] on the logit scale.
+On the normal this is \[0.018, 0.047\] which means that we would expect at most a change of 3% year over year.
+This is much larger than we would actually anticipate in a change but it allows the model to learn from the data to learn from the data more since we are unsure of exactly what the change beyond knowing that there is a change year over year.
 
@@
 
-# Results
+# Results {#sec-Results}
 
@@
 ```{r}
-plot(fit.poly, pars="xi")
+#| eval: false
+plot(fit.poly, pars="delta")
 ```
 
-# Figures
+```{r}
+#| eval: false
+print(fit.poly, pars="delta")
+```
+
+# Figures {#sec-Figures}
 
@@
        y = "Runs Scored")
+```
+
+```{r}
+HomeRuns.By.Year <- Lahman::Teams %>% 
+  group_by(yearID) %>% 
+  summarize(HR = sum(HR),
+            AB = sum(AB)) %>% 
+  mutate(p = HR / AB)
+```
+
+```{r}
+#| label: fig-HRByYear
+#| fig-cap: "Cumulative Home Runs in MLB by Year"
+ggplot(data = HomeRuns.By.Year)+
+  aes(x=yearID, y=HR)+
+  geom_line()+
+  scale_x_continuous(n.breaks = 20)+
+  labs(title = "Cumulative Home Runs in MLB by Year",
+       x = "Year",
+       y = "Home Runs - League Cumulative")
+```
+
+```{r}
+#| label: fig-pByYear
+#| fig-cap: "Cumulative Home Run Probability in MLB by Year"
+ggplot(data = HomeRuns.By.Year)+
+  aes(x=yearID, y=p)+
+  geom_line()+
+  scale_x_continuous(n.breaks = 20)+
+  labs(title = "Cumulative Home Run Probability in MLB by Year",
+       x = "Year",
+       y = "HR Proportion (HR/AB) - League Cumulative")
 ```

### 2023-12-01 Lahman Analysis/Formal Analysis/Draft1.qmd
Standardization of FF notation and mine

@@
 \begin{align}
-h_{ij}|\pi_{ij}&\sim Binomial(ab_{ij},\pi_{ij})\\
+HR_{nip}|\pi_{nip}&\sim Binomial(AB_{nip},\pi_{nip})\\
 \\
-logit(\frac{\pi_{ij}}{1-\pi_{ij}})&=x_{ij}\beta_{i}+\theta_i+\delta_j
+logit(\frac{\pi_{nip}}{1-\pi_{nip}})&=\beta_{n1}\cdot Age_{ni}+\beta_{n2}\cdot(Age_{ni})^2+\beta_{n3}\cdot(Age_{ni})^3+\beta_{n4}\cdot(Age_{ni})^4+\alpha_n+\delta_i
 +\xi_p
@@
 \begin{align}
-\theta_i&\sim Normal(\mu_d,\sigma^2_{d_i}), d\in \{1...14\}, i\in\{1,...,3735\}\\
+\alpha_n&\sim Normal(\mu_{d_n},\sigma^2_{d_n}), d\in \{1...14\}, n\in\{1,...,3735\}\\
 \\
-\mu_d&\sim Normal(-2.5,var=1), d_i\in\{1,...,14\}\\
+\mu_d&\sim Normal(-2.5,1), d\in\{1,...,14\}\\
 \\
-\sigma^2_{d_i}&\sim Gamma(3,3), d_i\in\{1,...14\} \\
+\sigma^2_{d}&\sim Gamma(3,3), d\in\{1,...14\} \\
 \\
-\delta_j&\sim Normal(-2.5, var=4),j\in \{1,...,138\}\\
+\delta_i&\sim Normal(-2.5, 2),i\in \{1,...,138\}\\
 \\
-\xi_p&\sim Normal(-2.5, var=4), p\in \{1,...,222\}\\
+\xi_p&\sim Normal(-2.5, 2), p\in \{1,...,222\}\\
 \end{align}
@@
 \begin{align}
-HR_{ni} &\sim Binomial(AB_{ni},\pi_{ni})\\
-logit(\frac{\pi_{ni}}{1-\pi_{ni}}) &= \alpha_{n}+\beta_h+\eta_h+\delta_p+\xi_i
+HR_{nip} &\sim Binomial(AB_{nip},\pi_{nip})\\
+\log(\pi_{nip}) &= \alpha_{n}+\beta_n\cdot (Age_{ni}-30)+\eta_n\cdot (Age_{ni}-30)^2+\delta_p+\xi_i
 \end{align}

### 2023-12-01 Lahman Analysis/Formal Analysis/Draft1.qmd
Added to Fellingham and Fisher's explanation of model

@@
 Fellingham and Fisher (2017) estimate a player's ability to hit home runs from his age in a season, decade born, season played, and home ball park.
-This leaves Fellingham and Fisher with the following model:
+This leaves Fellingham and Fisher with the following model, which we have adapted to make model parameters the same across both their model and my model (in @sec-Model):
 
@@
 \\
-logit(\frac{\pi_{nip}}{1-\pi_{nip}})&=\beta_{n1}\cdot Age_{ni}+\beta_{n2}\cdot(Age_{ni})^2+\beta_{n3}\cdot(Age_{ni})^3+\beta_{n4}\cdot(Age_{ni})^4+\alpha_n+\delta_i
+\log(\frac{\pi_{nip}}{1-\pi_{nip}})&=[\beta_{n1}\cdot Age_{ni}]+[\beta_{n2}\cdot(Age_{ni})^2]+[\beta_{n3}\cdot(Age_{ni})^3]+[\beta_{n4}\cdot(Age_{ni})^4]+\alpha_n+\delta_i
 +\xi_p
@@
 
-Where $h_{ij}$ is the HR by a player $i$ in year $j$, $\pi_{ij}$ is the probability of a HR for player $i$ in year $j$, $ab$ is the number of AB's for player $i$ in year $j$, $x_{ij}$ and $\beta_i$ are vectors of the associated weights of the orthogonal quartic polynomial for a player's age on HR probabilities, $\theta_i$ is the effect of decade of birth, $\delta_j$ is the effect of season of play, and $\xi_p$ is the effect of the home ballpark.
+Where $HR_{nip}$ is the HR's hit by a player $n$ in year $i$ played at stadium $p$.
+$AB_{nip}$ is the AB's for player $n$ in year $i$ at stadium $p$.
+$\pi_{nip}$ is the probability of a HR in an AB for player $n$ in year $i$ at stadium $p$.
 
-What makes this model so strange are the priors that Fellingham and Fisher (2017) chose.
+The parameter's $\beta_{n,1-4}$ describe the weights of the orthogonal polynomial for, up to the quartic, ages of players.
+We theorize, because Fellingham and Fisher (2017) did not explicitly articulate why, that they chose a quartic model because it has a closer fit to the data displayed in @fig-HR.By.Age.All and @fig-HR.By.Age.20to40.
+The properties of a squared term is that it captures the rise and fall of player's HR probability.
+However, the inclusion of the cubic and quartic term allow for for modeling two humps, or peaks, in a data set.
+This appears in @fig-HR.By.Age.20to40 where players see a peak at 27 and then their probability decreases a little, which is then followed by another peak at 31, and afterwards continues in a downward trend.
+This will be discussed in more depth in @sec-Model, but our model which aims to improve Fellingham and Fisher's (2017) model does not account for the cubic or quartic effects over fear of over-fitting the age data.
+
+The parameter $\alpha_n$ (which is $\theta_i$ in the actual paper) are the individual intercepts for each player.
+This parameter is the only one that Fellingham and Fisher (2017) use which takes advantage of multilevel modeling.
+In this modeling, Fellingham and Fisher (2017) set the mean of $\alpha_n$ to be $\mu_d$ where $d$ is the decade the player was born.
+That means that Fellingham and Fisher (2017) set the parameter which measures the player's "innate" HR probability based on the HR probability of other players, who were born in the same decade.
+This allows Fellingham and Fisher (2017) to fit make confident predictions about players who may have only played for 6 seasons because they know that their model is trained on more than 6 data points.
+This parameter, $\alpha_n$ along with $\beta_{n,1-4}$ represent the player specific terms of the model.
+That is, the parameters of the model which depend solely on the player being considered.
+
+In contrast, the parameter $\delta_i$ represents the effect that season $i$ has on player $n$'s ability to hit a HR at stadium $p$.
+This $\delta$ is calculated for each season considered where $i\in\{1,…,138\}$ for the 138 seasons considered.
+This $\delta_i$ is calculated independent of each player so it represents the different changes that we see in HR ability over all years considered.
+An example of why this effect is important refer to @fig-HRByYear.
+
+The parameter $\xi_p$ is the effect that playing in park $p$ has on the HR probability $\pi_{nip}$ in year $i$.
+Fellingham and Fisher (2017) decided that they want this estimate to independent of the hitter.
+Hence, they use no pooling or multilevel effects in the model.
+They estimate the effect that each park has on a player's ability to hit a HR and then add the associated effect to the players who play at that ballpark.
+
+What makes Fellingham and Fisher's model so strange are the priors that Fellingham and Fisher (2017) chose.
 Disregarding that there is no justification for these priors, Fellingham and Fisher (2017) set the following priors for these parameters:
@@
 On the normal probability scale, this corresponds to probability fluctuations of about (0.0015, 0.81).
-Surely there must be better priors than this which Fellingham and Fisher (2017) could have used.
+Furthermore, if you only consider the means that Fellingham and Fisher have used here, then you would be left with a player who, on the logit scale, hits HR's at a probability of -7.5.
+On the normal probability scale this becomes about 0.00055, which is 0.055% chance any AB ends in a HR.
+Confusingly, this is not even close to the 2.8% that Fellingham and Fisher (2017) even noted in their paper.
+Surely there must be better priors than this which Fellingham and Fisher (2017) could have used since it does not seem that even they believe these priors that they have set.
 Coincidentally, this is probably why in using MCMC to estimate their parameters they needed to use 200,000 iterations.
@@
 HR_{nip} &\sim Binomial(AB_{nip},\pi_{nip})\\
-\log(\pi_{nip}) &= \alpha_{n}+\beta_n\cdot (Age_{ni}-30)+\eta_n\cdot (Age_{ni}-30)^2+\delta_p+\xi_i
+\log(\frac{\pi_{nip}}{1-\pi_{nip}}) &= \alpha_{n}+\beta_n\cdot (Age_{ni}-30)+\eta_n\cdot (Age_{ni}-30)^2+\delta_p+\xi_i
 \end{align}

### 2023-12-01 Lahman Analysis/Formal Analysis/Draft1.qmd
Added notes for future self

@@
 
+In explore results, discuss both the trajectories of about 4-5 players and show how they change (probably want to sample about 10 just to be safe).
+But then, just for fun, look at which player has the highest $\alpha$ level since that would be my models prediction of best HR hitter independent of everything else.
+It will probably be Barry Bonds.
+
 ```{r}

### 2023-12-01 Lahman Analysis/Formal Analysis/Draft1.qmd
Created Trajectory for second player

@@
 This appears in @fig-HR.By.Age.20to40 where players see a peak at 27 and then their probability decreases a little, which is then followed by another peak at 31, and afterwards continues in a downward trend.
-This will be discussed in more depth in @sec-Model, but our model which aims to improve Fellingham and Fisher's (2017) model does not account for the cubic or quartic effects over fear of over-fitting the age data.
+This will be discussed in more depth in @sec-Model but our model, which aims to improve Fellingham and Fisher's (2017) model, does not account for the cubic or quartic effects over fear of over-fitting the age data.
 
@@
 
+```{r}
+#Using the weights of the model, lets predict how a couple of players would do and compare it to how they actually did
+
+#We want to set the seed of this chunk so that we're talking about the same players everytime we run the chunk
+set.seed(1)
+
+#This gives us a list of 10 players to consider so if there are a couple that look similar we can choose to not include them in the analysis
+select_players <- Batting %>% 
+  distinct(playerID) %>% 
+  pull(playerID) %>% 
+  sample(size=10, replace=FALSE)
+
+select_players
+```
+
+```{r}
+select_players_Batting <- Batting %>% 
+  filter(playerID %in% select_players) %>% 
+  select(playerID, player_index, yearID, year_index, Age, cAge, cAgeSq, AB, HR, p, park_index)
+
+#select_players_Batting
+```
+
+## Robin Yount
+
+```{r}
+set.seed(1)
+
+select_player <- Batting %>% 
+  distinct(playerID) %>% 
+  pull(playerID) %>% 
+  sample(size=1, replace=FALSE)
+
+select_player_Batting <- Batting %>% 
+  filter(playerID %in% select_player) %>% 
+  select(playerID, player_index, yearID, year_index, Age, cAge, cAgeSq, AB, HR, p, park_index)
+```
+
+```{r}
+# Robin Yount (651) Estimates cell
+
+alpha_i651 <- fit.poly %>% 
+  spread_draws(alpha[group]) %>% 
+  summarise_draws() %>% 
+  filter(group==651)
+
+beta_i651 <- fit.poly %>% 
+  spread_draws(beta[group]) %>% 
+  summarise_draws() %>% 
+  filter(group==651)
+
+eta_i651 <- fit.poly %>% 
+  spread_draws(eta[group]) %>% 
+  summarise_draws() %>% 
+  filter(group==651)
+
+#Meanwhile, this value corresponds to the park
+delta_i651 <- fit.poly %>% 
+  spread_draws(delta[group]) %>% 
+  summarise_draws() %>% 
+  filter(group==24)
+
+#And this value corresponds to the years that he played
+xi_i651 <- fit.poly %>% 
+  spread_draws(xi[group]) %>% 
+  summarise_draws() %>% 
+  filter(group >= 2 & group <= 21)
+```
+
+```{r}
+select_player_Batting <- select_player_Batting %>% 
+  mutate(alpha = alpha_i651 %>% pull(mean), .after = player_index) %>%
+  mutate(beta = beta_i651 %>% pull(mean), .after = cAge) %>% 
+  mutate(eta = eta_i651 %>% pull(mean), .after = cAgeSq) %>% 
+  mutate(delta = delta_i651 %>% pull(mean), .after = park_index) %>% 
+  mutate(xi = xi_i651 %>% pull(mean), .after = year_index) %>% 
+  mutate(pred = alpha + (beta*cAge) + (eta*cAgeSq) + (delta) + (xi) , .after = p) %>% 
+  mutate(inv_logit_pred = inv.logit(pred), .after=pred)
+```
+
+See @fig-p651Act and @fig-p651Pred.
+
+## Pat Tabler
+
+```{r}
+# Pat Tabler (582) Estimates cell
+
+#Find out what parks player 582 played in over his career
+i582_park <- Batting %>% 
+  filter(player_index == 582) %>% 
+  distinct(park_index) %>% 
+  pull(park_index)
+
+#Find our what years player 582 played in over his career
+i582_year <- Batting %>% 
+  filter(player_index == 582) %>% 
+  distinct(year_index) %>% 
+  pull(year_index)
+
+#Find the all the associated parameters needed for to predict model of player 582
+#These first three are specfic estimates for 582
+alpha_i582 <- fit.poly %>% 
+  spread_draws(alpha[group]) %>% 
+  summarise_draws() %>% 
+  filter(group==582)
+
+beta_i582 <- fit.poly %>% 
+  spread_draws(beta[group]) %>% 
+  summarise_draws() %>% 
+  filter(group==582)
+
+eta_i582 <- fit.poly %>% 
+  spread_draws(eta[group]) %>% 
+  summarise_draws() %>% 
+  filter(group==582)
+
+#Meanwhile, this value corresponds to the park
+delta_i582 <- fit.poly %>% 
+  spread_draws(delta[group]) %>% 
+  summarise_draws() %>% 
+  filter(group %in% i582_park)
+
+#And this value corresponds to the years that he played
+xi_i582 <- fit.poly %>% 
+  spread_draws(xi[group]) %>% 
+  summarise_draws() %>% 
+  filter(group %in% i582_year)
+```
+
+```{r}
+#Create the trajectory DF for player 582
+i582.Trajectory <- Batting %>% 
+  #Grab the player we're interested in
+  filter(player_index == 582) %>% 
+  #Only keep columns we're interested in
+  select(playerID, player_index, yearID, year_index, Age, cAge, cAgeSq, AB, HR, p, park_index) %>%
+  #Put the alpha, beta, and eta parameters in the DF
+  mutate(alpha = alpha_i582 %>% pull(mean), .after = player_index) %>%
+  mutate(beta = beta_i582 %>% pull(mean), .after = cAge) %>%
+  mutate(eta = eta_i582 %>% pull(mean), .after = cAgeSq) %>%
+  #Set-up to add "delta" to the table because 582 was traded
+  right_join(delta_i582 %>% select(group, mean), by = join_by(park_index==group)) %>% 
+  mutate(delta = mean, .after = park_index) %>% 
+  select(!mean) %>% 
+  #Set-up to add "xi" to the table because 582 was traded
+  right_join(xi_i582 %>% select(group, mean), by = join_by(year_index==group)) %>% 
+  mutate(xi = mean, .after = year_index) %>% 
+  select(!mean) %>% 
+  #Create the predictions for the player
+  mutate(pred = alpha + (beta*cAge) + (eta*cAgeSq) + (delta) + (xi) , .after = p) %>% 
+  mutate(inv_logit_pred = inv.logit(pred), .after=pred)
+```
+
+See @fig-p582Act and @fig-p582Pred.
+
+```{r}
+
+```
+
 # Figures {#sec-Figures}
@@
 ```{r}
-#The selected player (id=651) is Robin Yount
-
-set.seed(1)
-
-select_player <- Batting %>% 
-  distinct(playerID) %>% 
-  pull(playerID) %>% 
-  sample(size=1, replace=FALSE)
-
-select_player
-```
-
-```{r}
-select_player_Batting <- Batting %>% 
-  filter(playerID %in% select_player) %>% 
-  select(playerID, player_index, yearID, year_index, Age, cAge, cAgeSq, AB, HR, p, park_index)
-
-select_player_Batting
-```
-
-```{r}
-#These first three effects (alpha, beta, and eta) are all corresponding to the 651st player (Robin Yount)
-alpha_i651 <- fit.poly %>% 
-  spread_draws(alpha[group]) %>% 
-  summarise_draws() %>% 
-  filter(group==651)
-
-beta_i651 <- fit.poly %>% 
-  spread_draws(beta[group]) %>% 
-  summarise_draws() %>% 
-  filter(group==651)
-
-eta_i651 <- fit.poly %>% 
-  spread_draws(eta[group]) %>% 
-  summarise_draws() %>% 
-  filter(group==651)
-
-#Meanwhile, this value corresponds to the park
-delta_i651 <- fit.poly %>% 
-  spread_draws(delta[group]) %>% 
-  summarise_draws() %>% 
-  filter(group==24)
-
-#And this value corresponds to the years that he played
-xi_i651 <- fit.poly %>% 
-  spread_draws(xi[group]) %>% 
-  summarise_draws() %>% 
-  filter(group >= 2 & group <= 21)
-```
-
-```{r}
-#This outputs the associated means for each index for player 651
-alpha_i651 %>% pull(mean)
-beta_i651 %>% pull(mean)
-eta_i651 %>% pull(mean)
-delta_i651 %>% pull(mean)
-xi_i651 %>% pull(mean)
-```
-
-```{r}
-select_player_Batting <- select_player_Batting %>% 
-  mutate(alpha = alpha_i651 %>% pull(mean), .after = player_index) %>%
-  mutate(beta = beta_i651 %>% pull(mean), .after = cAge) %>% 
-  mutate(eta = eta_i651 %>% pull(mean), .after = cAgeSq) %>% 
-  mutate(delta = delta_i651 %>% pull(mean), .after = park_index) %>% 
-  mutate(xi = xi_i651 %>% pull(mean), .after = year_index) %>% 
-  mutate(pred = alpha + (beta*cAge) + (eta*cAgeSq) + (delta) + (xi) , .after = p) %>% 
-  mutate(inv_logit_pred = inv.logit(pred), .after=pred)
-```
-
-```{r}
+#| label: fig-p651Act
 ggplot(data = select_player_Batting)+
@@
 ```{r}
+#| label: fig-p651Pred
 ggplot(data = select_player_Batting)+
@@
   geom_point()+
-  geom_smooth()+
+  geom_smooth(color="red")+
   labs(title = "Predicted Career Trajectory for #651")
 ```
+
+```{r}
+#| label: fig-p582Act
+ggplot(data = i582.Trajectory)+
+  aes(x = cAge, y = p)+
+  geom_point()+
+  geom_smooth()+
+  labs(title = "Actual Trajectory for player #582")
+```
+
+```{r}
+#| label: fig-p582Pred
+ggplot(data = i582.Trajectory)+
+  aes(x = cAge, y=inv_logit_pred)+
+  geom_point()+
+  geom_smooth(color="red")+
+  labs(title = "Predicted Career Trajectory for #582")
+```

### 2023-12-02 Lahman Analysis/Formal Analysis/Draft1.qmd
Now have 4 (very different) player trajectories

@@
 library(tidyverse)
+library(scales)
 library(rstan)
@@
 
+```{r}
+#As a warning to all who are going to be running this in the future: even calculating the parameters from the model (alpha, beta, eta, and xi) are computationally expensive.
+They take a good chunk of time to calculate which makes rendering the document take a couple minutes.
+```
+
 ## Robin Yount
@@
 
+## Junior Felix
+
 ```{r}
+# Junior Felix (177) Estimates cell
 
+#Find out what parks player 177 played in over his career
+i177_park <- Batting %>% 
+  filter(player_index == 177) %>% 
+  distinct(park_index) %>% 
+  pull(park_index)
+
+#Find our what years player 177 played in over his career
+i177_year <- Batting %>% 
+  filter(player_index == 177) %>% 
+  distinct(year_index) %>% 
+  pull(year_index)
+
+#Find the all the associated parameters needed for to predict model of player 177
+#These first three are specfic estimates for 177
+alpha_i177 <- fit.poly %>% 
+  spread_draws(alpha[group]) %>% 
+  summarise_draws() %>% 
+  filter(group==177)
+
+beta_i177 <- fit.poly %>% 
+  spread_draws(beta[group]) %>% 
+  summarise_draws() %>% 
+  filter(group==177)
+
+eta_i177 <- fit.poly %>% 
+  spread_draws(eta[group]) %>% 
+  summarise_draws() %>% 
+  filter(group==177)
+
+#Meanwhile, this value corresponds to the park
+delta_i177 <- fit.poly %>% 
+  spread_draws(delta[group]) %>% 
+  summarise_draws() %>% 
+  filter(group %in% i177_park)
+
+#And this value corresponds to the years that he played
+xi_i177 <- fit.poly %>% 
+  spread_draws(xi[group]) %>% 
+  summarise_draws() %>% 
+  filter(group %in% i177_year)
 ```
+
+```{r}
+#Create the trajectory DF for player 177
+i177.Trajectory <- Batting %>% 
+  #Grab the player we're interested in
+  filter(player_index == 177) %>% 
+  #Only keep columns we're interested in
+  select(playerID, player_index, yearID, year_index, Age, cAge, cAgeSq, AB, HR, p, park_index) %>%
+  #Put the alpha, beta, and eta parameters in the DF
+  mutate(alpha = alpha_i177 %>% pull(mean), .after = player_index) %>%
+  mutate(beta = beta_i177 %>% pull(mean), .after = cAge) %>%
+  mutate(eta = eta_i177 %>% pull(mean), .after = cAgeSq) %>%
+  #Set-up to add "delta" to the table because 177 may have been traded
+  right_join(delta_i177 %>% select(group, mean), by = join_by(park_index==group)) %>% 
+  mutate(delta = mean, .after = park_index) %>% 
+  select(!mean) %>% 
+  #Set-up to add "xi" to the table because 177 may have been traded
+  right_join(xi_i177 %>% select(group, mean), by = join_by(year_index==group)) %>% 
+  mutate(xi = mean, .after = year_index) %>% 
+  select(!mean) %>% 
+  #Create the predictions for the player
+  mutate(pred = alpha + (beta*cAge) + (eta*cAgeSq) + (delta) + (xi) , .after = p) %>% 
+  mutate(inv_logit_pred = inv.logit(pred), .after=pred)
+```
+
+See @fig-p177Act and @fig-p177Pred.
+
+## Kendrys Morales
+
+```{r}
+# Kendrys Morales (420) Estimates cell
+
+#Find out what parks player 420 played in over his career
+i420_park <- Batting %>% 
+  filter(player_index == 420) %>% 
+  distinct(park_index) %>% 
+  pull(park_index)
+
+#Find our what years player 420 played in over his career
+i420_year <- Batting %>% 
+  filter(player_index == 420) %>% 
+  distinct(year_index) %>% 
+  pull(year_index)
+
+#Find the all the associated parameters needed for to predict model of player 420
+#These first three are specfic estimates for 420
+alpha_i420 <- fit.poly %>% 
+  spread_draws(alpha[group]) %>% 
+  summarise_draws() %>% 
+  filter(group==420)
+
+beta_i420 <- fit.poly %>% 
+  spread_draws(beta[group]) %>% 
+  summarise_draws() %>% 
+  filter(group==420)
+
+eta_i420 <- fit.poly %>% 
+  spread_draws(eta[group]) %>% 
+  summarise_draws() %>% 
+  filter(group==420)
+
+#Meanwhile, this value corresponds to the park
+delta_i420 <- fit.poly %>% 
+  spread_draws(delta[group]) %>% 
+  summarise_draws() %>% 
+  filter(group %in% i420_park)
+
+#And this value corresponds to the years that he played
+xi_i420 <- fit.poly %>% 
+  spread_draws(xi[group]) %>% 
+  summarise_draws() %>% 
+  filter(group %in% i420_year)
+```
+
+```{r}
+#Create the trajectory DF for player 420
+i420.Trajectory <- Batting %>% 
+  #Grab the player we're interested in
+  filter(player_index == 420) %>% 
+  #Only keep columns we're interested in
+  select(playerID, player_index, yearID, year_index, Age, cAge, cAgeSq, AB, HR, p, park_index) %>%
+  #Put the alpha, beta, and eta parameters in the DF
+  mutate(alpha = alpha_i420 %>% pull(mean), .after = player_index) %>%
+  mutate(beta = beta_i420 %>% pull(mean), .after = cAge) %>%
+  mutate(eta = eta_i420 %>% pull(mean), .after = cAgeSq) %>%
+  #Set-up to add "delta" to the table because 420 may have been traded
+  right_join(delta_i420 %>% select(group, mean), by = join_by(park_index==group)) %>% 
+  mutate(delta = mean, .after = park_index) %>% 
+  select(!mean) %>% 
+  #Set-up to add "xi" to the table because 420 may have been traded
+  right_join(xi_i420 %>% select(group, mean), by = join_by(year_index==group)) %>% 
+  mutate(xi = mean, .after = year_index) %>% 
+  select(!mean) %>% 
+  #Create the predictions for the player
+  mutate(pred = alpha + (beta*cAge) + (eta*cAgeSq) + (delta) + (xi) , .after = p) %>% 
+  mutate(inv_logit_pred = inv.logit(pred), .after=pred)
+```
+
+See @fig-p420Act and @fig-p420Pred.
 
@@
   scale_x_continuous(n.breaks = 20)+
+  scale_y_continuous(labels = label_percent())+
   labs(title = "Cumulative Home Run Probability in MLB by Year",
@@
                                 ymax = prop + prop.se))+
+  scale_y_continuous(labels = label_percent())+
   labs(title = "Proportion of Home Runs per At-Bat by Age",
@@
   geom_smooth()+
+  scale_y_continuous(labels = label_percent())+
   labs(title = "Actual Trajectory for player #651")
@@
   geom_smooth(color="red")+
+  scale_y_continuous(labels = label_percent())+
   labs(title = "Predicted Career Trajectory for #651")
@@
   geom_smooth()+
+  scale_y_continuous(labels = label_percent())+
   labs(title = "Actual Trajectory for player #582")
@@
   geom_smooth(color="red")+
+  scale_y_continuous(labels = label_percent())+
   labs(title = "Predicted Career Trajectory for #582")
 ```
+
+```{r}
+#| label: fig-p177Act
+ggplot(data = i177.Trajectory)+
+  aes(x = cAge, y = p)+
+  geom_point()+
+  geom_smooth()+
+  scale_y_continuous(labels = label_percent())+
+  labs(title = "Actual Trajectory for player #177")
+```
+
+```{r}
+#| label: fig-p177Pred
+ggplot(data = i177.Trajectory)+
+  aes(x = cAge, y=inv_logit_pred)+
+  geom_point()+
+  geom_smooth(color="red")+
+  scale_y_continuous(labels = label_percent())+
+  labs(title = "Predicted Career Trajectory for #177")
+```
+
+```{r}
+#| label: fig-p420Act
+ggplot(data = i420.Trajectory)+
+  aes(x = cAge, y = p)+
+  geom_point()+
+  geom_smooth()+
+  scale_y_continuous(labels = label_percent())+
+  labs(title = "Actual Trajectory for player #420")
+```
+
+```{r}
+#| label: fig-p420Pred
+ggplot(data = i420.Trajectory)+
+  aes(x = cAge, y=inv_logit_pred)+
+  geom_point()+
+  geom_smooth(color="red")+
+  scale_y_continuous(labels = label_percent())+
+  labs(title = "Predicted Career Trajectory for #420")
+```

### 2023-12-08 Lahman Analysis/Formal Analysis/Draft1.qmd
Added ideas to results section
Added code to allow for document render on my mac

@@
 #Lets me load in the model that I made. Will only work on my desktop.
-fit.poly <- readRDS("C:/Users/Sam/Desktop/RDS/PolyFit.rds", )
+#fit.poly <- readRDS("C:/Users/Sam/Desktop/RDS/PolyFit.rds", )
+
+#Lets me load in the model for my Mac
+fit.poly <- readRDS("~/Desktop/RDS/PolyFit.rds")
 ```
@@
 ```
+
+## General
+
+This section should be dedicated to talking about general fit of the model.
+Explore using the normals plots that we've been using with the confidence intervals to see if the parameters are near their priors.
+And, also explore using the trace-plots.
+The linegraphs that go all over the place.
 

### 2023-12-09 Lahman Analysis/Formal Analysis/Draft1.qmd
Rendered Figures and continued working on paper

@@
 
-As mentioned, Fellingham and Fisher (2017) have both considered and published their thoughts on this research question.
-This research question concerns predicting player home run performance using Bayesian multilevel modeling.
+As mentioned, Fellingham and Fisher (2017) have both considered and published their thoughts on this research question, which concerns predicting player home run performance using Bayesian multilevel modeling.
 To do this, Fellingham and Fisher (2017) used data from the Lahman data set spanning 1871 to 2008 to estimate the parametric weights for their model.
 Over this time frame they only considered players who had played for at least 6 seasons and who, in each of those seasons, had at least 50 AB's.
-They only considered seasons where the players were 18-years-old to 45-years-old.
+They further restricted their pool of players to only considered seasons where the players were 18-years-old to 45-years-old.
 
@@
 \\
-\log(\frac{\pi_{nip}}{1-\pi_{nip}})&=[\beta_{n1}\cdot Age_{ni}]+[\beta_{n2}\cdot(Age_{ni})^2]+[\beta_{n3}\cdot(Age_{ni})^3]+[\beta_{n4}\cdot(Age_{ni})^4]+\alpha_n+\delta_i
-+\xi_p
+\log(\frac{\pi_{nip}}{1-\pi_{nip}})&=[\beta_{n1}\cdot Age_{ni}]+[\beta_{n2}\cdot(Age_{ni})^2]+[\beta_{n3}\cdot(Age_{ni})^3]+[\beta_{n4}\cdot(Age_{ni})^4]+\alpha_n+\xi_i
++\delta_p
 \end{align}
@@
 $AB_{nip}$ is the AB's for player $n$ in year $i$ at stadium $p$.
-$\pi_{nip}$ is the probability of a HR in an AB for player $n$ in year $i$ at stadium $p$.
-
-The parameter's $\beta_{n,1-4}$ describe the weights of the orthogonal polynomial for, up to the quartic, ages of players.
-We theorize, because Fellingham and Fisher (2017) did not explicitly articulate why, that they chose a quartic model because it has a closer fit to the data displayed in @fig-HR.By.Age.All and @fig-HR.By.Age.20to40.
-The properties of a squared term is that it captures the rise and fall of player's HR probability.
-However, the inclusion of the cubic and quartic term allow for for modeling two humps, or peaks, in a data set.
-This appears in @fig-HR.By.Age.20to40 where players see a peak at 27 and then their probability decreases a little, which is then followed by another peak at 31, and afterwards continues in a downward trend.
-This will be discussed in more depth in @sec-Model but our model, which aims to improve Fellingham and Fisher's (2017) model, does not account for the cubic or quartic effects over fear of over-fitting the age data.
+$\pi_{nip}$ is the probability of a HR in an AB for player $n$ in year $i$ at stadium $p$ measured on the logistic scale.
+
+The parameter's $\beta_{n,1-4}$ describe the weights of the orthogonal polynomial, up to the quartic, for the ages of the players.
+We theorize, because Fellingham and Fisher (2017) did not explicitly articulate why, they chose a quartic model because it has a closer fit to the data displayed in @fig-HR.By.Age.All and @fig-HR.By.Age.20to40.
+The properties of a squared term capture the rise and fall of player's HR probability over their career.
+The inclusion of the cubic term allows for the modeling of multiple peaks, while the quartic term allows it to maintain that overall trend up and then down seen in the data.
+This appears in @fig-HR.By.Age.20to40 where players see a steady increase in their HR probability during their career with a peak when they are 27.
+Then, their probability decreases a little, followed quickly by another increase, which then leads to another peak when the player is 31.
+Afterwards, the probability continues in a downward trend.
+This will be discussed in more depth in @sec-Model but our model, which aims to improve Fellingham and Fisher's (2017) model, does not account for the cubic or quartic effects over fear of over-fitting the age data, since the distinction between the humps is rather small.
 
@@
 In this modeling, Fellingham and Fisher (2017) set the mean of $\alpha_n$ to be $\mu_d$ where $d$ is the decade the player was born.
-That means that Fellingham and Fisher (2017) set the parameter which measures the player's "innate" HR probability based on the HR probability of other players, who were born in the same decade.
-This allows Fellingham and Fisher (2017) to fit make confident predictions about players who may have only played for 6 seasons because they know that their model is trained on more than 6 data points.
-This parameter, $\alpha_n$ along with $\beta_{n,1-4}$ represent the player specific terms of the model.
-That is, the parameters of the model which depend solely on the player being considered.
-
-In contrast, the parameter $\delta_i$ represents the effect that season $i$ has on player $n$'s ability to hit a HR at stadium $p$.
-This $\delta$ is calculated for each season considered where $i\in\{1,…,138\}$ for the 138 seasons considered.
-This $\delta_i$ is calculated independent of each player so it represents the different changes that we see in HR ability over all years considered.
-An example of why this effect is important refer to @fig-HRByYear.
-
-The parameter $\xi_p$ is the effect that playing in park $p$ has on the HR probability $\pi_{nip}$ in year $i$.
+That means that Fellingham and Fisher (2017) set the parameter which measures the player's "innate" HR probability based on the HR probability of other players who were born in the same decade.
+This allows Fellingham and Fisher (2017) to make confident predictions about players who may only have 6 seasons of data because this model borrows data from all players who were born at around a similar time.
+Implicitly, this means that Fellingham and Fisher (2017) assume that when a player was born is a powerful predictor of their HR probability.
+
+In addition, $\alpha_n$ is not the only player specific parameter of this model.
+The parameters $\beta_{n,1-4}$ attempt to measure how a player's age might affect their ability to hit HR's.
+In contrast to $\alpha_n$, this parameter does not use multilevel modeling techniques.
+Instead, it relies on a more traditional statistical approach which considers how age affects players aggregately.
+That is, it relies on one fit for all the age data and then applying that uniformly to all players considered who are that age.
+
+In contrast, the parameter $\xi_i$ (which is $\delta_i$ in the actual paper) represents the effect that season $i$ has on player $n$'s ability to hit a HR at stadium $p$.
+This $\xi$ is calculated for each season considered where $i\in\{1,…,138\}$ for the 138 seasons considered.
+This $\xi_i$ is calculated independent of each player so it represents the different changes that we see in HR ability over all years considered.
+An example of why this effect is important refer to @fig-HRByYear, which clearly shows how HR rates have differed drastically over the years considered.
+
+The parameter $\delta_p$ (which is $\xi_p$ in the actual paper) is the effect that playing in park $p$ has on the HR probability $\pi_{nip}$ in year $i$.
 Fellingham and Fisher (2017) decided that they want this estimate to independent of the hitter.
-Hence, they use no pooling or multilevel effects in the model.
-They estimate the effect that each park has on a player's ability to hit a HR and then add the associated effect to the players who play at that ballpark.
-
-What makes Fellingham and Fisher's model so strange are the priors that Fellingham and Fisher (2017) chose.
-Disregarding that there is no justification for these priors, Fellingham and Fisher (2017) set the following priors for these parameters:
+Hence, they use no pooling or multilevel effects in this part of the model.
+They estimate the effect that each park has on a player's ability to hit a HR and then add the associated effect to the players who play at that ballpark in a given year.
+
+One of the main reasons that Fellingham and Fisher's (2017) model is so strange are the priors that they chose.
+Disregarding that there is no justification for these priors in their paper, Fellingham and Fisher (2017) set the following priors for these parameters:
 
@@
 \\
-\delta_i&\sim Normal(-2.5, 2),i\in \{1,...,138\}\\
+\xi_i&\sim Normal(-2.5, 2),i\in \{1,...,138\}\\
 \\
-\xi_p&\sim Normal(-2.5, 2), p\in \{1,...,222\}\\
+\delta_p&\sim Normal(-2.5, 2), p\in \{1,...,222\}\\
 \end{align}
@@
 
-These priors are quite strange because they are so broad.
-For example, consider the priors on either $\delta_j$ or $\xi_p$.
+These priors are quite strange because they are so uninformative.
+For example, consider the priors on either $\xi_i$ or $\delta_p$.
 The priors state that on a logistic scale, it is normal for the decade of birth and season of play parameters to vary along the interval \[-6.5, 1.5\].
@@
 Coincidentally, this is probably why in using MCMC to estimate their parameters they needed to use 200,000 iterations.
-Likely, their model could have estimated their parameters much more easily had they given stronger priors.
+Likely, their model could have estimated their parameters much more easily had they given it stronger priors.
 This is one of the areas this paper seeks to improve Fellingham and Fisher's (2017) methods.
@@
 In turn, this leads to better predictions about players where data is lacking.
-This "learning" is known as partial pooling, where the model identifies that parameters for each player (or year etc.) probably come from the same distribution but may not share the exact same estimates.
+This "learning" is known as partial pooling, where the model identifies that parameters for each player (or year etc.) probably comes from the same distribution but may not share the exact same estimates.
 
@@
 \begin{align}
-\alpha_n&\sim Normal(\mu_0, \sigma_0)\\
-\beta_h&\sim Normal(\mu_1, \sigma_1)\\
-\eta_h&\sim Normal(\mu_2, \sigma_2)\\
-\delta_p&\sim Normal(\mu_5, \sigma_5)\\
-\xi_i&\sim Normal(\mu_6, \sigma_6)\\
+\alpha_n&\sim Normal(\mu_0, \sigma_0),n\in\{1,...,657\}\\
+\beta_n&\sim Normal(\mu_1, \sigma_1),n\in\{1,...,657\}\\
+\eta_n&\sim Normal(\mu_2, \sigma_2),n\in\{1,...,657\}\\
+\delta_p&\sim Normal(\mu_5, \sigma_5),p\in\{1,...,88\}\\
+\xi_i&\sim Normal(\mu_6, \sigma_6),i\in\{1,...,47\}\\
 \end{align}
@@
 
-In this model, we say that the $HR$'s of player $n$ follow a binomial distribution with number of trials for player $n$ in season $i$ are $AB_{ni}$.
-The probability of success (hitting a HR) denoted as $\pi_{ni}$ for player $n$ in season $i$.
+In this model, we say that the $HR$'s of player $n$ follow a binomial distribution with number of trials for player $n$ in season $i$, where player $n$ played in park $p$, are $AB_{nip}$.
+The probability of success (hitting a HR) is denoted as $\pi_{nip}$ for player $n$ in season $i$ playing at park $p$.
 Furthermore, we measure $\pi$ on the logistic scale.
 
-As discussed in @sec-BaseBack, there are many factors that influence player $n$'s underlying probability of hitting a $HR$ in a given $AB$ during season $i$.
-The first parameter which affects $\pi$ is $\alpha_n$.
-In our model $\alpha_n$ can be generally thought of as the term which is used to account for most player's innate ability to hit HR's.
-This comes from our exploratory observations of the data.
-In @fig-HR.By.Age.20to40, @fig-pDist, and @fit-pDistSmall, we noticed that the average (median or mean) proportion of AB's which resulted in HR's to be about 0.03.
+As discussed in @sec-BaseBack, there are many factors that influence player $n$'s underlying probability of hitting a $HR$ in a given $AB$.
+The first parameter which affects this probability, $\pi_{nip}$, is $\alpha_n$.
+In our model $\alpha_n$ can be generally thought of as the term which is used to account for a player's innate ability to hit HR's.
+From our exploratory observations of the data, we noticed that the average (median or mean) proportion of AB's which resulted in HR's to be about 0.03.
 This means that about 3 in every 100 AB's result in HR's.
-On the logit scale, which is what our model relies on, this is about -3.5.
-However, we recognize this depends on the hitter which is where multilevel modeling becomes useful.
-Instead of simply using the average probability as our intercept, which is what a dummy variable would do in a GLM, we are able to allow our model choose an intercept which agrees with the data but can adapt depending on the player.
-Thus, we let $\alpha$ be normally distributed with mean $\mu_0$ (where $\mu_0$ is normally distributed with mean -3.5 and standard deviation 0.1) and standard deviation $\sigma_0$ (where the parameter $\sigma_0$ itself is exponentially distributed).
-By allowing $\mu_0$ and $\sigma_0$ to have their own distributions, we are able to have the distribution of $\alpha_n$ adapt to the player's specific underlying probability of hitting HR's.
-That is, we are able to consider what can be most closely considered their innate ability because this estimate for $\alpha_n$ does not account for other considered factors, such as age, park factors, or year played.
-
-The next parameter we considered which affects $\pi_{ni}$ was $\beta_h$ which represents the effect age $h$, on the centered scale, has on a player's ability to hit HR's.
-This topic was briefly touched upon in @sec-BaseBack, where we discussed how player age can impact their their strategy and physical prowess.
+This can be seen in @fig-HR.By.Age.20to40, @fig-pDist, and @fig-pDistSmall.
+On the logit scale, which is what our model relies on, a probability of 0.03 is about -3.5.
+We recognize that some players, for whatever reason, have better "innate" abilities to hit HR's than others.
+Therefore, we used a multilevel modeling technique for $\alpha_n$ to capture that player-to-player difference.
+Instead of simply using the average probability as our intercept, which is what a dummy variable would do in a GLM, we allow our model choose an intercept which adapts depending on the player and the data.
+That is, we give our model priors that we would expect to see in the data and then, as the model interprets additional observations, it updates these priors using Bayes theorem.
+For these priors, we let $\alpha$ be normally distributed with mean $\mu_0$ (where $\mu_0$ is normally distributed with mean -3.5 and standard deviation 0.1) and standard deviation $\sigma_0$ (where the parameter $\sigma_0$ itself is exponentially distributed).
+This allows us to have separate intercepts for each player $n$, but these intercepts come from the same hyper-priors $\mu_0$ and $\sigma_0$ that the model considers for all other players.
+Thus, the model constructs its posterior predictions based on both the priors and the data.
+In effect, this makes the model appear as though it is "learning" from the data because it has the same general sense of how a player should perform, but then adapts that prediction based on the data it takes into account to form its posterior.
+Specifically, for $\alpha_n$ we say that the means of all players roughly follow the distribution of $\mu_0$, but we also tell the model that it should look at how player $n$ performed to gauge its posterior prediction.
+A similar process is followed for $\alpha_n$'s standard deviation, $\sigma_0$.
+
+The next parameter we considered which affects $\pi_{nip}$ was $\beta_n$ which represents the effect centered age (30 minus the player's age) had on player $n$'s ability to hit HR's.
+This topic was briefly touched upon in @sec-BaseBack, where we discussed how a player's age can impact their their strategy and physical prowess.
 These ideas are well demonstrated in @fig-HR.By.Age.20to40.
-In @fig-HR.By.Age.20to40, one can easily see that, in general, players ability to hit HR's improves over the course of their career.
-As a note, we centered the the age of the all players at 30 because not only was that roughly the mean and the median, but also because it makes interpreting the $\beta_h$ easier.
-The value of $\beta_h$ represents how changes away from the centered age affect player $n$'s probability of hitting a HR, $\pi_{ni}$.
-In normal GLM's, $\beta_h$ would be general to age, meaning that there would be one intercept for each age that fit all players in the data the best.
-However, as with $\alpha_n$, we have employed a multilevel model to allow for each $\beta_h$ to be specific to each player.
-It becomes more difficult to interpret the values of $\beta_h$ because they correspond to the multiplicative effect of the centered age index.
-That is, in calculating the value of $\beta_h$, our model considers the priors we have set with $\mu_1$ and $\sigma_1$ to create a distribution for $\beta_h$ and then multiplies this $\beta$ with our centered age value $h$.
-In essence, this means that our $\beta_h$ describes the multiplicative effect of being a certain distance from the centered age, and how that effects player's $\pi$.
-Another important note, is that although this effect of $\beta$ is multiplicative with respect to the player's index, $h$, it is additive with respect to $\alpha$ and therefore $\pi$.
-This means that the whatever the model predicts for $\beta_h$ and its weight, it must be added to $\alpha_n$ to generate the correct $\pi$.
+In @fig-HR.By.Age.20to40, one can easily see that, in general, players' ability to hit HR's improves over the course of their career and peaks when players reach about 30 years old.
+After this peak, a player's ability starts to decline for the rest of their career.
+We chose to center the the age of the all players at 30 for several reasons.
+The first was that roughly the mean and the median ages of the data was at 30.
+And second, because having a centered age makes interpreting $\beta_n$ easier.
+The value of $\beta_n$ represents how changes away from the centered age affects player $n$'s probability of hitting a HR, $\pi_{nip}$.
+In a normal GLM's, $\beta_n$ would not be able to adapt to each player, $n$, as it does in our model.
+It would only be able to consider how aging affects the average player's HR probability and then apply that effect to all players.
+This implicitly assumes that each player is affected identically by changes in age.
+In contrast, we employ multilevel modeling techniques to our model which assumes that most players experience the same general effects of aging, but allows for subtle changes depending on the player.
+This is the job of our hyper-priors, $\mu_1$ and $\sigma_1$.
+The parameter $\mu_1$ acts as the general mean which we expect most players to follow.
+However, since $\mu_1$ is a distribution, it allows for our model to update its posterior with observed data so that it is not heavily influenced by either a few data points or our strong priors.
+Our standard deviation parameter, $\sigma_1$ acts similarly.
+Thus, we have estimates $\beta_n$ for player $n$ which describes how aging uniquely affects their probability of hitting a HR in a given AB, $\pi_{nip}$, but these unique estimates generally follow the same trend.
 
@@
 #Lets me load in the model that I made. Will only work on my desktop.
-#fit.poly <- readRDS("C:/Users/Sam/Desktop/RDS/PolyFit.rds", )
+fit.poly <- readRDS("C:/Users/Sam/Desktop/RDS/PolyFit.rds", )
 
 #Lets me load in the model for my Mac
-fit.poly <- readRDS("~/Desktop/RDS/PolyFit.rds")
+#fit.poly <- readRDS("~/Desktop/RDS/PolyFit.rds")
 ```
@@
        x = "Proportions (p)",
-       y = "Observations")
+       y = "Observations",
+       caption = "Median market in red.")
 ```
@@
        x = "Proportions (p)",
-       y = "Observations")
+       y = "Observations",
+       caption = "Median marked in red.")
 ```
@@
   geom_point()+
-  geom_smooth()+
+  geom_smooth(se=FALSE)+
   scale_y_continuous(labels = label_percent())+
-  labs(title = "Actual Trajectory for player #651")
+  labs(title = "Actual Trajectory for player #651 - Robin Yount",
+       x = "Centered Age (Player Age - 30)",
+       y = "HR's per AB",
+       caption = "General trend of data in blue.")
 ```
@@
   geom_point()+
-  geom_smooth(color="red")+
+  geom_smooth(color="red", se = FALSE)+
   scale_y_continuous(labels = label_percent())+
-  labs(title = "Predicted Career Trajectory for #651")
+  labs(title = "Predicted Career Trajectory for #651 - Robin Yount",
+       x = "Centered Age (Player Age - 30)",
+       y = "HR's per AB",
+       caption = "General trend of data in red.")
 ```
@@
   geom_point()+
-  geom_smooth()+
+  geom_smooth(se=FALSE)+
   scale_y_continuous(labels = label_percent())+
-  labs(title = "Actual Trajectory for player #582")
+  labs(title = "Actual Trajectory for player #582 - Pat Tabler",
+       x = "Centered Age (Player Age - 30)",
+       y = "HR's per AB",
+       caption = "General trend of data in blue.")
 ```
@@
   geom_point()+
-  geom_smooth(color="red")+
+  geom_smooth(color="red", se=FALSE)+
   scale_y_continuous(labels = label_percent())+
-  labs(title = "Predicted Career Trajectory for #582")
+  labs(title = "Predicted Career Trajectory for #582 - Pat Tabler",
+       x = "Centered Age (Player Age - 30)",
+       y = "HR's per AB",
+       caption = "General trend of data in red.")
 ```
@@
   geom_point()+
-  geom_smooth()+
+  geom_smooth(se=FALSE)+
   scale_y_continuous(labels = label_percent())+
-  labs(title = "Actual Trajectory for player #177")
+  labs(title = "Actual Trajectory for player #177 - Junior Felix",
+       x = "Centered Age (Player Age - 30)",
+       y = "HR's per AB",
+       caption = "General trend of data in blue.")
 ```
@@
   geom_point()+
-  geom_smooth(color="red")+
+  geom_smooth(color="red", se=FALSE)+
   scale_y_continuous(labels = label_percent())+
-  labs(title = "Predicted Career Trajectory for #177")
+  labs(title = "Predicted Career Trajectory for #177 - Junior Felix",
+       x = "Centered Age (Player Age - 30)",
+       y = "HR's per AB",
+       caption = "General trend of data in red.")
 ```
@@
   geom_point()+
-  geom_smooth()+
+  geom_smooth(se = FALSE)+
   scale_y_continuous(labels = label_percent())+
-  labs(title = "Actual Trajectory for player #420")
+  labs(title = "Actual Trajectory for player #420 - Kendrys Morales",
+       x = "Centered Age (Player Age - 30)",
+       y = "HR's per AB",
+       caption = "General trend of data in blue.")
 ```
@@
   geom_point()+
-  geom_smooth(color="red")+
+  geom_smooth(color="red", se=FALSE)+
   scale_y_continuous(labels = label_percent())+
-  labs(title = "Predicted Career Trajectory for #420")
-```
+  labs(title = "Predicted Career Trajectory for #420 - Kendrys Morales",
+       x = "Centered Age (Player Age - 30)",
+       y = "HR's per AB",
+       caption = "General trend of data in red.")
+```
+
+```{r}
+#| label: fig-alphaTrace
+#651, 582, 177, 420
+traceplot(fit.poly, pars = c("alpha[651]", "alpha[582]", "alpha[177]", "alpha[420]"))
+```

### 2023-12-11 Lahman Analysis/Formal Analysis/Draft1.qmd
Finished first draft of pres

@@
 \\
-\mu_1&\sim Normal(0,0.01)\\
+\mu_1&\sim Normal(0,0.1)\\
 \sigma_1&\sim Exponential(10)\\
@@
 #Lets me load in the model that I made. Will only work on my desktop.
-fit.poly <- readRDS("C:/Users/Sam/Desktop/RDS/PolyFit.rds", )
+#fit.poly <- readRDS("C:/Users/Sam/Desktop/RDS/PolyFit.rds", )
 
 #Lets me load in the model for my Mac
-#fit.poly <- readRDS("~/Desktop/RDS/PolyFit.rds")
+fit.poly <- readRDS("~/Desktop/RDS/PolyFit.rds")
 ```

### 2023-12-11 Lahman Analysis/Formal Analysis/Pres.qmd
Finished first draft of pres

---
title: "Bayesian Data Analysis"
subtitle: "MATH 470 Summary"
author: "Sam Turner"
format: 
  revealjs:
    theme: white
    embed-resources: true
editor: visual
---

## Fellingham and Fisher (2017)

::: {layout-nrow="2"}
![](Figures/FF%20Front.png)

Pros: Bayesian model

-   Bayesian model
:::

## Fellingham and Fisher (2017) Cont.

Cons:

-   Sparingly uses multilevel modeling

    -   Only use multilevel modeling on their orthogonal quartic polynomial

-   Priors are uninformative and don't fit data

    -   Expect average player's HR probability to be 0.0015

    -   Priors effectively suggest that players could have a HR probability between 0 and 1

-   Strange choices of parameters

    -   Age, decade of birth, season of play, home ballpark

## Model

$$
\begin{align}
HR_{nip} &\sim Binomial(AB_{nip},\pi_{nip})\\
\log(\frac{\pi_{nip}}{1-\pi_{nip}}) &= \alpha_{n}+\beta_n\cdot (Age_{ni}-30)+\eta_n\cdot (Age_{ni}-30)^2\\
&+\delta_p+\xi_i
\end{align}
$$

The amount of HR's hit by player $n$ in year $i$ played at park $p$ is binomially distributed, according to player $n$'s AB's and HR probability $\pi$ for year $i$ at park $p$.

HR probability $\pi$ is measured on the logistic scale.

## "Innate" HR Hitting Ability - $\alpha_n$

$$
\begin{align}
\alpha_n&\sim Normal(\mu_0, \sigma_0), n\in\{1,...,657\}\\
\\
\mu_0&\sim Normal(-3.5,0.1)\\
\sigma_0&\sim Exponential(1)
\end{align}
$$

-   Intercept term which represents "innate" hitting ability of players skilled enough to play in MLB

-   -3.5 on the logit scale is ≈0.029 or 2.9%

-   -3.5±0.1 = \[-3.6, -3.4\] = \[0.0266, 0.0323\] = \[2.7%, 3.2%\]

-   -3.5±2 = \[-5.5, -1.5\] = \[0.0041, 0.182\] = \[0.4%, 18.2%\]

## HR Hitting by Age

![](Figures/HRByAgeAll.png){fig-align="center"}

## HR Distribution

![](Figures/pDist.png){fig-align="center"}

## HR Distribution Cont.

![](Figures/pDistSmall.png){fig-align="center"}

## Centered Age Effect - $\beta_n$

$$
\begin{align}
\beta_n &\sim Normal(\mu_1,\sigma_1), n\in\{1,...,657\}\\
\\
\mu_1 &\sim Normal(0,0.1)\\
\sigma_1 &\sim Exponential(10)
\end{align}
$$

-   Multiplicative effect representing how deviation from centered age (Age - 30) affects HR hitting ability

-   Age plays a factor in hitting HR's, but it is likely not very large so we have the priors set near 0 to reflect this

## Centered Age Effect Squared - $\eta_n$

$$
\begin{align}
\eta_n &\sim Normal(\mu_2, \sigma_2),n\in\{1,...,657\}\\
\\
\mu_2 &\sim Normal(0, 0.01)\\
\sigma_2 &\sim Exponential(100)\\
\end{align}
$$

-   Multiplicative effect representing how deviation from centered age squared \[(Age - 30)²\] affects HR hitting ability

-   Used to capture the non-linearity of the data without risk of over-fitting

## HR Hitting by Age

![](Figures/HRByAge20to40.png){fig-align="center"}

## Park Effect - $\delta_p$

$$
\begin{align}
\delta_p &\sim Normal(\mu_5,\sigma_5),p\in\{1,...,88\}\\
\\
\mu_5 &\sim Normal(0,0.01)\\
\sigma_5 &\sim Exponential(10)
\end{align}
$$

-   Intercept term which captures the effect playing in different parks has on HR probabilities

-   Parks differ by both dimensions and altitude which affects HR rates

## Park Overlay

![](Figures/AllPark.png){fig-align="center"}

## Park Overlay Cont.

![](Figures/LongShortPark.png){fig-align="center"}

## Year Effect - $\xi_i$

$$
\begin{align}
\xi_i &\sim Normal(\mu_6,\sigma_6),i\in\{1,...,47\}\\
\\
\mu_6 &\sim Normal(0,0.25)\\
\sigma_6 &\sim Exponential(10)
\end{align}
$$

-   Intercept term which captures the effect playing in different years has on HR probability

-   Changes can occur because of rules, ownership goals, player goals, etc.

-   This term captures those changes without asking why there are changes

## HR's by Year in MLB

![](Figures/HRByYear.png){fig-align="center"}

## HR Proportion by Year in MLB

![](Figures/pByYear.png){fig-align="center"}

## Model (Again)

$$
\begin{align}
HR_{nip} &\sim Binomial(AB_{nip},\pi_{nip})\\
\log(\frac{\pi_{nip}}{1-\pi_{nip}}) &= \alpha_{n}+\beta_n\cdot (Age_{ni}-30)+\eta_n\cdot (Age_{ni}-30)^2\\
&+\delta_p+\xi_i
\end{align}
$$

-   There are $1.17\times10^{12}$ parameters for this model

-   The model predicts that the average player's $\pi$ is about 0.03 (on the normal scale) which is what we observe in the data

-   Uses Bayesian techniques to update estimates based on what the data says - allows for inference

## Robin Yount - Actual $\pi$

![](Figures/Actual651.png){fig-align="center"}

## Robin Yount - Predicted $\pi$

![](Figures/Pred651.png){fig-align="center"}

## Pat Tabler - Actual $\pi$

![](Figures/Actual582.png){fig-align="center"}

## Pat Tabler - Predicted $\pi$

![](Figures/Pred582.png){fig-align="center"}

## Kendrys Morales - Actual $\pi$

![](Figures/Actual420.png){fig-align="center"}

## Kendrys Morales - Predicted $\pi$

![](Figures/Pred420.png){fig-align="center"}

## Goodness of Fit - Trace-plots

![](Figures/alphaTrace.png){fig-align="center"}

## Conclusions

-   Will our model make the Hood math department excellent gamblers?

    -   No. But it did a fine job at adapting predictions of players and being "right" on average

-   Areas of future research

    -   Player archetypes and physical characteristics

    -   Considering more for a longer time interval (like Fellingham and Fisher (2017))

    -   Better data (advanced metrics or play-by-plays)

## Conclusions Cont.

-   We believe we did better than Fellingham and Fisher (2017), but without access to their model and computational resources it is difficult to determine

### 2023-12-12 Lahman Analysis/Formal Analysis/Draft1.qmd
Finished Presentation

@@
 
-Similarly $\eta_h$ is the effect that a player's age squared has on the probability of a HR in an AB.
-The purpose of this squared term is to capture the non-linearity of HR as player's age.
-It is apparent in both @fig-HR.By.Age.20to40 and @fig-HR.By.Age.All that there is a non-linear effect of age on a player's ability to hit HR's.
-This ability appears to peak about when the player is 30-years-old.
-We include $\eta_h$ in the model to capture this downward trend that we see after a player reaches 30-years-old.
-It is nearly impossible to interpret what $\eta_h$ is doing because it is the multiplicative effect of the player's centered age squared.
-This begs the question, what even is a squared age?
-How does one interpret it?
-There is not a clear answer on this question, so it is best understood that this term simply accounts for the non-linearity we see in the data and not become too concerned with what the actual estimates are.
-However, it is easy to justify our choice for the priors on $\eta_h$, $\mu_2$ and $\sigma_2$.
+Similarly $\eta_n$ is the effect that player $n$'s squared age has on his probability of hitting a HR in an AB.
+The purpose of this squared term is to capture the non-linear relationship between hitting a HR and player $n$'s age.
+It is apparent in both @fig-HR.By.Age.20to40 and @fig-HR.By.Age.All that there is a parabolic relationship between age and players' ability to hit HR's.
+This ability appears to peak when the player is roughly 30-years-old.
+We include $\eta_n$ in the model to capture this peak when a player reaches 30-years-old.
+It is nearly impossible to interpret what $\eta_n$ is doing because it describes the effect of centered age squared on HR probability $\pi$, and centered age squared on its own is already almost a meaningless concept.
+Therefore, it is best understood that this term simply accounts for the non-linearity we see in the data and not become too concerned with understanding its estimates.
+However, it is easy to justify our choice for the priors on $\eta_n$, $\mu_2$ and $\sigma_2$.
 We chose $\mu_2\sim Normal(0,0.01)$ because we assumed that there would be a negative effect so small that it would be very close to zero.
@@
 Furthermore, the exponential distribution with a parameter of 100 is very concentrated.
-Since we know that the coefficient of $\eta_h$ is likely to be very small since it is being multiplied by the age squared (which can be up over 100).
+Since we know that the coefficient of $\eta_n$ is likely to be very small since it is being multiplied by the age squared (which can be up over 100).
 Thus, we can be confident that our standard deviation will end up being small for our estimate.

### 2023-12-12 Lahman Analysis/Formal Analysis/Pres.qmd
Finished Presentation

@@
 ---
-title: "Bayesian Data Analysis"
-subtitle: "MATH 470 Summary"
+title: "Statistical Grand Slams"
+subtitle: "A Bayesian Approach to Modeling Home Run Production in Major League Baseball"
 author: "Sam Turner"
@@
 ---
+
+## Introduction
+
+-   In this research project, Dr. Parson and I sought to predict the home run (HR) production of Major League Baseball hitters.
+
+-   If you work for a MLB team, predicting HR's is important because it is the pinnacle outcome of an at-bat (AB)
+
+-   And even if you don't work for a MLB team, you could make a lot of money in Vegas if you can accurately predict HR's!
+
+## Introduction Cont.
+
+-   A problem with predicting player HR's is that there is often limited data available
+
+    -   For example, players who have played for more than 6 seasons are considered very "veteran" but only have 6 player-season data points
+
+-   Generalized linear models (GLM's) from classical statistics have a hard fitting accurate models given limited data points - such as the 6 data point scenario mentioned above
+
+-   This is where a Bayesian model shines - Bayesian models can "learn" from the data improving the effective observations it has for fitting
 
@@
 
-    -   Expect average player's HR probability to be 0.0015
+    -   Expect average player's HR probability to be 0.00015
 
@@
 HR_{nip} &\sim Binomial(AB_{nip},\pi_{nip})\\
-\log(\frac{\pi_{nip}}{1-\pi_{nip}}) &= \alpha_{n}+\beta_n\cdot (Age_{ni}-30)+\eta_n\cdot (Age_{ni}-30)^2\\
+\frac{\pi_{nip}}{1-\pi_{nip}} &= \alpha_{n}+\beta_n\cdot (Age_{ni}-30)+\eta_n\cdot (Age_{ni}-30)^2\\
 &+\delta_p+\xi_i
@@
 
-HR probability $\pi$ is measured on the logistic scale.
+The amount of AB's in a year is given for our prediction of HR's, but we say that the probability that the AB results in a HR, $\pi$ varies according to a number of factors.
+
+## Model Cont.
+
+We can both simulate and predict the amount of HR's of a player. For example, let's examine a player that has 100 AB's and a probability $\pi$ of 0.03.
+
+That is, $HR \sim Normal(100,0.03)$,
+
+Simulation:
+
+```{r}
+#| echo: true
+set.seed(1)
+rbinom(5, 100, 0.03)
+```
+
+Prediction:
+
+$E(HR)=AB\cdot \pi=100\cdot0.03=3$
+
+## Data
+
+Our data for the model comes from the Lahman data set. Lahman covers a variety of information on each player, including but not limited to, batting, fielding, team, and player statistics.
+
+In order to be considered in our analysis a player must have accumulated at least 6 seasons of play, with at least 50 AB's in each, and have played from 1973-2019.
 
 ## "Innate" HR Hitting Ability - $\alpha_n$
+
+$$
+\begin{align}
+\alpha_n&\sim Normal(\mu_0, \sigma_0), n\in\{1,...,657\}\\
+\\
+\mu_0&\sim Normal(-3.5,0.1)\\
+\sigma_0&\sim Exponential(1)
+\end{align}
+$$
+
+-   Intercept term which represents "innate" hitting ability of players skilled enough to play in MLB
+
+-   -3.5 on the logit scale is ≈0.029 or 2.9%
+
+-   -3.5±0.1 = \[-3.6, -3.4\] = \[0.0266, 0.0323\] = \[2.7%, 3.2%\]
+
+-   -3.5±2 = \[-5.5, -1.5\] = \[0.0041, 0.182\] = \[0.4%, 18.2%\]
+
+## Multilevel Trade-Off - No Pooling
+
+No pooling would mean that each player gets their own posterior $\alpha$ distribution. That is,
+
+$$
+\alpha_n\sim Normal(-3.5,1).
+$$
+
+This prior would be multiplied by the data to generate the posterior for each player.
+However, this disadvantages players with less seasons because there is not as much data to go off of.
+And we know that we could look a player, $m$, who performs similar to $n$ WLOG and generate more accurate predictions.
+
+## Multilevel Trade-Off - Total Pooling
+
+In a total pooling scenario $\alpha$ would just be a dummy variable in the model of the model,
+
+$$
+HR=\tau+\Gamma\cdot AB+\alpha_n+\beta_n+\eta_n+\delta_p+\xi_i+\epsilon
+$$
+
+But then we run into the problem of multicolinearity.
+That is, how do we tease out the relationship between $\alpha_n, \delta_p$, and $\xi_i$ since they all are intercept terms?
+This model also has a hard time allowing for players to stray from the mean for any parameter.
+
+In essence, we lose information from both no and total pooling.
+
+## "Innate" HR Hitting Ability - $\alpha_n$ (Again)
 
@@
 
-## Model (Again)
+## Model (Again I)
 
@@
 
--   There are $1.17\times10^{12}$ parameters for this model
+-   There are 2,116 parameters for this model
+
+    -   (3×657)+88+47+10
+
+-   This model is classically non-identifiable because of 3 intercept terms!
+
+## Model (Again II)
 

### 2023-12-13 Lahman Analysis/Formal Analysis/Draft1.qmd
Finished project, working on paper

@@
 ---
-title: "Bayesian Data Analysis, Rough Draft"
+title: "Statistical Grand Slams"
+subtitle: "A Bayesian Approach to Modeling Home Run Production in Major League Baseball"
 author: "Sam Turner"
@@
 \end{align}
-$$
+$$ {#eq-FFmodel}
 
@@
 \end{align}
-$$
+$$ {#eq-FFpriors}
 
@@
 \end{align}
-$$
+$$ {#eq-model}
 
@@
 \end{align}
-$$
+$$ {#eq-priors}
 
@@
 \end{align}
-$$
+$$ {#eq-hyperpriors}
 
@@
 The purpose of this squared term is to capture the non-linear relationship between hitting a HR and player $n$'s age.
-It is apparent in both @fig-HR.By.Age.20to40 and @fig-HR.By.Age.All that there is a parabolic relationship between age and players' ability to hit HR's.
-This ability appears to peak when the player is roughly 30-years-old.
-We include $\eta_n$ in the model to capture this peak when a player reaches 30-years-old.
-It is nearly impossible to interpret what $\eta_n$ is doing because it describes the effect of centered age squared on HR probability $\pi$, and centered age squared on its own is already almost a meaningless concept.
+It is apparent in both @fig-HR.By.Age.20to40 and @fig-HR.By.Age.All that there is a parabolic relationship between age and player $n$'s ability to hit HR's.
+This ability appears to rise steadily for the average player until it reaches its apex when the player is roughly 30-years-old.
+Then, in the seasons where the player is more than 30-years-old we see a decline in HR probabilities.
+The shape we are left with, as can be seen in @fig-HR.By.Age.20to40, is one of an upside-down parabola.
+Thus, by including this squared age effect, we are able to capture that unique relationship.
+However, as one can easily notice in @fig-HR.By.Age.20to40, the overall shape is parabolic but for ages past 37 the data begins to trend upwards.
+Likely, this trend upwards is due to players who have above average skill.
+That is, players who have earned massive contracts because they were so much better than the average player in their prime earned large multi-million dollar contracts and teams do not want to waste the this money so they play these players.
+So, we are then left with this group of outlying players who do not follow the trend of data at any point over their career, but it only became noticeable after the other, more normal, players dissapeared from the considered sample.
+Returning to the parameter $\eta_n$, it is nearly impossible to interpret what it is doing because it describes the effect of centered age squared on HR probability $\pi$, and centered age squared on its own is already almost a meaningless concept.
 Therefore, it is best understood that this term simply accounts for the non-linearity we see in the data and not become too concerned with understanding its estimates.
-However, it is easy to justify our choice for the priors on $\eta_n$, $\mu_2$ and $\sigma_2$.
+However, it is easy to justify our choice for the priors and hypterpriors of $\eta_n$, $\mu_2$, and $\sigma_2$, respectively.
+We chose $\eta_n\sim Normal(\mu_2,\sigma_2)$ for the same reasons discussed for both $\alpha_n$ and $\beta_n$: we wanted a multilevel model which described how individual players were impacted by the parameters.
 We chose $\mu_2\sim Normal(0,0.01)$ because we assumed that there would be a negative effect so small that it would be very close to zero.
@@
 We chose $\sigma_2\sim Exponential(100)$ because standard deviations cannot be negative, which the exponential distribution lends itself nicely to.
-Furthermore, the exponential distribution with a parameter of 100 is very concentrated.
+Furthermore, the exponential distribution with a parameter of 100 is very concentrated near 0.
 Since we know that the coefficient of $\eta_n$ is likely to be very small since it is being multiplied by the age squared (which can be up over 100).
 Thus, we can be confident that our standard deviation will end up being small for our estimate.
-
-Another parameter to consider the the effect that playing in certain ballparks has on HR probability $\pi$.
+This means that the estimates our model generates will be close to 0.
+
+Another parameter of our model is $\delta_p$, which describes the effect playing in ballpark $p$ in year $i$ has player $n$'s HR probability, $\pi_{nip}$.
+There are 88 different ballparks we considered, and we generated an estimate for each of them.
 This was discussed in @sec-BaseBack, where we highlighted that players in certain ballparks are more likely to hit HR's than in others because of the different dimensions of the outfield fences.
-Furthermore, we know that these parks will effect players on different teams differently because of how schedules are created.
-Players will play half of their games at their home-stadium and the other half at various other ballparks known as "on the road." Thus, the home stadium that player plays at is an important factor for their HR probability $\pi$.
-In our model, we have estimated the effect that each ballpark has on player's playing in that ballpark.
+Furthermore, we know that these parks will effect players differently because teams play at different ballparks with varying frequency.
+For example, players will play 81 of their games at their home-stadium and 81 at various other ballparks known as "on the road." Thus, the home-stadium that player $n$ plays at is an important factor for their HR probability, $\pi$, because player $n$ plays many more games at his home-stadium compared to any other ballpark.
+In our model, we have estimated $\delta_p$ as an intercept term which captures the effect that playing at stadium $p$ has on HR probability, $\pi$.
 Stadiums like Great American Ballpark (index=33) tend to have a better HR effect than stadiums such as Cleveland Stadium (index=19).
-We account for this difference in our model with the parameter $\delta_p$ where $\delta$ is the effect that park $p$ has on a player's HR probability $\pi$.
+By better, we mean that it the probability of hitting a HR is higher at Great American Ballpark compared to Cleaveland Stadium because our model interprets hitting a HR as a success.
 Interpreting $\delta_p$ is much easier than the previous two parameters.
 It simply states, on the logit scale, how the park effects the underlying probability of hitting a HR, $\pi$.
-Similar to $\alpha_n$ this effect is additive and can act almost like an intercept in the model.
-It also takes advantage of multilevel modeling since we have set priors which can adapt to the data on it, $\mu_5$ and $\sigma_5$.
-We have purposely set the parameter for $\mu_5\sim Normal(0,0.01)$ with mean 0 because we assume that parks in the aggregate will not have an effect on a player's ability to hit HR's.
-That is, for every Cleveland Stadium there exists a Great American Ballpark to counteract this.
+In addition, $\delta_p$ takes advantage of multilevel modeling since we have hyper-priors, $\mu_5$ and $\sigma_5$.
+We have purposely set the hyper-prior of $\mu_5\sim Normal(0,0.01)$ with mean 0 because we assume that parks in the aggregate will not have an effect on a player's ability to hit HR's.
+That is, on average, for every Cleveland Stadium there exists a Great American Ballpark which acts as a counterweight and makes the aggregate effect 0.
 There is no formal justification for this, it is just an assumption that we make based on our preliminary exploration and our understanding that ownership (who controls the dimensions of the stadium) are not particularly biased to any ballpark size which means that we should see good variation between parks.
@@
 At most we would see a 4% difference.
-We chose a standard deviation for $\delta_p$ of $\sigma_5\sim Exponential(10)$ for similar reasons to all of the other $\sigma$'s.
+We chose a standard deviation for $\delta_p$ to be $\sigma_5\sim Exponential(10)$ for similar reasons to all of the other $\sigma_i$'s.
 The exponential distribution cannot be negative, which we need for standard deviations.
-Further, we know that the effect of each park will probably not be huge because, although there is deviation, there is not extreme deviation between parks that makes some "unfair." Thus, with our parameter of 10 for the exponential distribution we are being fairly confident with our model about our estimates of the priors.
+Further, we know that the effect of each park will probably not be huge because, although there is deviation, there is not extreme deviation between parks that makes some "unfair." Thus, with our parameter of 10 for the exponential distribution we are being fairly confident with our model about our estimates of the priors, in that we do not allow much deviation from our mean of 0 unless the data strongly suggests otherwise.
 
@@
 This stems mostly from the observation in @fig-HRByYear which shows a steep increase in the amount of HR's hit per year across any subset of years considered.
+Further, we can tell from @fig-pByYear that these additional HR's are not just the product of exapnsion teams creating more opportunities for players to hit more HR's.
+@fig-pByYear shows that the proportion of AB's which are HR's have risen steadily over the period considered which suggests that players' probabilities have been changing.
 This means that there is an underlying change in the $\pi$ depending on the year considered.
 The reason for this change is irrelevant.
-It could be because batters are becoming better or pitchers are getting worse, or some combination of the two.
+It could be because batters are becoming better or pitchers are getting worse, or it could be some other unrelated reason.
 Regardless, we can see that there is an increase and we need to account for this in our estimates for $\pi$ of player $n$.
 Thus, we include $\xi_i$ to represent the effect that playing in a certain year will have on the player's ability to hit a HR.
-We set the mean of $\xi_i$ as $\mu_6\sim Normal(0, 0.25)$ because we can see that the HR proportion change year over year in @fig-pByYear by only fractions of a percent.
+We set the hyper-prior of $\xi_i$ as $\mu_6\sim Normal(0, 0.25)$ because we can see that the HR proportion change year over year in @fig-pByYear by only fractions of a percent.
 Thus, when we set our prior with mean 0 and standard deviation of 0.25, we are saying that we expect most year of year variation to be from \[-4, -3\] on the logit scale.
@@
 This is much larger than we would actually anticipate in a change but it allows the model to learn from the data to learn from the data more since we are unsure of exactly what the change beyond knowing that there is a change year over year.
-
-EXPLORE THE DATA HERE AND WHY A MULTILEVEL MODEL WORKS BEST
-
-CHOOSE A FEW PLAYERS TO SHOW HOW THE MODEL FIT CHANGES FOR EACH PLAYER
+Furthermore, we set the hyper-prior $\sigma_6\sim Exponential(10)$ for the same reasons mentioned during the other $\sigma_i$'s.
+The exponential distriubtion ensures that we never have a negative estimate for our standard deviation of $\xi_i$ and having a parameter value of 10 for $\sigma_6$ allows the data to be fairly concentrated near 0 which is what we would expect for a year over year effect.
+
+Compared to most GLM's, this model [@eq-model] we have created is massive.
+It has 2,116 parameters.
+It has 657 different $\alpha, \beta,$ and $\eta$ estimated for each player.
+It has 88 $\delta$'s for each park and
 
@@
 
+# Conclusions and Directions for Future Research
+
 # Figures {#sec-Figures}

### 2023-12-13 Lahman Analysis/Formal Analysis/Pres.qmd
Finished project, working on paper

@@
 
-That is, $HR \sim Normal(100,0.03)$,
+That is, $HR \sim Binomial(100,0.03)$,
 
@@
 
-No pooling would mean that each player gets their own posterior $\alpha$ distribution. That is,
+No pooling would mean that each player, $n$, gets their own $\alpha$ estimate. That is,
 
 $$
-\alpha_n\sim Normal(-3.5,1).
+\frac{\pi_n}{1-\pi_n}= \alpha_{PLAYER[n]}+...
 $$
 
-This prior would be multiplied by the data to generate the posterior for each player.
-However, this disadvantages players with less seasons because there is not as much data to go off of.
-And we know that we could look a player, $m$, who performs similar to $n$ WLOG and generate more accurate predictions.
+This intercept would just match the data we had on each player.
+It probably over-fits from relying on player data too much, which is sparse.
+You can think of this model suffers from amnesia because it assumes there is nothing in common moving from player to player.
 
@@
 
-In a total pooling scenario $\alpha$ would just be a dummy variable in the model of the model,
+In a total pooling scenario players share the same intercept $\alpha$, which is just the overall mean HR probability.
+There are lots of data, so we can be confidnet of the value that $\alpha$ would be, however it wouldn't be effective to apply on a case-per-case basis.
+This is because few players are likely to $\alpha$'s which exactly match the mean of the data.
 
 $$
-HR=\tau+\Gamma\cdot AB+\alpha_n+\beta_n+\eta_n+\delta_p+\xi_i+\epsilon
+\frac{\pi_n}{1-\pi_n}=\alpha+...=\overline{HR}+...
 $$
 
-But then we run into the problem of multicolinearity.
-That is, how do we tease out the relationship between $\alpha_n, \delta_p$, and $\xi_i$ since they all are intercept terms?
-This model also has a hard time allowing for players to stray from the mean for any parameter.
+You can think of this model suffering from over-sharing since it cannot distinguish the $\alpha$ from player to player.
 
@@
 
-## "Innate" HR Hitting Ability - $\alpha_n$ (Again)
+## "Innate" HR Hitting Ability - $\alpha_n$ (Again) {.scrollable}
 

### 2023-12-13 Lahman Analysis/Formal Analysis/Draft1.qmd
Finished Robin Yount

@@
     embed-resources: true
+    number-sections: true
 
@@
 \\
-\log(\frac{\pi_{nip}}{1-\pi_{nip}})&=[\beta_{n1}\cdot Age_{ni}]+[\beta_{n2}\cdot(Age_{ni})^2]+[\beta_{n3}\cdot(Age_{ni})^3]+[\beta_{n4}\cdot(Age_{ni})^4]+\alpha_n+\xi_i
+\log(\frac{\pi_{nip}}{1-\pi_{nip}})&=[\beta_{n1}\cdot Age_{ni}]+[\beta_{n2}\cdot(Age_{ni})^2]+[\beta_{n3}\cdot(Age_{ni})^3]+[\beta_{n4}\cdot(Age_{ni})^4]\\
+&+\alpha_n+\xi_i
 +\delta_p
@@
 Furthermore, we set the hyper-prior $\sigma_6\sim Exponential(10)$ for the same reasons mentioned during the other $\sigma_i$'s.
-The exponential distriubtion ensures that we never have a negative estimate for our standard deviation of $\xi_i$ and having a parameter value of 10 for $\sigma_6$ allows the data to be fairly concentrated near 0 which is what we would expect for a year over year effect.
-
-Compared to most GLM's, this model [@eq-model] we have created is massive.
+The exponential distribution ensures that we never have a negative estimate for our standard deviation of $\xi_i$ and having a parameter value of 10 for $\sigma_6$ allows the data to be fairly concentrated near 0 which is what we would expect for a year over year effect.
+
+Compared to most GLM's, this model ([@eq-model]) we have created is massive.
 It has 2,116 parameters.
-It has 657 different $\alpha, \beta,$ and $\eta$ estimated for each player.
-It has 88 $\delta$'s for each park and
+It has 657 different $\alpha, \beta,$ and $\eta$'s estimated for each player.
+It has 88 $\delta$'s for each park and 47 $\xi$'s for each year.
+Finally, it has 10 different $\mu$'s and $\sigma$'s for the different hyper-priors.
+All together, totaling 2,116 parameters.
+This gargantuan size is what allows such powerful inference for our model.
+Even though the data on each player individually is sparse, since there is so much data available for the model to take in beyond that player and the model uses many simulations to create estimates, this allows our model to make strong inferences to be made concerning future performance.
 
 # Results {#sec-Results}
+
+The goal of our model, as described in @eq-model, @eq-priors, and @eq-hyperpriors, was to be able to adaptively predict player $n$'s HR's.
+That is, we wished to create a model that would predict player $n$'s HR's and would adapt these predictions to be unique to player, $n$, the park played at, $p$, and the year played, $i$.
+To check the fit of their model, Fellingham and Fisher (2017) selected popular players (such as Hank Aaron and Barry Bonds) as a representative sample of the player population.
+However, the problem with this appraoch is that Fellingham and Fisher (2017) have selected players who are phenomenal hitters and generally outliers in the data.
+For example, Barry Bonds currently holds the single season and career HR records.
+Which means that he has the most HR's hit in one season and the most hit over the course of a career compared to all people who have every played in MLB.
+The career HR record he holds was previously held by Hank Aaron.
+Obviously, this sample of players is not representative of the population of players considered by Fellingham and Fisher (2017).
+So, while it may make the paper more enjoyable to read, it does not provide use with a rigourus understanding of the general fit of this model.
+We believe that a random sample should have been used to examine the fit of their model since it would give a better sense of overall model fit.
+Which, is exactly what we did in this paper.
+From the 657 players considered, we used R to select four random players to see how our model both fits and adapts to the different careers of players.
+These players are: Robin Yount ([@sec-yount]), Pat Tabler ([@sec-tabler]), Junior Felix ([@sec-felix]), and Kendrys Morales([@sec-morales]).
+
+In exploring the efficacy of our model, we chose to examine $\pi_{nip}$ for the players mentioned above instead of $HR_{nip}$.
+We did this because moving between the expected value of HR's ($E(HR_{nip})$) and the HR probability ($\pi_{nip}$) is rather elementary due to $E(HR_{nip})=AB_{nip}\times\pi_{nip}$, since $HR_{nip}$ is a binomial random variable.
+We assumed when making our model that $AB_{nip}$ were given which means moving between $E(HR_{nip})$ and $\pi_{nip}$ is simply a matter of multiplication or division depending on which term you would like to reference.
+Thus, because of the similarity between what $E(HR_{nip})$ and $\pi_{nip}$ are describing we decided to only focus on $\pi_{nip}$ without loss of generality.
+
+Furthermore, we chose to examine $\pi_{nip}$ relative to centered age for all figures.
+This is because centered age corresponds directly to the parameters $\beta, \eta,$ and $\xi$ in our model ([@eq-model]).
+It should be obvious why centered age corresponds to $\beta$ and $\eta$.
+However, the reason we know that centered age corresponds to $\xi$ directly is because players are only an age for one season.
+Thus, if we are given a year and player $n$'s corresponding age we can deduce the year for all ages of the player.
+This means that the year-over-year effect, which $\xi$ captures, is directly present in the figures.
+The parameters $\alpha$ and $\delta$ work more in the background of the predictions presented in the figures.
+We leave the discussion of each to when they appear in the figures of certain players.
 
@@
 This section should be dedicated to talking about general fit of the model.
-Explore using the normals plots that we've been using with the confidence intervals to see if the parameters are near their priors.
+Explore using the normal plots that we've been using with the confidence intervals to see if the parameters are near their priors.
 And, also explore using the trace-plots.
@@
 
-## Robin Yount
+## Robin Yount {#sec-yount}
 
@@
 
-See @fig-p651Act and @fig-p651Pred.
-
-## Pat Tabler
+Consider player $n=651$.
+That is, consider the player Robin Yount.
+In @fig-p651Act we can see how Yount's actual HR probability has changed over the years that he played.
+The x-axis of @fig-p651Act describes Yount's deviation from the centered age, where the centered age value of 0 represents a player being 30 years old.
+And the y-axis represents the actual HR probability of Yount.
+Thus, the point (0, 3.25) tells us that Yount's actual HR probability was 3.25% when he was 30 years old.
+Similarly, the points (-10, 1.5) and (6, 2) represent Yount's actual HR probability being 1.5% when he was 20 years old and 2% when he was 36 years old, respectively.
+Included in @fig-p651Act is a blue trend line that snakes through the data.
+This line does not represent anything meaningful, it was included simply to make seeing and interpreting the trend present in the scatter-plot easier.
+To that end, we can see in @fig-p651Act that during the first 10 years of his career, Yount saw a steady increase in his actual HR probability until he was 28 years old.
+From when he was 29 years old until he was about 35 years old, there some fluctuation in his HR probability by overall it stayed fairly consistent at around 2.5-3%.
+Finally, in the last three years of his career we can see Yount's HR probability decline to about 1.75% when he was 38 years old.
+Overall, Yount appears to have a typical career compared to what the told our model to expect.
+There was a steady increase in probability until he was about 30, followed by a decline afterwards as he got older.
+The shape of the trend-line through the scatter-plot follows a vaguely parabolic shape which is what we would expect to see.
+Thus, Yount will be a good benchmark for the model to see how it predicts "typical" players.
+
+In @fig-p651Pred we can examine how our model predicted Yount's HR probability, $\pi$, to change over his career.
+The x-axis is the same as it is in @fig-p651Act, where it describes Yount's deviation from the cetnered age (which is 30).
+The y-axis describes Yount's predicted HR probability $\pi$.
+An important note is that the y-axis is not measured on the logistic scale.
+We calculated it on the logistic scale but then converted it back to the normal scale, and use the normal scale in the figure, to prevent any confusion.
+Interpreting the points in @fig-p651Pred is very similar to the interpretation of points for @fig-p651Act.
+Each point in @fig-p651Pred represents the model's prediction of Yount's $\pi$ for a centered age value.
+For example, the point (0, 3.25) represents the model predicting that Yount's $\pi=3.25$ when he is 30 years old, which was very close to the actual value.
+Similarly, the points (-10, 1) and (6, 2) represent our model's prediction that Yount's $\pi=1$ when he is 20 years old and $\pi=2$ when he is 36 years old, respectively, which is also very close to what we saw in @fig-p651Act.
+Included in @fig-p651Pred is a red trend line for the same reasons mentioned in discussing @fig-p651Act.
+The red trend line represents the overall trend of the data in the scatter-plot.
+It is not nearly as necessary here, because the data is so clearly clustered in a parabolic pattern, but it is helpful in future plots.
+We can see in @fig-p651Pred that the predictions from our model follow closely to what we told it to expect.
+There is an increase in HR probability, $\pi$, for Yount from when he is 18 to when he is 32.
+This is quite close to what we see in the figure representing his actual HR probability.
+Then we see a sharp decline in predicted HR probability for Yount as he ages past 32 until he retires at the age of 38.
+
+Overall, it seems that our model fit Yount's trajectory well.
+In any given year we could see that the predicted HR probability, $\pi$, was closer to the mean value of 3% than Yount's actual HR probability, but this is a by-product of multilevel modeling and not a defect with the model.
+Our model is told to assume average and update expectations given past performance.
+Which is what it did, on any given year it assumes Yount is more average than he really is, for both good and bad years.
+In addition, since we can see that Yount's actual trajectory and his projected trajectory were quite similar, it seems fair to say that our model did a good job at predicted Yount's HR probability.
+Of course, this is not unexpected; Yount followed a trajectory that our model should do very well with.
+There were not many unique featuers to his career where he did astoundingly well or poorly.
+Thus, although our model did well with Yount, we expected it to do well because it was not a challenging player to model.
+It will be interesting to consider other players who do not have as normal of a career.
+
+## Pat Tabler {#sec-tabler}
 
@@
 
-## Junior Felix
+## Junior Felix {#sec-felix}
 
@@
 
-## Kendrys Morales
+Probably don't need to include this examination. Paper will probably be long enough and he's not that interesting.
+
+## Kendrys Morales {#sec-morales}
 
@@
 
+## Goodness of Fit - $\alpha$
+
 # Conclusions and Directions for Future Research
@@
 #| label: fig-pDist
+#| fig-cap: "Distribution of Players' HR Proportions"
 ggplot(data = Batting)+
@@
 #| label: fig-pDistSmall
+#| fig-cap: "Constricted View of Distribution of Players' HR Proportions"
 #I want this to have a little more breaks in the x-axis to highlight where the median is
@@
 #| label: fig-p651Act
+#| fig-cap: "Actual Trajectory for player #651 - Robin Yount"
 ggplot(data = select_player_Batting)+
@@
 #| label: fig-p651Pred
+#| fig-cap: "Predicted Career Trajectory for #651 - Robin Yount"
 ggplot(data = select_player_Batting)+
@@
 #| label: fig-p582Act
+#| fig-cap: "Actual Trajectory for player #582 - Pat Tabler"
 ggplot(data = i582.Trajectory)+
@@
 #| label: fig-p582Pred
+#| fig-cap: "Predicted Career Trajectory for #582 - Pat Tabler"
 ggplot(data = i582.Trajectory)+
@@
 #| label: fig-p177Act
+#| fig-cap: "Actual Trajectory for player #177 - Junior Felix"
 ggplot(data = i177.Trajectory)+
@@
 #| label: fig-p177Pred
+#| fig-cap: "Predicted Career Trajectory for #177 - Junior Felix"
 ggplot(data = i177.Trajectory)+
@@
 #| label: fig-p420Act
+#| fig-cap: "Actual Trajectory for player #420 - Kendrys Morales"
 ggplot(data = i420.Trajectory)+
@@
 #| label: fig-p420Pred
+#| fig-cap: "Predicted Career Trajectory for #420 - Kendrys Morales"
 ggplot(data = i420.Trajectory)+
@@
 #| label: fig-alphaTrace
+#| fig-cap: "Alpha Traceplots"
 #651, 582, 177, 420

### 2023-12-14 Lahman Analysis/Formal Analysis/Draft1.qmd
Finished through beta and eta section

@@
 
-## Baseball Background {#sec-BaseBack}
+# Baseball Background {#sec-BaseBack}
 
@@
 
-## Literature Review {#sec-LitReview}
+# Literature Review {#sec-LitReview}
 
@@
 To check the fit of their model, Fellingham and Fisher (2017) selected popular players (such as Hank Aaron and Barry Bonds) as a representative sample of the player population.
-However, the problem with this appraoch is that Fellingham and Fisher (2017) have selected players who are phenomenal hitters and generally outliers in the data.
+However, the problem with this approach is that Fellingham and Fisher (2017) have selected players who are phenomenal hitters and generally outliers in the data.
 For example, Barry Bonds currently holds the single season and career HR records.
@@
 Obviously, this sample of players is not representative of the population of players considered by Fellingham and Fisher (2017).
-So, while it may make the paper more enjoyable to read, it does not provide use with a rigourus understanding of the general fit of this model.
+So, while it may make the paper more enjoyable to read, it does not provide use with a rigorous understanding of the general fit of this model.
 We believe that a random sample should have been used to examine the fit of their model since it would give a better sense of overall model fit.
@@
 From the 657 players considered, we used R to select four random players to see how our model both fits and adapts to the different careers of players.
-These players are: Robin Yount ([@sec-yount]), Pat Tabler ([@sec-tabler]), Junior Felix ([@sec-felix]), and Kendrys Morales([@sec-morales]).
+These players are: Robin Yount ([@sec-yount]), Pat Tabler ([@sec-tabler]), and Kendrys Morales([@sec-morales]).
 
@@
 We leave the discussion of each to when they appear in the figures of certain players.
-
-In explore results, discuss both the trajectories of about 4-5 players and show how they change (probably want to sample about 10 just to be safe).
-But then, just for fun, look at which player has the highest $\alpha$ level since that would be my models prediction of best HR hitter independent of everything else.
-It will probably be Barry Bonds.
 
@@
   sample(size=10, replace=FALSE)
-
-select_players
 ```
@@
 ```
-
-## General
-
-This section should be dedicated to talking about general fit of the model.
-Explore using the normal plots that we've been using with the confidence intervals to see if the parameters are near their priors.
-And, also explore using the trace-plots.
-The linegraphs that go all over the place.
 
@@
 In @fig-p651Pred we can examine how our model predicted Yount's HR probability, $\pi$, to change over his career.
-The x-axis is the same as it is in @fig-p651Act, where it describes Yount's deviation from the cetnered age (which is 30).
+The x-axis is the same as it is in @fig-p651Act, where it describes Yount's deviation from the centered age (which is 30).
 The y-axis describes Yount's predicted HR probability $\pi$.
@@
 
-See @fig-p582Act and @fig-p582Pred.
-
-## Junior Felix {#sec-felix}
-
-```{r}
+Consider player $n=582$.
+That is, consider the player Pat Tabler.
+In @fig-p582Act we can see how Tabler's actual HR probability has changed over the years that he played.
+The x-axis of @fig-p582Act describes Tabler's deviation from the centered age, where the centered age value of 0 represents a player being 30 years old.
+And the y-axis represents the actual HR probability of Tabler.
+Thus, the point (-1, 2) tells us that Tabler's actual HR probability was 2% when he was 29 years old.
+In fact, the interpretation of @fig-p582Act is exactly the same as it was for @fig-p651Act, expect the probabilities are for Pat Tabler instead of Robin Yount.
+Observant readers may have noticed that there are two data points when $x=0$ and when $x=2$.
+This is because Pat Tabler was traded twice, once when he was 30 and another time when he was 32 years old.
+Since he was traded, it meant that he had two different home ballparks for that season.
+Thus, he had two different HR probabilities for that year with each representing his HR probability for a specific home ballpark.
+This "double observation" does not impact our impact our model at all.
+It is simply a quirk of the data.
+Included in @fig-p582Act is a blue trend line that snakes through the data.
+This line does not represent anything meaningful, it was included simply to make seeing and interpreting the trend present in the scatter-plot easier.
+To that end, we can see in @fig-p582Act that for the first 7 years of his career Tabler saw an increase in his actual HR probability, which lasted until he was 29 years old.
+Tabler's average HR probability at this time was about 1.5% during this period.
+Every year after he turned 29, with the exception of when he was 32, he performed starkly worse than he did in the beginning of his career.
+The average of these points being around 0.5%.
+The only outlier in this period was the season where he was 32, where he was traded to the New York Mets and had a higher HR probability for them that year then he did with the Kansas City Royals.
+However, this higher HR probability was probably due to sample size.
+He had the same amount of HR's with both teams but more than 4 times as many AB's with the Royals compared with the Mets.
+Overall, we can see that Tabler's career was different than Yount's in a couple key ways.
+The first is that Tabler stays well below the average HR probability for all of his career, whereas Yount says very close to it.
+The second way is that Tabler peaks before he is 30 and does so bad for the final years of his career that there is not much decline.
+Thus, it will be interesting to see how our adapts to this very interesting and unique player.
+
+In @fig-p582Pred we can examine how our model predicted Tabler's HR probability, $\pi$, to change over his career.
+The x-axis is the same as it is in @fig-p582Act, where it describes Tabler's deviation from the centered age (which is 30).
+The y-axis describes Tabler's predicted HR probability $\pi$.
+An important note is that the y-axis is not measured on the logistic scale.
+We calculated it on the logistic scale but then converted it back to the normal scale, and use the normal scale in the figure, to prevent any confusion.
+Interpreting the points in @fig-p582Pred is very similar to the interpretation of points for @fig-p582Act.
+Each point in @fig-p582Pred represents the model's prediction of Tabler's $\pi$ for a centered age value.
+For example, the point (-1, 1.55) represents the model predicting that Tabler's $\pi=1.55$ when he is 29 years old.
+Similarly, the points (-4, 1.2) and (4, 0.8) represent our model's prediction that Tabler's $\pi=1.2$ when he is 26 years old and $\pi=0.8$ when he is 34 years old, respectively.
+However, the $\pi$ values were not nearly as accurate for Tabler as they were for Yount.
+We can see that the high variation of the actual HR probabilities has forced our model rely more on the priors than on the actual data.
+That is, we can see our estimates for $\pi$ are often much closer to the average HR rate across players, which is 3%, than they are to Tabler's specific data because Tabler is such an outlier.
+Included in @fig-p582Pred is a red trend line for the same reasons mentioned in discussing @fig-p582Act.
+The red trend line represents the overall trend of the data in the scatter-plot.
+Coming into this, the worry was that our model would rely too heavily on its priors and not be able to update its parabolic shape to match the strange career that Tabler had.
+However, after quickly examining the red line anyone can see that it has clearly updated to fit Tabler's data much better than @fig-p651Pred.
+That is, we can see that the red line from @fig-p582Pred matches the blue line in @fig-p582Act fairly well, demonstrating that the model updates itself based on the data it is fed.
+We can see that red line captures Tabler's beginning career well, with a slight increase in HR probability until he is 29.
+This is then followed by a steep and drastic drop in predicted HR rate for all years including and past when Tabler is 30 years old.
+Thus, although the actual proportions themselves do not line up as well as we may have hoped, we can see that the model does match trends well, which is a good sign.
+Thus, it will be interesting to see how different extreme player trajectories challenge the model.
+
+```{r}
+#| eval: false
 # Junior Felix (177) Estimates cell
@@
 ```{r}
+#| eval: false
 #Create the trajectory DF for player 177
@@
 
-See @fig-p177Act and @fig-p177Pred.
-
-Probably don't need to include this examination. Paper will probably be long enough and he's not that interesting.
-
 ## Kendrys Morales {#sec-morales}
@@
 
-## Goodness of Fit - $\alpha$
-
-# Conclusions and Directions for Future Research
+Consider player $n=420$.
+That is, consider the player Kendrys Morales.
+In @fig-p420Act we can see how Morales' actual HR probability has changed over the years that he played.
+The x-axis of @fig-p420Act describes Morales' deviation from the centered age, where the centered age value of 0 represents a player being 30 years old.
+And the y-axis represents the actual HR probability of Morales.
+Thus, the point (0, 4.25) tells us that Morales' actual HR probability was 4.25% when he was 30 years old.
+In fact, the interpretation of @fig-p420Act is exactly the same as it was for both @fig-p651Act and @fig-p582Act, expect the probabilities are for Kendrys Morales instead of Pat Tabler or Robin Yount.
+Observant readers may have noticed that there are two data points when $x=1$ and when $x=6$.
+This is because Morales was traded twice, once when he was 31 and another time when he was 36 years old.
+Since he was traded, it meant that he had two different home ballparks for that season.
+Thus, he had two different HR probabilities for that year with each representing his HR probability for a specific home ballpark.
+This "double observation" does not impact our impact our model at all.
+It is simply a quirk of the data.
+Included in @fig-p420Act is a blue trend line that snakes through the data.
+This line does not represent anything meaningful, it was included simply to make seeing and interpreting the trend present in the scatter-plot easier.
+To that end, we can see in @fig-p420Act that for the first four years of his career Morales saw and increase in his actual HR probability, which lasted until he was 27 years old.
+Over this period, his actual HR probability rose from 2.25% when he was a rookie, when he was 23 years old, to 6% when he was 27 years old.
+For the next four years of his career, there was a clear and sharp decline in his actual HR probability.
+His HR probability dropped from about 5.5%, when he was 28 years old, to 3.1% or 0.5% when he was 31 years old.
+Then, he saw an average increase in his HR probability to about 5% from when he was 32 years old until he was 35 years old.
+During the last year of his career, Morales was traded a final time and performed the worst he had done over his career.
+Between both teams he had a HR probability below 2%.
+Morales' career has been the most atypical compared to Yount and Tabler.
+We see him perform astonishingly well to begin and then perform much more poorly in the middle of his career when the average player excels.
+Then we see him make a recovery towards the endof his career, only for the last year of his career to be quite poor.
+This makes Morales' career very different than compared to the typical player, such as Yount, so this prediction for Morales will be one of the best tests we can provide to see how our model handles updating posteriors on non-standard data.
+
+In @fig-p420Pred we can examine how our model predicted Morales' HR probability, $\pi$, to change over his career.
+The x-axis is the same as it is in @fig-p420Act, where it describes Morales' deviation from the centered age (which is 30).
+The y-axis describes Morales' predicted HR probability $\pi$.
+An important note is that the y-axis is not measured on the logistic scale.
+We calculated it on the logistic scale but then converted it back to the normal scale, and use the normal scale in the figure, to prevent any confusion.
+Interpreting the points in @fig-p420Pred is very similar to the interpretation of points for @fig-p420Act.
+Each point in @fig-p420Pred represents the model's prediction of Morales' $\pi$ for a centered age value.
+For example, the point (0, 4.25) represents the model predicting that Tabler's $\pi=4.25$ when he is 30 years old.
+Similarly, the points (-3, 4.75) and (4, 4.9) represent our model's prediction that Tabler's $\pi=4.75$ when he is 27 years old and $\pi=4.9$ when he is 34 years old, respectively.
+These $\pi$ values are fairly accurate, similar to what they were for Yount.
+However, similar to Tabler, we can see that for the seasons that Morales performed extraordinarily well or poorly, our model tends to fit more towards the average than how Morales actually performed.
+For example, when he was 27 years old Morales had an actual HR probability of 6%, yet our model only predicts for him to hit 4.75%.
+Similarly, when he was 36 years old Morales actually had a HR probability of less than 2%, yet our model predicts that he would have a HR probability of about 4%.
+This is likely because those extreme seasons were outliers both for Morales and across the data, so our model has a hard time "believing" that they are true more than they are a fluke.
+This is one of the properties of multilevel modeling, without a lot of evidence to the contrary it will assume that the player is more average than the observations suggests, which becomes helpful in the long run for prediction.
+
+Included in @fig-p420Pred is a red trend line for the same reasons mentioned in discussing @fig-p420Act.
+The red trend line represents the overall trend of the data in the scatter-plot.
+Coming into this, the worry was that our model would rely too heavily on its priors and not be able to update its parabolic shape to match the strange career that Morales had.
+However, after quickly examining the red line anyone can see that it has clearly updated to fit Morales' strange career.
+That is, we can see that the red line from @fig-p420Pred matches the blue line in @fig-p420Act fairly well, demonstrating that the model updates itself based on the data it is fed.
+We can see that red line captures Morales' beginning career well, with an increase in HR probability until he is about 27 years old.
+This is then followed by a steep and drastic drop in predicted HR rate until he is about 31 years old.
+Then there is a recovery in $\pi$ until Morales is 34 where we again see a drop in $\pi$ as Morales finishes out his career.
+Thus, we can see that the model does match trends well.
+It captures both the peaks that we saw for Morales throughout his career and matches them up to the correct time and gets rougly the same HR probabilities that we saw in the actual data.
+Therefore, we can be confident in saying that for Morales, our model updates its expectations based on the data well.
+
+## Goodness of Fit - $\alpha$ {#sec-alpha}
+
+There are many numerical measures of goodness-of-fit statistics because researchers are often searching for ways to determine which models of their many mondels are the best for their particular goals.
+Thus, many of these numerical estimates should be used in a comparative sense, with and against models which are trained on the same data set and attempt to answer the same resarch question.
+Examples of these statistics are the Bayesian Information Criterion (BIC), the Widely Applicable Information Criterion (WAIC), the Akaike Information Criterion (AIC), and many others.
+While it would be possible to quantify these statistics for our model, it would be frivolous.
+Due to computational constraints, we were only able to generate one model to answer our research question, and it has been the one discussed extensively in this paper.
+Hence, generating these goodness-of-fit statistics would be useless because they should be used ina comparative context.
+But, there are no models with which to compare our model to.
+Our model uses unique data compared to the models which attempt to answer the same research question.
+Thus, comparing models with these statistics would be foolish.
+
+However, that does not mean that there are no ways to check how well the model has fit.
+Beyond what has been discussed in @sec-yount, @sec-tabler, and @sec-morales, there is another method to check the goodness-of-fit for a model.
+Trace-plots (sometimes called trank-plots) are a way to check how individual parameters have fit.
+We have examined how the $\alpha$ parameter has fit for four players: Yount (651), Tabler (582), Morales (420), and Junior Felix (177).
+Junior Felix has not been discussed yet because he was the fourth of the random player's sampled and did not have as interesting of a career as the other three players discussed.
+Since he did not have a very interesting career, he did not provide much insight into the model fit and updating, so we decided to not display his career trajectory.
+However, we have included it @fig-alphaTrace as the last graph in the bottom right because it is a good example of what a trace-plot should look like, which contrasts nicely with the other trace-plots.
+
+First examine the trace-plot, in the bottom right of @fig-alphaTrace, labeled "alpha\[177\]".
+This plot represents our models Markov Chain Monte Carlo (MCMC) fitting process.
+What we should see, if the fit went well, is that the chains (the different colored lines) should remain at roughly the same place over the duration of the iterations.
+This is what we see in "alpha\[177\]".
+We can see that, in general, the chains stay between -3.5 and -3 over their 5,000 iterations.
+There are no sudden deviations so we can conclude that "alpha\[177\]" or $\alpha_{177}$ in our model, has a good fit.
+This stands in contrast to the other fits for $\alpha$.
+For example, "alpha\[651\]" and "alpha\[420\]" both see a slight upward trend in their $\alpha$ estimates from iterations 5,000 to 7,000.
+Then after about iteration 7,000 we see a sharp decrease in both estimates.
+For "alpha\[651\]" this drop goes from about -3.1 to about -3.3.
+While for "alpha\[420\]" this drop goes from about -2.75 to -3.
+Furthermore, we can see fluctuation in both "alpha\[651\]" and "alpha\[420\]" from iterations 7,000 to 10,000 suggesting that even after the shift the $\alpha$ parameter still does not find a good fit.
+The same problems persist for "alpha\[582\]" but are not as pronounced.
+At around iteration 7,000 our average for "alpha\[582\]" drops from about -4.1 to -4.3.
+The fluctuations for the average value of "alpha\[582\]" are not as pronounced as they were for either "alpha\[651\]" or "alpha\[420\]", and actually look quite similar to the estimates for "alpha\[177\]".
+
+Overall, we can see that the fit for our $\alpha$ parameter is quite poor for the consdiered players, even though the our model still fits the data well.
+This suggests that the other parameter fit well since our model did a good job at predicting and adapting to players trajectories.
+Moreover, in the future if we wish to get a better fit we should consider running our MCMC chain for more iterations.
+Running for more iterations should allow the model to find the best parameter value, which is what the trace-plots visualize.
+This is likely why Fellingham and Fisher (2017) ran their model for 200,000 iterations.
+Allowing the model to iterate for that long will let it find and settle on a parameter value.
+
+## Goodness of Fit - $\beta$ and $\eta$ {#sec-betaandeta}
+
+We can also examine the parameter fits for both $\beta_n$ and $\eta_n$, for $n\in\{651,582,420,177\}$.
+We chose to group $\beta$ and $\eta$ together in this section because in general they capture the same effect.
+That effect being how aging impacts player $n$'s ability to hit HR's.
+Furthermore, one can easily tell from @fig-betaTrace and @fig-etaTrace that the trace-plots for $\beta$ and $\eta$ are not nearly as interesting as they were for $\alpha$ in @sec-alpha.
+
+Consider first the fits for $\beta$ in @fig-betaTrace.
+The fits presented are for Robin Yount (651), Pat Tabler (582), Kendrys Morales (420) and Junior Felix (177).
+All fits in this figure look exactly as a good trace-plot should.
+There are no sudden shifts in the model's estimates for $\beta_n$.
+Nor are there any serious fluctuations throughout the 5,000 to 10,000 iterations.
+They all stay at roughly the same value throughout the iterations.
+We can tell this is not due to scaling on the y-axis because, compared to @fig-alphaTrace, the y-axis of @fig-betaTrace is much more concentrated and has a range of no more that 0.2.
+Thus, since all the chains stay at roughly the same value over the iterations we know that our model has a good fit for $\beta$ for our four players considered.
+
+Consider second the fits for $\eta$ in @fig-etaTrace.
+The discussion of these fits will be almost identical to the fits for $\beta$ because the model also fit very well here for the four players considered.
+The fits presented are for Robin Yount (651), Pat Tabler (582), Kendrys Morales (420) and Junior Felix (177).
+All fits in this figure look exactly as a good trace-plot should.
+There are no sudden shifts in the model's estimates for $\eta_n$.
+Nor are there any serious fluctuations throughout the 5,000 to 10,000 iterations.
+They all stay at roughly the same value throughout the iterations.
+We can tell this is not due to scaling on the y-axis because, compared to @fig-alphaTrace, the y-axis of @fig-etaTrace is much more concentrated and has a range of no more that 0.04.
+Thus, since all the chains stay at roughly the same value over the iterations we know that our model has a good fit for $\eta$ for our four players considered.
+
+Most other Bayesian papers would continue this discussion for their other parameters, which for our model would be discussing $\delta$ and $\xi$.
+However, these parameters in our model are not player specific, which means to consider them would mean examining 14 $\delta$'s and 33 $\xi$'s.
+Discussion of of 47 additional parameters in the context of each player seems both trivial and cumbersome.
+Examining these parameters in the context of the player would not improve our understanding of the model fit.
+In fact, after examining these parameters we would simply be left with how our model got its estimates which we have already reviewed in @sec-yount, @sec-tabler, and @sec-morales.
+Thus, because the labor is high and the pay-off is low we have elected to let the reader review any additional goodness-of-fit figures and statistics they would like.
+The code that we used to generate the model is available on [GitHub](https://github.com/Sturner21/MATH-470) and readers can use it to replicate our findings and investigate our model for themselves.
+
+# Conclusions and Directions for Future Research {#sec-conclusion}
+
+Overall, our model for predict the HR production of MLB hitters appears to do a fair job.
+It does a good job at updating its predictions based on the data for each player.
+But, still keeps its predictions close to the mean unless there is strong evidence, like with Pat Tabler, to suggest that it should deviate.
+Furthermore, as we can see with the contrast between Robin Yount and Kendrys Morales, our model does not have to rely solely on what the priors tell it.
+Our model can update itself to either follow a typical career trajectory, as we saw with Robin Yount, and it can update itself to follow an atypical career trajectory, as we saw with Kendrys Morales.
+Thus, we can conclude that our model does a good job at fitting overall.
+However, this model is far from perfect.
+We could see with predictions from Pat Tabler that our model would not predict HR probabilities $\pi$ at a rate where gambling would be a wise endeavor.
 
@@
 #| label: fig-BattingCor
+#| fig-cap: "Batting Correlation Table"
 corrplot(cor(Batting_cor), type = "upper", order = "hclust", tl.col = "black", tl.srt = 45)
@@
 #| fig-cap: "Actual Trajectory for player #177 - Junior Felix"
+#| eval: false
 ggplot(data = i177.Trajectory)+
@@
 #| fig-cap: "Predicted Career Trajectory for #177 - Junior Felix"
+#| eval: false
 ggplot(data = i177.Trajectory)+
@@
 #651, 582, 177, 420
-traceplot(fit.poly, pars = c("alpha[651]", "alpha[582]", "alpha[177]", "alpha[420]"))
-```
+traceplot(fit.poly, pars = c("alpha[651]", "alpha[582]", "alpha[420]", "alpha[177]"))
+```
+
+```{r}
+#| label: fig-betaTrace
+#| fig-cap: "Beta Traceplots"
+traceplot(fit.poly, pars = c("beta[651]", "beta[582]", "beta[420]", "beta[177]"))
+```
+
+```{r}
+#| label: fig-etaTrace
+#| fig-cap: "Eta Traceplots"
+traceplot(fit.poly, pars = c("eta[651]", "eta[582]", "eta[420]", "eta[177]"))
+```

### 2023-12-14 Lahman Analysis/Formal Analysis/Draft1.qmd
Finished through conclusion

@@
 Parson and I have embarked to investigate the same research question as them: can we use a multilevel Bayesian model to predict a player's home run performance?
+
+If modeling for the sake of learning was not reason enough to embark on this endeavor, there is the possibility of financial compensation also looming with this research question.
+If one is able to accurately predict the HR's that a player hits in a season, then they would probably be able to gain employment for a MLB team.
+HR's are the pinnacle of a hitter's trip to the plate, they are guaranteed to score at least one run and can lead to up to four being scored depending on how many people are on base.
+Thus, MLB teams which seek to maximize wins are likely looking to hire analysts who can accurately predict the HR's a player would hit so that they know who to target for drafting, signing, and trading.
+In addition, sports betting has recently become legal in many places in the United States.
+Thus, if someone is able to
 
@@
 
-Overall, our model for predict the HR production of MLB hitters appears to do a fair job.
-It does a good job at updating its predictions based on the data for each player.
+Overall, our model for predicting the HR production of MLB hitters appears to do a good job.
+It does a fine job at updating its predictions based on the data for each player.
 But, still keeps its predictions close to the mean unless there is strong evidence, like with Pat Tabler, to suggest that it should deviate.
@@
 Our model can update itself to either follow a typical career trajectory, as we saw with Robin Yount, and it can update itself to follow an atypical career trajectory, as we saw with Kendrys Morales.
-Thus, we can conclude that our model does a good job at fitting overall.
+Thus, we can conclude that our model does a good job at fitting overall since it appears to take extreme careers in stride.
+
 However, this model is far from perfect.
-We could see with predictions from Pat Tabler that our model would not predict HR probabilities $\pi$ at a rate where gambling would be a wise endeavor.
+We could see with predictions of $\pi$ for Pat Tabler that our model would not predict HR probabilities $\pi$ at a rate where gambling would be a wise endeavor.
+These estimates of $\pi$ for many players would be off by enough that the chances of a making money from a bet would be unlikely.
+But, gambling was never the ultimate motivation behind the prediction.
+This paper sought to improve on the methods of Fellingham and Fisher (2017), which we believe we did.
+While it would have been nice to have the computational resources to run this model for 200,000 iterations, we were able to get somewhat stable estimates for our parameters on $\frac{1}{20}$ of the iterations.
+That is, instead of needing 200,000 iterations to get stable estimates, we were able to get stable estimates using only 10,000 iterations.
+This suggests that our model is better Fellingham and Fisher's (2017) model and predicting HR's.
+
+Other than having better computational power, this model could be improved with better data.
+We implicitly assume in this paper that the probability of an AB resulting in a HR remains constant throughout an AB.
+However, we know that this is not true.
+A variety of factors that happen during the AB can effect the probability that any given pitch, and thus AB, results in a HR.
+These factors include, but are not limited to, the type of pitch, the speed of the pitch, crowd noise, state of the game, etc.
+Having access to this data would be interesting because it would allow our model to update its predictions of a player hitting a HR both before and during an AB.
+Some of this data is available to the public, and incorporating it into future research is an example of an area where this model could be improved upon.
+
+In the future, this model could easily be adapted to predict other outcomes too, such as hits, walks, strike outs, etc.
+Any batting outcome would lend itself nicely to this modeling structure.
+Thus, in the future, it would be interesting to see how this model could be applied to other areas of baseball for prediction.
 

### 2023-12-14 Lahman Analysis/Formal Analysis/Draft1.qmd
Revised through beta

@@
 Thus, MLB teams which seek to maximize wins are likely looking to hire analysts who can accurately predict the HR's a player would hit so that they know who to target for drafting, signing, and trading.
-In addition, sports betting has recently become legal in many places in the United States.
-Thus, if someone is able to
+In addition, sports betting has recently become legal in many states across the United States.
+Thus, if someone is able to accurately predict HR's before a season starts, they would be able to make a large amount of money from gambling.
+Dr.
+Parson and I are more interested in the former scenario concerning gainful employment, but other readers may not share our interests and it is fair to point out the obvious consequences of this research.
 
@@
 However, in a modeling context it is very important.
-When modeling a player's ability to hit home runs (HR), as this paper is concerned with, the distinction is important because it is unfair to consider all the PA's of a player because this set of outcomes includes outcomes where it is assumed that the player has not been given a fair chance to hit the ball.
-For similar reasons, considering the subset of outcomes where it is assumed that they player has been given a fair chance to hit the ball seems more appropriate, this subset is AB's.
+When modeling a player's ability to hit home runs (HR), as this paper is concerned with, the distinction is important because it is unfair to consider all the PA's of a player since this set of outcomes contains outcomes where it is assumed that the player has not been given a fair chance to hit the ball.
+For similar reasons, considering the subset of outcomes where it is assumed that they player has been given a fair chance to hit the ball seems more appropriate, this subset being AB's.
 Thus, this paper uses AB's as the main predictor for home runs hit by a certain player.
@@
 Furthermore, it is logical to ask: why not consider both?
-Incorporating more predictors can make a model more accurate at prediction, which both Fellingham and Fisher (2017) and this paper are concerned with.
-The reasons this paper does not consider both is because, especially in a Bayesian context, researchers must be concerned with multicollinearity.
-Without a doubt there is a strong correlation between the two because one is a subset of the other.
+Incorporating more predictors can often make a model more accurate at prediction; and both Fellingham and Fisher (2017) and this paper are concerned with maximizing predictive power of our respective models.
+The reason this paper does not consider both is because of multicollinearity.
+Multicollinearity is where two independent variables in a model are highly or perfectly correlated and the model is unable to determine which variable is responsible for the corresponding change in the depending variable.
+Without a doubt there is a strong correlation between PA's and AB's because one is a subset of the other.
 In fact, we can see that the correlation between the two is quite strong in @fig-BattingCor.
-When multicollinearity is introduced into a model, it can be difficult for the Markov Chain Monte Carlo (MCMC) algorithm to parse out which of the two variables is responsible for the changes in the dependent variable.
+If we included both PA's and AB's in our model, our model's alorithm Markov Chain Monte Carlo (MCMC) would have a difficult time determining which of the two variables is responsible for the changes in the dependent variable, HR's.
+For other generalized linaer models (GLM's) it becomes impossible to deal with mullticolinearity and either the model parameters need to be changed or the model is deemed imossible to calculate.
 Thus, we neglect PA's from our model both because theory stipulates it is an inferior predictor and used in tandem with our current predictor it would cause problems for the modeling software.
@@
 above sea level.
+These changes in eleveation allow the ball to fly farther in the air, and hence also impact HR rates.
 As illustrated there are not only vastly different dimensions that baseball fields can take on but also different environments which affect a player's ability to hit home runs.
@@
 
-MLB, as we know it, has been played since about 1871 and the origins of the game go back even father than that.
-Over this time the some rules of the game have undergone incredible transformations which vastly change the nature of the game.
+MLB has existed since about 1871.
+Since its inception the game has undergone some incredible transformations which have vastly changed the nature of the game, making it into what we know today.
 For example, from 1900-1920 baseball was in the "deadball era." During this period the materials used to create baseballs made it so that the ball absorbed much more force than it does today.
-This had a plethora of implications for the game and how it was played; HR's were much less common because they were harder to hit.
+This had a plethora of implications for the game and how it was played; for example, HR's were much less common then they are today because they were harder to hit.
 In contrast, the period from 1994-2004 is known informally as the "steroid era" because of the rampant use of performance enhancing drugs (PED) used by players.
 This, in combination with other factors such as harder materials used to make baseballs, led to players obliterating a variety of previous hitting records, such as the career and season home runs records.
-It is obvious from these facts that decade played for player's had a drastic impact on their ability to produce outcomes.
-In fact, this can be visualized by considering all the runs scored during a given season from 1871-2019 displayed in @fig-RunsByYear.
+It is obvious from these facts that the decade played for player's had a drastic impact on their ability to produce some outcomes, such as HR's.
+In fact, this can be visualized by considering all the runs scored during a given season from 1871-2022 displayed in @fig-RunsByYear, or considering the total HR's hit from 1871-2022 displayed in @fig-HRByYear.
 
 Finally, age is an important factor in athletic ability.
-To an extent in MLB, seniority can act both as a blessing and a curse.
-On the one hand, older players are less able to hit home runs because their body is starting to lose its peak athletic performance it had when it was, say, 30.
+In MLB, seniority can act both as both a blessing and a curse.
+On the one hand, older players are less able to hit home runs because their body is starting to lose its peak athletic performance compared to when it was, say, 30.
 On the other hand, these older players also have more experience than younger players and thus may be better able to hit home runs from strategies they have found during their time playing.
-Further we can visualize this by looking at the median home runs hit by each player from 1871-2022, which is displayed in @fig-HR.By.Age.All.
+Further we can visualize this by looking at the median HR's hit by each player from 1871-2022, which is displayed in @fig-HR.By.Age.All.
 However, this graph does not depict some of the nuances of the data well.
-For example, we have almost no data on how often 18 -year-olds hit HR's because they play in MLB infrequently.
+For example, we have almost no data on how often 18 -year-old players hit HR's because they play in MLB infrequently.
 As can be seen with the standard error bars on the ages above 40, older players suffer a similar problem of small sample size.
-Instead, consider @fig-HR.By.Age.20to40 which shows only considers players who are 20-40 years old during the season.
+Instead, consider @fig-HR.By.Age.20to40 which only displays players who are 20-40 years old during the season.
 This graph clearly shows the complex non-linear relationship between player performance and age.
 It shows that players are generally improving in their ability to hit HR's from 20-30 years old.
-Then, however, we see a slight downwards trend followed by an upward trend with increasing error bars.
+Then, however, we see a slight downwards trend followed by an upward trend with increasing error bars from ages 30-40.
 This shows that as players age past 30 their abilities start to decline.
@@
 However, if the player is not generally considered good, then a team will likely opt to try out a new young rookie who may not be as good in the moment but is more projectable.
-Thus, we see less players playing past 35 which explains the higher standard errors, but they are of a higher caliber than any other age group which is why they are still playing, thus we see their proportion as higher.
+Thus, we see less players playing past 35 which explains the higher standard errors, but they are of a higher caliber than any other age group which is why they are still playing.
+Thus, we see their proportion as higher than we would expect, given the trend of other 30-35 year old players.
 
@@
 
-As mentioned, Fellingham and Fisher (2017) have both considered and published their thoughts on this research question, which concerns predicting player home run performance using Bayesian multilevel modeling.
+As mentioned, Fellingham and Fisher (2017) have both considered and published their thoughts concerning prediction of player home run performance using Bayesian multilevel modeling.
 To do this, Fellingham and Fisher (2017) used data from the Lahman data set spanning 1871 to 2008 to estimate the parametric weights for their model.
 Over this time frame they only considered players who had played for at least 6 seasons and who, in each of those seasons, had at least 50 AB's.
-They further restricted their pool of players to only considered seasons where the players were 18-years-old to 45-years-old.
+They further restricted their pool of players to only considered seasons where the players were 18 years old to 45 years old.
 
 To predict a player's HR's they used a binomial logistic model.
-This model predicts HR's using the AB's of a player for the $n$ trials and the underlying ability of a player to hit a HR, $\pi$ as their probability of success.
-Fellingham and Fisher (2017) estimate a player's ability to hit home runs from his age in a season, decade born, season played, and home ball park.
-This leaves Fellingham and Fisher with the following model, which we have adapted to make model parameters the same across both their model and my model (in @sec-Model):
+This model predicts HR's using the AB's of a player for the number of trials and the underlying ability of a player to hit a HR, $\pi$, as their probability of success.
+Fellingham and Fisher (2017) estimate a player's ability to hit home runs using his age in a season, decade born, season played, and home ball park as predictors.
+This leaves Fellingham and Fisher (2017) with the following model, which we have adapted to make their model's parameter notation match our model's notation (in @sec-Model):
 
@@
 
-The parameter's $\beta_{n,1-4}$ describe the weights of the orthogonal polynomial, up to the quartic, for the ages of the players.
+The parameter's $\beta_{n,1-4}$ describe the weights of the orthogonal polynomial, up to the quartic degree, for the ages of the players.
 We theorize, because Fellingham and Fisher (2017) did not explicitly articulate why, they chose a quartic model because it has a closer fit to the data displayed in @fig-HR.By.Age.All and @fig-HR.By.Age.20to40.
 The properties of a squared term capture the rise and fall of player's HR probability over their career.
-The inclusion of the cubic term allows for the modeling of multiple peaks, while the quartic term allows it to maintain that overall trend up and then down seen in the data.
+The inclusion of the cubic term allows for the modeling of two peaks, while the quartic term allows it to maintain that overall trend of up and then down seen in the data.
 This appears in @fig-HR.By.Age.20to40 where players see a steady increase in their HR probability during their career with a peak when they are 27.
@@
 Afterwards, the probability continues in a downward trend.
-This will be discussed in more depth in @sec-Model but our model, which aims to improve Fellingham and Fisher's (2017) model, does not account for the cubic or quartic effects over fear of over-fitting the age data, since the distinction between the humps is rather small.
+This will be discussed in more depth in @sec-Model but our model, which aims to improve Fellingham and Fisher's (2017) model, does not account for the cubic or quartic effects due to fear of over-fitting the age data, since the distinction between the humps is rather small.
 
@@
 
-In addition, $\alpha_n$ is not the only player specific parameter of this model.
-The parameters $\beta_{n,1-4}$ attempt to measure how a player's age might affect their ability to hit HR's.
-In contrast to $\alpha_n$, this parameter does not use multilevel modeling techniques.
-Instead, it relies on a more traditional statistical approach which considers how age affects players aggregately.
-That is, it relies on one fit for all the age data and then applying that uniformly to all players considered who are that age.
-
 In contrast, the parameter $\xi_i$ (which is $\delta_i$ in the actual paper) represents the effect that season $i$ has on player $n$'s ability to hit a HR at stadium $p$.
@@
 An example of why this effect is important refer to @fig-HRByYear, which clearly shows how HR rates have differed drastically over the years considered.
-
-The parameter $\delta_p$ (which is $\xi_p$ in the actual paper) is the effect that playing in park $p$ has on the HR probability $\pi_{nip}$ in year $i$.
-Fellingham and Fisher (2017) decided that they want this estimate to independent of the hitter.
-Hence, they use no pooling or multilevel effects in this part of the model.
-They estimate the effect that each park has on a player's ability to hit a HR and then add the associated effect to the players who play at that ballpark in a given year.
-
-One of the main reasons that Fellingham and Fisher's (2017) model is so strange are the priors that they chose.
+Although, the inclusion of $\xi_i$ is confusing because it probably captures the same effect that $\alpha_n$ captures.
+Most players in the MLB are within 10 years of age of each other with the most extreme different in age being about 20 years.
+Thus, the $\alpha_n$ of players is unlikely to differ by very much and any difference in $\alpha_n$ which appears is likely also apparent in $\xi_i$.
+Thus, the inclusion of both $xi_i$ and $\alpha_n$ is strange because it introduces a strong possibility of multicollinearity since both parameters describe roughly the same trend in the data.
+
+The parameter $\delta_p$ (which is $\xi_p$ in the actual paper) is the effect that playing in park $p$ has on the HR probability $\pi_{nip}$.
+As discussed in @sec-BaseBack, there is strong precedent to include such a term because the variance between ballpark dimension and enviornment can be drastic.
+Fellingham and Fisher (2017) estimate the effect that each park has on a player's ability to hit a HR and then add the associated effect to the players who played at that ballpark in a given year.
+
+One of the main reasons that Fellingham and Fisher's (2017) model is so strange is because of the priors that they chose.
 Disregarding that there is no justification for these priors in their paper, Fellingham and Fisher (2017) set the following priors for these parameters:
@@
 For example, consider the priors on either $\xi_i$ or $\delta_p$.
-The priors state that on a logistic scale, it is normal for the decade of birth and season of play parameters to vary along the interval \[-6.5, 1.5\].
+The priors state that on a logistic scale, it is normal for the decade of birth and season of play parameters to vary along the logistic interval \[-6.5, 1.5\].
 On the normal probability scale, this corresponds to probability fluctuations of about (0.0015, 0.81).
@@
 On the normal probability scale this becomes about 0.00055, which is 0.055% chance any AB ends in a HR.
-Confusingly, this is not even close to the 2.8% that Fellingham and Fisher (2017) even noted in their paper.
+Confusingly, this is not even close to the 2.8% that Fellingham and Fisher (2017) noted in their paper.
 Surely there must be better priors than this which Fellingham and Fisher (2017) could have used since it does not seem that even they believe these priors that they have set.
@@
 Likely, their model could have estimated their parameters much more easily had they given it stronger priors.
-This is one of the areas this paper seeks to improve Fellingham and Fisher's (2017) methods.
+This is one of the areas this paper seeks to improve on from Fellingham and Fisher's (2017) methods.
 
@@
 Given the antiquity of FORTRAN, replication of Fellingham and Fisher (2017) is quite difficult.
-Not to mention that the 200,000 iterations makes the research so computationally expensive that it makes it nearly impossible for any other research team to replicate.
+Not to mention that the 200,000 iterations makes the research so computationally expensive that it is nearly impossible for any other research team to replicate.
 This is disappointing because the scientific method relies on replication for the advancement of knowledge.
 Thus, our research seeks to make the algorithms used in this research more accessible because we chose to use R to interface with the modeling software Stan, which is written in the (more contemporary) C++ programming language.
+Furthermore, our model will use less iterations and stronger priors so that other research teams can more easily repicate our findings.
 
@@
 While it is true that a GLM could be used to predict HR's of a player it would not nearly be as accurate as a Bayesian model.
-Nor would it be able to as accurately capture the non-linearities present in the data.
-This is because the greatest strength of using the Bayesian technique is that it allows for multilevel models to be created.
-That is, Bayesian models allow for models to "learn" from other players and group certain players when there is not as much data available for them.
-In turn, this leads to better predictions about players where data is lacking.
-This "learning" is known as partial pooling, where the model identifies that parameters for each player (or year etc.) probably comes from the same distribution but may not share the exact same estimates.
+This is because the greatest strength of using Bayesian modeling is that it allows for multilevel models to be created.
+That is, Bayesian models allow "learning" from clusters (sometimes known as "pools") of observations when there is not as much data available on a particular obersvation.
+Our model employs this method much more liberally than Fellingham and Fisher (2017).
+In turn, this should allow for us to make stronger inferences than Fellingham and Fisher (2017) because of this "learning".
 
@@
 
-In this model, we say that the $HR$'s of player $n$ follow a binomial distribution with number of trials for player $n$ in season $i$, where player $n$ played in park $p$, are $AB_{nip}$.
+In this model, we say that the $HR$'s of player $n$ follow a binomial distribution with number of trials, for player $n$ in season $i$ played at park $p$, are $AB_{nip}$.
 The probability of success (hitting a HR) is denoted as $\pi_{nip}$ for player $n$ in season $i$ playing at park $p$.
-Furthermore, we measure $\pi$ on the logistic scale.
+Furthermore, we estimate $\pi$ on the logistic scale.
 
@@
 The first parameter which affects this probability, $\pi_{nip}$, is $\alpha_n$.
-In our model $\alpha_n$ can be generally thought of as the term which is used to account for a player's innate ability to hit HR's.
-From our exploratory observations of the data, we noticed that the average (median or mean) proportion of AB's which resulted in HR's to be about 0.03.
-This means that about 3 in every 100 AB's result in HR's.
+In our model $\alpha_n$ can generally be thought of as the intercept term which is used to account for a player's innate ability to hit HR's.
+From our exploratory observations of the data, we noticed that the average (median or mean) proportion of AB's which resulted in HR's is about 0.03.
+This means that about 3% AB's result in HR's.
 This can be seen in @fig-HR.By.Age.20to40, @fig-pDist, and @fig-pDistSmall.
-On the logit scale, which is what our model relies on, a probability of 0.03 is about -3.5.
+On the logit scale, which is what our model relies on for estimation, a probability of 0.03 is about -3.5.
 We recognize that some players, for whatever reason, have better "innate" abilities to hit HR's than others.
 Therefore, we used a multilevel modeling technique for $\alpha_n$ to capture that player-to-player difference.
-Instead of simply using the average probability as our intercept, which is what a dummy variable would do in a GLM, we allow our model choose an intercept which adapts depending on the player and the data.
-That is, we give our model priors that we would expect to see in the data and then, as the model interprets additional observations, it updates these priors using Bayes theorem.
-For these priors, we let $\alpha$ be normally distributed with mean $\mu_0$ (where $\mu_0$ is normally distributed with mean -3.5 and standard deviation 0.1) and standard deviation $\sigma_0$ (where the parameter $\sigma_0$ itself is exponentially distributed).
-This allows us to have separate intercepts for each player $n$, but these intercepts come from the same hyper-priors $\mu_0$ and $\sigma_0$ that the model considers for all other players.
-Thus, the model constructs its posterior predictions based on both the priors and the data.
-In effect, this makes the model appear as though it is "learning" from the data because it has the same general sense of how a player should perform, but then adapts that prediction based on the data it takes into account to form its posterior.
-Specifically, for $\alpha_n$ we say that the means of all players roughly follow the distribution of $\mu_0$, but we also tell the model that it should look at how player $n$ performed to gauge its posterior prediction.
-A similar process is followed for $\alpha_n$'s standard deviation, $\sigma_0$.
-
-The next parameter we considered which affects $\pi_{nip}$ was $\beta_n$ which represents the effect centered age (30 minus the player's age) had on player $n$'s ability to hit HR's.
+To begin capturing this difference in $\alpha_n$ across players, we set the hyper-priors for $\alpha_n$ to be $\mu_0$ and $\sigma_0$.
+Where $\mu_0\sim Normal(-3.5,0.1)$ and $\sigma_0\sim Exponential(1)$.
+We set the mean of $\mu_0$ to be -3.5 because we estimate $\alpha_n$ on the logistic scale and we want it to be close to the average player's value.
+Since the parameters of $\alpha_n$ are distributions, we can think of $\alpha_n$ as a set of distributions, for the different values that either $\mu_0$ or $\sigma_0$ could take on.
+Further, the likelihood of the particular distribution that $\alpha_n$ takes on is depndent on the likelihood of either $\mu_0$ or $\sigma_0$ taking on a particular value.
+For example, $\mu_0=-3.0$ is extremely unlikely because it lies four standard deviations away from the mean of its normal distribution -3.5.
+Thus a distribution for $\alpha_n$ being $Normal(-3,\sigma_0)$ is highly unlikely for any value of $\sigma_0$.
+We chose $\mu_0$ to be normally distributed because there is precedent in Fellingham and Fisher (2017) to use normal distributions for their mean hyper-priors.
+Another reason it is easier to interpret normal distributions.
+We chose $\sigma_0$ to be exponentially distributed because the exponential distribution does not allow for negative values and the bulk of the possible values of $\sigma_0$ would be clustered near 0.
+Since we cannot have negative values for a standard deviation it seemed appropriate.
+This use of these hyper-priors is what makes the model appear to "learn".
+The model already comes in with certain expectations of what values $\alpha_n$ should take on, rather than simply coming into each new $n$ knowing nothing, which is what happens with more traditional GLM's.
+Finally, with this new set of distributions for $\alpha_n$ being the priors, our model constructs a posterior using Bayes theorem and the data on player $n$.
+It repeats this process for all 657 players that we considered for this research question to create a different $\alpha$ distribution for each player, $n$.
+Whatever value in $\alpha_n$ has the maximum likelihood, that's what we used to generate our predictions for inference in @sec-Results.
+
+The next parameter we considered which affects $\pi_{nip}$ was $\beta_n$ which represents the effect centered age (the player's age minus 30) had on player $n$'s ability to hit HR's.
 This topic was briefly touched upon in @sec-BaseBack, where we discussed how a player's age can impact their their strategy and physical prowess.
@@
 We chose to center the the age of the all players at 30 for several reasons.
-The first was that roughly the mean and the median ages of the data was at 30.
-And second, because having a centered age makes interpreting $\beta_n$ easier.
+The first was that roughly the mean and the median ages in the data was 30.
+Second, having a centered age makes interpreting $\beta_n$ easier.
 The value of $\beta_n$ represents how changes away from the centered age affects player $n$'s probability of hitting a HR, $\pi_{nip}$.
@@
 This implicitly assumes that each player is affected identically by changes in age.
-In contrast, we employ multilevel modeling techniques to our model which assumes that most players experience the same general effects of aging, but allows for subtle changes depending on the player.
-This is the job of our hyper-priors, $\mu_1$ and $\sigma_1$.
-The parameter $\mu_1$ acts as the general mean which we expect most players to follow.
-However, since $\mu_1$ is a distribution, it allows for our model to update its posterior with observed data so that it is not heavily influenced by either a few data points or our strong priors.
-Our standard deviation parameter, $\sigma_1$ acts similarly.
-Thus, we have estimates $\beta_n$ for player $n$ which describes how aging uniquely affects their probability of hitting a HR in a given AB, $\pi_{nip}$, but these unique estimates generally follow the same trend.
+In contrast, we employ multilevel modeling techniques to our model where $\beta_n\sim Normal(\mu_1,\sigma_1)$, $\mu_1\sim Normal(0,0.1)$, and $\sigma_1\sim Exponential(10)$.
+Similar to understanding $\alpha_n$, we can interpret $\beta_n$ as a collection of distributions corresponding to the different values of its hyper-priors $\mu_1$ and $\sigma_1$ and their corresponding likelihoods.
+We chose a normal distribution for $\mu_1$ for reasons similar to $\mu_0$.
+It makes interpretation easier and there is precedent in the literature.
+Again, for similar reasons to $\sigma_0$ we chose $\sigma_1$ to be exponentially distriubted.
+This ensured that $\sigma_1$ would not be negative and that the bulk of the possible values of $\sigma_1$ would be clustered near 0.
+With the distributions calculated for $\beta_n$ we can now use $\beta_n$ as the prior and combine it with the player data to generate the posterior using Bayes theorem.
+We repeat this process for all 657 players considered to get a unique distribution for each player $\beta_n$.
+This means that each player, $n$ has a unique aging coefficient which represents how aging effects them.
+Whatever value in $\beta_n$ has the maximum likelihood, that's what we used to generate our predictions for inference in @sec-Results.
 
@@
 It is apparent in both @fig-HR.By.Age.20to40 and @fig-HR.By.Age.All that there is a parabolic relationship between age and player $n$'s ability to hit HR's.
-This ability appears to rise steadily for the average player until it reaches its apex when the player is roughly 30-years-old.
+This ability appears to rise steadily for the average player until it reaches its apex when the player is roughly 30 years old.
 Then, in the seasons where the player is more than 30-years-old we see a decline in HR probabilities.
@@
 That is, players who have earned massive contracts because they were so much better than the average player in their prime earned large multi-million dollar contracts and teams do not want to waste the this money so they play these players.
-So, we are then left with this group of outlying players who do not follow the trend of data at any point over their career, but it only became noticeable after the other, more normal, players dissapeared from the considered sample.
+So, we are then left with this group of outlying players who do not follow the trend of data at any point over their career, but it only became noticeable after the other, more normal, players disappeared from the considered sample.
 Returning to the parameter $\eta_n$, it is nearly impossible to interpret what it is doing because it describes the effect of centered age squared on HR probability $\pi$, and centered age squared on its own is already almost a meaningless concept.
 Therefore, it is best understood that this term simply accounts for the non-linearity we see in the data and not become too concerned with understanding its estimates.
-However, it is easy to justify our choice for the priors and hypterpriors of $\eta_n$, $\mu_2$, and $\sigma_2$, respectively.
+However, it is easy to justify our choice for the priors and hyper-priors of $\eta_n$, $\mu_2$, and $\sigma_2$, respectively.
 We chose $\eta_n\sim Normal(\mu_2,\sigma_2)$ for the same reasons discussed for both $\alpha_n$ and $\beta_n$: we wanted a multilevel model which described how individual players were impacted by the parameters.
@@
 Stadiums like Great American Ballpark (index=33) tend to have a better HR effect than stadiums such as Cleveland Stadium (index=19).
-By better, we mean that it the probability of hitting a HR is higher at Great American Ballpark compared to Cleaveland Stadium because our model interprets hitting a HR as a success.
+By better, we mean that it the probability of hitting a HR is higher at Great American Ballpark compared to Cleveland Stadium because our model interprets hitting a HR as a success.
 Interpreting $\delta_p$ is much easier than the previous two parameters.
@@
 This stems mostly from the observation in @fig-HRByYear which shows a steep increase in the amount of HR's hit per year across any subset of years considered.
-Further, we can tell from @fig-pByYear that these additional HR's are not just the product of exapnsion teams creating more opportunities for players to hit more HR's.
+Further, we can tell from @fig-pByYear that these additional HR's are not just the product of expansion teams creating more opportunities for players to hit more HR's.
 @fig-pByYear shows that the proportion of AB's which are HR's have risen steadily over the period considered which suggests that players' probabilities have been changing.
@@
 Of course, this is not unexpected; Yount followed a trajectory that our model should do very well with.
-There were not many unique featuers to his career where he did astoundingly well or poorly.
+There were not many unique features to his career where he did astoundingly well or poorly.
 Thus, although our model did well with Yount, we expected it to do well because it was not a challenging player to model.
@@
 We see him perform astonishingly well to begin and then perform much more poorly in the middle of his career when the average player excels.
-Then we see him make a recovery towards the endof his career, only for the last year of his career to be quite poor.
+Then we see him make a recovery towards the end of his career, only for the last year of his career to be quite poor.
 This makes Morales' career very different than compared to the typical player, such as Yount, so this prediction for Morales will be one of the best tests we can provide to see how our model handles updating posteriors on non-standard data.
@@
 Thus, we can see that the model does match trends well.
-It captures both the peaks that we saw for Morales throughout his career and matches them up to the correct time and gets rougly the same HR probabilities that we saw in the actual data.
+It captures both the peaks that we saw for Morales throughout his career and matches them up to the correct time and gets roughly the same HR probabilities that we saw in the actual data.
 Therefore, we can be confident in saying that for Morales, our model updates its expectations based on the data well.
@@
 
-There are many numerical measures of goodness-of-fit statistics because researchers are often searching for ways to determine which models of their many mondels are the best for their particular goals.
-Thus, many of these numerical estimates should be used in a comparative sense, with and against models which are trained on the same data set and attempt to answer the same resarch question.
+There are many numerical measures of goodness-of-fit statistics because researchers are often searching for ways to determine which models of their many models are the best for their particular goals.
+Thus, many of these numerical estimates should be used in a comparative sense, with and against models which are trained on the same data set and attempt to answer the same research question.
 Examples of these statistics are the Bayesian Information Criterion (BIC), the Widely Applicable Information Criterion (WAIC), the Akaike Information Criterion (AIC), and many others.
@@
 Due to computational constraints, we were only able to generate one model to answer our research question, and it has been the one discussed extensively in this paper.
-Hence, generating these goodness-of-fit statistics would be useless because they should be used ina comparative context.
+Hence, generating these goodness-of-fit statistics would be useless because they should be used in a comparative context.
 But, there are no models with which to compare our model to.
@@
 
-Overall, we can see that the fit for our $\alpha$ parameter is quite poor for the consdiered players, even though the our model still fits the data well.
+Overall, we can see that the fit for our $\alpha$ parameter is quite poor for the considered players, even though the our model still fits the data well.
 This suggests that the other parameter fit well since our model did a good job at predicting and adapting to players trajectories.
@@
 Thus, in the future, it would be interesting to see how this model could be applied to other areas of baseball for prediction.
+
+# Work Cited
+
+Fellingham, G.
+W., & Fisher, J.
+D.
+(2017).
+Predicting Home Run Production in Major League Baseball Using a Bayesian Semiparametric Model. *The American Statistician*, *72*(3), 253--264.
+https://doi.org/10.1080/00031305.2017.1401959
 

### 2023-12-15 Lahman Analysis/Formal Analysis/Draft1.qmd
Edited through data

@@
 It makes interpretation easier and there is precedent in the literature.
+Furthermore, we chose it to have a mean of 0 because we assume that there is only a slighly poitive increase in HR probability year over year, which is close enough to zero that we let the data speak for itself.
 Again, for similar reasons to $\sigma_0$ we chose $\sigma_1$ to be exponentially distriubted.
@@
 We repeat this process for all 657 players considered to get a unique distribution for each player $\beta_n$.
-This means that each player, $n$ has a unique aging coefficient which represents how aging effects them.
+This means that each player, $n$ has a unique aging coefficient which represents how aging effects their HR probability.
 Whatever value in $\beta_n$ has the maximum likelihood, that's what we used to generate our predictions for inference in @sec-Results.
@@
 This ability appears to rise steadily for the average player until it reaches its apex when the player is roughly 30 years old.
-Then, in the seasons where the player is more than 30-years-old we see a decline in HR probabilities.
+Then, in the seasons where the player is more than 30 years old we see a decline in HR probabilities.
 The shape we are left with, as can be seen in @fig-HR.By.Age.20to40, is one of an upside-down parabola.
@@
 However, as one can easily notice in @fig-HR.By.Age.20to40, the overall shape is parabolic but for ages past 37 the data begins to trend upwards.
-Likely, this trend upwards is due to players who have above average skill.
-That is, players who have earned massive contracts because they were so much better than the average player in their prime earned large multi-million dollar contracts and teams do not want to waste the this money so they play these players.
+Likely, this trend upwards is due to players who have above average skill remaining in MLB beyond their average counterparts.
+That is, players who have earned massive contracts because they were so much better than the average player in their prime earned large, multimillion dollar, contracts and teams do not want to waste the this money so they play these players.
 So, we are then left with this group of outlying players who do not follow the trend of data at any point over their career, but it only became noticeable after the other, more normal, players disappeared from the considered sample.
-Returning to the parameter $\eta_n$, it is nearly impossible to interpret what it is doing because it describes the effect of centered age squared on HR probability $\pi$, and centered age squared on its own is already almost a meaningless concept.
-Therefore, it is best understood that this term simply accounts for the non-linearity we see in the data and not become too concerned with understanding its estimates.
+
+Returning to the parameter $\eta_n$, it is nearly impossible to interpret what its meaning because it describes the relationship between centered age squared and HR probability.
+Centered age squared on its own is already almost a meaningless concept.
+Therefore, it is best understood that this term simply accounts for the non-linearity we see in the data and not become too concerned with interpreting its estimates.
 However, it is easy to justify our choice for the priors and hyper-priors of $\eta_n$, $\mu_2$, and $\sigma_2$, respectively.
@@
 Furthermore, the exponential distribution with a parameter of 100 is very concentrated near 0.
-Since we know that the coefficient of $\eta_n$ is likely to be very small since it is being multiplied by the age squared (which can be up over 100).
+Since we know that the coefficient of $\eta_n$ is likely to be very small since it is being multiplied by the age squared (which can be over 100).
 Thus, we can be confident that our standard deviation will end up being small for our estimate.
-This means that the estimates our model generates will be close to 0.
-
-Another parameter of our model is $\delta_p$, which describes the effect playing in ballpark $p$ in year $i$ has player $n$'s HR probability, $\pi_{nip}$.
+This means that the estimates our model generates for $\eta_n$ will be distributions highly concentrated near 0 until the data for player $n$ informs otherwise through Bayes theorem.
+This leaves that each player, $n$ with another unique aging coefficient which represents how non-linear aging effects their HR probability.
+Whatever value in $\eta_n$ has the maximum likelihood, that's what we used to generate our predictions for inference in @sec-Results.
+
+Another parameter of our model is $\delta_p$, which describes the effect playing in ballpark $p$ has a player's HR probability.
 There are 88 different ballparks we considered, and we generated an estimate for each of them.
-This was discussed in @sec-BaseBack, where we highlighted that players in certain ballparks are more likely to hit HR's than in others because of the different dimensions of the outfield fences.
+This was discussed in @sec-BaseBack, where we highlighted that players in certain ballparks are more likely to hit HR's than in others because of the different dimensions of the outfield fences and the different altitude of the parks.
 Furthermore, we know that these parks will effect players differently because teams play at different ballparks with varying frequency.
 For example, players will play 81 of their games at their home-stadium and 81 at various other ballparks known as "on the road." Thus, the home-stadium that player $n$ plays at is an important factor for their HR probability, $\pi$, because player $n$ plays many more games at his home-stadium compared to any other ballpark.
-In our model, we have estimated $\delta_p$ as an intercept term which captures the effect that playing at stadium $p$ has on HR probability, $\pi$.
-Stadiums like Great American Ballpark (index=33) tend to have a better HR effect than stadiums such as Cleveland Stadium (index=19).
-By better, we mean that it the probability of hitting a HR is higher at Great American Ballpark compared to Cleveland Stadium because our model interprets hitting a HR as a success.
-Interpreting $\delta_p$ is much easier than the previous two parameters.
-It simply states, on the logit scale, how the park effects the underlying probability of hitting a HR, $\pi$.
-In addition, $\delta_p$ takes advantage of multilevel modeling since we have hyper-priors, $\mu_5$ and $\sigma_5$.
+In our model, we have estimated $\delta_p$ as an intercept term and it takes advantage of multilevel modeling since we have hyper-priors, $\mu_5$ and $\sigma_5$.
 We have purposely set the hyper-prior of $\mu_5\sim Normal(0,0.01)$ with mean 0 because we assume that parks in the aggregate will not have an effect on a player's ability to hit HR's.
-That is, on average, for every Cleveland Stadium there exists a Great American Ballpark which acts as a counterweight and makes the aggregate effect 0.
-There is no formal justification for this, it is just an assumption that we make based on our preliminary exploration and our understanding that ownership (who controls the dimensions of the stadium) are not particularly biased to any ballpark size which means that we should see good variation between parks.
-Further, we assume that a fair standard deviation for $\mu_5$ is 0.01 because on the logistic scale, this means that we could see about a two percent change in the probabilities between any given stadiums.
-At most we would see a 4% difference.
+There is no formal justification for this, it is just an assumption that we make based on our preliminary exploration and our understanding that ownership (who controls the dimensions of the stadium) are not particularly biased to any ballpark size or location which means that we should see good variation between parks.
 We chose a standard deviation for $\delta_p$ to be $\sigma_5\sim Exponential(10)$ for similar reasons to all of the other $\sigma_i$'s.
 The exponential distribution cannot be negative, which we need for standard deviations.
-Further, we know that the effect of each park will probably not be huge because, although there is deviation, there is not extreme deviation between parks that makes some "unfair." Thus, with our parameter of 10 for the exponential distribution we are being fairly confident with our model about our estimates of the priors, in that we do not allow much deviation from our mean of 0 unless the data strongly suggests otherwise.
-
-Our final parameter, $\xi_i$ represents the change in the underlying probability of player hitting a HR in the year $i$.
+Further, we know that the effect of each park will probably not be huge because, although there is deviation, there is not extreme deviation between parks that makes some "unfair" to play at.
+With the distributions calculated for $\delta_p$ (similar to every other multilevel parameter) we can now use $\delta_p$ as the prior and combine it with the player data to generate the posterior using Bayes theorem.
+We repeat this process for every park $p$ to generate the posterior distribution of possible effects that playing at park $p$ has on a player's HR probability.
+Whichever value in $\delta_p$ has the maximum likelihood, that is what we use to generate our predictions for inference in @sec-Results.
+
+Our final parameter, $\xi_i$ represents the change in the underlying probability of hitting a HR in the year $i$.
 This stems mostly from the observation in @fig-HRByYear which shows a steep increase in the amount of HR's hit per year across any subset of years considered.
@@
 This means that there is an underlying change in the $\pi$ depending on the year considered.
-The reason for this change is irrelevant.
-It could be because batters are becoming better or pitchers are getting worse, or it could be some other unrelated reason.
-Regardless, we can see that there is an increase and we need to account for this in our estimates for $\pi$ of player $n$.
-Thus, we include $\xi_i$ to represent the effect that playing in a certain year will have on the player's ability to hit a HR.
+Regardless of the reason for this change, we can see that there is an increase and we need to account for this in our estimates for $\pi$ for player $n$.
+Thus, we include $\xi_i$ to represent the effect that playing in year $i$ will have on a player's ability to hit a HR.
 We set the hyper-prior of $\xi_i$ as $\mu_6\sim Normal(0, 0.25)$ because we can see that the HR proportion change year over year in @fig-pByYear by only fractions of a percent.
-Thus, when we set our prior with mean 0 and standard deviation of 0.25, we are saying that we expect most year of year variation to be from \[-4, -3\] on the logit scale.
-On the normal this is \[0.018, 0.047\] which means that we would expect at most a change of 3% year over year.
-This is much larger than we would actually anticipate in a change but it allows the model to learn from the data to learn from the data more since we are unsure of exactly what the change beyond knowing that there is a change year over year.
+Thus, when we set our prior with mean 0, we are saying that we expect most year over year variation to be low and we would rather have the data speak through in the posterior compared to use setting a negligibly small prior.
 Furthermore, we set the hyper-prior $\sigma_6\sim Exponential(10)$ for the same reasons mentioned during the other $\sigma_i$'s.
-The exponential distribution ensures that we never have a negative estimate for our standard deviation of $\xi_i$ and having a parameter value of 10 for $\sigma_6$ allows the data to be fairly concentrated near 0 which is what we would expect for a year over year effect.
+The exponential distribution ensures that we never have a negative estimate for our standard deviation of $\xi_i$, and having a parameter value of 10 for $\sigma_6$ allows the data to be fairly concentrated near 0 which is what we would expect for a year over year effect.
+With the distributions calculated for $\xi_i$ (similar to every other multilevel parameter) we can now use $\xi_i$ as the prior and combine it with the player data to generate the posterior using Bayes theorem.
+We repeat this process for every year $i$ to generate the posterior distribution of possible effects that playing in year $i$ has on a player's HR probability.
+Whichever value in $\xi_i$ has the maximum likelihood, that is what we use to generate our predictions for inference in @sec-Results.
 
@@
 This gargantuan size is what allows such powerful inference for our model.
-Even though the data on each player individually is sparse, since there is so much data available for the model to take in beyond that player and the model uses many simulations to create estimates, this allows our model to make strong inferences to be made concerning future performance.
+Even though the data on each player individually is sparse, since there is so much data available for the model to take in beyond that player and the model uses many simulations to create estimates, this allows our model to make strong inferences concerning future performance.
 
@@
 For example, Barry Bonds currently holds the single season and career HR records.
-Which means that he has the most HR's hit in one season and the most hit over the course of a career compared to all people who have every played in MLB.
-The career HR record he holds was previously held by Hank Aaron.
+Which means that he has the most HR's hit in one season and the most hit over the course of a career compared to all people who have ever played in MLB.
+The career HR record he holds was previously held by Hank Aaron, who now sits second on the list.
 Obviously, this sample of players is not representative of the population of players considered by Fellingham and Fisher (2017).
@@
 
-Furthermore, we chose to examine $\pi_{nip}$ relative to centered age for all figures.
+Furthermore, we chose to examine $\pi_{nip}$ relative to centered age for all player.
 This is because centered age corresponds directly to the parameters $\beta, \eta,$ and $\xi$ in our model ([@eq-model]).
@@
 We leave the discussion of each to when they appear in the figures of certain players.
+
+## Data
+
+Similar to Fellingham and Fisher (2017), we have used the Lahman data set for player statistics.
+While Lahman has data on a wide range of MLB realted topics, we decided to focus on a singular data sheet.
+We focused on the "Batting" data set in Lahman which describes player-year-team batting statistics.
+That is, each entry in Batting is for a player $n$ who played in year $i$ for team $t$.
+This allows multiple entries for the same year for the same player if they were traded, which is important for our analysis because different teams implies different home ballparks.
+
+In order to be considered in our analysis a player had to be in Batting and have played for at least 6 seasons from 1973-2019.
+We chose these cut-offs because in 1973 the American League allowed the Designated Hitter, signaling that there would be an increased specialization in labor for hitting.
+We chose 2019 as the other cut-off because 2020 was the COVID season which was drastically shortened and very different compared to almost all other previous seasons.
+Furthermore, in order to be considered in our analysis players had to have had at least 50 AB's across those 6 seasons played.
+The effect of these cut-offs was that we were left with players who had longer careers and who hit HR's.
+There were many players dropped who never hit HR's because of their short stints which made prediction hard and prediction for these playes did not seem fruitful given their general inconsequential careers.
 
@@
 
+Friendly, M., Dalzell, C., Monkman, M., & Murphy, D. (2022). Lahman: Sean 'lahman' 27 baseball database. <https://CRAN.R-project.org/package=Lahman>
+
 # Figures {#sec-Figures}

### 2023-12-15 Lahman Analysis/Formal Analysis/Draft1.qmd
Finished rough draft 2

@@
 In fact, we can see that the correlation between the two is quite strong in @fig-BattingCor.
-If we included both PA's and AB's in our model, our model's alorithm Markov Chain Monte Carlo (MCMC) would have a difficult time determining which of the two variables is responsible for the changes in the dependent variable, HR's.
-For other generalized linaer models (GLM's) it becomes impossible to deal with mullticolinearity and either the model parameters need to be changed or the model is deemed imossible to calculate.
+If we included both PA's and AB's in our model, our model's algorithm Markov Chain Monte Carlo (MCMC) would have a difficult time determining which of the two variables is responsible for the changes in the dependent variable, HR's.
+For other generalized linear models (GLM's) it becomes impossible to deal with multicollinearity and either the model parameters need to be changed or the model is deemed impossible to calculate.
 Thus, we neglect PA's from our model both because theory stipulates it is an inferior predictor and used in tandem with our current predictor it would cause problems for the modeling software.
@@
 above sea level.
-These changes in eleveation allow the ball to fly farther in the air, and hence also impact HR rates.
+These changes in elevation allow the ball to fly farther in the air, and hence also impact HR rates.
 As illustrated there are not only vastly different dimensions that baseball fields can take on but also different environments which affect a player's ability to hit home runs.
@@
 The parameter $\delta_p$ (which is $\xi_p$ in the actual paper) is the effect that playing in park $p$ has on the HR probability $\pi_{nip}$.
-As discussed in @sec-BaseBack, there is strong precedent to include such a term because the variance between ballpark dimension and enviornment can be drastic.
+As discussed in @sec-BaseBack, there is strong precedent to include such a term because the variance between ballpark dimension and environment can be drastic.
 Fellingham and Fisher (2017) estimate the effect that each park has on a player's ability to hit a HR and then add the associated effect to the players who played at that ballpark in a given year.
@@
 Thus, our research seeks to make the algorithms used in this research more accessible because we chose to use R to interface with the modeling software Stan, which is written in the (more contemporary) C++ programming language.
-Furthermore, our model will use less iterations and stronger priors so that other research teams can more easily repicate our findings.
+Furthermore, our model will use less iterations and stronger priors so that other research teams can more easily replicate our findings.
 
@@
 This is because the greatest strength of using Bayesian modeling is that it allows for multilevel models to be created.
-That is, Bayesian models allow "learning" from clusters (sometimes known as "pools") of observations when there is not as much data available on a particular obersvation.
+That is, Bayesian models allow "learning" from clusters (sometimes known as "pools") of observations when there is not as much data available on a particular observation.
 Our model employs this method much more liberally than Fellingham and Fisher (2017).
@@
 Since the parameters of $\alpha_n$ are distributions, we can think of $\alpha_n$ as a set of distributions, for the different values that either $\mu_0$ or $\sigma_0$ could take on.
-Further, the likelihood of the particular distribution that $\alpha_n$ takes on is depndent on the likelihood of either $\mu_0$ or $\sigma_0$ taking on a particular value.
+Further, the likelihood of the particular distribution that $\alpha_n$ takes on is dependent on the likelihood of either $\mu_0$ or $\sigma_0$ taking on a particular value.
 For example, $\mu_0=-3.0$ is extremely unlikely because it lies four standard deviations away from the mean of its normal distribution -3.5.
@@
 It makes interpretation easier and there is precedent in the literature.
-Furthermore, we chose it to have a mean of 0 because we assume that there is only a slighly poitive increase in HR probability year over year, which is close enough to zero that we let the data speak for itself.
-Again, for similar reasons to $\sigma_0$ we chose $\sigma_1$ to be exponentially distriubted.
+Furthermore, we chose it to have a mean of 0 because we assume that there is only a slightly positive increase in HR probability year over year, which is close enough to zero that we let the data speak for itself.
+Again, for similar reasons to $\sigma_0$ we chose $\sigma_1$ to be exponentially distributed.
 This ensured that $\sigma_1$ would not be negative and that the bulk of the possible values of $\sigma_1$ would be clustered near 0.
@@
 Similar to Fellingham and Fisher (2017), we have used the Lahman data set for player statistics.
-While Lahman has data on a wide range of MLB realted topics, we decided to focus on a singular data sheet.
+While Lahman has data on a wide range of MLB related topics, we decided to focus on a singular data sheet.
 We focused on the "Batting" data set in Lahman which describes player-year-team batting statistics.
@@
 The effect of these cut-offs was that we were left with players who had longer careers and who hit HR's.
-There were many players dropped who never hit HR's because of their short stints which made prediction hard and prediction for these playes did not seem fruitful given their general inconsequential careers.
+There were many players dropped who never hit HR's because of their short stints which made prediction hard and prediction for these players did not seem fruitful given their general inconsequential careers.
 
@@
 Finally, in the last three years of his career we can see Yount's HR probability decline to about 1.75% when he was 38 years old.
-Overall, Yount appears to have a typical career compared to what the told our model to expect.
+Overall, Yount appears to have a typical career compared to what we told our model to expect.
 There was a steady increase in probability until he was about 30, followed by a decline afterwards as he got older.
@@
 We can see in @fig-p651Pred that the predictions from our model follow closely to what we told it to expect.
-There is an increase in HR probability, $\pi$, for Yount from when he is 18 to when he is 32.
+There is an increase in HR probability, $\pi$, for Yount from when he was 18 to when he was 32.
 This is quite close to what we see in the figure representing his actual HR probability.
@@
 It is simply a quirk of the data.
+
 Included in @fig-p582Act is a blue trend line that snakes through the data.
@@
 The second way is that Tabler peaks before he is 30 and does so bad for the final years of his career that there is not much decline.
-Thus, it will be interesting to see how our adapts to this very interesting and unique player.
+Thus, it will be interesting to see how our adapts to this very interesting and unique career.
 
@@
 That is, we can see our estimates for $\pi$ are often much closer to the average HR rate across players, which is 3%, than they are to Tabler's specific data because Tabler is such an outlier.
+
 Included in @fig-p582Pred is a red trend line for the same reasons mentioned in discussing @fig-p582Act.
@@
 Coming into this, the worry was that our model would rely too heavily on its priors and not be able to update its parabolic shape to match the strange career that Tabler had.
-However, after quickly examining the red line anyone can see that it has clearly updated to fit Tabler's data much better than @fig-p651Pred.
+However, after quickly examining the red line anyone can see that it has clearly updated to fit Tabler's data.
 That is, we can see that the red line from @fig-p582Pred matches the blue line in @fig-p582Act fairly well, demonstrating that the model updates itself based on the data it is fed.
@@
 
-See @fig-p420Act and @fig-p420Pred.
-
 Consider player $n=420$.
@@
 It is simply a quirk of the data.
+
 Included in @fig-p420Act is a blue trend line that snakes through the data.
@@
 
-There are many numerical measures of goodness-of-fit statistics because researchers are often searching for ways to determine which models of their many models are the best for their particular goals.
+There are many numerical measures of goodness-of-fit statistics because researchers are often searching for ways to determine which models, of their many, are the best for their particular research goals.
 Thus, many of these numerical estimates should be used in a comparative sense, with and against models which are trained on the same data set and attempt to answer the same research question.
@@
 Our model uses unique data compared to the models which attempt to answer the same research question.
-Thus, comparing models with these statistics would be foolish.
+Thus, comparing models with these statistics would be useless.
 
 However, that does not mean that there are no ways to check how well the model has fit.
-Beyond what has been discussed in @sec-yount, @sec-tabler, and @sec-morales, there is another method to check the goodness-of-fit for a model.
+Beyond what has been discussed in @sec-yount, @sec-tabler, and @sec-morales, there is another method to check the goodness-of-fit for a Bayesian model.
 Trace-plots (sometimes called trank-plots) are a way to check how individual parameters have fit.
-We have examined how the $\alpha$ parameter has fit for four players: Yount (651), Tabler (582), Morales (420), and Junior Felix (177).
+We have generated four plots to explore how the $\alpha$ parameter has fit for four players: Yount (651), Tabler (582), Morales (420), and Junior Felix (177).
 Junior Felix has not been discussed yet because he was the fourth of the random player's sampled and did not have as interesting of a career as the other three players discussed.
 Since he did not have a very interesting career, he did not provide much insight into the model fit and updating, so we decided to not display his career trajectory.
-However, we have included it @fig-alphaTrace as the last graph in the bottom right because it is a good example of what a trace-plot should look like, which contrasts nicely with the other trace-plots.
+However, we have included his $\alpha$ parameter estimation @fig-alphaTrace as the last graph in the bottom right because it is a good example of what a trace-plot should look like, which contrasts well with the other trace-plots.
 
@@
 We can see that, in general, the chains stay between -3.5 and -3 over their 5,000 iterations.
-There are no sudden deviations so we can conclude that "alpha\[177\]" or $\alpha_{177}$ in our model, has a good fit.
+There are no sudden deviations so we can conclude that "alpha\[177\]", or $\alpha_{177}$, in our model has a good fit.
 This stands in contrast to the other fits for $\alpha$.
@@
 
-Overall, we can see that the fit for our $\alpha$ parameter is quite poor for the considered players, even though the our model still fits the data well.
+Overall, we can see that the fit for our $\alpha$ parameter is quite poor for the considered players, even though the our model still adapts to the data well.
 This suggests that the other parameter fit well since our model did a good job at predicting and adapting to players trajectories.
@@
 That is, instead of needing 200,000 iterations to get stable estimates, we were able to get stable estimates using only 10,000 iterations.
-This suggests that our model is better Fellingham and Fisher's (2017) model and predicting HR's.
+This suggests that our model is better Fellingham and Fisher's (2017) model at predicting HR's.
 
@@
 Friendly, M., Dalzell, C., Monkman, M., & Murphy, D. (2022). Lahman: Sean 'lahman' 27 baseball database. <https://CRAN.R-project.org/package=Lahman>
+
+McElreath, R. (2020). *Statistical Rethinking: A Bayesian Course with Examples in R and Stan* (2nd ed.). CRC Press.
 

### 2023-12-15 Lahman Analysis/Formal Analysis/Draft1.qmd
FINISHED

@@
 The reason this paper does not consider both is because of multicollinearity.
-Multicollinearity is where two independent variables in a model are highly or perfectly correlated and the model is unable to determine which variable is responsible for the corresponding change in the depending variable.
+Multicollinearity is where two independent variables in a model are highly or perfectly correlated and the model is unable to determine which variable is responsible for the corresponding change in the dependent variable.
 Without a doubt there is a strong correlation between PA's and AB's because one is a subset of the other.
@@
 This, in combination with other factors such as harder materials used to make baseballs, led to players obliterating a variety of previous hitting records, such as the career and season home runs records.
-It is obvious from these facts that the decade played for player's had a drastic impact on their ability to produce some outcomes, such as HR's.
+It is obvious from these facts that the year played had a drastic impact on players' ability to produce some outcomes, such as HR's.
 In fact, this can be visualized by considering all the runs scored during a given season from 1871-2022 displayed in @fig-RunsByYear, or considering the total HR's hit from 1871-2022 displayed in @fig-HRByYear.
@@
 In MLB, seniority can act both as both a blessing and a curse.
-On the one hand, older players are less able to hit home runs because their body is starting to lose its peak athletic performance compared to when it was, say, 30.
-On the other hand, these older players also have more experience than younger players and thus may be better able to hit home runs from strategies they have found during their time playing.
+On the one hand, older players are less able to hit HR's because their body is starting to lose its peak athletic performance compared to when it was, say, 30.
+On the other hand, these older players also have more experience than younger players and thus may be better able to hit HR's from strategies they have found during their time playing.
 Further we can visualize this by looking at the median HR's hit by each player from 1871-2022, which is displayed in @fig-HR.By.Age.All.
 However, this graph does not depict some of the nuances of the data well.
-For example, we have almost no data on how often 18 -year-old players hit HR's because they play in MLB infrequently.
+For example, we have almost no data on how often 18 year old players hit HR's because they play in MLB infrequently.
 As can be seen with the standard error bars on the ages above 40, older players suffer a similar problem of small sample size.
@@
 This allows Fellingham and Fisher (2017) to make confident predictions about players who may only have 6 seasons of data because this model borrows data from all players who were born at around a similar time.
-Implicitly, this means that Fellingham and Fisher (2017) assume that when a player was born is a powerful predictor of their HR probability.
-
-In contrast, the parameter $\xi_i$ (which is $\delta_i$ in the actual paper) represents the effect that season $i$ has on player $n$'s ability to hit a HR at stadium $p$.
+Implicitly, this means that Fellingham and Fisher (2017) assume that when a player was born is a powerful predictor of their HR probability, which seems questionable.
+
+In contrast, the parameter $\xi_i$ (which is $\delta_i$ in the actual paper) represents the effect that season $i$ has on player $n$'s ability to hit a HR in year $i$.
 This $\xi$ is calculated for each season considered where $i\in\{1,…,138\}$ for the 138 seasons considered.
@@
 Thus, the $\alpha_n$ of players is unlikely to differ by very much and any difference in $\alpha_n$ which appears is likely also apparent in $\xi_i$.
-Thus, the inclusion of both $xi_i$ and $\alpha_n$ is strange because it introduces a strong possibility of multicollinearity since both parameters describe roughly the same trend in the data.
+Thus, the inclusion of both $\xi_i$ and $\alpha_n$ is strange because it introduces a strong possibility of multicollinearity since both parameters describe roughly the same trend in the data.
 
@@
 This is disappointing because the scientific method relies on replication for the advancement of knowledge.
-Thus, our research seeks to make the algorithms used in this research more accessible because we chose to use R to interface with the modeling software Stan, which is written in the (more contemporary) C++ programming language.
+Thus, our research seeks to make an algorithm that is more accessible because we chose to use R to interface with the modeling software Stan, which is written in the (more contemporary) C++ programming language.
 Furthermore, our model will use less iterations and stronger priors so that other research teams can more easily replicate our findings.
@@
 
-Exploring concretely how our model works will make it clearer than discussing it in the abstract. The current model that we have is the following:
+Our model is the following:
 
@@
 Further, the likelihood of the particular distribution that $\alpha_n$ takes on is dependent on the likelihood of either $\mu_0$ or $\sigma_0$ taking on a particular value.
-For example, $\mu_0=-3.0$ is extremely unlikely because it lies four standard deviations away from the mean of its normal distribution -3.5.
+For example, $\mu_0=-3.0$ is extremely unlikely because it lies five standard deviations away from the mean of its normal distribution -3.5.
 Thus a distribution for $\alpha_n$ being $Normal(-3,\sigma_0)$ is highly unlikely for any value of $\sigma_0$.
@@
 The value of $\beta_n$ represents how changes away from the centered age affects player $n$'s probability of hitting a HR, $\pi_{nip}$.
-In a normal GLM's, $\beta_n$ would not be able to adapt to each player, $n$, as it does in our model.
+In normal GLM's, $\beta_n$ would not be able to adapt to each player, $n$, as it does in our model.
 It would only be able to consider how aging affects the average player's HR probability and then apply that effect to all players.
